### 2015-12-21 paper/appendix3d.tex
upload paper and codes

\section{2D dispersion relation}\label{2ddisp}
The functions in Eq. \ref{thindisk} are 
\begin{align}
  A(S,K) =& \left(\frac{4}{3}\alpha+\alpha_b\right) K^2 + S +
  \mathcal{E}\mathcal{F}\label{bigA}
  - \frac{2|K|}{QS}, \\
  B(S,K) =& 2\left(\alpha q \mathcal{F} - 1\right),\\
  C(S,K) =& (2 - q) + \frac{\alpha q K^2(1+\mu)}{S} 
  + \alpha q \lambda \mathcal{E}\mathcal{F},\\
  D(S,K) = & \alpha K^2 + S + 2\alpha^2q^2\lambda\mathcal{F},
\end{align}
with
\begin{align}
  \mathcal{E} = \frac{\alpha q^2(1+\mu)}{S} +
    \frac{\gamma}{\gamma-1} + \frac{\theta}{\beta S(\gamma-1)},\quad
  \mathcal{F} = \frac{K^2(\gamma-1)\beta}{1 + \beta S - \alpha\beta
    q^2\lambda(\gamma-1)}\label{bigF}. 
\end{align}

\section{2D viscous GI} \label{gammie_check}
We recover \cite{gammie96}'s dispersion relation 
for viscous gravitational instability as follows. First we set
$\mu=-1, \lambda=0$ to obtain the same viscosity models. Next we
simplify Eq. \ref{bigA}---\ref{bigF} by assuming $|\beta S |\gg 1$  
%This will turn out to be equivalent to $\beta\gg 1$, i.e. the
%adiabatic limit. 
and $S\lesssim O(K^2)$ as $|K|\to 0$. Applying these considerations to
Eq. \ref{thindisk} gives 
\begin{align}
  S = \frac{\alpha K^2\left[\frac{2|K|}{Q} - \gamma K^2 +
      2q(2-q)(\gamma-1)\right]}{2(2-q) + \gamma K^2 - \frac{2K}{Q}}, 
\end{align}
which does not explicitly depend on cooling. 
Setting $\gamma=1$ then gives \citeauthor{gammie96}'s Eq. 18. The above
is expression thus generalizes \citeauthor{gammie96}'s result to
non-isothermal disks.

%cooling is not the physical cause of instability

\section{3D corrections in 2D theory}\label{3dcorr}
The simplest way to mimic the effect of finite disk thickness
on gravitational instabilities is to weaken self-gravity by reducing 
the gravitational constant  
\begin{align}
  G \to G\left(1+|k|H_\mathrm{sg}\right)^{-1}, 
\end{align}
or equivalently $Q\to Q\left(1+|k|H_\mathrm{sg}\right)$. This
prescription, derived by \cite{shu84}, is widely applied
\citep[e.g.][]{youdin11,takahashi14}. Here $H_\mathrm{sg}$ is a
measure of the disk thickness. We intuitively expect
$H_\mathrm{sg}\sim H$, but its precise value is not known a priori.   
We regard $H_\mathrm{sg}$ as a free parameter of the problem.

### 2015-12-21 paper/application.tex
upload paper and codes

\section{Application to protoplanetary disks}\label{2dppd}
We consider the stability of realistic, self-gravitating
protoplanetary disks. Specifically, we will apply our linear framework
to the gravito-turbulent disk model recently developed by 
\cite{rafikov15}. This 2D, Keplerian disk orbits a Solar mass star and 
is defined by the following parameters,

\begin{itemize}
  \item $\dot{M}$, the global radial accretion rate;
  \item $Q_0$, the value of the 2D Toomre parameter where the disk is
    gravito-turbulent;
  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
  \item $\alpha_m$, the dimensionless viscosity associated with other
    sources of turbulence.
\end{itemize} 
These are used in the statements of thermal equilibrium, mass and
angular momentum conservation, together with an opacity law, 
to construct a global disk model with surface density $\Sigma(R)$ and
temperature $T(R)$ where $R$ is the global cylindrical radius from the
star. This gives the $Q(R)$ value required for input
into our linear framework. We derive other dimensionless
parameters below.

Note that we are considering the local stability at each radius. We
are thus neglecting the slow, background viscous radial flow in the
linear problem.

%proper treatment of cooling  in 3d
%vertically isothermal basic state for simplicity (also appropriate for
%irradiated disks) --- this formally requires zero viscosity  
%or have viscosity but assume it does not affect basic state (dominated
%by irradiation) but affect perturbations

\subsection{Effective $\alpha$}
In a steady, viscously accreting Keplerian disk we have,
approximately, $\nu\Sigma = \dot{M}/3\pi$. Hence
\begin{align}
  \alpha = \frac{\dot{M}}{3\pi}\frac{\Omega(R)}{c_{s}^2(T)\Sigma(R)},  
\end{align} 
which sets the viscosity coefficient to be used at each radius.

\subsection{PPD cooling}
Energy loss in \citeauthor{rafikov15}'s model is given
by 
\begin{align}\label{real_cool}
  \Lambda = \frac{2\sigma}{f(\tau)}\left(T^4 - T_\mathrm{irr}^4\right)
\end{align}
per unit area, 
\begin{align}
  f(\tau) = \tau + \frac{1}{\tau},
\end{align}
and
\begin{align}
  \tau = \kappa_d(T)\Sigma
\end{align}
is the optical depth and recall Eq. \ref{opacity_law} 
is our opacity model. 
%and $\kappa_d$ is the opacity. 
%We parameterize
%\begin{align}
%  \kappa_d = \kappa_{d0} \Sigma^a T^b,
%\end{align}
%where $\kappa_{d0}, a, b$ are constants.

Since we are interested in the linear response of the system, the
corresponding parameters $\beta$ and $\theta$ are found by
comparing the linearized form of the cooling function
(Eq. \ref{real_cool}) to that of the model function
(Eq. \ref{beta_cool}).

%{\bf as opposed to comparing their unperturbed forms, as done in literature}

Linearizing Eq. \ref{real_cool} gives   
\begin{align}\label{linear_real_cool}
  (\gamma-1)\frac{\delta\Lambda}{\Sigma} = \frac{2\sigma(\gamma-1)
    T^4C_1}{f(\tau)c_{s}^2\Sigma}\left(\frac{\delta P}{\Sigma} -
  \frac{C_2}{C_1}c_{s}^2\frac{\delta\Sigma}{\Sigma}\right), 
\end{align}
where
\begin{align}
C_1(\tau, T) &= 4 - b\times g(\tau, T),\\ 
C_2(\tau, T) &= 4 + (1-b)\times g(\tau, T),\\ 
  g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
  \frac{T_\mathrm{irr}^4}{T^4}\right). 
\end{align}
Comparing Eq. \ref{linear_real_cool} with Eq. \ref{beta_cool}, we identify
\begin{align}
  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\\
  &\theta = \frac{C_2}{C_1}, 
\end{align}
to be used in the 2D linear framework.

This `realistic' beta cooling prescription can behave differently to 
our previous, simple beta cooling (Eq. \ref{beta_cool}) for the same
physical system. In particular, for a non-irradiated disk
($T_\mathrm{irr}=0$), simple beta 
cooling has  $\theta=0$, but realistic beta cooling has $\theta\neq
0$. However, fully-irradiated disks with $T=T_\mathrm{irr}$ can be
modeled by simple beta cooling since in both cases $\theta=1$.

%{\bf caution in applying numerical results obtained from beta cooling to
%systems modeled with real cooling. above example shows that to model
%real, non-irradiated disk with beta cooling actually requires irradiation in
%the latter.}

%Notice that $\theta\to 1$ as 
%$T\toT_\mathrm{irr}$ the disk becomes passive.

\subsection{Inviscid stability condition} 
%We specialize to $a=0, b=2$ for the opacity law. 
We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
\begin{align*}
  \theta = \frac{4-g}{4-2g},
\end{align*}
implying $5/6<\theta<3/2$. 
%For a non-irradiated disk, $T_\mathrm{irr}=0$, $\theta\to5/6$ as
%$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
From our discussion in \S\ref{2d_inviscid}, we conclude that this
system cannot be unstable without
viscous effects acting on the perturbations if 
\begin{align}
  \gamma > \frac{3}{2} \quad \text{and} \quad Q > \sqrt{\frac{6}{5}} \label{ppd_invisc_cond}
\end{align} 
are both satisfied.

\subsection{Example 2D calculation}\label{pp2d_example}
We consider a disk with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. For the adopted opacity
law we follow \citeauthor{rafikov15} and take $\kappa_{d0} =
5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$. Here we use
$\gamma=1.6$ so the inviscid stability condition
(Eq. \ref{ppd_invisc_cond}) is satisfied. 
%{\bf actually this opacity law only valid $>30$au for this disk}

Fig. \ref{rafikov_model} shows the equilirbium disk profile in terms
of $Q$, $\alpha$, $\beta$, and $\theta$. These are set as input to the
2D dispersion relation to yield a growth timescale, in units of the
local orbital period, shown in Fig. \ref{rafikov_growth}. We also plot
an analytic estimate based on Eq. \ref{gammie_smallk} (instead of
Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) which gives 
\begin{align}
  S = \frac{27\alpha}{16\theta^3Q^4}, 
\end{align}
and the optimum wavenumber is
\begin{align}
  |K| = \frac{3}{2\theta Q}.
\end{align}
%For the
%stability calculation we use $\mu=-1, \lambda=0$ for the viscosity
%prescription.

The disk becomes dynamically unstable with growth timescales $\lesssim
10$ orbits for $R\gtrsim60$AU. Coincidentally, beyond this radius
$\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is often quoted as 
criteria for disk fragmentation \citep[e.g.][]{rafikov15}. 
%{\bf despite apply beta cooling results to realistic cooling disks}
Our result suggest an alternative interpretation of disk fragmentation
is where the gravito-turbulent state becomes dynamically unstable due to
both viscosity and self-gravity.

Fig. \ref{rafikov_model} shows that the most unstable wavelength is a
few times the disk thickness. The increase in $|K|$ from $\sim
10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is due to the decrease in $Q$;
while that from $\sim 60\mathrm{AU}$ to $\sim 100\mathrm{AU}$ occurs
as the disk transitions from optically thick to optically thin
regimes. The mismatch at large distances is not surprising since the
above expressions assume $|K|\ll 1$. Nevertheless the analytic
estimates reproduce the correct behavior.

%{\bf note: disk might be stablized by extremely large alpha at large
%  distances but `turbulent' framework no longer valid }

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.0cm]{figures/ppd_2d_basic}
  \caption{Equilibrium profile obtained from the disk model developed
    by \cite{rafikov15}, with parameters $\dot{M} =
    10^{-6}M_\sun\yr^{-1}$, $Q_0=1.5$, $\tirr=10\mathrm{K}$, and
    $\alpha_m=10^{-3}$.   
    \label{rafikov_model}}
\end{figure}

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0.0cm]{figures/ppd_2d_growth}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_2d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous,
    self-gravitational modes in the 2D PPD model shown in
    Fig. \ref{rafikov_model}. 
    The black curve is obtained numerically
    from the 2D dispersion Eq. \ref{thindisk}, while green curves 
    are analytic results based on Eq. \ref{gammie_smallk}. 
    \label{rafikov_growth}}
\end{figure}

\subsection{3D PPD}
%{\bf radiative diffusion only for optically thick disk. how to connect
%to optically thin regions?}
We briefly consider 3D PPDs subject to explicit radiative diffusion 
(\S\ref{rad_cool}). At each radius $R$ we obtain the vertical
structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux, given
the $\alpha(R)$ and $Q(R)$ profiles obtained from the 2D model
above. We then solve the 3D eigenvalue problem as in
\S\ref{3ddisk}, with the additional boundary condition for the temperature 
perturbation $\delta T^\prime(0) = \delta T(\zmax) = 0$.

We use the same disk model as in \S\ref{pp2d_example} but without
irradiation ($T_\mathrm{irr}=0$), and consider a smaller domain
with $R\in[10,100]AU$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0.0cm]{figures/ppd_3d_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.8cm]{figures/ppd_3d_maxk}
  \caption{Growth timescales (top) of the most unstable
    wavenumber (bottom) for viscous,
    self-gravitational modes.  
    \label{rafikov_growth3d}}
\end{figure}

%3d rad diffusion is mathematically different problem (more
%derivatives) so worth considering briefly
%optically thin cooling in 3D is the same problem as before

%use 2D model as midplane values to get 3D structure assuming radiative
%diffusion and constant alpha w.r.t. height. allow cooling time to vary
%with height. can't assume polytropic vertical structure. need to
%reformulate linear framework (some transformations needed). 
%skip 2D eigen prob if 3D works out

### 2015-12-21 paper/linear3d.tex
upload paper and codes

%On the other hand, if a time-independent heat source is invoked, then
%we may choose $\alpha, \beta$ and $\theta$ separately.  
\section{Linear problem}
We consider infinitestimal axisymmetric Eulerian perturbations in the
form \begin{align}
  \delta \rho= \widetilde{\delta \rho}(z)\exp{\left(\ii kx +
    s t\right)},  
\end{align}
and similarly for other variables. Here, $k$ 
is a real horizontal wavenumber and $s$ is a (generally) complex
growth rate. 
%The gravitational potential perturbation $\delta \Phi$ depends on $z$
%but only its midplane value $\delta \Phi(z=0)$ is
%used in the dynamical equations, which has the same form as above. 
For clarity, hereafter we drop the tilde.  
%and note that equilibrium
%quantities without the $\delta$ prefix.

The linearized continuity, momentum and energy equations are
\begin{align}
   &s \delta \rho = - \ii k \rho\delta v_x -
  \left(\rho\delta v_z\right)^\prime \label{lin_mass} \\ 
   &s  \delta v_x = - \ii k\frac{\delta P}{\rho} - \ii k \delta 
  \Phi + 2\Omega \delta v_y + \delta F_x,\\
   &s \delta v_y = (q-2)\Omega\delta v_x + \delta F_y,\\
%   &s \delta v_z = - \left(\frac{\delta P}{\rho}\right)^\prime -
%  \delta\Phi^\prime +
%  \left(\ln{\rho}\right)^\prime\left(c_s^2\frac{\delta\rho}{\rho} -
%  \frac{\delta P}{\rho}\right) + \delta F_z,\\
  &s \delta v_z =
  \left(\ln{P}\right)^\prime c_s^2\frac{\delta\rho}{\rho} -
  \left(\frac{\delta P}{\rho}\right)^\prime -
  \left(\ln{\rho}\right)^\prime\frac{\delta P}{\rho} + \delta F_z, \\ 
  &s \frac{\delta P}{\rho} = 
%- \ii \gamma k \frac{c_s^2}{\Gamma}\delta v_x 
 - \ii k \gamma c_s^2\delta v_x
%- c_s^2\left[
%  \left(\ln{\rho}\right)^\prime\delta v_z + 
%  \frac{\gamma}{\Gamma}\delta v_z^\prime\right] \notag\\
- c_s^2\left[
 \left(\ln{P}\right)^\prime\delta v_z + 
  \gamma\delta v_z^\prime\right] \notag\\
   & \phantom{s \frac{\delta P}{\rho}=}
  + (\gamma-1)\frac{\delta\mathcal{H}_\mathrm{visc}}{\rho} 
  -(\gamma-1)\frac{\delta\Lambda}{\rho}
%  - \frac{1}{t_c}\left(\frac{\delta P}{\rho} -
%  \frac{\theta}{\Gamma}c_s^2\frac{\delta\rho}{\rho}\right) 
  \label{lin_energy},\\
  &\delta\Phi^{\prime\prime} - k^2\delta\Phi =
  \frac{\Omega^2}{Q_\mathrm{3D}}\left(\frac{\delta\rho}{\rho_0}\right)\label{lin_gravity},  
\end{align}
with
\begin{align}
  \delta F_x =& \nu \left[\delta v_x^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_x^\prime - 
     \frac{4}{3}k^2\delta v_x \right] - \nu_bk^2\delta v_x \notag\\
   &+ \ii \nu k \left[\frac{1}{3}\delta v_z^\prime+ 
    \left(\ln{\rho\nu}\right)^\prime\delta v_z\right] + \ii k \nu_b
  \delta v_z^\prime,\\
  \delta F_y = & \nu  \left[\delta v_y^{\prime\prime} + 
    \left(\ln{\rho\nu}\right)^\prime \delta v_y^\prime - 
     k^2\delta v_y \right] - \ii\nu kq\Omega\delta\ln{\rho\nu},\\
  \delta F_z =& \nu\left[\frac{4}{3}\delta v_z^{\prime\prime} + 
    \frac{4}{3}\left(\ln{\rho\nu}\right)^\prime \delta v_z^\prime - 
    k^2\delta v_z \right] \notag\\
  & +\nu_b\left[\delta v_z^{\prime\prime} + \left(\ln{\rho\nu_b}\right)^\prime\delta v_z^\prime\right]\notag\\
  & + \ii \nu k  \left[\frac{1}{3}\delta v_x^\prime -  
    \frac{2}{3}\left(\ln{\rho\nu}\right)^\prime\delta v_x\right]\notag\\
  &+\ii \nu_b k \left[\delta v_x^\prime + \left(\ln{\rho\nu_b}\right)^\prime\delta v_x\right],\\
%  &\delta f_x = -\frac{4}{3} \nu k^2\Sigma \delta v_x,\\
%  &\delta f_y = -\ii q k \Omega \delta(\nu\Sigma) - \nu k^2
%  \Sigma\delta v_y,\\
%  &\delta Q_\mathrm{visc} = (q\Omega)^2\delta(\nu\Sigma) - 2 \ii \nu 
%  k q\Omega\Sigma\delta v_y,\\
%  &\delta(\nu\Sigma) =
%  \frac{\alpha}{\Omega}\left[c_s^2(1+\mu)\delta\Sigma + \lambda\delta P\right].
%  &\delta f_y = - \ii \nu q k \Omega \delta \Sigma - \nu k^2\Sigma\delta
%  v_y,\\
%  &\delta Q_\mathrm{visc} = \nu
%  \left(q\Omega\right)^2\delta\Sigma - 2\ii\nu q k \Omega\Sigma\delta
%  v_y. 
  \frac{\delta\mathcal{H}_\mathrm{visc}}{\rho}=& \nu (q\Omega)^2
  \delta\ln{\rho\nu} - 2\ii\nu k q \Omega \delta v_y,\\
  \delta\ln{\rho\nu} =& (1+\mu)\frac{\delta\rho}{\rho} +
  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}, 
\end{align}\label{linear_beta_cool}
where $^\prime$ denotes $d/dz$. The linearized cooling functions
$\delta\Lambda$ are given below. 
Eq. \ref{lin_mass}---\ref{lin_gravity}, supplemented with appropriate
boundary conditions, constitutes an eigenvalue problem for the growth
rate $s$.

\subsection{Linearized beta cooling}
For the standard beta cooling prescription, linearizing
Eq. \ref{beta_cool} gives

\begin{align}
  (\gamma-1)\frac{\delta \Lambda}{\rho} = \frac{1}{t_c}\left(\frac{\delta P}{\rho} -
  \theta c_s^2 \frac{\delta\rho}{\rho}\right), \label{linear_beta}
\end{align}
where we have used $\delta T/T  = \delta P/P - \delta\rho/\rho$ from
the ideal gas law.

\subsection{Linearized radiative cooling}
Linearizing Eq. \ref{rad_cool1}---\ref{rad_cool2} with the
temperature-dependent opacity law in Eq. \ref{opacity_law} gives
\begin{align}
  \frac{\delta \Lambda}{\rho} 
  =& \frac{16\sigma T^3}{3\kappa_d\rho^2} k^2 \delta T \notag\\
  & - \frac{16\sigma }{3\rho}\frac{d}{dz}\left\{
  \frac{T^3}{\kappa_d\rho}\left[\delta T^\prime +
    (3-b)\left(\ln{T}\right)^\prime \delta T - T^\prime \frac{\delta\rho}{\rho}\right]
  \right\}.\label{linear_rad_cool}
\end{align}
Since Eq. \ref{linear_rad_cool} contain vertical derivatives of the
perturbations, it is generally not possible to map radiative cooling 
to the beta cooling prescription, except for special problems
\citep[e.g.][]{lin15}.

### 2015-12-21 paper/model3d.tex
upload paper and codes

\section{Basic equations}
We consider a three-dimensional (3D), self-gravitating,
viscous disk with heating and cooling. We use the shearing box framework to
study a small patch of the disk \citep{goldreich65}. The local frame
co-rotates with a fiducial point in the unperturbed disk at angular
frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ corresponds
to the radial, azimuthal and verical directions in the global disk. 
The fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right) = 0, \\
  & \frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi - 2\Omega\hat{\bm{z}}\times\bm{v}\notag\\ & 
  \phantom{\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =}
  +2\Omega^2qx\hat{\bm{x}} -\Omega_z^2z\hat{\bm{z}} 
   + \frac{1}{\rho}\nabla\cdot\bm{T},\label{momentum_eq}\\
  & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} +
  \mathcal{H}_\mathrm{ext} - \Lambda,\label{energy_eq}
\end{align} 
where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
the velocity field.% $q$ is the shear parameter and $\Omega_z$ is the
%local vertical oscillation frequency. 
We assume an ideal has so that the pressure $P$ and thermal energy
density $E$ are related by  
\begin{align}
  P = (\gamma-1)E = \mathcal{R}\rho T, 
\end{align}
where $\gamma$ is the adiabatic index, $\mathcal{R}$ is the gas
constant, and $T$ is the temperature. We use $\gamma=1.4$ unless
otherwise stated. The gas 
gravitatational potential $\Phi$ is given via the Poisson equation,  
\begin{align}
 \nabla^2\Phi = 4 \pi G \rho, 
\end{align}
where $G$ is the gravitational constant.

In the momentum equation (Eq. \ref{momentum_eq}), the third,
fourth/fifth, and last term on the right-hand side represent the Coriolis, 
tidal, and viscous forces (see below), respectively. We consider    
Keplerian disks with shear paramter $q=3/2$ and vertical oscillation 
frequency $\Omega_z=\Omega$.

In the energy equation (Eq. \ref{energy_eq}) the source terms 
$\mathcal{H}_\mathrm{visc}$, $\mathcal{H}_\mathrm{ext}$ and $\Lambda$
represent viscous heating, time-independent external heating and
thermal losses, respectively, described below. We normally set
$\mathcal{H}_\mathrm{ext}=0$ unless otherwise stated.

\subsection{Viscosity and heating}\label{visc_model}
The Cartesian components of the viscous stress tensor $\bm{T}$ is
defined by 
\begin{align}
  T_{ij} \equiv \rho \left[\nu \left(\p_j v_i + \p_i v_j\right) +
    \left(\nu_b-\frac{2}{3}\nu\right)\delta_{ij}\nabla\cdot\bm{v}\right], 
\end{align}
%bulk visc ignored 
where $\rho\nu$ is the shear viscosity, and we include a bulk
viscosity $\rho\nu_b$ for completeness.  
The associated viscous heating given by
\begin{align}
  \mathcal{H}_\mathrm{visc} \equiv \left(\p_jv_i\right)T_{ij}, 
\end{align}
where summation over repeated indices is implied.

We adopt the viscosity law  
\begin{align}
  \nu = \alpha
  \frac{c_{s0}^2}{\Omega}\left(\frac{\rho}{\rho_\mathrm{eq}}\right)^\mu\left(\frac{P}{P_\mathrm{eq}}\right)^\lambda,            
\end{align}
where subscript `eq' denotes the equilibrium state and  
%$c_{s0}$ is a modified sound-speed defined below. 
$c_{s0}^2\equiv P_\mathrm{eq}(0)/\rho_\mathrm{eq}(0)$. 
The dimensionless coefficient
$\alpha=\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
magnitude of the shear viscosity and is time-independent.  
The indices $\mu,\,\lambda$ are chosen to model how 
the viscosity behaves in the perturbed state. Given the uncertain
properties of gravito-turbulence, we will treat $\mu,\lambda$ as free
parameters. We adopt the same prescription for the bulk viscosity but
with $\alpha\to\alpha_b$.

Our numerical calculations mostly use $\mu=-1,\lambda=0$ so
that $\rho\nu$ remains constant, as considered
in previous gravitational instabilities with 
viscosity \citep{hunter83,gammie96}. 
This model conveniently eliminates  
viscous over-stability \citep{schmit95,latter06}, which is unrelated
to self-gravity, that would otherwise contaminate our results.

While the choice $\mu=-1,\lambda=0$ may seem artifical, we emphasize
this is done in order to bring out the relevant physical process ---
viscosity-assisted gravitational instability. Interestingly, 
\cite{laughlin96b} had already suggested a $1/\rho$ dependence for the 
shear viscosity when modeling the evoluting of self-gravitating disks
as a viscous process.  
{\bf note paardekooper 2012: alpha goes down as Sigma goes up}

\subsection{Steady states and cooling models}

We consider equilibrium solutions (omitting the `eq' subscripts
hereafter for clarity)
\begin{align}
  \bm{v} &= -q\Omega x \hat{\bm{y}}, \\
  \rho   &= \rho(z),\\
  P      &= P(z) \equiv c_s^2(z)\rho.
\end{align} 
%and assume a polytropic relation for the equilibrium pressure,
%\begin{align}
%  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma
%  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
%\end{align}
%where $\rho_0$ is the equilibrium midplane density and $\Gamma$ is the
%constant polytropic index.  
The equilibrium density and pressure fields are obtained by solving
the vertical momentum equation with self-gravity,
\begin{align}
  &\frac{1}{\rho}\frac{dP}{dz} +
  \Omega_z^2z + \frac{d\Phi}{dz} = 0, \label{vert_eq1}\\
 &\frac{d^2\Phi}{dz^2} = 4 \pi G \rho,\label{vert_eq2}
\end{align}
%A numerical solution is usually required to obtain the equilibrium
%state, except for special values of $\Gamma$. 
together with thermodynamic equilibrium,
%\subsection{Thermal equilibrium}
%In equilibrium the energy equation implies
\begin{align}\label{thermal_eq}
%  \mathcal{H}_\mathrm{visc} + \mathcal{H}_\mathrm{ext} = \Lambda.   
(q\Omega)^2\rho\nu + \mathcal{H}_\mathrm{ext} = \Lambda,
\end{align}
where the first term represents viscous heating. For the viscous
problem we may set $\mathcal{H}_\mathrm{ext}=0$. However, if we wish
to neglect viscosity but include cooling, we must invoke
$\mathcal{H}_\mathrm{ext}\neq0$ to define an equilibrium state.  
To proceed further, we separately describe the two cooling models
considered in this work. 
%It is possible to have both models present - but too complicated

\subsubsection{Beta cooling}
In this case the energy loss per unit volume is specified 
as an explicit function of the thermodynamic variables. 
A prototypical example is 
\begin{align}\label{beta_cool}
  \Lambda =
  %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
\end{align}
where 
\begin{align}
  \tcool = \beta\Omega^{-1} 
\end{align}
is the cooling timescale with $\beta$ a constant input parameter, and
\begin{align}
  \tirr = \theta T_\mathrm{eq}
\end{align}
is a reference temperature field parameterized as a constant fraction
$\theta$ of the equilibrium temperature $T_\mathrm{eq}$. We will refer
to $\theta$ as the `irradiation level'.

This type of cooling is widely applied in 2D and 3D
numerical simulations of self-gravitating disks \citep{gammie01,
  rice05,baehr15}. It permits numerical experiments to be carried out in a
controlled manner as a function of the cooling time $\beta$.

We remark that for the linear problem, 
any explicit cooling function $\Lambda=\Lambda(\rho,T)$ can be
mapped to that obtained from the standard function in Eq.
\ref{beta_cool}. We do this in \S\ref{2dppd} where we adopt a more
realistic beta cooling prescription for 2D PPDs.

For models with beta cooling we assume an equilibrium polytropic relation
\begin{align} 
  P  =
%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma,
  %  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
c_{s0}^2\rho_0\left(\frac{\rho}{\rho_0}\right)^\Gamma,
\end{align}
where $\Gamma$ is the constant polytropic index and $\rho_0 =
\rho(0)$. The vertical
structure is first obtained from Eq. \ref{vert_eq1}---\ref{vert_eq2}
(usually numerically), then inserted into
Eq. \ref{thermal_eq} to infer the required viscosity profile

\begin{align}\label{alpha_beta_relation}
\alpha(z) = \frac{1}{(\gamma-1)\beta
   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
\end{align}

\subsubsection{Radiative cooling}\label{rad_cool}
Here we assume energy transfer by radiative diffusion. Then
\begin{align}
  \Lambda &= \nabla\cdot\bm{F},\label{rad_cool1}\\
  \bm{F}   &= -\frac{16\sigma T^3}{3\kappa_d\rho}\nabla T, \label{rad_cool2}
\end{align}
where $\sigma$ is the Stefan-Boltzmann constant and 
$\kappa_d$ is the (dust) opacity. We adopt
\begin{align}\label{opacity_law}
  \kappa_d = \kappa_{d0}T^b,
\end{align}
and take the constant index $b=2$ as appropriate for {\bf blah},
but retain the general notation $b$ to keep track of the opacity.   
%We use radiative cooling when considering 3D PPDs in \S.

While radiative cooling is arguably more realistic than 
parametrized beta cooling, it complicates our 
problem because it generally implies a vertically non-isothermal
equilibrium disk, and increases the order of the linearized equations.

In this case we specify a constant viscosity coefficient $\alpha$ and
solve Eq. \ref{vert_eq1}---\ref{thermal_eq}, together with
Eq. \ref{rad_cool1}---\ref{rad_cool2}, as a fourth order system of
ordinary differential equations to obtain $P(z)$, $T(z)$, and hence
$\rho(z)$.

%\subsection{Steady states}
%\subsubsection{Equilibria with beta cooling}   
%\subsubsection{Equilibria with radiative cooling}
%If $\mathcal{H}_\mathrm{ext}=0$ then viscosity and cooling
%cannot be prescribed independently and we have 
%\begin{align}\label{alpha_beta_relation}
%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
%\end{align}
%On the other hand, if a time-independent heat source/sink is invoked
%($\mathcal{H}_\mathrm{ext}\neq 0$), then we may choose $\alpha, \beta$
%and $\theta$ separately.

\subsection{Parameters}
{\bf temporary, maybe move to intro}
For the 3D problem it is convenient to parameterize the strength of
self-gravity via 
\begin{align}
  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
\end{align}
It is also useful to define the
two-dimensional (2D) Toomre parameter 
\begin{align}
  Q = \frac{c_{s0}\Omega}{\pi G \Sigma}, 
\end{align}
where $\Sigma = \int_{-\infty}^\infty\rho dz$ is the surface density.

%We are interested in comparing systems with different $\beta$. 
%However, this gives unphysically large values of $\alpha$ for small
%values of $\beta$. In order to compare systems with different $\beta$,
%we adopt the following procedure. We choose parameters  
%$\theta_\mathrm{min}$ and $\alpha_\mathrm{max}$, and evaluate
%Eq. \ref{alpha_beta_relation} at the midplane to define a cooling time
%$\beta_\mathrm{min}$. Then for $\beta\geq\beta_\mathrm{min}$ we use 
%Eq. \ref{alpha_beta_relation} to set $\alpha$ at fixed
%$\theta=\theta_\mathrm{min}$; while for $\beta < \beta_\mathrm{min}$
%we use Eq. \ref{alpha_beta_relation} to set $\theta$ at fixed
%$\alpha(0)=\alpha_\mathrm{max}$.  That is, we raise the reference
%temperature (i.e. irradiation levels) when the imposed cooling is too
%rapid to be sustaind by viscous heating.

### 2015-12-21 paper/paper3d.tex
upload paper and codes

%\documentclass[12pt]{aastex}
\documentclass[iop, numberedappendix]{emulateapj}
%\documentclass{article}
%\usepackage{draftwatermark}
%\usepackage{paralist}
\shorttitle{TBD}
\shortauthors{M-K.\ Lin, K. Kratter}

\usepackage{amsmath}
\usepackage{paralist}
%\usepackage{url}
\usepackage{bm}
%\usepackage[font=small,format=plain,labelfont=bf,up,textfont=it,up]{caption}
\newcommand{\p}{\partial}
\newcommand{\zmax}{z_\mathrm{max}}
\newcommand{\lmax}{l_\mathrm{max}}
\newcommand{\sbar}{\bar{\sigma}} 
\newcommand{\He}{\operatorname{He}}
\newcommand{\imgi}{\mathrm{i}} 
\newcommand{\ii}{\mathrm{i}}
\newcommand{\C}{\mathcal{C}^\lambda}
\newcommand{\wop}[1]{\mathcal{V}^{(#1)}}
\newcommand{\avg}[1]{\langle{#1}\rangle}
%\newcommand{\max}[1]{\mathrm{max}({#1})}
%\newcommand{\min}[1]{\mathrm{min}({#1})}
\newcommand{\zhat}{\hat{z}}
\newcommand{\khat}{\hat{k}}
\newcommand{\rhat}{\hat{r}}
\newcommand{\qthree}{Q_\mathrm{3D}}
\newcommand{\qtwo}{Q}
\newcommand{\he}{\mathcal{H}}
\newcommand{\amp}{\mathcal{A}}
\newcommand{\dd}{\delta}
\newcommand{\real}{\operatorname{Re}}
\newcommand{\imag}{\operatorname{Im}}
\newcommand{\sgn}{\operatorname{sgn}}
\newcommand{\wbar}{\tilde{W}}
\newcommand{\qbar}{\tilde{Q}}
\newcommand{\dbigH}{\frac{H^\prime}{H}}
\newcommand{\Rp}{\frac{\rho_0^\prime}{\rho_0}}
\newcommand{\Fp}{\frac{g^\prime}{g}}
\newcommand{\Rpp}{\frac{\rho_0^{\prime\prime}}{\rho_0}}
\newcommand{\Fpp}{\frac{g^{\prime\prime}}{g}}
\newcommand{\ils}{L_s^{-1}}
\newcommand{\ilp}{L_p^{-1}}
\newcommand{\ihp}{H_p^{-1}}
\newcommand{\ihs}{H_s^{-1}}
\newcommand{\qcrit}{Q_\mathrm{crit}}
\newcommand{\ciso}{c_\mathrm{iso}}
\newcommand{\csmid}{c_{s0}} 
\newcommand{\hiso}{h_\mathrm{iso}}
\newcommand{\Hiso}{H_\mathrm{iso}}
\newcommand{\zeus}{\texttt{ZEUS-MP }}
\newcommand{\rsph}{r_\mathrm{sph}}
\newcommand{\dvx}{\dd v_x}
\newcommand{\dvy}{\dd v_y}
\newcommand{\dvz}{\dd v_z}
\newcommand{\dbx}{\dd B_x}
\newcommand{\dby}{\dd B_y}
\newcommand{\dbz}{\dd B_z}
\newcommand{\tcool}{t_\mathrm{c}}
\newcommand{\tirr}{T_\mathrm{irr}}
\newcommand{\w}{ \widetilde{W}}
\newcommand{\dphi}{\dd \Phi}
\newcommand{\actaa}{Acta Astron.}
\newcommand{\hsg}{H_\mathrm{sg}}
\newcommand{\yr}{\mathrm{yr}}
\newcommand{\qvert}{Q_\mathrm{3D}}
\defcitealias{lin12}{L12} 
\defcitealias{nelson13}{N13}
\defcitealias{barker15}{BL15}
\defcitealias{mcnally14}{MP14}
\defcitealias{goldreich67}{GS67}

\def \OmK {\Omega_{\rm K}}

\begin{document}

%\title{untitled linear problem}
\title{On the Gravitational stability of gravito-turbulent disks}
\author{Min-Kai Lin\altaffilmark{1} \& Kaitlin M. Kratter}
\affil{Department of Astronomy and Steward Observatory, \\
University of
  Arizona, 933 North Cherry Avenue, Tucson, AZ 85721, USA}
\altaffiltext{1}{Steward Theory Fellow}
\email{minkailin@email.arizona.edu}

\begin{abstract}
stuff
%on the gravitational stability of gravito-turbulent disks
\end{abstract}

\section{Introduction}\label{intro}
%normal GI and viscous GI. appplicability of simple stability criteria
%and dispersion relation (toomre and cooling) to complex
%systems.

%non-existence of gravito-turbulent state for fast cooling
%because such states are dynamically unstable. secondary gi in two
%ways: classic, pure gravity; viscosity-assisted.

%provide generalized dispersion relations.

%3D calculations.

%apply linear theory
%to real disk models.

%notion of max alpha comes from simulations. 
%max alpha criteria applies in local simulations, so not to do with how
%fast gi can transport material in a global disk.

%combine independent fragmentation conditions Q less than one and beta
%less than a few in a single framework: these determine growth rate of
%viscous GI.

%toomre Q and standard dispersion relation here

%real disks have cooling (removes pressure support)
%and turbulence (removes rotational support ---  if it's like a
%viscosity)

%indeed fragmenting disks have cooling and turbulence

%so discussion of GI in real disks should have both

Understanding the gravitational stability (GI) of rotating disks   
is fundamental to many astrophysical problems. In the context of 
protoplanetary disks (PPDs), GI has two applications. First, it can
provide gravitational torques to transport angular momentum outwards
and thus enable mass accretion inwards. Secondly, it may lead to disk
fragmentation, which has been invoked to explain planet formation at
large radii. Any realistic study of these phenomena require direct
numerical simulations. Nevertheless, significant progress can be
made through analytical understanding of GI.  %physical insight

It is now customary to begin any discussion of disk GI with 
the \emph{Toomre criterion}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}<1 
\end{align}
for instability \citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic 
frequency (which equals the rotation frequency $\Omega$ in a Keplerian
disk), $\Sigma$ is the surface density and $G$ is the
gravitational constant. This condition describes how self-gravity
($\Sigma$) destabilizes the disk against 
rotation ($\kappa$) and pressure ($c_s$). Eq. \ref{toomreQ_criterion}
formally applies to local, axisymmetric disturbances in a
two-dimensional (2D, razor-thin), inviscid and adiabatic (including
isothermal) disk.

%assess GI in general

Despite being a result from ideal disk models, the Toomre $Q$  
parameter is widely applied to assess GI in more general 
situtations. This includes non-ideal, realistic disk structures
obtained with additional physics not encapsulated by $Q$
\citep[e.g.][]{kimura12}.  It is, however, not obvious that such additions
leave the stability condition unaltered. Indeed, we show that GI in more
realistic disk models cannot be described by a single `Toomre-like' 
condition.

%disk structure may be computed with sophisticated physics but
%stability condition still based on ideal disk models. 
%three main ingredients - gravity, pressure, rotation 
%a standard way to apply this condition is to compute the disk
%structure  based on realistic physics

The goal of this work is to generalize the analytic treatment of disk 
GI to allow a self-consistent discussion of GI in 
realistic PPD models. Specifically, we will examine the effect of
cooling and viscosity on GI. Their relevance to PPDs are described
below.

%stuff aboout gravito turbulence 
%Th
%we focus on cooling and viscosity but also do 3D 
%but real disks have cooling and viscosity/turbulence
%(e.g. gravito-turbulent disks)

\subsection{Cooling removes pressure support}
A net thermal loss\footnote{Isothermal disks implicitly assume 
  heating and cooling processes are exactly balanced at all
  times, so these do not have cooling in the current context.} (hereafter `cooling')
reduces pressure (thermal) support against self-gravity. The cooling time 
$\tcool$ is the time-scale over which the disk temperature is relaxed
to some pre-defined value, which may be zero. It is 
often written as 
\begin{align} 
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time.

{\bf physical origin of cooling in PPDs}

The Toomre $Q$ parameter does not include cooling, and thus cannot be 
used to assess GI in realistic disks. Indeed, there exists a separate, 
\emph{empirical} cooling time criterion which governs stability against disk
fragmentation (see below), but its precise value is uncertain.

We shall quantify, using linear stability analyses, how cooling 
destablize disks. We will find that cooling in fact changes the 
qualitative nature of GI, in that its mere presence can
formally lead to instability.

\subsection{Viscosity removes rotational support}
It is well known that viscous, self-gravitating disks can develop 
GI even when $Q>1$. This is because viscosity compromises rotational
support against self-gravity for long-wavelength disturbances
\citep{lynden-bell74,willerding92,gammie96}.

Now, accretion disks are often modeled with a Navier-Stokes viscosity
to mimic hydrodynamic or magneto-hydrodynamic turbulence
\citep{shakura73}. This involves writing the
kinematic viscosity $\nu$ as  
\begin{align}\label{alpha_param}
  \nu  = \alpha \frac{c_s^2}{\Omega}, %dimensional argument 
\end{align}
where $\alpha$ is the dimensionless viscosity
coefficient. This
parameterization can be modified to include more complex dependencies on the
fluid variables, but Eq. \ref{alpha_param} is the general form.

How well turbulence can be modeled as an effective, Navier-Stokes
viscosity is a separate issue \citep{balbus99}. However, it is 
reasonable to speculate that turbulence frustates rotational 
support. (This is, after all, a physical cause for mass accretion.)

Again, viscosity is absent in the Toomre $Q$ parameter. The above
studies account for viscosity but consider isothermal disks. On the
other hand, proper inclusion of visosity should also account for
viscous heating. We thus extend the study of viscous GI to 
include an energy equation with viscous heating and explicit
cooling. This allows us to consider GI in realistic PPDs.

%effect of viscosity 
%secular instability 
%but adding viscosity means viscous heating should also be included

\subsection{Relation to gravito-turbulent disk fragmentation}
The two effects described above are immediately relevant to
`gravito-turbulence'. This is a steady disk state in which cooling is
balanced by turbulent heating, itself caused by GI. In this case 
the $\alpha$ and $\beta$ parameters are inversely related.

Experimentally, however, it is found that if the imposed $\beta$ is
too small, 
\begin{align}
  \beta < \beta_\mathrm{c},
\end{align}
then the disk fragments. There is considerable debate on the exact
value of $\beta_\mathrm{c}$ and whether or not a critical cooling time can
be defined at all. However, it is generally accepted that \emph{
  steady, gravito-turbulent disks do not exist for sufficiently rapid
cooling}.

We do not attempt to model disk fragmentation from first
principles. Instead, we simply ask: \emph{when does a 
  gravito-turbulent disk become dynamically unstable?} 
%The two destabilizing agents are cooling and viscosity  
We will find the answer to be consistent with the current consensus on
disk fragmentation.

%important to note that this is an experimental results. we refrain
%from using input from numerical simulations in our modeling 
%because we want to explain simulations 
%we will find the regimes where viscous, cooling PPDs are dynamically
%unstbale coincides with where fragmentation occurs 
% we find instability formally possible for long cooling but takes a
% long time

\subsection{Plan of this paper}
%plan of this paper 
%aim to analyse GI in realistic disks with minimal input from numerical
%simulation

This paper is organized as follows.

\input{model3d}
\input{linear3d}
\input{results2d}
\input{results3d}
\input{application} 
\input{summary}

\appendix
\input{appendix3d}

\bibliographystyle{apj}
\bibliography{ref}

\end{document}

### 2015-12-21 paper/results2d.tex
upload paper and codes

\section{Two-dimensional disks with beta cooling}
We begin in the 2D limit with beta cooling  to make connection 
with previous studies. The disk
material assumed to be confined to the midplane so $\delta v_z=0$. 
We make the replacements 
$\rho \to \Sigma$ and $P\to\Pi$, the vertically-integrated
pressure, and set $\Gamma=1$ without loss of generality. 
%The perturbations $\delta\Sigma$, 
%$\delta\Pi$, $\delta v_x$ and $\delta v_y$ have no vertical
%dependence and $\delta v_z=0$. 
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

We obtain from the linearized equations 
an algebraic dispersion relation $s = s(k)$. It is simplest
to write this  
in terms of the dimensionless growth rate $S = s/\Omega$ and
wavenumber $K=kH = k c_{s0}/\Omega$ as
\begin{align}\label{thindisk}
  f(S,K)\equiv AD - BC = 0,   
\end{align}
where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}.

\subsection{Inviscid limit}\label{2d_inviscid}
Here we neglect viscosity by setting 
$\alpha = \alpha_b = 0$. This implies an external,
time-independent heat source must be invoked to balance the imposed
cooling rate in equilibrium ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are assuming 
viscosity is only important for setting up thermal equilibrium, and
can be neglected for the pertubations. This is useful as the resulting
instabilties are only due to self-gravity.  In any case, we are
considering the \emph{sole} effect of cooling on stability.

Eq. \ref{thindisk} simplifies to  
\begin{align}\label{inviscid}
%  &\beta S^3 + S^2 - \beta \left\{ \frac{2|K|}{\qtwo} - \left[2(2-q) +
%    \gamma K^2\right]\right\}S \notag\\ &- \left[\frac{2|K|}{\qtwo} - 2(2-q)\right]
%  = 0. 
  S^2 = \frac{2|K|}{Q} - 2(2-q) - \left(\frac{\theta + \beta \gamma
    S}{1+\beta S}\right)K^2, 
\end{align}
similar to the classic Lin-Shu dispersion relation. The first term on
the right-hand-side represents destabilization by self-gravity; 
the second and third terms represent stabilization by rotation and
pressure, respectively. The imposed cooling term only affects the
pressure response.

Eq. \ref{inviscid} is a cubic equation in $S$. Applying the
Routh-Hurwitz criterion, we find that stability is ensured if 
\begin{align}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. Thus, the question of stability does not depend on
the cooling rate $\beta$. The system becomes more unstable with
decreasing $\theta$.

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p |K| =
0$ and $S = 
S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
General solutions to this cubic for $S_*$ are unwieldy, but simple
results may be obtained in special cases, discussed next. 
%We plot examples of the maximimum growth rate as a function of cooling
%times in Fig. \ref{2d_inviscid}. General expressions for the solution
%of Eq. \ref{inviscid_max} are unwieldy, but simple results may 
%be obtained for special cases, as discussed below.  
%\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm 0cm]{figures/inviscsg}
%  \caption{Maximum growth rate $s$ in the inviscid limit,
%    obtained from Eq. \ref{inviscid}, as a function of the cooling
%    time $\beta$ for $Q/Q_\mathrm{crit}=1,\,2,\,10$, where $\qcrit$ is
%    defined in Eq. \ref{qcrit_def} as 
%    the value of $Q$ such that the disk is marginally stable in the
%    adiabatic limit. 
%    \label{2d_inviscid}}
%\end{figure}

\subsubsection{$\theta = 0$ and $Q=Q_\mathrm{crit}$}  
Beta cooling with $\theta=0$ is typically employed in numerical
simulations of self-gravitating disks \citep{gammie01}. There, 
a steady state corresponds to the imposed cooling entirely balanced by heat generated through 
gravito-turbulence. In the present framework, this gravito-turbulence
is assumed to provide a constant 
background heating ($\mathcal{H}_\mathrm{ext}$) and does
not enter the linearized equations. Cooling, however, is still imposed in the perturbed state. 
From the discussion above, this system is unconditionally unstable for
finite $Q$. The most unstable wavelength decreases with decreasing
$\beta$. The $\beta\to0$ limit corresponds to a pressureless disk.

% it is always
%possible to find a sufficiently small scale such that pressure effects
%are

Specifically, let us consider a disk with 
\begin{align}\label{qcrit_def}
  \qtwo = \frac{1}{\sqrt{2\gamma(2-q)}}\equiv Q_{\mathrm{crit}},
\end{align} 
so that it is marginally stable in the absence of cooling.  
How does introducing finite cooling destabilize the system? 
Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
$\theta=0$, we find 
%In this case the coefficient of 
%$S_*$ in Eq. \ref{inviscid_max} vanish, so that  
\begin{align}\label{sstar}
  S_*^3 = \left(\frac{1}{\gamma Q_\mathrm{crit}^2 \beta}\right). 
\end{align}
The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
increases with decreasing $\beta$. Instability does not require the
cooling time to be below some 
critical value.

Nevertheless, we can define a characteristic cooling
time $\beta_*$ as that which removes pressure support against
self-gravity over the only natural lengthscale in the problem, $H$. We
thus set $|K_*|=1$ and find, for the Keplerian disk,

\begin{align}\label{betastar}
  \beta_* = \frac{1}{\left(\sqrt{\gamma} - 1\right)^{3/2}}. 
\end{align}
%Table \ref{bstar_compare} compares Eq. \ref{betastar} to 
%previously reported values of $\beta_c$ 
%for which fragmentation
%occurs.
It is interesting to note that this $\beta_*$ is similar to previously reported 
values of the cooling times, $\beta_c$, below which fragmentation
occurs on a dynamical timescale. We compare these in Table
\ref{bstar_compare}, showing rough agreement.

%{\bf brief physical interpretation. gravito-turbulent heating is a
%  constant background heating - it does not affect perturbations about
%  the turbulent state.  
%  imposed beta cooling works on all
%  scales. but `gravito-turbulence' only makes sense on scales larger
%  than averaging procedure (probably about $H$). }
%{\bf wild guess: fragmentation had high resolution because internal
%  structure of forming-clumps become resolved. internal turbulence
%  helps collapse. 
%}

\begin{deluxetable}{rrrr}
  \tablecolumns{4}
  \tablecaption{Characteristic cooling times as a function of
    $\gamma$. \label{bstar_compare}
  }
  \tablehead{
    %\colhead{}    &  \multicolumn{3}{c}{Non-shell Stars} &   \colhead{}   &
    %\multicolumn{3}{c}{Shell Stars} \\
    %\cline{2-4} \cline{6-8} \\
    \colhead{$\gamma$}   & \colhead{Eq. \ref{betastar}, $\beta_*$} &
    \colhead{Simulation, $\beta_c$} &  \colhead{Reference}
  }
\startdata
 $7/5$ & 12.75 & 12---13 & \cite{rice05}\\
$1.6$  &  7.33 & 8 & \cite{rice11}\\
$5/3$  &  6.37 & 6---7 & \cite{rice05}\\
$2$    &  3.75 & 3 & \cite{gammie01}
\enddata
\end{deluxetable}

\subsubsection{Passive disk with $\theta = 1$}
Beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.

%nothing interesting here

%\subsubsection{Expected maximum stress}
%{\bf for notes only} 
%A reasonable assumption is that in the non-linear regime 
%there is an associated turbulent viscosity that may be estimated  
%as $\nu \sim l^2s$, where the lengthscale $l\sim 1/k$; which
%translates to $\alpha\sim S/K^2$.  
%
%Using Eq. \ref{pressureless}, we find that 
%\begin{align}\label{max_alpha}
%  \mathrm{max}\left(\frac{S}{K^2}\right) = \frac{3^{3/2}}{2^{7/2}\left(2-q\right)^{3/2}Q^2}
%\end{align}
%in the limit $\beta\to 0$. This maximum is occurs at wavenumber $|K| =
%4Q(2-q)/3$, which is $O(1)$ for typical disk parameters we consider. 
%Eq. \ref{max_alpha} indicates that $\alpha\propto Q^{-2}$, as
%suggested by \cite{lin87}. Note that although
%$\mathrm{max}(S)\to\infty$ as $\beta\to0$ (Eq. \ref{sstar}), 
%the associated stress is bound because the
%most unstable modes occur at the smallest scales. In fact, for
%$Q=Q_\mathrm{crit}$, we find $s(K_*)/K_*^2 \propto \beta $ as
%$\beta\to 0$, so transport by the most unstable mode is expected to be
%unimportant. 
%\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0.8cm]{figures/inviscsg_alpha}
%  \caption{A measure of the
%    associated stress that can be expected in the non-linear
%    regime. The horizontal dotted lines
%    correspond to Eq. \ref{max_alpha}.}
%\end{figure}

\subsection{Viscous disk}\label{2dvisc}
We now consider a viscous disk with 
$\mu=-1,\,\lambda=0$ as discussed in \S\ref{visc_model}. We first
check in Appendix \ref{gammie_check} that our dispersion relation
reduces to that of \cite{gammie96} in the appropriate limit (namely
$\beta\to\infty$).

To see the effect of cooling, we simplify the dispersion relation,
Eq. \ref{thindisk}, by assuming $|\beta S|\ll 1$. Then  
for $|K| \to 0$ we find
\begin{align}\label{gammie_smallk}
%  S \sim \frac{\alpha |K|^3}{(2-q)Q},%derived assuming |beta*S|<<1 
%  S \sim \frac{\alpha K^2\left[2|K|Q^{-1} - \gamma K^2 + 2\alpha q
%    (2-q)(\gamma-1)K^2\right]}{2(2-q)+\gamma K^2 - 2|K|Q^{-1}} %derived
                                %assuming |\beta S|>>1
  S\simeq \frac{\alpha K^2}{2(2-q)}\left(\frac{2|K|}{Q} - \theta
  K^2\right), 
\end{align}
which coincides with \citeauthor{gammie96}'s Eq. 18 for vanishing
wavenumber. For $|K|\to\infty$ we find
\begin{align}\label{gammie_bigk}
  %  S \sim \frac{2}{Q|K|}\left(\frac{4}{3}\alpha + \gamma\beta\right)^{-1}.
  S \simeq\left(\frac{2}{Q|K|} - \theta\right)\left(\frac{4}{3}\alpha + 
  \alpha_b + \gamma\beta\right)^{-1}.
\end{align}

For $\theta\ll1$ a rough measure of the maximum growth rate can be obtained by
equating Eq. \ref{gammie_smallk} and \ref{gammie_bigk}\footnote{If
  $\theta$ is not small and/or $Q$  is large then one may just use Eq. \ref{gammie_smallk}
  to maximize $S$ over $K$, see the $\theta=0.3,\beta=100$ curve in the bottom
  panel of Fig. \ref{gammie_rate_plot}}.  
This exercise yields 
\begin{align}\label{gammie_maxrate_simple}
  %S_* \simeq
  %\frac{1}{Q}\left(\frac{\alpha}{2-q}\right)^{1/4}\left(\frac{6}{4\alpha
%+ 3\gamma\beta}\right)^{3/4}. 
  S_*\simeq \frac{
    6^{3/4}\left[\alpha\left(4\alpha +
      3\alpha_b + 3\gamma\beta\right)\right]^{1/4} - 3\theta
    Q(2-q)^{1/4}}{Q\left(4\alpha + 3\alpha_b +
    3\gamma\beta\right)(2-q)^{1/4}}. 
\end{align} 
%gravito-viscous instability

As a numerical example, we consider a model with $\alpha_b=0$ and
$\alpha=\alpha(\beta)$ given by thermal equilibrium,
Eq. \ref{alpha_beta_relation}. Furthermore, we set
\begin{align}
  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
\end{align}
to mimic a basic state maintained by gravito-turbulence where one can 
expect the dimensionless stress $\alpha \sim Q^{-2}$ \citep{lin87}.

Fig. \ref{gammie_rate_plot} show growth rates as a function of the
wavenumber $k$ obtained from the dispersion relation
Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
well-captured by Eq. \ref{gammie_smallk} and
\ref{gammie_bigk}. Comparing the two panels shows that increasing the
irradiation level ($\theta$) surpress  
small-scale perturbations.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.4cm
    0.0cm]{figures/viscsg_modes}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.4cm
    0.0cm]{figures/viscsg_modes_theta0d3}
  \caption{Growth rates for the 2D viscous problem as a function of
    the radial wavenumber $k$ for a range of cooling times
    $\beta$. The dashed and dotted lines correspond to asympototic
    behaviors for small and large $k$, respectively, computed from
    Eq. \ref{gammie_smallk} and \ref{gammie_bigk}. Top: without
    irradiation, bottom: partially irradiated disk. 
    \label{gammie_rate_plot}}
\end{figure}

Fig. \ref{gammie_maxrate_plot} show the maximum growth rate as a
function of the cooling time $\beta$ for the case $\theta=0$. 
%We set $\alpha_b=\theta=0$ and 
%consider three cases (top to bottom panels): 
%\begin{enumerate}[(a)]
%  \item $\alpha$, $\beta$ and $Q$ are independent; \label{casea}
%  \item $\alpha=\alpha(\beta)$ from thermal equlibrium
%    (Eq. \ref{alpha_beta_relation}) and $Q$ specified separately; \label{caseb}
%  \item $\alpha(\beta)$ from Eq. \ref{alpha_beta_relation} and
%    assuming $Q = Q_\mathrm{crit}/\sqrt{\alpha}$. This mimics a
%    basic state maintained by 
%    gravito-turbulence where the viscosity coefficient $\alpha \sim
%    Q^{-2}$ \citep{lin87}.
%    \label{gvisc}
%\end{enumerate} 
There is good agreement between numerical results and
Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. The assumption
of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
explains the mismatch there.

Eq. \ref{gammie_maxrate_simple} gives the limiting behavior (for
negligible $\theta$) as 
\begin{align*}
  S_*\propto \begin{cases}
    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} &  \beta
    \gg \alpha, \\
    \frac{1}{Q\sqrt{\alpha}}  \propto \mathrm{const.} & \beta \ll \alpha,
  \end{cases}
\end{align*}
where the second proportionality applies Eq. \ref{alpha_beta_relation}
and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
the growth timescales are long ($>10$ orbits) for $\beta \gtrsim
20$. This region is marked by the vertical dashed-dotted lines in
Fig. \ref{gammie_maxrate_plot}.

As $\beta\to 0$ the required viscous heating to set up an equilibrium
state diverges ($\alpha\to\infty$). 
In the present model the increased
viscous damping, which is stabilizing, is balanced by the decreasing
(and hence destablizing) $Q$ required to achieve large $\alpha$
(Eq. \ref{Qalpha}).

Physically, however, we expect $\alpha < 1$. In fact, numerical
simulations of self-gravitating disks show there is a maximum $\alpha$
($\sim 0.06$) that can be sustained before fragmentation
\citep{rice11}. We can interpret this result in our linear framework.
%We can use our case (\ref{gvisc}) to interpret this
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
shows that such a system is dynamically unstable with 
growth rate $s = O(\Omega)$.  Thus we do not expect rapidly-cooled,
self-gravitating steady disks to exist for extended periods of
time. This is consistent with previous numerical simulations performed
by \cite{lodato05}.

%Case (\ref{gvisc}) is consistent with

\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \caption{Growth rates maximized over wavenumber as a function of cooling time for the
    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
    line marks the region with $\alpha > 1$, and the vertical
    dashed-dot line marks the region with growth timescales longer
    than dynamical. 
    \label{gammie_maxrate_plot}}
\end{figure}

### 2015-12-21 paper/results3d.tex
upload paper and codes

\section{Three-dimensional disks with beta cooling}\label{3ddisk}
We now extend the previous results to 3D disks with vertical
structure. This is expected to weaken gravitational instabilities by
`diluting' self-gravity along the vertical direction. It is possible to incorportate
this effect in the previous 2D framework, but this introduces
additional parameters. We discuss this in Appendix \ref{3dcorr}. 
It is more direct to solve the 3D eigenvalue problem to avoid such 
uncertainties.

{\bf note: fast beta-cooling in 3D destablizes vertical modes (enables
convection)}

%take k>0 wlog
%consider Gamma=1
%height s.t. rho(zmax) = 0.05 rho(0) avoid very low densities 
\subsection{Numerical procedure}
We use a pseudo-spectral method to solve the set of ordinary
differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
  \delta v_x, \delta 
  v_y]$ in even Chebyshev polynomials,
\begin{align}
  \bm{U}(z) = \sum_{j=1}^N \bm{a}_jT_{2(j-1)}(z/\zmax), 
\end{align}
and the vertical velocity in odd Chebyshev polynomials,
\begin{align}
  \delta v_z(z) = \sum_{j=1}^N b_jT_{2j-1}(z/\zmax),   
\end{align}
where $\bm{a}_j$ and $b_j$ are the spectral coefficients. 
The basis functions are chosen to satisfy a reflecting boundary
condition at the midplane, %GI has even symmetry across midplane 
\begin{align}
  \bm{U}^\prime(0) = \bm{0}, \quad \delta v_z(0) = 0.
\end{align}
%At the upper disk boundary we also apply reflection,
We explicitly apply a reflecting upper disk boundary, 
\begin{align}
  \delta v_z(\zmax) = \delta v_x^\prime(\zmax) = \delta
  v_y^\prime(\zmax) = 0,
\end{align}
and the potential satisfies
\begin{align}
  \delta \Phi^\prime (\zmax) + k \delta\Phi(\zmax) = 0, 
\end{align}
as derived by \cite{goldreich65a} and used in similar studies  
\citep{kim12,lin14c}. %reflection is standard in direct simulations

We discretize the equations, including upper disk boundary conditions,
over the $N$ positive abscissae of the extrema of $T_{2N-1}$. This
procedure converts the differential equations into a generalized
eigenvalue problem, for which we use the standard matrix package
LAPACK to solve.

In the following examples we consider a vertically isothermal disk
($\Gamma=1$) with a numerical surface such 
that $\rho(\zmax)=0.05\rho_0$, and use $N=65$.

%in practice eliminate density with poisson
\subsection{Inviscid disk}

Fig. \ref{3d_inviscid} show growth rates and the most unstable wavenumber
obtained for a 3D inviscid disk without irradiation. We also plot 2D
results with the 3D correction as described in Appendix
\ref{3dcorr}. While the chosen value of $H_\mathrm{sg}$ results in a close
match between 2D and 3D growth rates, the 3D wavenumber is somewhat
smaller. This reflects self-gravity being weakened in the verical
direction: a larger horizontal scale is required to achieve the same
strength of self-gravity (and hence growth rate) as the 2D case.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.23cm
    0.0cm]{figures/3d_inviscid_rates}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.23cm
    0.88cm]{figures/3d_inviscid_kmax}
  \caption{Growth rates (top) and optimal wavenumber (bottom) obtained
    from the 3D eigenvalue problem (solid line). Asterisks are
    corresponding values from the 2D dispersion relation
    (Eq. \ref{thindisk}) but with a modified gravitational constant
    as described in Appendix \ref{3dcorr}. \label{3d_inviscid}}
\end{figure}

\subsection{Viscous disk} %growth rates, kmax, eigenfunction
For the 3D viscous problem we use the same set up as that in 2D
(\S\ref{2dvisc}), but with 
\begin{align*}
  \qvert = \frac{Q_\mathrm{3D,crit}}{\sqrt{\alpha}}, 
\end{align*}
where $Q_\mathrm{3D,crit}\simeq0.36$ is the 3D equivalent to the 2D
critical value, $Q_\mathrm{crit}$.

Fig. \ref{3d_visc} show growth rates, maximized over $k$, as a
function of the cooling time $\beta$. We also plot 2D results with 3D
corrections. Softening the self-gravity in 2D captures the correct 
qualitative behavior as in the full 3D case, but it is clear that a
single, constant value of $H_\mathrm{sg}$ cannot re-produce 3D growth 
rates for all $\beta$. 
This suggests that the exact value of
$H_\mathrm{sg}$ is problem-dependent (cooling and viscosity), although taking
$H_\mathrm{sg}\sim O(H)$ should give the correct 3D growth rate within
a factor of two. In this respect it is easier to
directly solve the 3D problem.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0.23cm
    0.0cm]{figures/growth_visc3d}
  \caption{Growth rates from the viscous 3D eigenvalue problem (black solid
    line). The other curves are obtained from the 2D dispersion
    relation (Eq. \ref{thindisk}) but with a modified gravitational
    constant as described in Appendix \ref{3dcorr}. \label{3d_visc}}
\end{figure}

### 2015-12-21 paper/summary.tex
upload paper and codes

\section{Summary and discussion}
In this paper, we study the implications on the stability of
turbulent, cooling accretion disks if said turbulence is modeled as an
effective, Navier-Stokes viscosity. This viscosity provides a 
background heating and may also be active in the perturbed 
state. We pay particular attention to 
protoplanetary disks subject to `gravito-turbulence'. 
Our framework provide an alternative interpretation of existing
results on PPD fragmentation established from direct numerical
simulations. 
{\bf cooling removes pressure support, viscosity removes rotational support}
{\bf our conclusions do require input of emperical resuls from numerical
sims}

%generalized dispersion relation (inviscid)
%no magic beta - connection with stochastic

We derive the disperion relation for local axisymmetric density waves 
subject to external heating and cooling (Eq. \ref{inviscid}), but
without viscous effects on the waves. We find
the sufficient condition for stability 
is \emph{independent} of the cooling rate. Long cooling times can
formally lead to instability, but the growth timescales may be
uninterestingly long. We obtain an analytic expression for the
dimensionless cooling time, $\beta_*$, required to remove pressure
support against self-gravity on radial lengthscales comparable to the 
disk scale height. We find $\beta_*$ reproduces the fragmentation
boundary found in previous numerical simulations (Table
\ref{bstar_compare}).

%viscous GI ad connection with frag

We then include viscosity in the linear problem and generalize the
`viscous gravitational instability', previously discussed in 
simpler disk models \citep{lynden-bell74,gammie96,hunter83}, to include
heating/cooling and three-dimensionality. Our generalized framework is
readily applied to realistic protoplanetary disk models. For
gravito-turbulent PPDs, we find viscous GI grows on orbital time-scales
beyond $\sim 100\mathrm{AU}$. This means the assumed gravito-turbulent
steady state should not exist beyond dynamical time-scales in the
outer parts of PPDs.

%viscosity > 0.1, beta < 3 - 4

Our results bear striking similarities with the current consensus on
PPD fragmentation. It is generally accepted that PPDs can fragment
beyond a few tens of $\mathrm{AU}$. This conclusion results from
applying a Toomre stability condition (e.g. $Q\lesssim 1.5$) and a
cooling time or viscosity criteria (e.g. $\beta\lesssim 3$ or
$\alpha\gtrsim 0.1$), the latter of which is
based on numerical simulations. We find these
emperical conditions essentially translates to viscous GI on dynamical
time-scales. This suggests that gravito-turbulent
PPDs fragment because the associated viscous stresses further
destabilize the disk against self-gravity.

%ultimately: why do disks fragment? viscosity may play a role

% also conclude this region of

### 2015-12-21 codes/2d/viscsg.f90

@@ -1,0 +1,412 @@
+module global
+  implicit none
+  character*5 :: method 
+  real*8 :: bigQ, k, bcool, gmma, smallq, mu, lambda, alpha, eta 
+  real*8 :: Htrue, bvisc, vbulk, tirr, theta, bigQ_marg2d, bigQ_marg3d
+end module global
+
+program viscsg
+  use global
+  implicit none 
+  logical :: gsoft, refine, filter, gvisc 
+  integer, parameter :: order = 4
+  integer :: nk, nb, i, j, loc(1), m, iflag
+  real*8 :: tol = 1d-15
+  real*8 :: bigQ_fix, fixalpha, alim, grav3d 
+  real*8 :: kmin, kmax, bmin, bmax, dk, dlogb, btrans
+  real*8 :: c0, c1, c2, c3, c4, growth(order), freq(order), bigA, bigB, bigC, bigD, bigF
+  real*8 :: x(2), fvec(2), err(order)
+  complex*16 :: coeff(order+1), roots(order), sfreq, det 
+  complex*16, allocatable :: rate_fixb(:)
+  real*8, allocatable :: baxis(:), kaxis(:), max_rate(:), max_freq(:),  max_k(:), alpha_arr(:), error(:), error_fixb(:), Q_arr(:) 
+  real*8, external :: zmag, gviscosity 
+  external :: fcn 
+  namelist /params/ bigQ_fix, gsoft, Htrue, gmma, smallq, mu, lambda, fixalpha, gvisc, vbulk, tirr, refine, filter, method
+  namelist /grid/ kmin, kmax, nk, bmin, bmax, nb
+
+  !read input parameters.
+  open(7, file="input")
+  read(7, nml=params)
+  read(7, nml=grid)
+  close(7)
+
+  if(gsoft .eqv. .true.) then 
+     print*, 'soften self-grav due to 3D'
+     grav3d = Htrue
+  else
+     grav3d = 0d0 
+  endif
+
+  !input bigQ_fix is initial Q in units of the marginally stable value
+     k = -(1d0/6d0) + gmma/( &
+          3d0*2d0**(2d0/3d0)*(432d0*gmma**2 - 2d0*gmma**3 - 216d0*gmma**2*smallq + &
[…]

### 2015-12-21 codes/3d/viscsg3d.f90

@@ -1,0 +1,804 @@
+module global
+  logical :: maximize, no_decay, no_oscil, gvisc 
+  character*4 :: vbc 
+  integer, parameter :: nvar = 5
+  integer :: nz, bignz, nb, nk, lmax, nzout 
+  real*8, parameter :: omegaz=1d0, pi = 2d0*acos(0d0)
+  real*8 :: Q3d, Q2d, smallq, gmma, bgmma, zmax, bigH, mu, lambda, bmin, bmax, kmin, kmax  
+  real*8 :: fixalpha, tirr, bulkv 
+  real*8, allocatable :: bvals(:), kvals(:) 
+  real*8, allocatable :: zaxis(:), dens(:), dlogdens(:), csq(:), dcsq(:)
+  real*8, allocatable :: alpha2d(:,:), dalpha2d(:,:), Qarr(:) 
+  real*8, allocatable :: T(:,:), Tp(:,:), Tpp(:,:)
+  real*8, allocatable :: T_odd(:,:), Tp_odd(:,:), Tpp_odd(:,:)
+  real*8, allocatable :: growth(:,:), growth_fixb(:), freq(:,:)
+  complex*16, parameter :: ii = (0d0, 1d0) 
+  complex*16, allocatable :: drho(:), dpres(:), dpot(:), vx(:), vy(:), vz(:), invT(:,:)
+  complex*16, allocatable :: bigLmatrix(:,:), bigRmatrix(:,:), eigen_vec(:)
+  complex*16, allocatable :: L11(:,:), L12(:,:), L13(:,:), L14(:,:), L15(:,:) 
+  complex*16, allocatable :: L21(:,:), L22(:,:), L23(:,:), L24(:,:), L25(:,:) 
+  complex*16, allocatable :: L31(:,:), L32(:,:), L33(:,:), L34(:,:), L35(:,:) 
+  complex*16, allocatable :: L41(:,:), L42(:,:), L43(:,:), L44(:,:), L45(:,:) 
+  complex*16, allocatable :: L51(:,:), L52(:,:), L53(:,:), L54(:,:), L55(:,:) 
+  complex*16, allocatable :: R11(:,:), R12(:,:), R13(:,:), R14(:,:), R15(:,:) 
+  complex*16, allocatable :: R21(:,:), R22(:,:), R23(:,:), R24(:,:), R25(:,:)
+  complex*16, allocatable :: R31(:,:), R32(:,:), R33(:,:), R34(:,:), R35(:,:)
[…]

### 2015-12-21 codes/rad2d/viscsg.f90

@@ -1,0 +1,599 @@
+module global
+  implicit none
+  character*5 :: method 
+  real*8, parameter :: pi = 2d0*acos(0d0)
+  real*8, parameter :: stefan = 5.67d-5, bigG = 6.67d-8, mstar = 1.99d33, au=1.50d13, kboltz = 1.38d-16, mh=1.66d-24
+  real*8, parameter :: mean_mu = 2.3, rstar = kboltz/mh, opacity_b2 = 5d-4, opacity_b0=0.24, year = 3.16d7 
+  real*8 :: fixQ, opscale, k, bcool, gmma, smallq, mu, lambda, alpha, eta, alpha_m, opacity0, opacity_law   
+  real*8 :: Htrue, bvisc, vbulk, tirr, theta, fixalpha, bigQ, mdot, radius 
+end module global
+
+program viscsg
+  use global
+  implicit none 
+  logical :: gsoft, refine, filter, no_oscil
+  integer, parameter :: order = 4
+  integer :: nk, nr, i, j, loc(1), m, iflag
+  real*8, parameter :: tol=1d-15
+  real*8, parameter :: osc_lim = 1d-9
+  real*8 :: grav3d, bigQ_fix
+  real*8 :: kmin, kmax, rmin, rmax, dk, dlogr, surf, temp  
+  real*8 :: c0, c1, c2, c3, c4, growth(order), freq(order), bigA, bigB, bigC, bigD, bigF
+  real*8 :: x(2), fvec(2), err(order)
+  complex*16 :: coeff(order+1), roots(order), sfreq, det 
+  complex*16, allocatable :: rate_fixr(:)
+  real*8, allocatable :: raxis(:), kaxis(:), max_rate(:), max_freq(:),  max_k(:), alpha_arr(:), theta_arr(:), Q_arr(:), error(:), error_fixr(:), bcool_arr(:)
+  real*8, allocatable :: temp_arr(:), surf_arr(:)  
+  real*8, external :: zmag, toomreQ, optical_depth 
+  external :: fcn
+  namelist /params/ fixQ, mdot, opscale, opacity_law, gsoft, Htrue, gmma, smallq, mu, lambda, fixalpha, alpha_m, vbulk, tirr, refine, filter, no_oscil, method
+  namelist /grid/ kmin, kmax, nk, rmin, rmax, nr
+
+  !read input parameters.
+  open(7, file="input")
+  read(7, nml=params)
[…]

### 2015-12-21 codes/rad3d/basic2d.f90

@@ -1,0 +1,179 @@
+real*8 function bigOmega(rad)
+  use global
+  implicit none
+  real*8, intent(in) :: rad
+
+  bigOmega = sqrt(bigG*mstar/au**3d0)*rad**(-3d0/2d0)
+end function bigOmega
+
+real*8 function cs2(tmp)
+  use global
+  implicit none
+  real*8,intent(in) :: tmp
+
+  cs2 = rstar*tmp/mean_mu
+end function cs2
+
+real*8 function T_Q(rad, surface_density)
+  use global
+  implicit none
+  real*8, intent(in) :: rad, surface_density
+  real*8 :: Omega
+  real*8, external :: bigOmega
+
+  Omega = bigOmega(rad)
+  T_Q = (mean_mu/rstar)*(pi*bigG*surface_density*fixQ/Omega)**2d0
+end function T_Q
+
+real*8 function FJ_gt(rad, surface_density)
+  use global
+  implicit none
+  real*8, intent(in) :: rad, surface_density
+  real*8 :: Omega, tauQ, ftauQ, TQ
+  real*8, external :: bigOmega, T_Q, optical_depth, ftau
+
+  Omega = bigOmega(rad)
+  TQ   = T_Q(rad, surface_density)
+  tauQ = optical_depth(surface_density,TQ)
+  ftauQ= ftau(tauQ)
+
+  FJ_gt = 8d0*pi*rad*rad*au*au/(3d0*Omega)
+  FJ_gt = FJ_gt*stefan*(TQ**4d0-Tirr**4d0)/ftauQ
+end function FJ_gt
+
+real*8 function FJ_m(rad, surface_density, tmp)
+  use global
+  implicit none
+  real*8, intent(in) :: rad, surface_density, tmp
+  real*8, external :: cs2
+  
+  FJ_m = 3d0*pi*alpha_m*cs2(tmp)*surface_density*rad*rad*au*au
+end function FJ_m
+
+real*8 function opacity(tmp)
+  use global
+  implicit none
+  real*8, intent(in) :: tmp
+  
+  opacity = opacity0*opscale*tmp**opacity_law
+end function opacity
+
+real*8 function optical_depth(surface_density, tmp)
+  use global
+  implicit none
+  real*8, intent(in) :: tmp, surface_density
+  real*8, external :: opacity
[…]

### 2015-12-21 codes/rad3d/basic3d.f90

@@ -1,0 +1,184 @@
+subroutine get_thickness_and_Q3d
+  use global
+  implicit none
+  integer, parameter :: n = 2
+  integer, parameter :: lwa = n*(3*n+13)
+  integer            :: info 
+  real*8,  parameter :: guess = 1d0, tol=1d-15
+  real*8             :: x(n), fvec(n), wa(lwa)
+  external           :: fcn 
+  
+  x(1) = guess
+  x(2) = Q2d/4d0
+
+  call hybrd1(fcn,n,x,fvec,tol,info,wa,lwa)
+  
+  bigH = x(1)
+  Q3d  = x(2) 
+end subroutine get_thickness_and_Q3d
+
+subroutine fcn(n,x,fvec,iflag)
+  use global
+  implicit none
+  integer, parameter :: neq = 4, itol = 1, itask=1, iopt=0, MF = 10 
+  integer, parameter :: lrw =2*(20 + 16*NEQ), liw = 60
+  integer            :: n, iflag, istate, iwork(liw), IPAR
+  real*8, parameter  :: rtol = 1d-15, atol = 1d-15
+  real*8             :: x(n), fvec(n)
+  real*8             :: Y(neq), Z, ZOUT, rwork(lrw), RPAR 
+  real*8             :: final_density
+  external           :: F, JAC
+ 
+  Z    = 0d0 !start from midplane
+  ZOUT = x(1)!guess end point 
+  Q3d  = x(2)!3D Q parameter  
+
+  Y(1) = 1d0 !normalized pressure at z=0
+  Y(2) = 1d0 !normalized temperature at z=0
+  Y(3) = 0d0 !flux at the midplane
+  Y(4) = 0d0 !self-gravity force at midplane 
+  
+  istate = 1 
+  call DVODE (F, NEQ, Y, Z, ZOUT, ITOL, RTOL, ATOL, ITASK, &
+       ISTATE, IOPT, RWORK, LRW, IWORK, LIW, JAC, MF, &
+       RPAR, IPAR)
+
+  !test final density and Q2d 
+  final_density = Y(1)/Y(2)
+  fvec(1) = final_density - (1d0 - zmax)
+  fvec(2) = Y(4) - 2d0/Q2d 
+end subroutine fcn
+
+subroutine F (NEQ, Z, Y, YDOT, RPAR, IPAR)
+  use global
+  implicit none
[…]

### 2015-12-21 codes/rad3d/viscsg3d.f90

@@ -1,0 +1,787 @@
+module global
+  logical :: maximize, no_decay, no_oscil, betacool 
+  character*4 :: vbc, tbc  
+  integer, parameter :: nvar = 6
+  integer :: nz, bignz, nr, nk, lmax,  nzout 
+  real*8, parameter :: omegaz=1d0, pi = 2d0*acos(0d0)
+  real*8, parameter :: stefan = 5.67d-5, bigG = 6.67d-8, mstar = 1.99d33, au=1.50d13, kboltz = 1.38d-16, mh=1.66d-24
+  real*8, parameter :: mean_mu = 2.3, rstar = kboltz/mh, sig0=2200d0, T0=120d0, opacity_b2 = 5d-4, opacity_b0 = 0.24, year = 3.16d7
+  real*8 :: fixQ, mdot, opscale, alpha_m, tirr, rmin, rmax, radius, alpha, opacity_law, opacity0, Tmid
+  real*8, allocatable :: Q2d_arr(:), alpha_arr(:), raxis(:), temp_radial(:), bcool2d(:), theta2d(:), tau2d(:)
+
+  real*8 :: Q3d, Q2d, smallq, gmma, bigH, mu, lambda, kmin, kmax, bulkv, zmax
+  real*8, allocatable :: kvals(:) 
+  real*8, allocatable :: zaxis(:), dens(:), dlogdens(:), pres(:), dlogpres(:), temp(:), dlogtemp(:), d2logtemp(:), sg(:), dsg(:), radflux(:)
+  
+  real*8, allocatable :: T(:,:), Tp(:,:), Tpp(:,:)
+  real*8, allocatable :: T_odd(:,:), Tp_odd(:,:), Tpp_odd(:,:)
+  real*8, allocatable :: growth(:,:), growth_fixr(:), freq(:,:)
+  complex*16, parameter :: ii = (0d0, 1d0) 
+  complex*16, allocatable :: drho(:), dpres(:), dpot(:), dtemp(:), vx(:), vy(:), vz(:) 
+  complex*16, allocatable :: bigLmatrix(:,:), bigRmatrix(:,:), eigen_vec(:), eigen_fixr(:)
+  complex*16, allocatable :: L11(:,:), L12(:,:), L13(:,:), L14(:,:), L15(:,:), L16(:,:)
[…]

### 2015-12-21 codes/rad3d/vode.f

@@ -1,0 +1,4347 @@
+*DECK DVODE
+      SUBROUTINE DVODE (F, NEQ, Y, T, TOUT, ITOL, RTOL, ATOL, ITASK,
+     1            ISTATE, IOPT, RWORK, LRW, IWORK, LIW, JAC, MF,
+     2            RPAR, IPAR)
+      EXTERNAL F, JAC
+      DOUBLE PRECISION Y, T, TOUT, RTOL, ATOL, RWORK, RPAR
+      INTEGER NEQ, ITOL, ITASK, ISTATE, IOPT, LRW, IWORK, LIW,
+     1        MF, IPAR
+      DIMENSION Y(*), RTOL(*), ATOL(*), RWORK(LRW), IWORK(LIW),
+     1          RPAR(*), IPAR(*)
+C-----------------------------------------------------------------------
+C DVODE: Variable-coefficient Ordinary Differential Equation solver,
+C with fixed-leading-coefficient implementation.
+C This version is in double precision.
+C
+C DVODE solves the initial value problem for stiff or nonstiff
+C systems of first order ODEs,
+C     dy/dt = f(t,y) ,  or, in component form,
+C     dy(i)/dt = f(i) = f(i,t,y(1),y(2),...,y(NEQ)) (i = 1,...,NEQ).
+C DVODE is a package based on the EPISODE and EPISODEB packages, and
+C on the ODEPACK user interface standard, with minor modifications.
+C-----------------------------------------------------------------------
+C Authors:
+C               Peter N. Brown and Alan C. Hindmarsh
+C               Center for Applied Scientific Computing, L-561
+C               Lawrence Livermore National Laboratory
+C               Livermore, CA 94551
+C and
+C               George D. Byrne
+C               Illinois Institute of Technology
+C               Chicago, IL 60616
+C-----------------------------------------------------------------------
+C References:
+C
+C 1. P. N. Brown, G. D. Byrne, and A. C. Hindmarsh, "VODE: A Variable
+C    Coefficient ODE Solver," SIAM J. Sci. Stat. Comput., 10 (1989),
+C    pp. 1038-1051.  Also, LLNL Report UCRL-98412, June 1988.
+C 2. G. D. Byrne and A. C. Hindmarsh, "A Polyalgorithm for the
+C    Numerical Solution of Ordinary Differential Equations,"
+C    ACM Trans. Math. Software, 1 (1975), pp. 71-96.
+C 3. A. C. Hindmarsh and G. D. Byrne, "EPISODE: An Effective Package
[…]

### 2015-12-21 paper/makepdf.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+latex paper3d.tex
+dvips -o paper3d.ps paper3d.dvi
+ps2pdf -sPAPERSIZE=a4 paper3d.ps
+echo 'done'

### 2015-12-22 paper/appendix3d.tex
re-doing appendix B

@@
 \section{2D viscous GI} \label{gammie_check}
-We recover \cite{gammie96}'s dispersion relation 
-for viscous gravitational instability as follows. First we set
-$\mu=-1, \lambda=0$ to obtain the same viscosity models. Next we
-simplify Eq. \ref{bigA}---\ref{bigF} by assuming $|\beta S |\gg 1$  
+We obtain the dispersion relation for viscous GI described by
+previous authors \citep{lynden-bell74,willerding92,gammie96} as 
+follows. We set $\mu=-1, \lambda=0$ to obtain the same viscosity
+models. Next we consider $|\beta S|\to \infty$, i.e. no cooling. Then
+the condition $AD = BC$ imply
+\begin{align}
+  S^3 + \left(\frac{7}{3}\alpha + \alpha_b\right)K^2S + \left[2(2-q) -
+    \frac{2|K|}{Q} + \gamma K^2 + \alpha K^4 \left(\frac{4}{3}\alpha +
+    \alpha_b\right)\right] + \alpha K^2 \left[\gamma K^2 -
+    \frac{2|K|}{Q} - 2q(2-q)(\gamma-1)\right]=0,
+\end{align}
+which agrees with \citet[Eq. 11]{willerding92} in the isothermal
+limit ($\gamma=1$). The non-isothermal term $\propto (\gamma-1)$
+originates from viscous dissipation which was excluded in the
+aforementioned works. Its effect is to increase 
+(decrease) the maximum wavenumber (wavelength) allowed for viscous
+GI.
+
+An
+
+%We recover \cite{gammie96}'s dispersion relation 
+%for viscous gravitational instability as follows.  Next we
+%simplify Eq. \ref{bigA}---\ref{bigF} by assuming $|\beta S |\gg 1$  
 %This will turn out to be equivalent to $\beta\gg 1$, i.e. the
 %adiabatic limit. 
-and $S\lesssim O(K^2)$ as $|K|\to 0$. Applying these considerations to
-Eq. \ref{thindisk} gives 
-\begin{align}
-  S = \frac{\alpha K^2\left[\frac{2|K|}{Q} - \gamma K^2 +
-      2q(2-q)(\gamma-1)\right]}{2(2-q) + \gamma K^2 - \frac{2K}{Q}}, 
-\end{align}
-which does not explicitly depend on cooling. 
-Setting $\gamma=1$ then gives \citeauthor{gammie96}'s Eq. 18. The above
-is expression thus generalizes \citeauthor{gammie96}'s result to
-non-isothermal disks.
+%and $S\lesssim O(K^2)$ as $|K|\to 0$. Applying these considerations to
+%Eq. \ref{thindisk} gives 
+%\begin{align}
+%  S = \frac{\alpha K^2\left[\frac{2|K|}{Q} - \gamma K^2 +
+%      2q(2-q)(\gamma-1)\right]}{2(2-q) + \gamma K^2 - \frac{2K}{Q}}, 
+%\end{align}
+%which does not explicitly depend on cooling. 
+%Setting $\gamma=1$ then gives \citeauthor{gammie96}'s Eq. 18. The above
+%is expression thus generalizes \citeauthor{gammie96}'s result to
+%non-isothermal disks.
 

### 2015-12-22 paper/paper3d.tex
re-doing appendix B

@@
 support against self-gravity for long-wavelength disturbances
-\citep{lynden-bell74,willerding92,gammie96}.
+\citep{lynden-bell74,willerding92,gammie96}. %also in dusty fluids -
+                                %youdin 11 and the inutsuka paper
 
@@
 viscosity is a separate issue \citep{balbus99}. However, it is 
-reasonable to speculate that turbulence frustates rotational 
+reasonable to assume that turbulence frustates rotational 
 support. (This is, after all, a physical cause for mass accretion.)
@@
 studies account for viscosity but consider isothermal disks. On the
-other hand, proper inclusion of visosity should also account for
-viscous heating. We thus extend the study of viscous GI to 
+other hand, proper inclusion of viscosity should also account for
+viscous heating which, as discussed above, also affect stability. 
+We thus extend the study of viscous GI to 
 include an energy equation with viscous heating and explicit
@@
 
-\subsection{Relation to gravito-turbulent disk fragmentation}
+\subsection{Relation to gravito-turbulent disk fragmentation}%and ring
+                                %formation 
+%can the continous formation/destruction of clumps (as seen in sims)
+%be interpreted as `overstability'? 
+%for overstable modes over-density oscilates 
 The two effects described above are immediately relevant to
@@
 disk fragmentation.
+
+%viscous gi speculated by gammie 01 but now found in those sims. but
+%longer sims of the same system by paardekooper eventually fragment
 

### 2015-12-23 paper/appendix3d.tex
minor appendix B update

@@
 follows. We set $\mu=-1, \lambda=0$ to obtain the same viscosity
-models. Next we consider $|\beta S|\to \infty$, i.e. no cooling. Then
-the condition $AD = BC$ imply
+models. Next we consider $|\beta S|\to \infty$, i.e. no explicit
+cooling on the perturbations. Then the condition $AD = BC$ imply 
 \begin{align}
-  S^3 + \left(\frac{7}{3}\alpha + \alpha_b\right)K^2S + \left[2(2-q) -
+  S^3 + \left(\frac{7}{3}\alpha + \alpha_b\right)K^2S^2 + \left[2(2-q) -
     \frac{2|K|}{Q} + \gamma K^2 + \alpha K^4 \left(\frac{4}{3}\alpha +
-    \alpha_b\right)\right] + \alpha K^2 \left[\gamma K^2 -
+    \alpha_b\right)\right]S + \alpha K^2 \left[\gamma K^2 -
     \frac{2|K|}{Q} - 2q(2-q)(\gamma-1)\right]=0,
@@
 originates from viscous dissipation which was excluded in the
-aforementioned works. Its effect is to increase 
-(decrease) the maximum wavenumber (wavelength) allowed for viscous
-GI.
+aforementioned works. Its effect is to increase the maximum wavenumber 
+for viscous GI. %it acts to stabilize
 
-An
-
+An approximate solution for the growth rate may be obtained for small
+$|K|$ by balancing the last two terms,
+\begin{align}
+  S \simeq \frac{\alpha K^2\left[2|K|/Q- \gamma K^2 + 
+     2q(2-q)(\gamma-1)\right]}{2(2-q)  - 2|K|/Q + \gamma K^2 }, 
+\end{align}
+which coincides with \citeauthor{gammie96}'s Eq. 18 for $\gamma=1$. 
 %We recover \cite{gammie96}'s dispersion relation 

### 2016-01-09 paper/application.tex
update, rough draft

@@
 We briefly consider 3D PPDs subject to explicit radiative diffusion 
-(\S\ref{rad_cool}). At each radius $R$ we obtain the vertical
+(\S\ref{rad_cool}). Given the $\alpha(R)$ and $Q(R)$ profiles obtained
+from the 2D model above, at each radius $R$ we obtain the vertical
 structure from Eq. \ref{vert_eq1}---\ref{thermal_eq}, with 
-Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux, given
-the $\alpha(R)$ and $Q(R)$ profiles obtained from the 2D model
-above. We then solve the 3D eigenvalue problem as in
-\S\ref{3ddisk}, with the additional boundary condition for the temperature 
-perturbation $\delta T^\prime(0) = \delta T(\zmax) = 0$.
+Eq. \ref{rad_cool1}---\ref{rad_cool2} for the radiative flux.
+We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
+additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
+We use the same disk model as in \S\ref{pp2d_example} but without 
+irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
+domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density
 
-We use the same disk model as in \S\ref{pp2d_example} but without
-irradiation ($T_\mathrm{irr}=0$), and consider a smaller domain
-with $R\in[10,100]AU$.
+Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
+wavenumber for $R\in[10,100]$AU. We also plot the corresponding 2D
+results with softened self-gravity. There is a close match for
+$R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
+1$) and both models apply. However, beyond $60$AU where the disk is
+optically thin, radiative diffusion (the 3D curve) does not apply and
+under-estimates the growth rates. Nevertheless, the comparison shows
+that the transition radius $R\sim60$AU can be calculated within the 2D
+framework. This is not surprising since the instability is 
+fundamentally driven by viscous and self-gravity forces in the
+momentum equations, not the energy equation.
+
+%the instability is due to viscosity/gravity in momenta equation so
+%details of energy eq doesn't matter
 
@@
     wavenumber (bottom) for viscous, 
-    self-gravitational modes.  
+    self-gravitating unstable modes. The black curves are obtained
+    from the 3D eigenvalue problem with radiative diffusion, while the
+    green curves are 
     \label{rafikov_growth3d}}

### 2016-01-09 paper/paper3d.tex
update, rough draft

@@
 %\usepackage{paralist}
-\shorttitle{TBD}
+\shorttitle{Non-ideal GI}
 \shortauthors{M-K.\ Lin, K. Kratter}
@@
 %\title{untitled linear problem}
-\title{On the Gravitational stability of gravito-turbulent disks}
+\title{On the Gravitational stability of gravito-turbulent accretion disks}
 \author{Min-Kai Lin\altaffilmark{1} \& Kaitlin M. Kratter}
@@
 \begin{abstract}
-stuff
-%on the gravitational stability of gravito-turbulent disks
+Low mass, self-gravitating accretion disks are able to sustain a  
+quasi-steady, `gravito-turbulent' state provided that thermal 
+time-scales are sufficiently long, or equivalently for sufficiently
+weak turbulent stresses. Otherwise, numerical simulations show that
+disk fragmentation ensues.   
+We explore the physical interpretation of this result as the
+compromise of gravitational stability by non-ideal effects: 
+\begin{inparaenum}[1)] 
+\item cooling, which reduces pressure support; and
+\item viscosity, which reduces rotational support. 
+\end{inparaenum}
+We derive a flexible dispersion relation for linear waves in a 
+self-gravitating, viscous, non-adiabatic disk with internal and
+external heating, and explicit cooling. We 
+consider parameterized cooling functions in two and three dimensions,
+as well as full radiative diffusion in 3D. 
+We apply linear theory to 
+global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
+viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
+destabilize a realistic PPD. However, inclusion of viscosity renders
+the disk beyond $\sim 60$AU dynamically unstable on radial 
+lengthscales a few times the local disk thickness. 
+Thus, a steady gravito-turbulent state cannot persist for more and a
+few orbits in the outer disk, coincident with the expected region for
+fragmentation based on emperical conditions inferred from numerical
+simulations. 
+%This is understood within our framework as the disk's
+%inability to sustain too large of a stress as it undermines
+%rotational support against self-gravity. 
+%This motivates an interpretation of disk 
+%fragmentation as a secondary gravitational instability enabled by 
+%the underlying gravito-turbulent viscosity when it exceeds a critical value.         
 \end{abstract}
+%opacity
 
 \section{Introduction}\label{intro}
-%normal GI and viscous GI. appplicability of simple stability criteria
-%and dispersion relation (toomre and cooling) to complex
-%systems.
-
-%non-existence of gravito-turbulent state for fast cooling
-%because such states are dynamically unstable. secondary gi in two
-%ways: classic, pure gravity; viscosity-assisted.
-
-%provide generalized dispersion relations.
-
-%3D calculations.
-
-%apply linear theory
-%to real disk models.
-
-%notion of max alpha comes from simulations. 
-%max alpha criteria applies in local simulations, so not to do with how
-%fast gi can transport material in a global disk.
-
-%combine independent fragmentation conditions Q less than one and beta
-%less than a few in a single framework: these determine growth rate of
-%viscous GI.
-
-%toomre Q and standard dispersion relation here
-
-%real disks have cooling (removes pressure support)
-%and turbulence (removes rotational support ---  if it's like a
-%viscosity)
-
-%indeed fragmenting disks have cooling and turbulence
-
-%so discussion of GI in real disks should have both
-
 Understanding the gravitational stability (GI) of rotating disks   
 is fundamental to many astrophysical problems. In the context of 
-protoplanetary disks (PPDs), GI has two applications. First, it can
+protoplanetary disks (PPDs), GI has two applications. It can 
 provide gravitational torques to transport angular momentum outwards
-and thus enable mass accretion inwards. Secondly, it may lead to disk
-fragmentation, which has been invoked to explain planet formation at
-large radii. Any realistic study of these phenomena require direct
-numerical simulations. Nevertheless, significant progress can be
-made through analytical understanding of GI.  %physical insight
+and thus enable mass accretion inwards \citep{armitage10,turner14}.   
+GI may also lead to disk fragmentation, which has been invoked to explain planet formation at
+large radii \citep{helled14}. Any realistic study of these phenomena require direct
+numerical simulations. Nevertheless, physical insight can be  
+obtained through analytical understanding of GI.  %physical insight
 
-It is now customary to begin any discussion of disk GI with 
-the \emph{Toomre criterion}, 
+The standard metric for the (inverse) strength of disk self-gravity is the 
+\emph{Toomre parameter}, 
 \begin{align}\label{toomreQ_criterion}
-  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}<1 
+  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
 \end{align}
-for instability \citep{toomre64}. Here, $c_s$ is the sound-speed,
-$\kappa$ is the epicyclic 
-frequency (which equals the rotation frequency $\Omega$ in a Keplerian
-disk), $\Sigma$ is the surface density and $G$ is the
-gravitational constant. This condition describes how self-gravity
-($\Sigma$) destabilizes the disk against 
-rotation ($\kappa$) and pressure ($c_s$). Eq. \ref{toomreQ_criterion}
-formally applies to local, axisymmetric disturbances in a
-two-dimensional (2D, razor-thin), inviscid and adiabatic (including
-isothermal) disk.
+\citep{toomre64}. Here, $c_s$ is the sound-speed,
+$\kappa$ is the epicyclic frequency (which equals the rotation
+frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
+density and $G$ is the gravitational constant. The Toomre $Q$ is a
+measure of the destabilizing effect of self-gravity ($\Sigma$) against 
+the stabilzing effect of rotation ($\kappa$) and pressure
+($c_s$). When $Q\leq 1$ the disk is formally unstable to local,
+axisymmetric disturbances in a two-dimensional (2D, razor-thin),
+inviscid and adiabatic (including isothermal) disk.
 
@@
 
-Despite being a result from ideal disk models, the Toomre $Q$  
-parameter is widely applied to assess GI in more general 
-situtations. This includes non-ideal, realistic disk structures
-obtained with additional physics not encapsulated by $Q$
-\citep[e.g.][]{kimura12}.  It is, however, not obvious that such additions
+The Toomre  parameter is widely applied to assess GI in more general 
+situtations (cf. the ideal disk models from which $Q$ is derived
+from).  This includes realistic disk structures
+obtained with non-ideal physics that are not explicitly contained in
+$Q$ \citep[e.g.][]{kimura12}.  It is, however, not obvious that such additions
 leave the stability condition unaltered. Indeed, we show that GI in more
@@
 The goal of this work is to generalize the analytic treatment of disk 
-GI to allow a self-consistent discussion of GI in 
-realistic PPD models. Specifically, we will examine the effect of
-cooling and viscosity on GI. Their relevance to PPDs are described
-below.
+GI to allow a self-consistent discussion of GI in realistic accretion
+disk models. Specifically, we will examine the effect of cooling and
+viscosity on GI in PPDs.  
+%Their relevance to PPDs are described below.
 
@@
 reduces pressure (thermal) support against self-gravity. The cooling time 
-$\tcool$ is the time-scale over which the disk temperature is relaxed
-to some pre-defined value, which may be zero. It is 
-often written as 
+$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
+to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
+-(T-T_0)/\tcool$. It is often written as 
 \begin{align} 
@@
 \end{align}
-where $\beta$ is the corresponding dimensionless cooling time.
-
+where $\beta$ is the corresponding dimensionless cooling time. This
+type of cooling function, first applied by \cite{gammie01}, is a
+simple parameterization that allows the full range of thermodynamic
+response to be explored. 
 {\bf physical origin of cooling in PPDs}
@@
 The Toomre $Q$ parameter does not include cooling, and thus cannot be
-used to assess GI in realistic disks. Indeed, there exists a separate, 
-\emph{empirical} cooling time criterion which governs stability against disk
-fragmentation (see below), but its precise value is uncertain.
+formally applied to assess GI in realistic disks. Indeed, there exists
+a separate, \emph{empirical} cooling time criterion which governs
+stability against disk fragmentation (see below), but this is not well
+understood.
 
-We shall quantify, using linear stability analyses, how cooling 
-destablize disks. We will find that cooling in fact changes the 
+We shall quantify how cooling removes pressure
+support to enable GI even when $Q>1$. 
+We will find that cooling in fact changes the 
 qualitative nature of GI, in that its mere presence can
@@
 support against self-gravity for long-wavelength disturbances
-\citep{lynden-bell74,willerding92,gammie96}. %also in dusty fluids -
-                                %youdin 11 and the inutsuka paper
+\citep{lynden-bell74,willerding92,gammie96}. A similar effect 
+occurs in dusty fluids where the required frictional forces are 
+provided by dust-gas drag forces \citep{youdin11,takahashi14}.
 
-Now, accretion disks are often modeled with a Navier-Stokes viscosity
+Now, gaseous accretion disks are often modeled with a Navier-Stokes viscosity
 to mimic hydrodynamic or magneto-hydrodynamic turbulence
@@
 reasonable to assume that turbulence frustates rotational 
-support. (This is, after all, a physical cause for mass accretion.)
+support. (This is, after all, the physical cause for turbulent mass
+accretion.)
 
@@
 viscous heating which, as discussed above, also affect stability. 
-We thus extend the study of viscous GI to 
-include an energy equation with viscous heating and explicit
-cooling. This allows us to consider GI in realistic PPDs.
+We thus extend the study of viscous GI to include an energy equation
+with viscous heating and explicit cooling. This allows us to consider
+GI in realistic PPDs.
 
@@
 
-\subsection{Relation to gravito-turbulent disk fragmentation}%and ring
+\subsection{Relevance to gravito-turbulent disk fragmentation}%and ring
                                 %formation 
@@
 balanced by turbulent heating, itself caused by GI. In this case 
-the $\alpha$ and $\beta$ parameters are inversely related.
+the $\alpha$ and $\beta$ parameters are inversely related
+\citep{gammie01}.
 
-Experimentally, however, it is found that if the imposed $\beta$ is
-too small, 
+Experimentally, however, it is found that if $\beta$ ($\alpha$) is
+too small (large), 
 \begin{align}
-  \beta < \beta_\mathrm{c},
+  \beta < \beta_\mathrm{c} \quad
+  \left(\alpha>\alpha_\mathrm{c}\right), 
 \end{align}
-then the disk fragments. There is considerable debate on the exact
-value of $\beta_\mathrm{c}$ and whether or not a critical cooling time can
-be defined at all. However, it is generally accepted that \emph{
-  steady, gravito-turbulent disks do not exist for sufficiently rapid
-cooling}.
+then the disk fragments \citep{gammie01,rice05,rice11}. 
+There is considerable debate on the exact value of $\beta_\mathrm{c}$
+and whether or not a critical cooling time can 
+be defined at all \citep{meru11,meru12,paardekooper12}. However, it is
+generally accepted that \emph{steady, gravito-turbulent disks do not
+  exist for sufficiently rapid cooling}.
 
-We do not attempt to model disk fragmentation from first
-principles. Instead, we simply ask: \emph{when does a 
-  gravito-turbulent disk become dynamically unstable?} 
-%The two destabilizing agents are cooling and viscosity  
-We will find the answer to be consistent with the current consensus on
-disk fragmentation.
+The non-existence of steady gravito-turbulence for rapid cooling, or
+equivalently large viscosity, is intriguing because, as discussed
+above, these effects reduce pressure and rotational support 
+against self-gravity, respectively.
+
+In this paper, we consider the stability of PPDs subject to 
+gravito-turbulence modeled as an effective viscosity. 
+We perform standard stability analyses with explicit inclusion of the
+relevant physics. 
+We will find that the
+regions of a gravito-turbulent PPD that is dynamically unstable,
+agrees with the current consensus on where, and the conditions under
+which, a PPD is expected to fragment. This motivates an alternative 
+interpretation of disk fragmentation as the inability to maintain a
+stable, gravito-turbulent state due to the action of cooling and
+viscosity.
+
+%We do not attempt to model disk fragmentation from first
+%principles. Instead, we simply ask: \emph{when does a 
+%  gravito-turbulent disk become dynamically unstable?} 
+%We will find the answer to be consistent with the current consensus on
+%disk fragmentation.
 
@@
 
-\subsection{Plan of this paper}
+\subsection{Plan}
 %plan of this paper 

### 2016-01-11 paper/linear3d.tex
regular update

@@
 %we may choose $\alpha, \beta$ and $\theta$ separately.  
-\section{Linear problem}
+\section{Linear problem}\label{linear}
 We consider infinitestimal axisymmetric Eulerian perturbations in the
@@
   \delta\ln{\rho\nu} =& (1+\mu)\frac{\delta\rho}{\rho} +
-  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}, 
-\end{align}\label{linear_beta_cool}
-where $^\prime$ denotes $d/dz$. The linearized cooling functions
-$\delta\Lambda$ are given below. 
+  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}, \label{linear_beta_cool}
+\end{align}
+where $^\prime$ denotes $d/dz$, and the 3D self-gravity parameter is
+\begin{align}
+  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
+\end{align}
+The linearized cooling functions $\delta\Lambda$ are given below. 
 Eq. \ref{lin_mass}---\ref{lin_gravity}, supplemented with appropriate

### 2016-01-11 paper/model3d.tex
regular update

@@
-\section{Basic equations}
+\section{Basic equations}\label{basic}
 We consider a three-dimensional (3D), self-gravitating,
@@
 %local vertical oscillation frequency. 
-We assume an ideal has so that the pressure $P$ and thermal energy
+We assume an ideal gas so that the pressure $P$ and thermal energy
 density $E$ are related by  
@@
 $\mathcal{H}_\mathrm{visc}$, $\mathcal{H}_\mathrm{ext}$ and $\Lambda$
-represent viscous heating, time-independent external heating and
-thermal losses, respectively, described below. We normally set
+represent viscous heating, any time-independent heat
+source/sinks, and time-dependent thermal losses, respectively. These
+are described below. We normally set
 $\mathcal{H}_\mathrm{ext}=0$ unless otherwise stated.
@@
 $\alpha=\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
-magnitude of the shear viscosity and is time-independent.  
-The indices $\mu,\,\lambda$ are chosen to model how 
-the viscosity behaves in the perturbed state. Given the uncertain
-properties of gravito-turbulence, we will treat $\mu,\lambda$ as free
-parameters. We adopt the same prescription for the bulk viscosity but
+magnitude of the shear viscosity \emph{in steady state} 
+and is time-independent. The indices $\mu,\,\lambda$ are free
+parameters chosen to model how the viscosity behaves in the perturbed
+state.  
+%Given the uncertain
+%properties of gravito-turbulence, we will treat $\mu,\lambda$ as free
+%parameters. 
+We adopt the same prescription for the bulk viscosity but
 with $\alpha\to\alpha_b$.
 
-Our numerical calculations mostly use $\mu=-1,\lambda=0$ so
-that $\rho\nu$ remains constant, as considered
-in previous gravitational instabilities with 
-viscosity \citep{hunter83,gammie96}. 
-This model conveniently eliminates  
-viscous over-stability \citep{schmit95,latter06}, which is unrelated
-to self-gravity, that would otherwise contaminate our results.
+Our numerical calculations use $\mu=-1,\lambda=0$ so 
+that $\rho\nu$ remains constant, as adopted in previous, pioneering
+studies of viscous GI 
+\citep{lynden-bell74,hunter83,willerding92,gammie96}.  
+This choice also conveniently eliminates viscous over-stability
+\citep{schmit95,latter06}, which is unrelated to self-gravity, that
+would otherwise contaminate our results.
 
-While the choice $\mu=-1,\lambda=0$ may seem artifical, we emphasize
-this is done in order to bring out the relevant physical process ---
-viscosity-assisted gravitational instability. Interestingly, 
-\cite{laughlin96b} had already suggested a $1/\rho$ dependence for the 
-shear viscosity when modeling the evoluting of self-gravitating disks
-as a viscous process.  
-{\bf note paardekooper 2012: alpha goes down as Sigma goes up}
+%the alpha(rho_eq, P_eq) function can be derived with some rigor 
+%but this assumes steady state. don't know how to model
+%time-dependence.
+
+While steady-state viscosity values can be determined analytically
+\citep[e.g.][]{rafikov15}, its time-dependent behavior is not
+well-explored. \cite{laughlin96b} suggested a $\nu \propto 1/\Sigma$
+dependence when modeling the evoluting of self-gravitating disks 
+as a viscous process. \cite{paardekooper12} measured
+an increase in the effective viscosity when their gravito-turbulent
+disk models underwent fragmentation (i.e. as $\Sigma$ increases).  
+These motivate the choice $\mu=-1,\lambda=0$, but we emphasize this is
+done to bring out the physical process of interest --- viscosity-assisted
+gravitational instability.
+
+%{\bf note paardekooper 2012: alpha goes down as Sigma goes up}
 
@@
 
-\subsection{Parameters}
-{\bf temporary, maybe move to intro}
-For the 3D problem it is convenient to parameterize the strength of
-self-gravity via 
-\begin{align}
-  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
-\end{align}
-It is also useful to define the
-two-dimensional (2D) Toomre parameter 
-\begin{align}
-  Q = \frac{c_{s0}\Omega}{\pi G \Sigma}, 
-\end{align}
-where $\Sigma = \int_{-\infty}^\infty\rho dz$ is the surface density.
+%\subsection{Parameters}
+%{\bf temporary, maybe move to intro}
+%For the 3D problem it is convenient to parameterize the strength of
+%self-gravity via 
+%\begin{align}
+%  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
+%\end{align}
+%It is also useful to define the
+%two-dimensional (2D) Toomre parameter 
+%\begin{align}
+%  Q = \frac{c_{s0}\Omega}{\pi G \Sigma}, 
+%\end{align}
+%where $\Sigma = \int_{-\infty}^\infty\rho dz$ is the surface density.
 

### 2016-01-11 paper/paper3d.tex
regular update

@@
 %\usepackage{paralist}
-\shorttitle{Non-ideal GI}
-\shortauthors{M-K.\ Lin, K. Kratter}
+\shorttitle{Non-ideal Gravitational instability}
+\shortauthors{M-K.\ Lin \& K. Kratter}
 
@@
 \begin{abstract}
-Low mass, self-gravitating accretion disks are able to sustain a  
+Low mass, self-gravitating accretion disks can remain in a  
 quasi-steady, `gravito-turbulent' state provided that thermal 
@@
 disk fragmentation ensues. 
-We explore the physical interpretation of this result as the
-compromise of gravitational stability by non-ideal effects: 
+We quantify the interpretation of this result as the
+compromise of gravitational stability by non-ideal physics: 
 \begin{inparaenum}[1)] 
@@
 \end{inparaenum}
-We derive a flexible dispersion relation for linear waves in a 
+To this end, we analyse the gravitational stability of accretion disks
+including these effects explicitly.   
+We derive a versatile dispersion relation for linear waves in a 
 self-gravitating, viscous, non-adiabatic disk with internal and
@@
 the disk beyond $\sim 60$AU dynamically unstable on radial 
-lengthscales a few times the local disk thickness. 
-Thus, a steady gravito-turbulent state cannot persist for more and a
-few orbits in the outer disk, coincident with the expected region for
-fragmentation based on emperical conditions inferred from numerical
-simulations. 
-%This is understood within our framework as the disk's
-%inability to sustain too large of a stress as it undermines
-%rotational support against self-gravity. 
-%This motivates an interpretation of disk 
-%fragmentation as a secondary gravitational instability enabled by 
-%the underlying gravito-turbulent viscosity when it exceeds a critical value.         
+lengthscales a few times the local disk thickness. This is coincident 
+with the numerically-established empirical condition for disk fragmentation based on a
+maximum sustainable stress. We 
+thus suggest that disk fragmentation can be physically attributed,
+at least partially, to the removal of rotational stabilization against
+self-gravity by turbulence.   
 \end{abstract}
-%opacity
+%alternative abstract
 
@@
 
-This paper is organized as follows.
+The basic equations are listed in \S\ref{basic}, along with
+descriptions of the disk 
+equilibra and the two cooling models 
+considered --- parametrized `beta' cooling and radiative diffusion. 
+The linear stability problem is defined in
+\S\ref{linear}. We present results with beta cooling in
+\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks, respectively. In
+\S\ref{2dppd} we apply linear theory to PPDs with realistic cooling
+models, including radiative diffusion in 3D. We find regions in a
+PPD that cannot sustain steady gravito-turbulence to coincide with
+that expected for disk fragmentation, based on previous empirical
+criteria. We summarize our results in \S\ref{summary}, including a
+discussion of how our model may aid the physical understanding of  
+fragmentation in realistic PPDs. 
+% as the inability to sustain gravito-turbulence
+%because the associated stress removes rotational support
 
@@
 
+%\acknowlegement
+%S. Stahler
+
 \appendix

### 2016-01-11 paper/summary.tex
regular update

@@
-\section{Summary and discussion}
+\section{Summary and discussion}\label{summary}
 In this paper, we study the implications on the stability of

### 2016-01-11 paper/model3d.tex
regular update

@@
 The dimensionless coefficient
-$\alpha=\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
-magnitude of the shear viscosity \emph{in steady state} 
-and is time-independent. The indices $\mu,\,\lambda$ are free
-parameters chosen to model how the viscosity behaves in the perturbed
-state.  
+$\alpha=~\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
+magnitude of the shear viscosity \emph{in steady state}. The indices
+$\mu,\,\lambda$ are free parameters chosen to model how the viscosity
+behaves in the perturbed state.  
 %Given the uncertain
@@
 Our numerical calculations use $\mu=-1,\lambda=0$ so 
-that $\rho\nu$ remains constant, as adopted in previous, pioneering
+that $\rho\nu$ is time-indpendent, as adopted in previous 
 studies of viscous GI 
@@
 \citep{schmit95,latter06}, which is unrelated to self-gravity, that
-would otherwise contaminate our results.
-
-%the alpha(rho_eq, P_eq) function can be derived with some rigor 
-%but this assumes steady state. don't know how to model
-%time-dependence.
+would otherwise contaminate our results. %obscure
 
@@
 \citep[e.g.][]{rafikov15}, its time-dependent behavior is not
-well-explored. \cite{laughlin96b} suggested a $\nu \propto 1/\Sigma$
+well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made to
+bring out the physically relevant process --- viscosity-assisted
+gravitational instability. %expect to be generally present, but will
+                           %be obscured by viscous over-stability with
+                           %other indices
+Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
 dependence when modeling the evoluting of self-gravitating disks 
-as a viscous process. \cite{paardekooper12} measured
+as a viscous process. Furthermore, \cite{paardekooper12} measured
 an increase in the effective viscosity when their gravito-turbulent
 disk models underwent fragmentation (i.e. as $\Sigma$ increases).
-These motivate the choice $\mu=-1,\lambda=0$, but we emphasize this is
-done to bring out the physical process of interest --- viscosity-assisted
-gravitational instability.
 
@@
 \subsection{Steady states and cooling models}
-
-We consider equilibrium solutions (omitting the `eq' subscripts
-hereafter for clarity)
+We consider equilibrium solutions (here omitting the `eq' subscripts
+for clarity) 
 \begin{align}
@@
 where the first term represents viscous heating. For the viscous
-problem we may set $\mathcal{H}_\mathrm{ext}=0$. However, if we wish
-to neglect viscosity but include cooling, we must invoke
+problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$. However,
+if we wish to neglect viscosity but include cooling, we must invoke
 $\mathcal{H}_\mathrm{ext}\neq0$ to define an equilibrium state.  
@@
 to $\theta$ as the `irradiation level'.  
+%This cooling function restores 
+%the disk temperature to $\tirr$ over a time-scale $\tcool$.
 
-This type of cooling is widely applied in 2D and 3D
+   Beta cooling is widely applied in 2D and 3D
 numerical simulations of self-gravitating disks \citep{gammie01,
@@
 controlled manner as a function of the cooling time $\beta$.
-
-We remark that for the linear problem, 
-any explicit cooling function $\Lambda=\Lambda(\rho,T)$ can be
-mapped to that obtained from the standard function in Eq.
-\ref{beta_cool}. We do this in \S\ref{2dppd} where we adopt a more
-realistic beta cooling prescription for 2D PPDs.
 
@@
 structure is first obtained from Eq. \ref{vert_eq1}---\ref{vert_eq2}
-(usually numerically), then inserted into
-Eq. \ref{thermal_eq} to infer the required viscosity profile
+then inserted into Eq. \ref{thermal_eq} to infer the required
+viscosity profile,
 
@@
 \end{align}
+
+We remark that for the linear problem  
+any explicit cooling function $\Lambda=\Lambda(\rho,T)$ can be
+mapped to that obtained from the standard function in Eq.
+\ref{beta_cool}. We do this in \S\ref{2dppd} where we adopt a more
+realistic beta cooling prescription for 2D PPDs.
 
@@
 
-While radiative cooling is arguably more realistic than 
-parametrized beta cooling, it complicates our 
-problem because it generally implies a vertically non-isothermal
-equilibrium disk, and increases the order of the linearized equations.
-
 In this case we specify a constant viscosity coefficient $\alpha$ and
@@
 
-%\subsection{Steady states}
-%\subsubsection{Equilibria with beta cooling}   
-%\subsubsection{Equilibria with radiative cooling}
-%If $\mathcal{H}_\mathrm{ext}=0$ then viscosity and cooling
-%cannot be prescribed independently and we have 
-%\begin{align}\label{alpha_beta_relation}
-%  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
-%    q^2}\left(\frac{\rho}{\rho_0}\right)^{\Gamma-1} \left(1 - \theta\right). 
-%\end{align}
-%On the other hand, if a time-independent heat source/sink is invoked
-%($\mathcal{H}_\mathrm{ext}\neq 0$), then we may choose $\alpha, \beta$
-%and $\theta$ separately.
-
-%\subsection{Parameters}
-%{\bf temporary, maybe move to intro}
-%For the 3D problem it is convenient to parameterize the strength of
-%self-gravity via 
-%\begin{align}
-%  Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 
-%\end{align}
-%It is also useful to define the
-%two-dimensional (2D) Toomre parameter 
-%\begin{align}
-%  Q = \frac{c_{s0}\Omega}{\pi G \Sigma}, 
-%\end{align}
-%where $\Sigma = \int_{-\infty}^\infty\rho dz$ is the surface density.
-
-%We are interested in comparing systems with different $\beta$. 
-%However, this gives unphysically large values of $\alpha$ for small
-%values of $\beta$. In order to compare systems with different $\beta$,
-%we adopt the following procedure. We choose parameters  
-%$\theta_\mathrm{min}$ and $\alpha_\mathrm{max}$, and evaluate
-%Eq. \ref{alpha_beta_relation} at the midplane to define a cooling time
-%$\beta_\mathrm{min}$. Then for $\beta\geq\beta_\mathrm{min}$ we use 
-%Eq. \ref{alpha_beta_relation} to set $\alpha$ at fixed
-%$\theta=\theta_\mathrm{min}$; while for $\beta < \beta_\mathrm{min}$
-%we use Eq. \ref{alpha_beta_relation} to set $\theta$ at fixed
-%$\alpha(0)=\alpha_\mathrm{max}$.  That is, we raise the reference
-%temperature (i.e. irradiation levels) when the imposed cooling is too
-%rapid to be sustaind by viscous heating.
+While radiative cooling is arguably more realistic than
+parametrized beta cooling, it generally implies a vertically
+non-isothermal equilibrium disk, and increases the order of the
+linearized equations. It also only applies to optically-thick 
+disks. It is possible to modify the flux function to account for
+optically-thin disks, but this complication is beyond the scope of
+this work.

### 2016-01-11 paper/results2d.tex
regular update

@@
-\section{Two-dimensional disks with beta cooling}
+\section{Two-dimensional disks with beta cooling}\label{result_2d}
 We begin in the 2D limit with beta cooling  to make connection 
@@
 viscosity is only important for setting up thermal equilibrium, and
-can be neglected for the pertubations. This is useful as the resulting
-instabilties are only due to self-gravity.  In any case, we are
-considering the \emph{sole} effect of cooling on stability.
+can be neglected for the pertubations. This eliminates viscous GI and 
+allows us to quantify the \emph{sole} effect of cooling.
 
@@
 are both satisfied. Thus, the question of stability does not depend on
-the cooling rate $\beta$. The system becomes more unstable with
-decreasing $\theta$.
+the cooling rate $\beta$ (provided it is non-zero and finite). The
+system becomes more unstable with decreasing $\theta$.
 
@@
 \subsubsection{$\theta = 0$ and $Q=Q_\mathrm{crit}$}  
-Beta cooling with $\theta=0$ is typically employed in numerical
+Beta cooling with $\theta=0$ is the standard choice for numerical
 simulations of self-gravitating disks \citep{gammie01}. There, 
@@
 not enter the linearized equations. Cooling, however, is still imposed in the perturbed state. 
-From the discussion above, this system is unconditionally unstable for
+For $\theta=0$ the system is unconditionally unstable for
 finite $Q$. The most unstable wavelength decreases with decreasing
-$\beta$. The $\beta\to0$ limit corresponds to a pressureless disk.
+$\beta$. The $\beta\to0$ limit corresponds to a pressureless disk (not
+merely isothermal).
 
@@
 so that it is marginally stable in the absence of cooling.  
-How does introducing finite cooling destabilize the system? 
+How does finite cooling destabilize the disk?  
 Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
@@
 increases with decreasing $\beta$. Instability does not require the
-cooling time to be below some 
-critical value.
+cooling time to be below some critical value.
 
@@
 check in Appendix \ref{gammie_check} that our dispersion relation
-reduces to that of \cite{gammie96} in the appropriate limit (namely
-$\beta\to\infty$).
+reduces previous results in the adiabatic/isothermal limit.
 
-To see the effect of cooling, we simplify the dispersion relation,
+To see the effect of finite cooling, we simplify the dispersion relation,
 Eq. \ref{thindisk}, by assuming $|\beta S|\ll 1$. Then  

### 2016-01-14 paper/model3d.tex
regular update

@@
 \end{align}
-is a reference temperature field parameterized as a constant fraction
+is a reference temperature field, here parameterized as a constant fraction
 $\theta$ of the equilibrium temperature $T_\mathrm{eq}$. We will refer
@@
 %the disk temperature to $\tirr$ over a time-scale $\tcool$.
-
-   Beta cooling is widely applied in 2D and 3D
-numerical simulations of self-gravitating disks \citep{gammie01,
-  rice05,baehr15}. It permits numerical experiments to be carried out in a
-controlled manner as a function of the cooling time $\beta$.
 
@@
 
+ Beta cooling is widely applied in 2D and 3D
+numerical simulations of self-gravitating disks \citep{gammie01,
+  rice05,baehr15}. It permits numerical experiments to be carried out in a
+controlled manner as a function of the cooling time $\beta$. 
 We remark that for the linear problem  
@@
 \end{align}
-and take the constant index $b=2$ as appropriate for {\bf blah},
-but retain the general notation $b$ to keep track of the opacity.   
+and take the constant index $b=2$ as appropriate for the cold outer regions
+of a realistic PPD \citep{bell94}, but retain the general notation $b$
+to keep track of the opacity.    
 %We use radiative cooling when considering 3D PPDs in \S.
 
-In this case we specify a constant viscosity coefficient $\alpha$ and
+In this case, we specify a constant viscosity coefficient $\alpha$ and
 solve Eq. \ref{vert_eq1}---\ref{thermal_eq}, together with
 Eq. \ref{rad_cool1}---\ref{rad_cool2}, as a fourth order system of
-ordinary differential equations to obtain $P(z)$, $T(z)$, and hence
+ordinary differential equations to obtain equilibrium profiles $P(z)$, $T(z)$, and hence
 $\rho(z)$.
@@
 non-isothermal equilibrium disk, and increases the order of the
-linearized equations. It also only applies to optically-thick 
-disks. It is possible to modify the flux function to account for
-optically-thin disks, but this complication is beyond the scope of
+linearized equations. It formally applies to optically-thick 
+disks, but it is possible to modify the flux function to account
+for optically-thin disks. However, this complication is beyond the scope of
 this work.

### 2016-01-14 paper/paper3d.tex
regular update

@@
 including these effects explicitly.   
-We derive a versatile dispersion relation for linear waves in a 
+We derive a general dispersion relation for linear waves in a 
 self-gravitating, viscous, non-adiabatic disk with internal and
@@
 lengthscales a few times the local disk thickness. This is coincident 
-with the numerically-established empirical condition for disk fragmentation based on a
+with the empirical condition for disk fragmentation based on a
 maximum sustainable stress. We 
@@
 \section{Introduction}\label{intro} 
-Understanding the gravitational stability (GI) of rotating disks   
-is fundamental to many astrophysical problems. In the context of 
+The gravitational instability (GI) of rotating disks   
+plays a fundamental role in many astrophysical problems. In the context of 
 protoplanetary disks (PPDs), GI has two applications. It can 
@@
 \subsection{Viscosity removes rotational support}
-It is well known that viscous, self-gravitating disks can develop 
-GI even when $Q>1$. This is because viscosity compromises rotational
-support against self-gravity for long-wavelength disturbances
+Viscous disks can develop GI even when $Q>1$. This is because
+viscosity compromises rotational support against self-gravity for
+long-wavelength disturbances
 \citep{lynden-bell74,willerding92,gammie96}. A similar effect  
 occurs in dusty fluids where the required frictional forces are 
-provided by dust-gas drag forces \citep{youdin11,takahashi14}.
+provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
+is a mechanism for planetesimational formation %frag of dust
+\citep{youdin11}. Then it is not unreasonable to expect an analogous
+fragmentation of gaseous disks due to viscosity.
 
-Now, gaseous accretion disks are often modeled with a Navier-Stokes viscosity
+For gaseous accretion disks, a Navier-Stokes viscosity is often used to 
 to mimic hydrodynamic or magneto-hydrodynamic turbulence
@@
 
-How well turbulence can be modeled as an effective, Navier-Stokes
-viscosity is a separate issue \citep{balbus99}. However, it is 
+How well turbulence can be modeled as an effective viscosity is a
+separate issue \citep{balbus99}. However, it is  
 reasonable to assume that turbulence frustates rotational 
 support. (This is, after all, the physical cause for turbulent mass
-accretion.)
+accretion.) In this work, we use viscosity to represent this physical
+effect.
 
-Again, viscosity is absent in the Toomre $Q$ parameter. The above
+Viscosity is absent in the standard Toomre parameter. The above
 studies account for viscosity but consider isothermal disks. On the
@@
 \S\ref{linear}. We present results with beta cooling in
-\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks, respectively. In
+\S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks,
+respectively. %key result: generalized dispersion relation 
+In
 \S\ref{2dppd} we apply linear theory to PPDs with realistic cooling

### 2016-01-16 paper/paper3d.tex
regular update

@@
 We quantify the interpretation of this result as the
-compromise of gravitational stability by non-ideal physics: 
+compromise of gravitational stability by: 
 \begin{inparaenum}[1)] 
-\item cooling, which reduces pressure support; and
-\item viscosity, which reduces rotational support. 
+\item 
+cooling, which reduces pressure support; and/or
+\item 
+viscosity, which reduces rotational support. 
 \end{inparaenum}
-To this end, we analyse the gravitational stability of accretion disks
-including these effects explicitly.   
+We analyse the gravitational stability of accretion disks
+including these non-ideal effects explicitly.   
 We derive a general dispersion relation for linear waves in a 
 self-gravitating, viscous, non-adiabatic disk with internal and
-external heating, and explicit cooling. We 
+external heating, and cooling. We 
 consider parameterized cooling functions in two and three dimensions,

### 2016-01-16 paper/results3d.tex
regular update

@@
 \section{Three-dimensional disks with beta cooling}\label{3ddisk}
-We now extend the previous results to 3D disks with vertical
-structure. This is expected to weaken gravitational instabilities by
-`diluting' self-gravity along the vertical direction. It is possible to incorportate
-this effect in the previous 2D framework, but this introduces
-additional parameters. We discuss this in Appendix \ref{3dcorr}. 
-It is more direct to solve the 3D eigenvalue problem to avoid such 
-uncertainties.
+We briefly confirm the above results in 3D disks with vertical
+structure. This is expected to weaken gravitational instabilities
+because the disk mass is spread across the vertical dimension. It is
+possible to incorportate this effect in the previous 2D framework, but this
+introduces additional parameters. We discuss this in Appendix
+\ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
+avoid such uncertainties. 
+%We will find that while softening the
+%self-gravity in 2D can produce qualitatively similar results to 3D, a
+%quantitative match is
 
-{\bf note: fast beta-cooling in 3D destablizes vertical modes (enables
-convection)}
+%{\bf note: fast beta-cooling in 3D destablizes vertical modes (enables
+%convection)}
 
@@
 \subsection{Numerical procedure}
+{\bf move to appendix?}
 We use a pseudo-spectral method to solve the set of ordinary
@@
 In the following examples we consider a vertically isothermal disk
-($\Gamma=1$) with a numerical surface such 
-that $\rho(\zmax)=0.05\rho_0$, and use $N=65$.
+($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
+constant (Eq. \ref{alpha_beta_relation}). We use $N=65$  and apply
+numerical disk surface such that $\rho(\zmax)=0.05\rho_0$.
 
@@
 Fig. \ref{3d_inviscid} show growth rates and the most unstable wavenumber
-obtained for a 3D inviscid disk without irradiation. We also plot 2D
+obtained for a 3D inviscid disk with $\qthree=0.71$
+($Q=2Q_\mathrm{crit}$) and $\theta=0$. We also plot 2D
 results with the 3D correction as described in Appendix
-\ref{3dcorr}. While the chosen value of $H_\mathrm{sg}$ results in a close
-match between 2D and 3D growth rates, the 3D wavenumber is somewhat
-smaller. This reflects self-gravity being weakened in the verical
+\ref{3dcorr}. The (empirically) chosen value of
+$H_\mathrm{sg}=0.64H$ results in a close match between 2D and 3D
+growth rates, but the most unstable wavenumber in 3D is somewhat smaller. 
+This reflects self-gravity being weakened in the verical 
 direction: a larger horizontal scale is required to achieve the same
-strength of self-gravity (and hence growth rate) as the 2D case.
+strength of self-gravity as the 2D case. Similarly, 
+choosing $H_\mathrm{sg}=0.53H$ matches the optimum wavenumbers, but
+growth rates are over-estimated in 2D.
 
@@
 This suggests that the exact value of
-$H_\mathrm{sg}$ is problem-dependent (cooling and viscosity), although taking
+$H_\mathrm{sg}$ is problem-dependent, although taking
 $H_\mathrm{sg}\sim O(H)$ should give the correct 3D growth rate within
-a factor of two. In this respect it is easier to
-directly solve the 3D problem.
+a factor of two. 
+%A smaller $H_\mathrm{sg}$ is needed with faster 
+%cooling (larger viscosity) and vice versa. 
+%because most unstable wavenumber is smaller with high viscosity
+
+%smaller scales with increasing beta
+
+%, because viscosity makes the
+%disk behave more two-dimensional.
+
+%by eliminating vertical velocity gradients.
 

### 2016-01-16 paper/application.tex

@@
 \section{Application to protoplanetary disks}\label{2dppd}
-We consider the stability of realistic, self-gravitating
+We now consider the stability of realistic, self-gravitating
 protoplanetary disks. Specifically, we will apply our linear framework
@@
 \begin{itemize}
-  \item $\dot{M}$, the global radial accretion rate;
+  \item $\dot{M}$, the global radial mass accretion rate;
   \item $Q_0$, the value of the 2D Toomre parameter where the disk is
@@
 temperature $T(R)$ where $R$ is the global cylindrical radius from the
-star. This gives the $Q(R)$ value required for input
-into our linear framework. We derive other dimensionless
-parameters below.
+star. \citep[See][for details.]{rafikov15} This gives the $Q(R)$
+value required for input into our linear framework. We derive other
+dimensionless parameters below.
 
@@
 %$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
-From our discussion in \S\ref{2d_inviscid}, we conclude that this
-system cannot be unstable without
+From the results in \S\ref{2d_inviscid}, we conclude that this
+system cannot be unstable anywhere without
 viscous effects acting on the perturbations if 
@@
 are both satisfied.
+
+A PPD becomes optically-thin ($\tau\to 0$) at
+large radii, and $\theta$ decreases. The first inequality in
+Eq. \ref{ppd_invisc_cond} then relaxes to $\gamma > 5/6$ in the
+outer disk, which is always true. On the other hand, gravito-turbulent
+simulations typically show $Q\simeq 1.5$---2, which satisfies the
+second critrion. This means that in the
+outer regions of a realistic PPD, cooling itself cannot cause instability.
+
+%conditions are likely satisfied in the outer disk 
+%Condition \ref{ppd_invisc_cond} must be violated in order for cooling
+%to destablize
 
@@
 $\gamma=1.6$ so the inviscid stability condition
-(Eq. \ref{ppd_invisc_cond}) is satisfied. 
-%{\bf actually this opacity law only valid $>30$au for this disk}
+(Eq. \ref{ppd_invisc_cond}) is satisfied. Then cooling alone cannot
+destabilize the disk. Fig. \ref{rafikov_model} shows the equilirbium
+disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
+These are set as input to the 2D dispersion relation.
 
-Fig. \ref{rafikov_model} shows the equilirbium disk profile in terms
-of $Q$, $\alpha$, $\beta$, and $\theta$. These are set as input to the
-2D dispersion relation to yield a growth timescale, in units of the
-local orbital period, shown in Fig. \ref{rafikov_growth}. We also plot
-an analytic estimate based on Eq. \ref{gammie_smallk} (instead of
-Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) which gives 
+Fig. \ref{rafikov_growth} show growth timescales and
+optimum wavenumbers for viscous GI in this disk model.  
+We also plot analytic estimates based on Eq. \ref{gammie_smallk} (instead of
+Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) 
+which gives the optimum wavenumber as
 \begin{align}
-  S = \frac{27\alpha}{16\theta^3Q^4}, 
+  |K| = \frac{3}{2\theta Q},
 \end{align}
-and the optimum wavenumber is
+with growth rate
 \begin{align} 
-  |K| = \frac{3}{2\theta Q}.
+  S = \frac{27\alpha}{16\theta^3Q^4}. 
 \end{align}
-%For the
-%stability calculation we use $\mu=-1, \lambda=0$ for the viscosity
-%prescription.
+These are similar to the isothermal results of
+\citet[][their Eq. 19 and 21, respectively]{sterzik95}, and identical if one
+takes $\theta=1$.
 
-The disk becomes dynamically unstable with growth timescales $\lesssim
-10$ orbits for $R\gtrsim60$AU. Coincidentally, beyond this radius
+The most unstable wavelength is a few times the disk thickness. The
+increase in $|K|$ from $\sim 10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is
+due to the decrease in $Q$; while that from $\sim 60\mathrm{AU}$ to
+$\sim 100\mathrm{AU}$ occurs as the disk transitions from optically
+thick to optically thin regimes. The mismatch at large distances is
+not surprising since the above expressions assume $|K|\ll
+1$. Nevertheless the analytic estimates reproduce the correct
+behavior.
+
+The disk is subject to viscous GI on dynamical timescales ($\lesssim
+10$ orbits) for $R\gtrsim60$AU. Coincidentally, beyond this radius
 $\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is often quoted as 
 criteria for disk fragmentation \citep[e.g.][]{rafikov15}. 
+This suggests that fragmentation in the outer parts of a realistic PPD
+occurs is physically due to large turbulent stresses.
+
 %{\bf despite apply beta cooling results to realistic cooling disks}
-Our result suggest an alternative interpretation of disk fragmentation
-is where the gravito-turbulent state becomes dynamically unstable due to
-both viscosity and self-gravity.
-
-Fig. \ref{rafikov_model} shows that the most unstable wavelength is a
-few times the disk thickness. The increase in $|K|$ from $\sim
-10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is due to the decrease in $Q$;
-while that from $\sim 60\mathrm{AU}$ to $\sim 100\mathrm{AU}$ occurs
-as the disk transitions from optically thick to optically thin
-regimes. The mismatch at large distances is not surprising since the
-above expressions assume $|K|\ll 1$. Nevertheless the analytic
-estimates reproduce the correct behavior.
+%Our result suggest an alternative interpretation of disk fragmentation
+%is where the gravito-turbulent state becomes dynamically unstable due to
+%both viscosity and self-gravity.
 
@@
 
-\subsection{3D PPD}
+\subsection{3D PPD with radiative diffusion}
 %{\bf radiative diffusion only for optically thick disk. how to connect

### 2016-01-16 paper/results2d.tex

@@
 with previous studies. The disk
-material assumed to be confined to the midplane so $\delta v_z=0$. 
-We make the replacements 
-$\rho \to \Sigma$ and $P\to\Pi$, the vertically-integrated
+material is assumed to be confined to the midplane so $\delta v_z=0$. 
+We make the replacement  
+$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
 pressure, and set $\Gamma=1$ without loss of generality. 
-%The perturbations $\delta\Sigma$, 
-%$\delta\Pi$, $\delta v_x$ and $\delta v_y$ have no vertical
-%dependence and $\delta v_z=0$. 
 The gravitational potential perturbation remains 3D and its midplane
@@
 
-We obtain from the linearized equations 
-an algebraic dispersion relation $s = s(k)$. It is simplest
-to write this  
+The linearized equations yield an algebraic dispersion relation $s =
+s(k)$. We write this  
 in terms of the dimensionless growth rate $S = s/\Omega$ and
@@
 Here we neglect viscosity by setting 
-$\alpha = \alpha_b = 0$. This implies an external,
-time-independent heat source must be invoked to balance the imposed
-cooling rate in equilibrium ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are assuming 
-viscosity is only important for setting up thermal equilibrium, and
-can be neglected for the pertubations. This eliminates viscous GI and 
-allows us to quantify the \emph{sole} effect of cooling.
+$\alpha = \alpha_b = 0$. This eliminates viscous GI and 
+allows us to quantify the \emph{sole} effect of cooling on the
+perturbations. An external, time-independent heat source should be
+invoked to balance the imposed cooling to allow an equilibrium to be
+defined ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are
+assuming viscosity only provides a background heating and does not
+play an active role in the perturbed state. %may or may not be true in a real
+                                %disk because we are using viscosity
+                                %as proxy for turbulence
 
@@
 \end{align}
-similar to the classic Lin-Shu dispersion relation. The first term on
+similar to the classic Lin-Shu dispersion relation, which may be
+obtained by taking the limit $|\beta S|\to\infty$. 
+The first term on
 the right-hand-side represents destabilization by self-gravity; 
 the second and third terms represent stabilization by rotation and
-pressure, respectively. The imposed cooling term only affects the
+pressure, respectively. The imposed cooling only affects the
 pressure response.
 
-Eq. \ref{inviscid} is a cubic equation in $S$. Applying the
-Routh-Hurwitz criterion, we find that stability is ensured if 
+Eq. \ref{inviscid} is a cubic equation in $S$. The 
+Routh-Hurwitz criteria imply that stability is ensured if 
 \begin{align}
@@
 \end{align}
-are both satisfied. Thus, the question of stability does not depend on
-the cooling rate $\beta$ (provided it is non-zero and finite). The
-system becomes more unstable with decreasing $\theta$.
+are both satisfied. Thus, whether or not the system is stable is
+independent of the cooling rate $\beta$ (provided it is non-zero and
+finite). The system becomes more unstable with decreasing $\theta$.
 
@@
 \end{align}
+%Eq. \ref{kstar} describes how the most unstable  mode is affected by
+%cooling ($\beta$) and external heating ($\theta$). 
 Inserting this into Eq. \ref{inviscid}, we find the maximum growth
@@
 results may be obtained in special cases, discussed next.
-%We plot examples of the maximimum growth rate as a function of cooling
-%times in Fig. \ref{2d_inviscid}. General expressions for the solution
-%of Eq. \ref{inviscid_max} are unwieldy, but simple results may 
-%be obtained for special cases, as discussed below.  
-%\begin{figure}
-%  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm 0cm]{figures/inviscsg}
-%  \caption{Maximum growth rate $s$ in the inviscid limit,
-%    obtained from Eq. \ref{inviscid}, as a function of the cooling
-%    time $\beta$ for $Q/Q_\mathrm{crit}=1,\,2,\,10$, where $\qcrit$ is
-%    defined in Eq. \ref{qcrit_def} as 
-%    the value of $Q$ such that the disk is marginally stable in the
-%    adiabatic limit. 
-%    \label{2d_inviscid}}
-%\end{figure}
 
-\subsubsection{$\theta = 0$ and $Q=Q_\mathrm{crit}$}  
-Beta cooling with $\theta=0$ is the standard choice for numerical
-simulations of self-gravitating disks \citep{gammie01}. There, 
-a steady state corresponds to the imposed cooling entirely balanced by heat generated through 
-gravito-turbulence. In the present framework, this gravito-turbulence 
-is assumed to provide a constant 
-background heating ($\mathcal{H}_\mathrm{ext}$) and does
-not enter the linearized equations. Cooling, however, is still imposed in the perturbed state. 
-For $\theta=0$ the system is unconditionally unstable for
-finite $Q$. The most unstable wavelength decreases with decreasing
-$\beta$. The $\beta\to0$ limit corresponds to a pressureless disk (not
-merely isothermal).
+\subsubsection{$\theta = 0$}  
+%Beta cooling with $\theta=0$ is commonly employed in 
+%numerical simulations of self-gravitating disks \citep{gammie01}. 
+For beta cooling with $\theta=0$, the disk is unconditionally
+unstable for finite $Q$. The most unstable wavelength decreases with
+decreasing $\beta$. The $\beta\to0$ limit corresponds to a
+pressureless disk (not merely isothermal).
 
-% it is always
-%possible to find a sufficiently small scale such that pressure effects
-%are
-
-Specifically, let us consider a disk with 
+Let us consider a disk with 
 \begin{align}\label{qcrit_def}
@@
 \begin{align}\label{sstar}
-  S_*^3 = \left(\frac{1}{\gamma Q_\mathrm{crit}^2 \beta}\right). 
+  S_*^3 = \frac{1}{\gamma Q_\mathrm{crit}^2 \beta}. 
 \end{align}
@@
 time $\beta_*$ as that which removes pressure support against
-self-gravity over the only natural lengthscale in the problem, $H$. We
+self-gravity over the natural lengthscale in the problem, $H$. We
 thus set $|K_*|=1$ and find, for the Keplerian disk,
@@
 %occurs.
-It is interesting to note that this $\beta_*$ is similar to previously reported 
-values of the cooling times, $\beta_c$, below which fragmentation
-occurs on a dynamical timescale. We compare these in Table
-\ref{bstar_compare}, showing rough agreement.
+This $\beta_*$ is similar to values of the cooling times, $\beta_\mathrm{c}$,
+below which fragmentation occur on dynamical timescales in
+numerical simulations of self-gravitating disks
+\citep{gammie01,rice05,rice11}, which employ the same beta cooling prescription.    
+We compare these in Table \ref{bstar_compare}, showing rough
+agreement; and further discuss the potential significance of this result in
+\S\ref{prev_works}.
 
@@
 
-\subsubsection{Passive disk with $\theta = 1$}
+\subsubsection{$\theta = 1$}
 Beta cooling with $\theta=1$ corresponds to 
@@
 As a numerical example, we consider a model with $\alpha_b=0$ and
-$\alpha=\alpha(\beta)$ given by thermal equilibrium,
-Eq. \ref{alpha_beta_relation}. Furthermore, we set
+$\alpha=\alpha(\beta)$ given by thermal equilibrium (
+Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
 \begin{align}
@@
 \end{align}
-to mimic a basic state maintained by gravito-turbulence where one can 
-expect the dimensionless stress $\alpha \sim Q^{-2}$ \citep{lin87}.
+to mimic a basic state maintained by gravito-turbulence where one
+might expect the dimensionless stress $\alpha \sim Q^{-2}$
+\citep{lin87}.
 
@@
 \ref{gammie_bigk}. Comparing the two panels shows that increasing the
-irradiation level ($\theta$) surpress  
-small-scale perturbations.
+irradiation level ($\theta$) surpresses small-scale perturbations.
 
@@
 Fig. \ref{gammie_maxrate_plot} show the maximum growth rate as a
-function of the cooling time $\beta$ for the case $\theta=0$. 
+function of the cooling time $\beta$ for $\theta=0$. 
 %We set $\alpha_b=\theta=0$ and 
@@
 There is good agreement between numerical results and
-Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. The assumption
-of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
-explains the mismatch there.
-
-Eq. \ref{gammie_maxrate_simple} gives the limiting behavior (for
-negligible $\theta$) as 
+Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. 
+%The assumption
+%of $|\beta S|\ll 1$ is not well satisfied at smaller $\beta$ which
+%explains the mismatch there. 
+Eq. \ref{gammie_maxrate_simple} gives the
+limiting behavior for this case as   
 \begin{align*}
   S_*\propto \begin{cases}
-    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} &  \beta
-    \gg \alpha, \\
-    \frac{1}{Q\sqrt{\alpha}}  \propto \mathrm{const.} & \beta \ll \alpha,
+%    \frac{\alpha^{1/4}}{Q\beta^{3/4}} \propto \beta^{-3/2} 
+    \alpha^{1/4}\beta^{-3/4}Q^{-1}\propto \beta^{-3/2} 
+    &  \beta \gg \alpha, \\
+%    \frac{1}{Q\sqrt{\alpha}}  
+    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha,    
   \end{cases}
 \end{align*}
-where the second proportionality applies Eq. \ref{alpha_beta_relation}
-and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
+where we have applied Eq. \ref{alpha_beta_relation}
+and \ref{Qalpha} to obtain the final result. 
+The disk is formally unstable for all $\beta$, but
 the growth timescales are long ($>10$ orbits) for $\beta \gtrsim
-20$. This region is marked by the vertical dashed-dotted lines in
+20$. This region is marked by the vertical dashed-dotted line in
 Fig. \ref{gammie_maxrate_plot}.
 
-As $\beta\to 0$ the required viscous heating to set up an equilibrium
-state diverges ($\alpha\to\infty$). 
-In the present model the increased
-viscous damping, which is stabilizing, is balanced by the decreasing
-(and hence destablizing) $Q$ required to achieve large $\alpha$
-(Eq. \ref{Qalpha}).
+%As $\beta\to 0$ the required viscous heating to set up an equilibrium
+%state diverges ($\alpha\to\infty$). 
+%In the present model the increased
+%viscous damping, which is stabilizing, is balanced by the decreasing
+%(and hence destablizing) $Q$ required to achieve large $\alpha$
+%(Eq. \ref{Qalpha}).
 
-Physically, however, we expect $\alpha < 1$. In fact, numerical
-simulations of self-gravitating disks show there is a maximum $\alpha$
-($\sim 0.06$) that can be sustained before fragmentation
-\citep{rice11}. We can interpret this result in our linear framework.
+{\bf maybe move following to discussion}
+Numerical simulations of self-gravitating disks show there is a
+maximum $\alpha$ ($\sim 0.06$) that can be sustained before
+fragmentation \citep{rice11}. We can interpret this result in our
+linear framework. 
 %We can use our case (\ref{gvisc}) to interpret this
@@
 ($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
-shows that such a system is dynamically unstable with 
-growth rate $s = O(\Omega)$.  Thus we do not expect rapidly-cooled,
-self-gravitating steady disks to exist for extended periods of
-time. This is consistent with previous numerical simulations performed
-by \cite{lodato05}.
-
+shows that such a system would be dynamically unstable with 
+growth rate $s = O(\Omega)$. This due to the direct effect of viscous 
+stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
+and hence highly turbulent, self-gravitating disks to persist beyond
+dynamical timescales. This is consistent with previous numerical
+simulations performed by \cite{lodato05}. 
 %Case (\ref{gvisc}) is consistent with 
-
 \begin{figure}
@@
 %    0.0cm]{figures/result2d_fixQ}\\
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.cm
+    0.0cm]{figures/result2d_gvisc}
   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
-    0.0cm]{figures/result2d_gvisc}
-  \caption{Growth rates maximized over wavenumber as a function of cooling time for the
+    1.0cm]{figures/result2d_gvisc_kmax}
+  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
+    viscous GI as a function of cooling time for the
     case shown in Fig. \ref{gammie_rate_plot}. Solid lines are

### 2016-01-17 paper/application.tex
regular update

@@
 \section{Application to protoplanetary disks}\label{2dppd}
-We now consider the stability of realistic, self-gravitating
-protoplanetary disks. Specifically, we will apply our linear framework
-to the gravito-turbulent disk model recently developed by 
-\cite{rafikov15}. This 2D, Keplerian disk orbits a Solar mass star and 
-is defined by the following parameters,
+We now apply our linear framework to assess the stability of realistic 
+protoplanetary disks. We consider the gravito-turbulent disk models recently
+developed by \cite{rafikov15}. This 2D, Keplerian disk orbits a Solar 
+mass star and is defined by the following parameters,
 
@@
 \subsection{PPD cooling}
-Energy loss in \citeauthor{rafikov15}'s model is given
-by 
+Energy loss in \citeauthor{rafikov15}'s model is given by 
 \begin{align}\label{real_cool}
@@
 \begin{align}
-  f(\tau) = \tau + \frac{1}{\tau},
+  f(\tau) = \tau + \frac{1}{\tau}, \label{ftau} 
 \end{align}
@@
 is the optical depth and recall Eq. \ref{opacity_law} 
-is our opacity model. 
-%and $\kappa_d$ is the opacity. 
-%We parameterize
-%\begin{align}
-%  \kappa_d = \kappa_{d0} \Sigma^a T^b,
-%\end{align}
-%where $\kappa_{d0}, a, b$ are constants.
+is our opacity model. Eq. \ref{ftau} accounts for cooling in the
+optically-thin ($\tau\ll 1$) and optically-thick ($\tau\gg1$)
+regimes.
 
 Since we are interested in the linear response of the system, the
-corresponding parameters $\beta$ and $\theta$ are found by
+corresponding cooling parameters $\beta$ and $\theta$ are found by
 comparing the linearized form of the cooling function
@@
 (Eq. \ref{beta_cool}).  
-
 %{\bf as opposed to comparing their unperturbed forms, as done in literature}
-
 Linearizing Eq. \ref{real_cool} gives   
@@
 
-%{\bf caution in applying numerical results obtained from beta cooling to
-%systems modeled with real cooling. above example shows that to model
-%real, non-irradiated disk with beta cooling actually requires irradiation in
-%the latter.}
-
-%Notice that $\theta\to 1$ as 
-%$T\toT_\mathrm{irr}$ the disk becomes passive.
-
 \subsection{Inviscid stability condition}
-%We specialize to $a=0, b=2$ for the opacity law. 
 We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
@@
 \begin{align}
-  \gamma > \frac{3}{2} \quad \text{and} \quad Q > \sqrt{\frac{6}{5}} \label{ppd_invisc_cond}
+  \gamma > \frac{3}{2} \quad \text{and} \quad Q >
+  \sqrt{\frac{6}{5}} \label{ppd_invisc_cond} 
 \end{align} 
@@
 
-A PPD becomes optically-thin ($\tau\to 0$) at
-large radii, and $\theta$ decreases. The first inequality in
-Eq. \ref{ppd_invisc_cond} then relaxes to $\gamma > 5/6$ in the
-outer disk, which is always true. On the other hand, gravito-turbulent
-simulations typically show $Q\simeq 1.5$---2, which satisfies the
-second critrion. This means that in the
-outer regions of a realistic PPD, cooling itself cannot cause instability.
+PPDs become optically-thin ($\tau\to 0$) with 
+increasing distance from its host star. Then $\theta$ decreases, and
+the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to $\gamma >
+5/6$ in the (gravito-turbulent) outer disk, which is always satisfied.  
+On the other hand, numerical simulations of gravito-turbulence
+show that the second inequality is generally satisfied \citep{rice11}.  
+Taken together, this means that the outer regions of a realistic PPD,
+cooling itself cannot cause instability.
 
@@
 $\gamma=1.6$ so the inviscid stability condition
-(Eq. \ref{ppd_invisc_cond}) is satisfied. Then cooling alone cannot
-destabilize the disk. Fig. \ref{rafikov_model} shows the equilirbium
+(Eq. \ref{ppd_invisc_cond}) is satisfied
+everywhere. Fig. \ref{rafikov_model} shows the equilirbium 
 disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  

### 2016-01-17 paper/results2d.tex
regular update

@@
 
-Fig. \ref{gammie_maxrate_plot} show the maximum growth rate as a
-function of the cooling time $\beta$ for $\theta=0$. 
-%We set $\alpha_b=\theta=0$ and 
-%consider three cases (top to bottom panels): 
-%\begin{enumerate}[(a)]
-%  \item $\alpha$, $\beta$ and $Q$ are independent; \label{casea}
-%  \item $\alpha=\alpha(\beta)$ from thermal equlibrium
-%    (Eq. \ref{alpha_beta_relation}) and $Q$ specified separately; \label{caseb}
-%  \item $\alpha(\beta)$ from Eq. \ref{alpha_beta_relation} and
-%    assuming $Q = Q_\mathrm{crit}/\sqrt{\alpha}$. This mimics a
-%    basic state maintained by 
-%    gravito-turbulence where the viscosity coefficient $\alpha \sim
-%    Q^{-2}$ \citep{lin87}.
-%    \label{gvisc}
-%\end{enumerate} 
-There is good agreement between numerical results and
+Fig. \ref{gammie_maxrate_plot} shows the maximum
+growth rate (top panel) and the corresponding wavenumber (bottom
+panel) as a function of the cooling time $\beta$ for $\theta=0$.  
+There is good agreement between numerical growth rates and
 Eq. \ref{gammie_maxrate_simple} for $\beta \gtrsim 1$. 
@@
 where we have applied Eq. \ref{alpha_beta_relation}
-and \ref{Qalpha} to obtain the final result. 
-The disk is formally unstable for all $\beta$, but
-the growth timescales are long ($>10$ orbits) for $\beta \gtrsim
+and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
+growth timescales are long ($>10$ orbits) for $\beta \gtrsim
 20$. This region is marked by the vertical dashed-dotted line in
-Fig. \ref{gammie_maxrate_plot}.
-
-%As $\beta\to 0$ the required viscous heating to set up an equilibrium
-%state diverges ($\alpha\to\infty$). 
-%In the present model the increased
-%viscous damping, which is stabilizing, is balanced by the decreasing
-%(and hence destablizing) $Q$ required to achieve large $\alpha$
-%(Eq. \ref{Qalpha}).
+Fig. \ref{gammie_maxrate_plot}. The optimum wavenumber decreases with
+the cooling time for $\beta\lesssim O(1)$ because larger scales are
+more resistant to the associated increase in viscous damping.
 

### 2016-01-17 paper/results3d.tex
regular update

@@
 \section{Three-dimensional disks with beta cooling}\label{3ddisk}
-We briefly confirm the above results in 3D disks with vertical
+We confirm the above results in 3D disks with vertical
 structure. This is expected to weaken gravitational instabilities
@@
 possible to incorportate this effect in the previous 2D framework, but this
-introduces additional parameters. We discuss this in Appendix
+introduces additional parameters as discussed in Appendix
 \ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to

### 2016-01-17 paper/application.tex
reg update

@@
 protoplanetary disks. We consider the gravito-turbulent disk models recently
-developed by \cite{rafikov15}. This 2D, Keplerian disk orbits a Solar 
+developed by \cite[][hereafter \citetalias{rafikov15}]{rafikov15}. 
+This 2D, Keplerian disk orbits a Solar 
 mass star and is defined by the following parameters,
@@
 temperature $T(R)$ where $R$ is the global cylindrical radius from the
-star. \citep[See][for details.]{rafikov15} This gives the $Q(R)$
+star. \citepalias[See][for details.]{rafikov15} This gives the $Q(R)$
 value required for input into our linear framework. We derive other
@@
 \subsection{PPD cooling}
-Energy loss in \citeauthor{rafikov15}'s model is given by 
+Energy loss in \citetalias{rafikov15}'s model is given by 
 \begin{align}\label{real_cool}
@@
 PPDs become optically-thin ($\tau\to 0$) with 
-increasing distance from its host star. Then $\theta$ decreases, and
+increasing distance from the star. Then $\theta$ decreases, and
 the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to $\gamma >
-5/6$ in the (gravito-turbulent) outer disk, which is always satisfied.  
+5/6$ in the outer disk, which is always satisfied.  
 On the other hand, numerical simulations of gravito-turbulence
 show that the second inequality is generally satisfied \citep{rice11}.  
-Taken together, this means that the outer regions of a realistic PPD,
+Taken together, this means that in the outer regions of a realistic PPD,
 cooling itself cannot cause instability.
 
-%conditions are likely satisfied in the outer disk 
-%Condition \ref{ppd_invisc_cond} must be violated in order for cooling
-%to destablize
-
 \subsection{Example 2D calculation}\label{pp2d_example}
-We consider a disk with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
-$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. For the adopted opacity
-law we follow \citeauthor{rafikov15} and take $\kappa_{d0} =
-5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$. Here we use
+We consider a model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
+$\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. 
+We adopt the opacity scale $\kappa_{d0} =
+5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$  as in
+\citetalias{rafikov15}. We use 
 $\gamma=1.6$ so the inviscid stability condition
@@
 disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
-These are set as input to the 2D dispersion relation.
+These are set as input to the 2D dispersion relation
+(Eq. \ref{thindisk}).
 
@@
 due to the decrease in $Q$; while that from $\sim 60\mathrm{AU}$ to
-$\sim 100\mathrm{AU}$ occurs as the disk transitions from optically
-thick to optically thin regimes. The mismatch at large distances is
+$\sim 100\mathrm{AU}$ occurs as the disk transitions from
+optically-thick to optically-thin regimes. The mismatch at large
+distances is  
 not surprising since the above expressions assume $|K|\ll
@@
 %both viscosity and self-gravity. 
-
 %{\bf note: disk might be stablized by extremely large alpha at large
@@
 %to optically thin regions?} 
-We briefly consider 3D PPDs subject to explicit radiative diffusion 
-(\S\ref{rad_cool}). Given the $\alpha(R)$ and $Q(R)$ profiles obtained
+We briefly consider 3D PPDs with explicit radiative diffusion
+(\S\ref{rad_cool}). 
+Given the $\alpha(R)$ and $Q(R)$ profiles obtained
 from the 2D model above, at each radius $R$ we obtain the vertical
@@
 Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
-wavenumber for $R\in[10,100]$AU. We also plot the corresponding 2D
+wavenumber for $R\in[10,100]$AU; along with corresponding 2D 
 results with softened self-gravity. There is a close match for
 $R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
-1$) and both models apply. However, beyond $60$AU where the disk is
-optically thin, radiative diffusion (the 3D curve) does not apply and
-under-estimates the growth rates. Nevertheless, the comparison shows
-that the transition radius $R\sim60$AU can be calculated within the 2D
-framework. This is not surprising since the instability is 
-fundamentally driven by viscous and self-gravity forces in the
-momentum equations, not the energy equation.
-
-%the instability is due to viscosity/gravity in momenta equation so
-%details of energy eq doesn't matter
+1$) and both models apply. However, beyond $60$AU where the disk
+becomes optically-thin, radiative diffusion (the 3D curve) is not
+valid and under-estimates the growth rates. Nevertheless, the
+comparison shows that the transition radius $R_t\sim60$AU can be
+calculated within the 2D framework. 
+%This is not surprising since the
+%instability is fundamentally driven by viscous and self-gravity forces in the
+%momentum equations, not the energy equation.
 
@@
     wavenumber (bottom) for viscous, 
-    self-gravitating unstable modes. The black curves are obtained
-    from the 3D eigenvalue problem with radiative diffusion, while the
-    green curves are 
+    self-gravitating modes. The black curves are obtained
+    from the 3D eigenvalue problem with radiative diffusion,
+    applicable to optically-thick regions ($R\lesssim 60$AU); 
+    while the green curves are obtained from the 2D corresponding
+    model, which is applicable to both optically-thick and
+    optically-thin regimes.  
     \label{rafikov_growth3d}}
 \end{figure}
-
-%3d rad diffusion is mathematically different problem (more
-%derivatives) so worth considering briefly
-%optically thin cooling in 3D is the same problem as before
-
-%use 2D model as midplane values to get 3D structure assuming radiative
-%diffusion and constant alpha w.r.t. height. allow cooling time to vary
-%with height. can't assume polytropic vertical structure. need to
-%reformulate linear framework (some transformations needed). 
-%skip 2D eigen prob if 3D works out

### 2016-01-17 paper/summary.tex
regular update
reg update

--- 2015-12-21 paper/summary.tex
@@
-\section{Summary and discussion}
-In this paper, we study the implications on the stability of
-turbulent, cooling accretion disks if said turbulence is modeled as an
-effective, Navier-Stokes viscosity. This viscosity provides a 
-background heating and may also be active in the perturbed 
-state. We pay particular attention to 
-protoplanetary disks subject to `gravito-turbulence'. 
-Our framework provide an alternative interpretation of existing
-results on PPD fragmentation established from direct numerical
-simulations. 
-{\bf cooling removes pressure support, viscosity removes rotational support}
-{\bf our conclusions do require input of emperical resuls from numerical
-sims}
+\section{Summary and discussion}\label{summary}
+In this paper, we develop the linear theory of cooling,
+turbulent accretion disks in order to understand gravitational
+instability (GI) in realistic protoplanetary disks (PPDs). 
+We model turbulence as an effective, 
+Navier-Stokes viscosity. This viscosity provides a 
+background heating to balance the imposed cooling rate,  
+and may also be active in the perturbed state. We provide physical
+interpretations of disk fragmentation, observed in 
+numerical simulations, as a secondary GI from a 
+gravito-turbulent state, driven by cooling and/or viscosity. 
+%{\bf cooling removes pressure support, viscosity removes rotational support}
+%{\bf our conclusions do require input of emperical resuls from numerical
+%sims}
+%going beyond toomre 
+%move to subsection?
+While viscosity and cooling are inversely related through thermal
+balance, they independently enhance GI: cooling reduces thermal
+support; and viscous forces compromise rotational support.
 
@@
 
-We derive the disperion relation for local axisymmetric density waves 
-subject to external heating and cooling (Eq. \ref{inviscid}), but
-without viscous effects on the waves. We find
-the sufficient condition for stability 
-is \emph{independent} of the cooling rate. Long cooling times can
+The effect of cooling on GI is quantified by the dispersion
+relation Eq. \ref{inviscid}. 
+%We derive the disperion relation for local axisymmetric density waves 
+%subject to external heating and cooling (Eq. \ref{inviscid}), but
+%without viscous effects. 
+We find
+the sufficient condition for stability is \emph{independent} of the
+cooling rate, provided it is non-zero and finite.   
+Long cooling times can
 formally lead to instability, but the growth timescales may be
-uninterestingly long. We obtain an analytic expression for the
-dimensionless cooling time, $\beta_*$, required to remove pressure
-support against self-gravity on radial lengthscales comparable to the 
-disk scale height. We find $\beta_*$ reproduces the fragmentation
-boundary found in previous numerical simulations (Table
-\ref{bstar_compare}).
+uninterestingly long.
 
-%viscous GI ad connection with frag
+%We obtain an analytic expression for the
+%dimensionless cooling time, $\beta_*$, required to remove pressure
+%support against self-gravity on radial lengthscales comparable to the 
+%disk scale height. We find $\beta_*$ reproduces the fragmentation
+%boundary found in previous numerical simulations (Table
+%\ref{bstar_compare}).
 
-We then include viscosity in the linear problem and generalize the
-`viscous gravitational instability', previously discussed in 
+%viscous GI (new: 3D, cooling, ext. heating)
+
+We generalize the `viscous gravitational instability', previously discussed in 
 simpler disk models \citep{lynden-bell74,gammie96,hunter83}, to include
-heating/cooling and three-dimensionality. Our generalized framework is
+heating/cooling and three-dimensionality.
+
+%application to PPD
+
+Our generalized framework is
 readily applied to realistic protoplanetary disk models. For
@@
 
-%viscosity > 0.1, beta < 3 - 4
-
-Our results bear striking similarities with the current consensus on
-PPD fragmentation. It is generally accepted that PPDs can fragment
-beyond a few tens of $\mathrm{AU}$. This conclusion results from
-applying a Toomre stability condition (e.g. $Q\lesssim 1.5$) and a
-cooling time or viscosity criteria (e.g. $\beta\lesssim 3$ or
-$\alpha\gtrsim 0.1$), the latter of which is
-based on numerical simulations. We find these
-emperical conditions essentially translates to viscous GI on dynamical
-time-scales. This suggests that gravito-turbulent
-PPDs fragment because the associated viscous stresses further
-destabilize the disk against self-gravity.
+Our results bear striking similarities with the current
+consensus on PPD fragmentation. It is generally ac-
+cepted that PPDs can fragment beyond a few tens of
+AU. This conclusion results from applying a Toomre
+stability condition (e.g. $Q\lesssim 1.5$) and a cooling time or
+viscosity criteria (e.g. $\beta\lesssim3$ or $\alpha\gtrsim 0.1$), the latter of
+which is based on numerical simulations. We find these
+emperical conditions essentially translates to viscous GI
+on dynamical time-scales. This suggests that gravito-
+turbulent PPDs fragment because the associated viscous
+stresses further destabilize the disk against self-gravity.
 
 %ultimately: why do disks fragment? viscosity may play a role
+%details of thermo doesn't matter if it's viscosity that leads to
+%fragmentation
 
-% also conclude this region of
+\subsection{Relation to numerical simulations}\label{prev_works}
+
+%frag due to cooling -> structures less than H
+%frag due to viscosity -> structures more than H
+%stochastic frag
+%non-convergence with resolution
+
+\subsection{MHD turbulence}
+%application to gravito-turbulence but actually linear theory does not assume
+%actually application to 
+%MHD turbulence prob makes more sense 
+%require `scale separation' for GI turbulence (small-scale GI provide
+%turbulence, large-scale GI for fragmentation)
+
+\subsection{Outstanding issues}
+%non-axisymmtric problem 
+%viscosity model (maybe it doesn't behave like a power law)
+%chosen a particular mu and lambda

### 2016-01-18 paper/paper3d.tex
local copy
reg update

@@
 \newcommand{\dphi}{\dd \Phi}
-\newcommand{\actaa}{Acta Astron.}
 \newcommand{\hsg}{H_\mathrm{sg}}
@@
 \defcitealias{goldreich67}{GS67}
+\defcitealias{rafikov15}{R15}
 
@@
 with the empirical condition for disk fragmentation based on a
-maximum sustainable stress. We 
-thus suggest that disk fragmentation can be physically attributed, 
+maximum sustainable stress. We suggest that realistic disk fragmentation can be
+physically attributed,  
 at least partially, to the removal of rotational stabilization against
@@
 \subsection{Viscosity removes rotational support}
+%there is also viscous damping which stabilizes
 Viscous disks can develop GI even when $Q>1$. This is because

### 2016-01-18 paper/summary.tex
reg update

@@
 \section{Summary and discussion}\label{summary}
-In this paper, we develop the linear theory of cooling,
-turbulent accretion disks in order to understand gravitational
+In this paper, we develop the linear theory of cooling, irradiated,
+and turbulent accretion disks in order to understand gravitational
 instability (GI) in realistic protoplanetary disks (PPDs). 
@@
 Navier-Stokes viscosity. This viscosity provides a 
-background heating to balance the imposed cooling rate,  
-and may also be active in the perturbed state. We provide physical
-interpretations of disk fragmentation, observed in 
-numerical simulations, as a secondary GI from a 
-gravito-turbulent state, driven by cooling and/or viscosity. 
+background heating to balance the imposed cooling,  
+and may also be active in the perturbed state.  
+We suggest that disk fragmentation, observed in numerical simulations,
+could be understood as the non-linear outcome of a secondary GI of the
+gravito-turbulent basic state, driven by cooling and/or viscosity.   
+%This %allows us to assess realistic disk fragmentation without using
+%uncertain empirical conditions based on numerical simulations.   
 %{\bf cooling removes pressure support, viscosity removes rotational support}
@@
 %move to subsection?
-While viscosity and cooling are inversely related through thermal
+While viscosity and cooling are related through thermal
 balance, they independently enhance GI: cooling reduces thermal
-support; and viscous forces compromise rotational support.
+stabilization; and viscous forces compromise rotational
+stabilization. This provides a physical explanation as to why
+rapidly-cooled, gravito-turbulent disks cannot exist.
 
+%these reduce Q 
+%note: frag due to increase in sigma by turbulent fluctionations not modeled here 
 %generalized dispersion relation (inviscid)
 %no magic beta - connection with stochastic 
+%cooling function includes irradiation
 
-The effect of cooling on GI is quantified by the dispersion
-relation Eq. \ref{inviscid}. 
-%We derive the disperion relation for local axisymmetric density waves 
-%subject to external heating and cooling (Eq. \ref{inviscid}), but
-%without viscous effects. 
-We find
-the sufficient condition for stability is \emph{independent} of the
-cooling rate, provided it is non-zero and finite.   
-Long cooling times can
-formally lead to instability, but the growth timescales may be
-uninterestingly long.
+The effect of cooling and irradiation on GI is quantified by the
+dispersion relation 
+Eq. \ref{inviscid}. The sufficient condition for stability 
+depends on the irradiation level (Eq. \ref{stable_condition}), and 
+\emph{independent} of the cooling rate, provided it is non-zero and
+finite. This means that long cooling times can still  
+formally lead to instability. However, growth timescales may be 
+uninterestingly long for cooling times $\tcool\Omega\gtrsim
+O(10)$. Because cooling affects pressure support, GI driven by cooling
+occur on small scales, $kH\lesssim O(1)$.
 
@@
 %\ref{bstar_compare}).  
-
 %viscous GI (new: 3D, cooling, ext. heating)
 
-We generalize the `viscous gravitational instability', previously discussed in 
-simpler disk models \citep{lynden-bell74,gammie96,hunter83}, to include
-heating/cooling and three-dimensionality.
+We generalize the `viscous gravitational instability', previously 
+studied in isothermal disks
+\citep{lynden-bell74,willerding92,gammie96}, to include 
+cooling, viscous heating and irradiation. Our fiducial example,  
+where the viscosity and self-gravity are inversely
+related ($\alpha\propto Q^{-2}$), is a toy model for a
+gravito-turbulent disk. We find viscous GI occurs on orbital  
+timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
+of a maximum stress sustainable by gravito-turbulence established in
+previous numerical simulations. Because viscosity affects
+rotational support, viscous GI occurs on large scales, $kH\gtrsim
+O(1)$.  Furthermore, irradiation preferentially
+stabilizes small-scale perturbations.
 
-%application to PPD
+%application to PPD %realistic relation between alpha and Q
 
-Our generalized framework is
-readily applied to realistic protoplanetary disk models. For
-gravito-turbulent PPDs, we find viscous GI grows on orbital time-scales
-beyond $\sim 100\mathrm{AU}$. This means the assumed gravito-turbulent
-steady state should not exist beyond dynamical time-scales in the
-outer parts of PPDs.
+We apply our linear framework to protoplanetary disks with 
+realistic models for cooling and gravito-turbulence
+\citep{rafikov15}. We show that, for linear dynamics, realistic 
+cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
+actually modeled by `beta' cooling with a  
+non-zero irradiation parameter ($\theta\neq0$). However, 
+numerical simulations typically employ beta cooling with 
+$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
+are such simulations to realistic PPDs must then be further 
+investigated. The important implication of having 
+$\theta\sim 1$ in a PPD is that cooling alone cannot 
+destabilize the disk at large radii. Instead, viscous GI occurs on
+dynamical timescales for $R\gtrsim 60$AU because 
+$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
+a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
+\emph{empirical} conditions cited in the literature to determine disk
+fragmentation 
+\citep[e.g.][]{rafikov15}. By contrast, we attribute a \emph{physical}
+cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
+fragment when turbulent stresses are large enough to further
+destabilize the disk against self-gravity.    
+%collary: details of cooling unimportant as long as the associated
+%stress is the same
 
-Our results bear striking similarities with the current
-consensus on PPD fragmentation. It is generally ac-
-cepted that PPDs can fragment beyond a few tens of
-AU. This conclusion results from applying a Toomre
-stability condition (e.g. $Q\lesssim 1.5$) and a cooling time or
-viscosity criteria (e.g. $\beta\lesssim3$ or $\alpha\gtrsim 0.1$), the latter of
-which is based on numerical simulations. We find these
-emperical conditions essentially translates to viscous GI
-on dynamical time-scales. This suggests that gravito-
-turbulent PPDs fragment because the associated viscous
-stresses further destabilize the disk against self-gravity.
-
+%small scale transport 
 %ultimately: why do disks fragment? viscosity may play a role
@@
 
+%we do not look at fragment survival
+%fragment due to cooling or viscosity?
+%they should produce diff size frags
 \subsection{Relation to numerical simulations}\label{prev_works}
+Our framework may shed light on 
+%may offer an alternative understanding of 
+some recent numerical results concerning disk
+fragmentation.  \cite{paardekooper12} found that fragmentation can occur for 
+slowly-cooled disks, $\tcool\Omega\gg O(1)$, but that this requires  
+simulations to run for significantly longer than dynamical 
+timescales. This is consistent with our finding
+that for either cooling-driven or viscous GI, there is no critical 
+cooling rate (viscosity) beyond (below) which the disk is  
+formally stable. Instead, growth rates smoothly decrease with
+increasing (decreasing) $\tcool$ ($\alpha$), implying that
+instabilities take longer to develop for slowly-cooled disks.        
+%hopkins paper
 
+%resolution convergence 
+Another issue is convergence with numerical resolution. Notably, 
+\cite{baehr15} found that for fixed cooling rates, disks eventually 
+fragment with sufficient spatial resolution. %This could be due to 
+%small-scale, turbulent angular momentum transport being enabled at
+%high resolutions, which aids clump formation. 
+This might be related to viscous GI in the following sense: high
+resolutions enable small-scale turbulent angular momentum transport
+away from over-dense regions, aiding clump formation. This 
+mechanism is unavailable at low resolutions, for which one can only
+resort to rapid cooling for destabilization.
+
+We note that power in gravito-turbulence is dominated by scales with
+$kH\sim 1$ \citep{cossins09}, and this is always  
+well-resolved by modern numerical simulations. However, this does not
+necessarily imply small-scale dynamics are unimportant, as is evident
+from the non-converging simulations described above.
+
+%it is true that most unstable wavelength is H is this is always
+%resolved
+%but that doesn't mean small-scale structures are unimportant 
+%note: no turbulence observed inside clumps
+%consistent with rho*nu = const  
+%this alpha should be that associated with small-scale transport 
 %frag due to cooling -> structures less than H
@@
 %non-convergence with resolution 
+%no magic beta when non-ideal effects are included
 
-\subsection{MHD turbulence}
+\subsection{MHD turbulence}%turbulence and viscosity 
+It is important to note that while we consider gravito-turbulent disks  
+as an application, the linear theory itself does not assume any 
+specific origin of turbulence.
+
+%effect of MHD turbulence on GI 
+Our models may thus apply to self-gravitating disks dominated by, for
+example, magneto-hydrodynamic (MHD) turbulence.   
+%In fact, the viscous approach may be more
+%applicable here than gravito-turbulence, because the characteristic
+%wavelength of the magneto-rotational instability (which leads to MHD
+%turbulence) is $\ll H$, at least in the ideal MHD limit. Then there is
+%a scale separation between the underlying, small-scale MHD 
+%turbulence and the large-scale effects of self-gravity. 
+%fromang is isothermal sims - no cooling effect 
+Explicit numerical simulations of magnetized, massive disks have been performed 
+by \cite{fromang05}. He finds disk fragmentation with increasing numerical
+resolution, attributing this to resolving the most unstable MRI
+wavelength, which enables small-scale angular momentum removal by MHD
+turbulence to aid fragmentation. %this supports using viscosity to
+                                %mimic turbulence
+This would be a physical mechnism represented by the viscous GI
+discussed in this paper; which lends some support for  
+the use of a viscosity to represent turbulence.
+
+%these sims are isothermal so cooling is not responsible for frag 
+%note fromang -> comparing HD and MHD at low res: 
+%mhd turb lowers T_grav -> no frag
+%suggest frag associated with stress 
+%turb enhances frag 
+%turb due to MHD 
+%maybe even better model for (ideal) MHD turbulence 
+%because mri scales are << H 
+%scale-separation 
 %application to gravito-turbulence but actually linear theory does not assume
@@
 \subsection{Outstanding issues}
+%the link between the linear GI's discussed in this paper 
+%and frag 
+True disk fragmentation is a non-linear process characterized by clumps   
+reaching densities orders of magnitude above the ambient value, and
+which survives tidal shear and shocks \citep{paardekooper12}.  
+%cite the young papers 
+Clearly, our linear models cannot address fragmentation
+directly. Technically, we have only demonstrated that a 
+gravito-turbulent state becomes dynamically unstable, and thus should
+not persist, when cooling (viscosity) is too rapid (large).  
+However, numerical simulations show that there are only two
+possible outcomes of cooling, self-gravitating disks: steady
+gravito-turbulence or fragmentation. It seems then reasonable to 
+speculate that the non-existence of a gravito-turbulent state, here due to
+dynamical instability, implies disk fragmentation.
+
+%modeling turb as visc %explore other visc laws 
+The most important assumption in this work is modeling turbulence as a  
+Navier-Stokes viscosity. Furthermore, we have chosen a particular  
+viscosity law (see \S\ref{visc_model}) to mimic the effect of 
+turbulence in reducing rotational support (in the sense that it
+provides small-scale angular momentum transport). 
+%to enhance clumping 
+%this visc law is used in prev studies (by much smarter people)
+How to quantitatively model turbulence (be it due to self-gravity or
+MHD) as a viscosity, especially on dynamical timescales, should be
+clarified with direct numerical simulations. The present linear models
+should then be modified accordingly. 
+%could consider other visc parameters mu and lambda (but this
+%introduces visc overstability, or new functional forms of the
+%viscosity law altogether
+
+%non-axisymmetry and transient dynamics 
+Another possible generalization is to non-axisymmetric
+disturbances. In the absence of cooling and viscosity, 
+non-axisymmetric GI can develop for $Q$ somewhat larger than unity
+\citep{papaloizou89,papaloizou91}. It would be interesting to study
+how non-axisymmetric perturbations are affected by cooling and
+viscosity, and thus
+
+%papaloizou91, papaloizou lin 87
+
+%our linear model includes the same physics 
+%we show 
 %non-axisymmtric problem 
@@
 %chosen a particular mu and lambda 
+%link between linear instability and frag is issue
+%true successful frag also depend on frag survival

### 2016-01-18 paper/summary.tex
reg update

@@
 We suggest that disk fragmentation, observed in numerical simulations,
-could be understood as the non-linear outcome of a secondary GI of the
-gravito-turbulent basic state, driven by cooling and/or viscosity.   
+can be understood as the non-linear outcome of a secondary
+GI of the gravito-turbulent basic state, driven by cooling and/or viscosity.   
 %This %allows us to assess realistic disk fragmentation without using
@@
 %non-axisymmetry and transient dynamics 
-Another possible generalization is to non-axisymmetric
-disturbances. In the absence of cooling and viscosity, 
+Another possible generalization is non-axisymmetric
+disturbances. In barotropic, inviscid disks  
 non-axisymmetric GI can develop for $Q$ somewhat larger than unity
@@
 how non-axisymmetric perturbations are affected by cooling and
-viscosity, and thus
+viscosity in order to improve %or verify 
+the link between disk fragmentation and the
+stability of gravito-turbulent disks.
 
 %papaloizou91, papaloizou lin 87 
-
 %our linear model includes the same physics 

### 2016-01-19 paper/paper3d.tex
reg update

@@
 \end{abstract}
-%alternative abstract
+%we suggest turbulent stresses play an active, dynamical role in disk fragmentation by 
+%removing rotational stabilization against self-gravity. 
+%in the sense of small-scale transport
 
@@
 provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
-is a mechanism for planetesimational formation %frag of dust
+is a mechanism to enhance particle clumping for planetesimational formation 
+%frag of dust 
 \citep{youdin11}. Then it is not unreasonable to expect an analogous
@@
 and whether or not a critical cooling time can 
-be defined at all \citep{meru11,meru12,paardekooper12}. However, it is
+be defined at all \citep{meru11,meru12,paardekooper12,hopkins13}. However, it is
 generally accepted that \emph{steady, gravito-turbulent disks do not

### 2016-01-19 paper/summary.tex
reg update

@@
 O(10)$. Because cooling affects pressure support, GI driven by cooling
-occur on small scales, $kH\lesssim O(1)$.
+occur on small scales, $kH\gtrsim O(1)$.
 
@@
 \citep{lynden-bell74,willerding92,gammie96}, to include 
-cooling, viscous heating and irradiation. Our fiducial example,  
-where the viscosity and self-gravity are inversely
-related ($\alpha\propto Q^{-2}$), is a toy model for a
-gravito-turbulent disk. We find viscous GI occurs on orbital  
+cooling, viscous heating and irradiation. In our fiducial disk,  
+where viscosity and self-gravity are inversely
+related ($\alpha\propto Q^{-2}$) to model a
+gravito-turbulent background, viscous GI occurs on orbital  
 timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
 of a maximum stress sustainable by gravito-turbulence established in
-previous numerical simulations. Because viscosity affects
-rotational support, viscous GI occurs on large scales, $kH\gtrsim
+previous numerical simulations \citep{rice05}. Because viscosity affects 
+rotational support, viscous GI occurs on large scales, $kH\lesssim
 O(1)$.  Furthermore, irradiation preferentially
-stabilizes small-scale perturbations.
+stabilizes small-scale perturbations. %maybe that's why irradiated disks produce smaller alpha
 
@@
 actually modeled by `beta' cooling with a  
-non-zero irradiation parameter ($\theta\neq0$). However, 
-numerical simulations typically employ beta cooling with 
+non-zero irradiation parameter ($\theta\neq0$), in contrast to 
+numerical simulations which typically employ beta cooling with  
 $\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
 are such simulations to realistic PPDs must then be further 
-investigated. The important implication of having 
+investigated.
+
+The important implication of having 
 $\theta\sim 1$ in a PPD is that cooling alone cannot 
@@
 fragmentation 
-\citep[e.g.][]{rafikov15}. By contrast, we attribute a \emph{physical}
+\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
 cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
@@
 well-resolved by modern numerical simulations. However, this does not
-necessarily imply small-scale dynamics are unimportant, as is evident
-from the non-converging simulations described above.
+necessarily imply small-scale dynamics are unimportant \citep[especially for non-linear evolution, see][]{young15}, 
+as is evident from the non-converging simulations described above.
 
@@
 reaching densities orders of magnitude above the ambient value, and
-which survives tidal shear and shocks \citep{paardekooper12}.  
+which survives tidal shear and shocks \citep{paardekooper12,young16}.  
 %cite the young papers 
@@
 dynamical instability, implies disk fragmentation.
+
+{\bf comment on stochasitc frag. we only consider 
+how to enhance GI by reducing rotation and pressure. but a third route is 
+increasing sigma. according to toomre Q these are the 3 effects one should consider
+for SG. turbulent flucations may occasionally produce high sigma dens perts. 
+possible relation to current framework? high sigma perts arise because of viscous GI? 
+high alpha $\to$ small-scale transport $\to$ enhance over-dense regions}
 
@@
 %true successful frag also depend on frag survival
+
+%large stresses -> viscous GI operates quickly -> get non-linear den perts quickly

### 2016-01-21 paper/appendix3d.tex
reg udpate

@@
 \end{align}
-which coincides with \citeauthor{gammie96}'s Eq. 18 for $\gamma=1$. 
+which coincides with \citeauthor{gammie96}'s Eq. 18 for $\gamma=1$. In
+that case the condition for instability is
+\begin{align}
+%  \frac{2|K|}{Q} - K^2>0. 
+  |K|< \frac{2}{Q}.
+\end{align} 
+This is the same stability condition as if rotation was neglected in
+the standard, isothermal dispersion relation for density waves
+(obtained from Eq. \ref{thindisk} by setting $\gamma=\theta=1$). Thus
+the effect of viscosity is to remove 
+rotational stabilization against self-gravity, as originally explained
+by \cite{lynden-bell74}.
+
 %We recover \cite{gammie96}'s dispersion relation 
@@
 We regard $H_\mathrm{sg}$ as a free parameter of the problem.
+
+\section{Numerical method for the 3D eigenvalue problem}
+We use a pseudo-spectral method to solve the set of ordinary
+differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
+the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
+We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
+  \delta v_x, \delta 
+  v_y]$ in even Chebyshev polynomials,
+\begin{align}
+  \bm{U}(z) = \sum_{j=1}^N \bm{a}_jT_{2(j-1)}(z/\zmax), 
+\end{align}
+and the vertical velocity in odd Chebyshev polynomials,
+\begin{align}
+  \delta v_z(z) = \sum_{j=1}^N b_jT_{2j-1}(z/\zmax),   
+\end{align}
+where $\bm{a}_j$ and $b_j$ are the spectral coefficients. 
+The basis functions are chosen to satisfy a reflecting boundary
+condition at the midplane, %GI has even symmetry across midplane 
+\begin{align}
+  \bm{U}^\prime(0) = \bm{0}, \quad \delta v_z(0) = 0.
+\end{align}
+%At the upper disk boundary we also apply reflection,
+We explicitly apply a reflecting upper disk boundary, 
+\begin{align}
+  \delta v_z(\zmax) = \delta v_x^\prime(\zmax) = \delta
+  v_y^\prime(\zmax) = 0,
+\end{align}
+and the potential satisfies
+\begin{align}
+  \delta \Phi^\prime (\zmax) + k \delta\Phi(\zmax) = 0, 
+\end{align}
+as derived by \cite{goldreich65a} and used in similar studies  
+\citep{kim12,lin14c}. %reflection is standard in direct simulations
+
+We discretize the equations, including upper disk boundary conditions,
+over the $N$ positive abscissae of the extrema of $T_{2N-1}$. This
+procedure converts the differential equations into a generalized
+eigenvalue problem, for which we use the standard matrix package
+LAPACK to solve. We use $N=65$.

### 2016-01-21 paper/paper3d.tex
reg udpate

@@
 \begin{abstract}
- Low mass, self-gravitating accretion disks can remain in a  
-quasi-steady, `gravito-turbulent' state provided that thermal 
-time-scales are sufficiently long, or equivalently for sufficiently
-weak turbulent stresses. Otherwise, numerical simulations show that
-disk fragmentation ensues. 
+%Low mass, self-gravitating accretion disks can remain in a  
+%quasi-steady, `gravito-turbulent' state provided that thermal 
+%time-scales are sufficiently long, or equivalently for sufficiently
+%weak turbulent stresses. Otherwise, numerical simulations show that
+%disk fragmentation ensues. 
+%{\bf
+Low mass, self-gravitating accretion disks admit a quasi-steady,
+`gravito-turbulent' state in which cooling balances turbulent
+heating. However, numerical simulations show that gravito-turbulence
+cannot be sustained when the cooling rate or  
+turbulent viscosity is too large, resulting in disk fragmentation.     
+%} 
 We quantify the interpretation of this result as the
@@
 \end{inparaenum}
-We analyse the gravitational stability of accretion disks
+We analyze the gravitational stability of accretion disks 
 including these non-ideal effects explicitly.    
@@
 We apply linear theory to 
-global models of protoplanetary disks (PPDs) subject to gravito-turbulence, modeled as an effective
-viscosity, and cooling via dust opacity. We find the sole effect of cooling is insufficient to
+global models of protoplanetary disks (PPDs) subject to gravito-turbulence modeled as an effective
+viscosity and cooling via dust opacity. We find the sole effect of cooling is insufficient to
 destabilize a realistic PPD. However, inclusion of viscosity renders
@@
 with the empirical condition for disk fragmentation based on a
-maximum sustainable stress. We suggest that realistic disk fragmentation can be
-physically attributed,  
-at least partially, to the removal of rotational stabilization against
-self-gravity by turbulence.   
+maximum sustainable stress. 
+%We suggest that realistic disk fragmentation can be
+%physically attributed,  
+%at least partially, to the removal of rotational stabilization against
+%self-gravity by turbulence.   
+We suggest turbulent stresses play an active role in realistic disk
+fragmentation by removing rotational stabilization against 
+self-gravity. 
+%}
 \end{abstract}
@@
 plays a fundamental role in many astrophysical problems. In the context of 
-protoplanetary disks (PPDs), GI has two applications. It can 
+gaseous protoplanetary disks (PPDs), GI has two applications. It can  
 provide gravitational torques to transport angular momentum outwards
 and thus enable mass accretion inwards \citep{armitage10,turner14}.   
-GI may also lead to disk fragmentation, which has been invoked to explain planet formation at
-large radii \citep{helled14}. Any realistic study of these phenomena require direct
+GI may also lead to disk fragmentation, which has been invoked to 
+explain giant planet formation at large radii \citep{helled14}. Any 
+realistic study of these phenomena require direct   
 numerical simulations. Nevertheless, physical insight can be  
@@
 density and $G$ is the gravitational constant. The Toomre $Q$ is a
-measure of the destabilizing effect of self-gravity ($\Sigma$) against 
+measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
 the stabilzing effect of rotation ($\kappa$) and pressure 
@@
 from). This includes realistic disk structures 
-obtained with non-ideal physics that are not explicitly contained in
-$Q$ \citep[e.g.][]{kimura12}.  It is, however, not obvious that such additions
-leave the stability condition unaltered. Indeed, we show that GI in more
-realistic disk models cannot be described by a single `Toomre-like' 
-condition.
+that account for non-ideal physics not 
+explicitly reflected by $Q$ \citep[e.g.][]{kimura12}.  It is, however,
+not obvious that such additions leave the stability condition
+unaltered.
 
+Indeed, we show that when cooling and viscosity are  
+included, as expected in realistic disks, GI cannot be assessed from a
+single `Toomre-like' condition.      
 %disk structure may be computed with sophisticated physics but
@@
 %structure  based on realistic physics 
-
 The goal of this work is to generalize the analytic treatment of disk 
 GI to allow a self-consistent discussion of GI in realistic accretion 
-disk models. Specifically, we will examine the effect of cooling and
-viscosity on GI in PPDs.  
+disk models. 
+%Specifically, we will examine the effect of cooling and  
+%viscosity on GI in PPDs.
+
 %Their relevance to PPDs are described below.

### 2016-01-21 paper/results3d.tex
reg udpate

@@
 %height s.t. rho(zmax) = 0.05 rho(0) avoid very low densities
-\subsection{Numerical procedure}
-{\bf move to appendix?}
-We use a pseudo-spectral method to solve the set of ordinary
-differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
-the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
-We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
-  \delta v_x, \delta 
-  v_y]$ in even Chebyshev polynomials,
-\begin{align}
-  \bm{U}(z) = \sum_{j=1}^N \bm{a}_jT_{2(j-1)}(z/\zmax), 
-\end{align}
-and the vertical velocity in odd Chebyshev polynomials,
-\begin{align}
-  \delta v_z(z) = \sum_{j=1}^N b_jT_{2j-1}(z/\zmax),   
-\end{align}
-where $\bm{a}_j$ and $b_j$ are the spectral coefficients. 
-The basis functions are chosen to satisfy a reflecting boundary
-condition at the midplane, %GI has even symmetry across midplane 
-\begin{align}
-  \bm{U}^\prime(0) = \bm{0}, \quad \delta v_z(0) = 0.
-\end{align}
-%At the upper disk boundary we also apply reflection,
-We explicitly apply a reflecting upper disk boundary, 
-\begin{align}
-  \delta v_z(\zmax) = \delta v_x^\prime(\zmax) = \delta
-  v_y^\prime(\zmax) = 0,
-\end{align}
-and the potential satisfies
-\begin{align}
-  \delta \Phi^\prime (\zmax) + k \delta\Phi(\zmax) = 0, 
-\end{align}
-as derived by \cite{goldreich65a} and used in similar studies  
-\citep{kim12,lin14c}. %reflection is standard in direct simulations
-
-We discretize the equations, including upper disk boundary conditions,
-over the $N$ positive abscissae of the extrema of $T_{2N-1}$. This
-procedure converts the differential equations into a generalized
-eigenvalue problem, for which we use the standard matrix package
-LAPACK to solve.
 

### 2016-01-22 paper/paper3d.tex
new eigenvector fig
starting another revision

@@
 \begin{abstract}
-%Low mass, self-gravitating accretion disks can remain in a  
-%quasi-steady, `gravito-turbulent' state provided that thermal 
-%time-scales are sufficiently long, or equivalently for sufficiently
-%weak turbulent stresses. Otherwise, numerical simulations show that
-%disk fragmentation ensues. 
-%{\bf
-Low mass, self-gravitating accretion disks admit a quasi-steady,
-`gravito-turbulent' state in which cooling balances turbulent
+Low mass, self-gravitating accretion disks admit quasi-steady,
+`gravito-turbulent' states in which cooling balances turbulent
 heating. However, numerical simulations show that gravito-turbulence
-cannot be sustained when the cooling rate or  
-turbulent viscosity is too large, resulting in disk fragmentation.     
-%} 
-We quantify the interpretation of this result as the
-compromise of gravitational stability by: 
+cannot be sustained beyond dynamical timescales when the cooling rate
+or turbulent viscosity is too large, resulting in disk fragmentation.        
+%We quantify the interpretation of this result as the
+%compromise of gravitational stability by: 
+We quantify the interpretation of this result as a 
+secondary gravitational instability due to: 
 \begin{inparaenum}[1)] 
@@
 We apply linear theory to 
-global models of protoplanetary disks (PPDs) subject to gravito-turbulence modeled as an effective
+global models of protoplanetary disks (PPDs) subject to
+gravito-turbulence modeled as an effective 
 viscosity and cooling via dust opacity. We find the sole effect of cooling is insufficient to
@@
 the disk beyond $\sim 60$AU dynamically unstable on radial
+%this is for Mdot = 1d-6  
 lengthscales a few times the local disk thickness. This is coincident 
@@
 maximum sustainable stress. 
-%We suggest that realistic disk fragmentation can be
-%physically attributed,  
-%at least partially, to the removal of rotational stabilization against
-%self-gravity by turbulence.   
 We suggest turbulent stresses play an active role in realistic disk
@@
 self-gravity. 
-%}
 \end{abstract}
-%we suggest turbulent stresses play an active, dynamical role in disk fragmentation by 
-%removing rotational stabilization against self-gravity. 
-%in the sense of small-scale transport
 
@@
 provide gravitational torques to transport angular momentum outwards
-and thus enable mass accretion inwards \citep{armitage10,turner14}.   
+and thus enable mass accretion \citep{armitage10,turner14}.   
 GI may also lead to disk fragmentation, which has been invoked to 
-explain giant planet formation at large radii \citep{helled14}. Any 
-realistic study of these phenomena require direct   
+explain giant planet formation at large radii
+\citep{helled14}. Studying these non-linear phenomena require direct      
 numerical simulations. Nevertheless, physical insight can be  
-obtained through analytical understanding of GI.  %physical insight
+obtained through analytical understanding of GI.
 
-The standard metric for the (inverse) strength of disk self-gravity is the 
-\emph{Toomre parameter}, 
+The standard metric for the (inverse) strength of disk self-gravity is
+the \emph{Toomre parameter}, 
 \begin{align}\label{toomreQ_criterion}
@@
 the stabilzing effect of rotation ($\kappa$) and pressure 
-($c_s$). When $Q\leq 1$ the disk is formally unstable to local,
-axisymmetric disturbances in a two-dimensional (2D, razor-thin),
-inviscid and adiabatic (including isothermal) disk.
+($c_s$). This is evident from the dispersion relation for local,
+axisymmetric waves in a two-dimensional (2D, razor-thin), 
+inviscid and adiabatic (including isothermal) disk, 
+\begin{align} 
+  s^2 = \kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|, 
+\end{align}
+relating the growth rate $s$ and the radial wavenumber $k$. When
+$Q\leq 1$ the disk is formally unstable.
 
@@
 
-\subsection{Cooling removes pressure support}
+\subsection{Cooling-driven gravitational instability}
 A net thermal loss\footnote{Isothermal disks implicitly assume 
@@
 
-\subsection{Viscosity removes rotational support}
+\subsection{Viscosity-driven gravitational instability}
 %there is also viscous damping which stabilizes

### 2016-01-22 paper/paper3d.tex
starting another revision
re-writing intro

@@
 the stabilzing effect of rotation ($\kappa$) and pressure 
-($c_s$). This is evident from the dispersion relation for local,
-axisymmetric waves in a two-dimensional (2D, razor-thin), 
-inviscid and adiabatic (including isothermal) disk, 
-\begin{align} 
-  s^2 = \kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|, 
+($c_s$). This is evident from the dispersion relation, 
+\begin{align}\label{classic_dr}
+  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
 \end{align}
-relating the growth rate $s$ and the radial wavenumber $k$. When
-$Q\leq 1$ the disk is formally unstable.
+which relates the growth rate $s$ and radial wavenumber $k$ for local,
+axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
+adiabatic (including isothermal) disk. Such disturbances are formally
+unstable when the Toomre condition, $Q<1$, is satisfied.
+Non-axisymmetric modes can develop for  larger but still order-unity
+values of $Q$ . {\bf ref} 
+%strictly speaking non axi *local* modes are stable -> only transient
+%growth
+%true non-axi instability needs global disk  model 
+%non-axi GI at slightly larger Q, but still characterized by Q
 
+%qualitative intro to gravito-turbulence 
+%role of cooling and resulting viscosity
+
+It is important to remember that the dispersion relation
+Eq. \ref{classic_dr}  is derived from \emph{idealized} conditions: 
+the disk is laminar and does not experience any net thermal
+losses\footnote{Isothermal disks implicitly assume  
+  heating and thermal losses are exactly balanced at all 
+  times, so these do not cool in the current context.} 
+(hereafter `cooling').  
+However, real disks can cool, for example,
+due to radiative losses. Accretion disks may also be turbulent, for
+which its dynamic and thermodynamic effect is often modeled through an
+effective viscosity \citep{shakura73,lin87,rafikov15}. {\bf 
+  more refs here? non-sg refs} 
+%in terms of AM transport and heating 
 %assess GI in general
 
-The Toomre parameter is widely applied to assess GI in more general 
-situtations (cf. the ideal disk models from which $Q$ is derived
-from). This includes realistic disk structures 
-that account for non-ideal physics not 
-explicitly reflected by $Q$ \citep[e.g.][]{kimura12}.  It is, however,
-not obvious that such additions leave the stability condition
-unaltered.
+Despite this, the Toomre parameter, and its associated
+dispersion relation, are still widely applied to realistic disk models
+with viscosity and cooling \citep[e.g.][]{cossins09, kimura12}, 
+even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
+derived from inviscid disks without cooling.
 
-Indeed, we show that when cooling and viscosity are  
-included, as expected in realistic disks, GI cannot be assessed from a
-single `Toomre-like' condition.      
-%disk structure may be computed with sophisticated physics but
-%stability condition still based on ideal disk models. 
-%three main ingredients - gravity, pressure, rotation 
-%a standard way to apply this condition is to compute the disk
-%structure  based on realistic physics 
-The goal of this work is to generalize the analytic treatment of disk 
-GI to allow a self-consistent discussion of GI in realistic accretion 
-disk models. 
+However, as we emphasize in this work, the inclusion of non-ideal
+physics, such as cooling and viscosity, modifies the dispersion
+relation, Eq. \ref{classic_dr}, and hence the condition for GI. This 
+means that the Toomre condition alone is insufficient to assess GI
+in realistic disks. The goal of this work is to generalize the
+analytic treatment of disk GI to allow a self-consistent discussion of
+GI in realistic accretion disk models. Specifically, we consider the non-ideal
+effects of viscosity and cooling on GI.
+
 %Specifically, we will examine the effect of cooling and  
@@
 
-%Their relevance to PPDs are described below.
-
-%stuff aboout gravito turbulence 
-%Th
-%we focus on cooling and viscosity but also do 3D 
-%but real disks have cooling and viscosity/turbulence
-%(e.g. gravito-turbulent disks)
-
 \subsection{Cooling-driven gravitational instability}
-A net thermal loss\footnote{Isothermal disks implicitly assume 
-  heating and cooling processes are exactly balanced at all
-  times, so these do not have cooling in the current context.} (hereafter `cooling')
-reduces pressure (thermal) support against self-gravity. The cooling time 
+Cooling reduces pressure (thermal) support against self-gravity. The cooling time 
 $\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
@@
 
-The Toomre $Q$ parameter does not include cooling, and thus cannot be
-formally applied to assess GI in realistic disks. Indeed, there exists
-a separate, \emph{empirical} cooling time criterion which governs
-stability against disk fragmentation (see below), but this is not well
-understood.
+%The Toomre $Q$ parameter does not include cooling, and thus cannot be
+%formally applied to assess GI in realistic disks. Indeed, there exists
+%a separate, \emph{empirical} cooling time criterion which governs
+%stability against disk fragmentation (see below), but this is not well
+%understood.
 
-We shall quantify how cooling removes pressure
-support to enable GI even when $Q>1$. 
+We shall quantify how cooling removes pressure support to enable GI
+even when $Q>1$.
+
 We will find that cooling in fact changes the 
@@
 
-\subsection{Relevance to gravito-turbulent disk fragmentation}%and ring
-                                %formation 
-%can the continous formation/destruction of clumps (as seen in sims)
-%be interpreted as `overstability'? 
-%for overstable modes over-density oscilates 
+\subsection{Relevance to gravito-turbulent disk fragmentation}
+
+% Then the disk temperature will be lowered 
+% until $Q\sim 1$, whence non-axisymmetric modes grow 
+% and heats the disk through dissipation of spiral shocks. This may 
+% lead to a quasi-steady turbulent state with $Q\sim 1$ in
+% which cooling is balanced by shock heating to maintain thermal
+% equilibrium \citep{gammie01}.
+
 The two effects described above are immediately relevant to

### 2016-01-22 paper/paper3d.tex
re-writing intro

@@
 \shorttitle{Non-ideal Gravitational instability}
-\shortauthors{M-K.\ Lin \& K. Kratter}
+\shortauthors{M.-K.\ Lin \& K. Kratter}
 
@@
 unstable when the Toomre condition, $Q<1$, is satisfied.
-Non-axisymmetric modes can develop for  larger but still order-unity
+Non-axisymmetric modes can develop for larger, but still order-unity
 values of $Q$ . {\bf ref} 
@@
 
-It is important to remember that the dispersion relation
-Eq. \ref{classic_dr}  is derived from \emph{idealized} conditions: 
+It is important to remember that the popular dispersion relation, 
+Eq. \ref{classic_dr},  is derived from \emph{idealized} conditions: 
 the disk is laminar and does not experience any net thermal
@@
 Despite this, the Toomre parameter, and its associated
-dispersion relation, are still widely applied to realistic disk models
+dispersion relation, are still widely applied to accretion disk models
 with viscosity and cooling \citep[e.g.][]{cossins09, kimura12}, 
@@
 analytic treatment of disk GI to allow a self-consistent discussion of
-GI in realistic accretion disk models. Specifically, we consider the non-ideal
-effects of viscosity and cooling on GI.
+GI in realistic accretion disk models. Specifically, we will examine how the
+effects of cooling and viscosity can lead to GI by removing pressure
+and rotational support, respectively, even when the Toomre condition
+for instability is not satisfied.
 
@@
 $\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
-to some pre-defined value $T_0$, which may be zero. That is, $\p_t T =
+to some floor value $T_0$, which may be zero. That is, $\p_t T =
 -(T-T_0)/\tcool$. It is often written as 
@@
 
-We shall quantify how cooling removes pressure support to enable GI
-even when $Q>1$.
+We will quantify how cooling removes pressure support to enable GI
+even when $Q>1$. For example, if perturbations are
+cooled towards zero tempertaure, we find the above dispersion relation is
+modified to read 
+\begin{align}\label{cool_dr}
+s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
+  \tcool s}\right)c_s^2k^2,
+\end{align}
+where $\gamma$ is the adiabatic index. This equation shows that
+cooling increases the growth rate by reducing the magnitude of the
+pressure term. In fact, Eq. \ref{cool_dr} implies that instability is
+possible for any finite $Q$. This shows that cooling can change the
+qualitative nature of GI compared to the classic Toomre condition.
 
-We will find that cooling in fact changes the 
-qualitative nature of GI, in that its mere presence can
-formally lead to instability.
+%Eq. \ref{cool_dr} is a cubic in $s$ and differs qualitatively from the
+%classic result, Eq. \ref{classic_dr}, which is quadratic in $s$. In
+%fact, 
+%We will find that cooling in fact changes the 
+%qualitative nature of GI, in that its mere presence can
+%formally lead to instability.
 
@@
 %there is also viscous damping which stabilizes
-Viscous disks can develop GI even when $Q>1$. This is because
-viscosity compromises rotational support against self-gravity for
-long-wavelength disturbances
+Viscous disks can develop GI even when $Q>1$. This is because, as demonstrated below, viscosity compromises rotational support against
+self-gravity for long-wavelength disturbances
 \citep{lynden-bell74,willerding92,gammie96}. A similar effect   
@@
 effect.
+
+\begin{align}
+s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
+  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}
+\end{align}
+The condition for instability,  
+%\begin{align}
+$  2\pi G \Sigma > c_s^2|k|$, is identical that obtained from 
+Eq. \ref{classic_dr} with 
+%\end{align}
 

### 2016-01-22 paper/appendix3d.tex
intro version 2

@@
 \end{align}
-which coincides with \citeauthor{gammie96}'s Eq. 18 for $\gamma=1$. In
-that case the condition for instability is
-\begin{align}
-%  \frac{2|K|}{Q} - K^2>0. 
-  |K|< \frac{2}{Q}.
-\end{align} 
-This is the same stability condition as if rotation was neglected in
-the standard, isothermal dispersion relation for density waves
-(obtained from Eq. \ref{thindisk} by setting $\gamma=\theta=1$). Thus
-the effect of viscosity is to remove 
-rotational stabilization against self-gravity, as originally explained
-by \cite{lynden-bell74}.
+which coincides with \citeauthor{gammie96}'s Eq. 18 for $\gamma=1$.
+
+%In
+%that case the condition for instability is
+%\begin{align}
+%%  \frac{2|K|}{Q} - K^2>0. 
+%  |K|< \frac{2}{Q}.
+%\end{align} 
+%This is the same stability condition as if rotation was neglected in
+%the standard, isothermal dispersion relation for density waves
+%(obtained from Eq. \ref{thindisk} by setting $\gamma=\theta=1$). Thus
+%the effect of viscosity is to remove 
+%rotational stabilization against self-gravity, as originally explained
+%by \cite{lynden-bell74}.
 

### 2016-01-22 paper/paper3d.tex
intro version 2

@@
 or turbulent viscosity is too large, resulting in disk fragmentation.        
-%We quantify the interpretation of this result as the
-%compromise of gravitational stability by: 
-We quantify the interpretation of this result as a 
-secondary gravitational instability due to: 
+% `alternative' or `physical' here?
+%quantify or develop
+We motivate and quantify an alternative interpretation of disk
+fragmentation as the inability to maintain gravito-turbulence due to  
+secondary gravitational instabilities driven by:  
 \begin{inparaenum}[1)] 
@@
 We derive a general dispersion relation for linear waves in a 
-self-gravitating, viscous, non-adiabatic disk with internal and
-external heating, and cooling. We 
+self-gravitating, viscous, non-adiabatic disk with internal heating,
+external irradiation, and cooling. We  
 consider parameterized cooling functions in two and three dimensions,
-as well as full radiative diffusion in 3D. 
+as well as full radiative diffusion in 3D. We show that 
+cooling/viscosity can destabilize the disk even when the classic
+Toomre stability condition is satisfied; and generally there is
+no critical cooling rate/viscosity below which the disk is formally
+stable. %there is no special/precise cooling/visc that marks the
+        %stable/unstable boundary
 We apply linear theory to 
@@
 gravito-turbulence modeled as an effective 
-viscosity and cooling via dust opacity. We find the sole effect of cooling is insufficient to
-destabilize a realistic PPD. However, inclusion of viscosity renders
-the disk beyond $\sim 60$AU dynamically unstable on radial
-%this is for Mdot = 1d-6  
+viscosity and cooling via dust opacity. 
+%We find the sole effect of
+%cooling is insufficient to 
+%destabilize a realistic PPD. 
+We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
+unstable on radial %this is for Mdot = 1d-6  
 lengthscales a few times the local disk thickness. This is coincident 
@@
 maximum sustainable stress. 
-We suggest turbulent stresses play an active role in realistic disk
+We suggest turbulent stresses can play an active role in realistic disk 
 fragmentation by removing rotational stabilization against 
-self-gravity. 
+self-gravity. {\bf shorten abstract more?}
 \end{abstract}
@@
 adiabatic (including isothermal) disk. Such disturbances are formally
-unstable when the Toomre condition, $Q<1$, is satisfied.
-Non-axisymmetric modes can develop for larger, but still order-unity
-values of $Q$ . {\bf ref} 
+unstable when $Q<1$. Non-axisymmetric modes can develop for larger,
+but still order-unity values of $Q$
+\citep{papaloizou89,papaloizou91}. 
 %strictly speaking non axi *local* modes are stable -> only transient
@@
 %non-axi GI at slightly larger Q, but still characterized by Q
-
-%qualitative intro to gravito-turbulence 
-%role of cooling and resulting viscosity
 
@@
 (hereafter `cooling').  
-However, real disks can cool, for example,
+However, real disks can cool, for example, %radiative, convective cooling
 due to radiative losses. Accretion disks may also be turbulent, for
@@
 effective viscosity \citep{shakura73,lin87,rafikov15}. {\bf 
-  more refs here? non-sg refs} 
+  more refs here?} 
 %in terms of AM transport and heating 
@@
 
-Despite this, the Toomre parameter, and its associated
-dispersion relation, are still widely applied to accretion disk models
-with viscosity and cooling \citep[e.g.][]{cossins09, kimura12}, 
+Despite this, the Toomre parameter and its associated
+dispersion relation are still widely applied to viscous, cooling
+accretion disk models \citep[e.g.][]{cossins09, kimura12}, 
 even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
@@
 analytic treatment of disk GI to allow a self-consistent discussion of
-GI in realistic accretion disk models. Specifically, we will examine how the
-effects of cooling and viscosity can lead to GI by removing pressure
-and rotational support, respectively, even when the Toomre condition
-for instability is not satisfied.
+GI in realistic accretion disk models. Specifically, we will examine
+how  cooling and viscosity can lead to GI even when $Q>1$, by 
+removing pressure and rotational support, respectively.
 
@@
 
-\subsection{Cooling-driven gravitational instability}
-Cooling reduces pressure (thermal) support against self-gravity. The cooling time 
+\subsection{Cooling-driven gravitational instability}\label{cool_gi}
+Cooling reduces pressure support against self-gravity. The
+cooling time  
 $\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
@@
 
-%The Toomre $Q$ parameter does not include cooling, and thus cannot be
-%formally applied to assess GI in realistic disks. Indeed, there exists
-%a separate, \emph{empirical} cooling time criterion which governs
-%stability against disk fragmentation (see below), but this is not well
-%understood.
-
 We will quantify how cooling removes pressure support to enable GI
@@
 pressure term. In fact, Eq. \ref{cool_dr} implies that instability is
-possible for any finite $Q$. This shows that cooling can change the
-qualitative nature of GI compared to the classic Toomre condition.
+possible for any finite $Q$. %any finite pair of Q and beta 
+This shows that cooling can change the
+qualitative nature of GI compared to the simple Toomre condition.
 
-%Eq. \ref{cool_dr} is a cubic in $s$ and differs qualitatively from the
-%classic result, Eq. \ref{classic_dr}, which is quadratic in $s$. In
-%fact, 
-%We will find that cooling in fact changes the 
-%qualitative nature of GI, in that its mere presence can
-%formally lead to instability.
+%current understanding of frag is all due to cooling (viscosity is
+%just a reflection of cooling)
 
-\subsection{Viscosity-driven gravitational instability}
+\subsection{Viscosity-driven gravitational instability}\label{visc_gi}
 %there is also viscous damping which stabilizes
-Viscous disks can develop GI even when $Q>1$. This is because, as demonstrated below, viscosity compromises rotational support against
+Viscous disks can also develop GI even when $Q>1$. This is because, as 
+demonstrated below, viscosity compromises rotational support against  
 self-gravity for long-wavelength disturbances 
@@
 provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
-is a mechanism to enhance particle clumping for planetesimational formation 
-%frag of dust 
+is a mechanism to enhance particle clumping for planetesimational
+formation %frag of dust 
 \citep{youdin11}. Then it is not unreasonable to expect an analogous
@@
 
-For gaseous accretion disks, a Navier-Stokes viscosity is often used to 
-to mimic hydrodynamic or magneto-hydrodynamic turbulence
-\citep{shakura73}. This involves writing the
-kinematic viscosity $\nu$ as  
+It is conventional to write the kinematic viscosity $\nu$ as   
 \begin{align}\label{alpha_param}
@@
 fluid variables, but Eq. \ref{alpha_param} is the general form. 
+{\bf maybe no need to intro alpha param here?}
 
-How well turbulence can be modeled as an effective viscosity is a
-separate issue \citep{balbus99}. However, it is  
-reasonable to assume that turbulence frustates rotational 
-support. (This is, after all, the physical cause for turbulent mass
-accretion.) In this work, we use viscosity to represent this physical
-effect.
-
+For an isothermal, viscous, self-gravitating disk in 2D,
+\cite{gammie96} finds the approximate disperion relation   
 \begin{align}
 s \simeq \frac{\nu k^2\left(2\pi G\Sigma |k| -
-  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}
+  c_s^2k^2\right)}{\kappa^2 + c_s^2k^2 - 2\pi G\Sigma|k|}. 
 \end{align}
-The condition for instability,  
-%\begin{align}
-$  2\pi G \Sigma > c_s^2|k|$, is identical that obtained from 
-Eq. \ref{classic_dr} with 
-%\end{align}
+Assuming $Q>1$, instability occurs if 
+\begin{align*}
+  2\pi G \Sigma > c_s^2|k|,
+\end{align*}
+provided that $\nu k^2\neq0$. The above condition for viscous GI is 
+identical that would be obtained from Eq. \ref{classic_dr} with
+rotation neglected ($\Omega\to 0$). That is, a classically stable
+disk can be made unstable by viscosity as it effectively removes
+rotational stabilization \citep{lynden-bell74}.
 
-Viscosity is absent in the standard Toomre parameter. The above
-studies account for viscosity but consider isothermal disks. On the
-other hand, proper inclusion of viscosity should also account for
-viscous heating which, as discussed above, also affect stability. 
-We thus extend the study of viscous GI to include an energy equation
-with viscous heating and explicit cooling. This allows us to consider
-GI in realistic PPDs.
+%turbulence <-> viscosity assumption 
+For gaseous accretion disks, a Navier-Stokes viscosity is often
+implemented (as above) to mimic hydrodynamic or magneto-hydrodynamic
+(MHD) turbulence \citep{shakura73}. How well turbulence can be modeled
+as an effective viscosity is a separate issue 
+\citep{balbus99}. However, it is reasonable to assume that turbulence
+frustates rotational support. (This is, after all, the physical
+mechanism to turbulent mass accretion.) 
+For example, \cite{fromang05} found numerical evidence that MHD
+turbulence can enhance GI and lead to disk fragmentation (see also
+\S\ref{MHD}).
+
+In this work, we use viscosity to model such a 
+physical effect due to turbulence, but we note that viscous 
+GI does not assume any origin for the viscosity. We will extend theory
+of viscous GI to include an energy equation  with viscous heating,
+irradiation, explicit cooling, as well as three-dimensionality. This
+allows us to consider viscous GI in more realistic PPD models.
 
@@
 \subsection{Relevance to gravito-turbulent disk fragmentation}
+We develop a general framework for viscous disks without  
+assuming a specific origin of turbulence that this viscosity 
+represents. We will, however, apply the theory to `gravito-turbulent'
+disks in which the viscosity associated with some underlying (classic)
+GI, as described below.
 
-% Then the disk temperature will be lowered 
-% until $Q\sim 1$, whence non-axisymmetric modes grow 
-% and heats the disk through dissipation of spiral shocks. This may 
-% lead to a quasi-steady turbulent state with $Q\sim 1$ in
-% which cooling is balanced by shock heating to maintain thermal
-% equilibrium \citep{gammie01}.
+Consider a disk initially laminar without a heating mechanism and
+cools. Then the disk temperature will be lowered until $Q=O(1)$, whence
+non-axisymmetric modes grow and heats the disk through the dissipation
+of spiral shocks. %new rafikov 16 
+This permits a quasi-steady, turbulent state with
+$Q =O(1)$ in which cooling is balanced by shock heating to maintain
+thermal equilibrium \citep{gammie01}. 
+The dynamic and thermodynamic effect of this turbulence will be
+represented by a viscosity. Then the $\alpha$ and $\beta$ parameters
+defined above will be inversely related.
 
-The two effects described above are immediately relevant to
-`gravito-turbulence'. This is a steady disk state in which cooling is
-balanced by turbulent heating, itself caused by GI. In this case 
-the $\alpha$ and $\beta$ parameters are inversely related
-\citep{gammie01}.
-
-Experimentally, however, it is found that if $\beta$ ($\alpha$) is
-too small (large), 
-\begin{align}
+Experimentally, however, it is found that if the cooling rate or
+viscosity is too large, say,   
+\begin{align}\label{frag_cond}
   \beta < \beta_\mathrm{c} \quad
@@
 \end{align}
-then the disk fragments \citep{gammie01,rice05,rice11}. 
+then the disk fragments \citep{gammie01,rice05,rice11}. In fact, in
+the absence of global effects, numerical simulations show there are
+two --- and only two --- possible outcomes for self-gravitating,
+cooling disks: gravito-turbulence or fragmentation.  
+%more accurately: given a total simulation time, there are two
+%outcomes 
+%sufficiently long sims -> always frag?
+
 There is considerable debate on the exact value of $\beta_\mathrm{c}$ 
-and whether or not a critical cooling time can 
-be defined at all \citep{meru11,meru12,paardekooper12,hopkins13}. However, it is
-generally accepted that \emph{steady, gravito-turbulent disks do not
-  exist for sufficiently rapid cooling}.
+and whether or not a critical cooling time can be defined at all 
+\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. 
+%mention numerical effects? 
+However, it is generally accepted that \emph{steady, gravito-turbulent
+  disks do not  
+  exist for sufficiently rapid cooling or large viscosity}.  
+This intriguing because, as highlighted in
+\S\ref{cool_gi}---\ref{visc_gi}, these effects can reduce
+gravitational stability. These considerations motivate a physical    
+interpretation of disk fragmentation as the inability to maintain a
+gravito-turbulent state due to the action of cooling and/or viscosity.
 
-The non-existence of steady gravito-turbulence for rapid cooling, or
-equivalently large viscosity, is intriguing because, as discussed
-above, these effects reduce pressure and rotational support 
-against self-gravity, respectively.
-
-In this paper, we consider the stability of PPDs subject to 
-gravito-turbulence modeled as an effective viscosity. 
-We perform standard stability analyses with explicit inclusion of the
-relevant physics. 
-We will find that the
-regions of a gravito-turbulent PPD that is dynamically unstable,
-agrees with the current consensus on where, and the conditions under
-which, a PPD is expected to fragment. This motivates an alternative 
-interpretation of disk fragmentation as the inability to maintain a
-stable, gravito-turbulent state due to the action of cooling and
-viscosity.
-
-%We do not attempt to model disk fragmentation from first
-%principles. Instead, we simply ask: \emph{when does a 
-%  gravito-turbulent disk become dynamically unstable?} 
-%We will find the answer to be consistent with the current consensus on
-%disk fragmentation.
-
-%viscous gi speculated by gammie 01 but now found in those sims. but
-%longer sims of the same system by paardekooper eventually fragment
-
-%important to note that this is an experimental results. we refrain
-%from using input from numerical simulations in our modeling 
-%because we want to explain simulations 
-%we will find the regimes where viscous, cooling PPDs are dynamically
-%unstbale coincides with where fragmentation occurs 
-% we find instability formally possible for long cooling but takes a
-% long time
+Indeed, we will find that the region where a gravito-turbulent PPD is
+dynamically unstable agrees with the current consensus on  where, and
+the conditions under which, a PPD is expected to fragment based on
+Eq. \ref{frag_cond}, with typical critical values cited in the
+literature \citep[e.g. $\beta_c\sim
+  3,\,\alpha_c\sim0.1$;][]{gammie01,rice05}.
 
@@
 
-The basic equations are listed in \S\ref{basic}, along with
-descriptions of the disk 
-equilibra and the two cooling models 
-considered --- parametrized `beta' cooling and radiative diffusion. 
+The basic equations, disk equilibria, and the two cooling models  
+considered --- parametrized `beta' cooling and radiative diffusion, 
+are given in \S\ref{basic}. 
 The linear stability problem is defined in
@@
 respectively. %key result: generalized dispersion relation 
-In
-\S\ref{2dppd} we apply linear theory to PPDs with realistic cooling
-models, including radiative diffusion in 3D. We find regions in a
-PPD that cannot sustain steady gravito-turbulence to coincide with
-that expected for disk fragmentation, based on previous empirical
-criteria. We summarize our results in \S\ref{summary}, including a
-discussion of how our model may aid the physical understanding of  
-fragmentation in realistic PPDs. 
-% as the inability to sustain gravito-turbulence
-%because the associated stress removes rotational support
+In \S\ref{2dppd} we consider PPDs with realistic viscosity/cooling   
+models, including radiative diffusion in 3D, and determine where such
+disks are gravitationally unstable. We summarize our results in
+\S\ref{summary} with a discussion of how our models may aid the 
+physical understanding of fragmentation in realistic PPDs.
 
@@
 %S. Stahler 
+%A. Youdin
 

### 2016-01-22 paper/summary.tex
intro version 2

@@
 
-\subsection{MHD turbulence}%turbulence and viscosity 
+\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
 It is important to note that while we consider gravito-turbulent disks  

### 2016-01-23 paper/model3d.tex
reg update
model version 2

@@
 \section{Basic equations}\label{basic}
-We consider a three-dimensional (3D), self-gravitating,
+We consider a 3D, self-gravitating,
 viscous disk with heating and cooling. We use the shearing box framework to
@@
   & \frac{\p E}{\p t} + \nabla\cdot\left(E\bm{v}\right) = - P
-  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} +
-  \mathcal{H}_\mathrm{ext} - \Lambda,\label{energy_eq}
+  \nabla\cdot\bm{v} + \mathcal{H}_\mathrm{visc} - \Lambda +
+  \mathcal{H}_\mathrm{ext} ,\label{energy_eq}
 \end{align} 
@@
 \end{align}
-where $\gamma$ is the adiabatic index, $\mathcal{R}$ is the gas
-constant, and $T$ is the temperature. We use $\gamma=1.4$ unless
-otherwise stated. The gas 
+where $\mathcal{R}$ is the gas constant and $T$ is the temperature. We
+use $\gamma=1.4$ unless otherwise stated. The gas 
 gravitatational potential $\Phi$ is given via the Poisson equation,  
 \begin{align}
- \nabla^2\Phi = 4 \pi G \rho, 
+ \nabla^2\Phi = 4 \pi G \rho. 
 \end{align}
-where $G$ is the gravitational constant.
+%where $G$ is the gravitational constant.
 
@@
 In the energy equation (Eq. \ref{energy_eq}) the source terms 
-$\mathcal{H}_\mathrm{visc}$, $\mathcal{H}_\mathrm{ext}$ and $\Lambda$
-represent viscous heating, any time-independent heat
-source/sinks, and time-dependent thermal losses, respectively. These
-are described below. We normally set
-$\mathcal{H}_\mathrm{ext}=0$ unless otherwise stated.
+$\mathcal{H}_\mathrm{visc}$ and $\Lambda$
+represent viscous heating and time-dependent cooling, 
+respectively, and $\mathcal{H}_\mathrm{ext}$ represents any
+time-independent heat source/sinks. We normally set
+$\mathcal{H}_\mathrm{ext}=~0$ unless otherwise stated.
 
@@
 %bulk visc ignored 
-where $\rho\nu$ is the shear viscosity, and we include a bulk
-viscosity $\rho\nu_b$ for completeness.  
-The associated viscous heating given by
+where $\rho\nu$ is the shear viscosity. We also include a bulk
+viscosity $\rho\nu_b$ for completeness, but will neglect it in
+numerical calculations. The associated viscous heating given by 
 \begin{align}
@@
 We adopt the viscosity law  
-\begin{align}
+\begin{align}\label{visc_law}
   \nu = \alpha
@@
 where subscript `eq' denotes the equilibrium state and  
-%$c_{s0}$ is a modified sound-speed defined below. 
-$c_{s0}^2\equiv P_\mathrm{eq}(0)/\rho_\mathrm{eq}(0)$. 
-The dimensionless coefficient
+$c_{s0}^2\equiv P_\mathrm{eq}(z=0)/\rho_\mathrm{eq}(z=0)$. 
+The dimensionless viscosity coefficient 
 $\alpha=~\alpha(\rho_\mathrm{eq},P_\mathrm{eq})$ characterizes the
@@
 behaves in the perturbed state.  
-%Given the uncertain
-%properties of gravito-turbulence, we will treat $\mu,\lambda$ as free
-%parameters. 
 We adopt the same prescription for the bulk viscosity but
@@
 Our numerical calculations use $\mu=-1,\lambda=0$ so 
-that $\rho\nu$ is time-indpendent, as adopted in previous 
+that $\rho\nu$ is time-indpendent, folling previous 
 studies of viscous GI  
@@
 
-While steady-state viscosity values can be determined analytically
-\citep[e.g.][]{rafikov15}, its time-dependent behavior is not
-well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made to
-bring out the physically relevant process --- viscosity-assisted
-gravitational instability. %expect to be generally present, but will
+While steady-state viscosity values can be determined 
+analytically \citep[e.g.][]{rafikov15}, its time-dependent behavior is
+not well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made
+to bring out the physical process of interest --- viscous GI. %expect
+                                %to 
+%be generally present, but will 
 %be obscured by viscous over-stability with
@@
 Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
-dependence when modeling the evoluting of self-gravitating disks 
-as a viscous process. Furthermore, \cite{paardekooper12} measured
-an increase in the effective viscosity when their gravito-turbulent
-disk models underwent fragmentation (i.e. as $\Sigma$ increases).
+dependence when modeling the evoluting of 2D self-gravitating disks 
+as a viscous process. We note that if $\rho\nu$ is constant in time
+then increasing the density leads to reduction in the viscosity. 
+This is perhaps not inconsistent with numerical simulations of disk
+fragmentation which show the internal structure of clumps, which
+represent increased density, is laminar \citep{gammie01}.  
+{\bf is this true?}
 
+%Furthermore, \cite{paardekooper12} measured
+%an increase in the effective viscosity when their gravito-turbulent
+%disk models underwent fragmentation (i.e. as $\Sigma$ increases).  
 %{\bf note paardekooper 2012: alpha goes down as Sigma goes up}
@@
 We consider equilibrium solutions (here omitting the `eq' subscripts
-for clarity) 
+for simplicity)  
 \begin{align}
@@
 \end{align} 
-%and assume a polytropic relation for the equilibrium pressure,
-%\begin{align}
-%  P  =
-%  \frac{c_{s0}^2\rho_0}{\Gamma}\left(\frac{\rho}{\rho_0}\right)^\Gamma
-%  \equiv \frac{c_s^2(z)}{\Gamma}\rho,
-%\end{align}
-%where $\rho_0$ is the equilibrium midplane density and $\Gamma$ is the
-%constant polytropic index.  
 The equilibrium density and pressure fields are obtained by solving
@@
 \end{align}
-%A numerical solution is usually required to obtain the equilibrium
-%state, except for special values of $\Gamma$. 
 together with thermodynamic equilibrium,
-%\subsection{Thermal equilibrium}
-%In equilibrium the energy equation implies
 \begin{align}\label{thermal_eq}
-%  \mathcal{H}_\mathrm{visc} + \mathcal{H}_\mathrm{ext} = \Lambda.   
 (q\Omega)^2\rho\nu + \mathcal{H}_\mathrm{ext} = \Lambda,
@@
   %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
-  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
+  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}. 
 \end{align}
-where 
-\begin{align}
-  \tcool = \beta\Omega^{-1} 
-\end{align}
-is the cooling timescale with $\beta$ a constant input parameter, and
+Recall from Eq. \ref{beta_def} that 
+%\begin{align}
+  $\tcool = \beta\Omega^{-1}$ is 
+%\end{align}
+the cooling timescale with $\beta$ a constant input parameter. Here, 
 \begin{align}
@@
 \end{align}
-is a reference temperature field, here parameterized as a constant fraction
-$\theta$ of the equilibrium temperature $T_\mathrm{eq}$. We will refer
-to $\theta$ as the `irradiation level'.  
-%This cooling function restores 
-%the disk temperature to $\tirr$ over a time-scale $\tcool$.
+is a reference temperature field parameterized as a constant 
+fraction $\theta$ of the equilibrium temperature
+$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
+for example, stellar irradiation. We will thus refer to $\theta$ as
+the `irradiation level'. %stellar illum or env rad
 
-For models with beta cooling we assume an equilibrium polytropic relation
+We will formulate the linear problem with beta cooling using the
+functional form for $\Lambda$ as given by
+Eq. \ref{beta_cool}. However, any explicit cooling function, say
+$\Lambda=\Lambda_1(\rho,T)$, can be mapped to that obtained from
+Eq. \ref{beta_cool} by equating their linearized forms, i.e. $\delta
+\Lambda \equiv \delta \Lambda_1$, which then defines the $\beta$ and
+$\theta$ parameters to be used in the framework we develop. This is
+done in \S\ref{2dppd} where we adopt a more realistic beta cooling 
+prescription for 2D PPDs. In this case, $\theta$ may or may not
+represent a physical irradiation.
+
+For models with beta cooling given by Eq. \ref{beta_cool} we assume an 
+equilibrium polytropic relation 
 \begin{align} 
@@
 where $\Gamma$ is the constant polytropic index and $\rho_0 =
-\rho(0)$. The vertical
-structure is first obtained from Eq. \ref{vert_eq1}---\ref{vert_eq2}
-then inserted into Eq. \ref{thermal_eq} to infer the required
-viscosity profile,
-
+\rho(z=0)$ is the equilibrium midplane density. 
+The vertical structure is first obtained from
+Eq. \ref{vert_eq1}---\ref{vert_eq2} then inserted into
+Eq. \ref{thermal_eq} to infer the required viscosity profile for
+thermal equilibrium,   
 \begin{align}\label{alpha_beta_relation}
@@
   rice05,baehr15}. It permits numerical experiments to be carried out in a
-controlled manner as a function of the cooling time $\beta$. 
-We remark that for the linear problem  
-any explicit cooling function $\Lambda=\Lambda(\rho,T)$ can be
-mapped to that obtained from the standard function in Eq.
-\ref{beta_cool}. We do this in \S\ref{2dppd} where we adopt a more
-realistic beta cooling prescription for 2D PPDs.
+controlled manner as a function of the cooling time $\beta$. An
+adiabatic disk corresponds to $\beta\to \infty$. The
+physical meaning of the limit $\beta\to0$ depends on $\theta$, as
+discussed in \S\ref{theta0}---\ref{theta1}.
 

### 2016-01-23 paper/paper3d.tex
model version 2

@@
 fragmentation by removing rotational stabilization against 
-self-gravity. {\bf shorten abstract more?}
+self-gravity. {\bf shorten abstract?}
 \end{abstract}
@@
 -(T-T_0)/\tcool$. It is often written as 
-\begin{align} 
+\begin{align} \label{beta_def}
   \tcool = \beta \Omega^{-1},
@@
 physical effect due to turbulence, but we note that viscous 
-GI does not assume any origin for the viscosity. We will extend theory
-of viscous GI to include an energy equation  with viscous heating,
-irradiation, explicit cooling, as well as three-dimensionality. This
+GI does not assume any origin for the viscosity. We will generalized
+the theory of viscous GI to include an energy equation  with viscous
+heating, irradiation, explicit cooling, as well as
+three-dimensionality (3D). This 
 allows us to consider viscous GI in more realistic PPD models.
@@
 
-Indeed, we will find that the region where a gravito-turbulent PPD is
+Indeed, we find that the region where a gravito-turbulent PPD is
 dynamically unstable agrees with the current consensus on  where, and
@@
 
-The basic equations, disk equilibria, and the two cooling models  
-considered --- parametrized `beta' cooling and radiative diffusion, 
-are given in \S\ref{basic}. 
-The linear stability problem is defined in
-\S\ref{linear}. We present results with beta cooling in
+The basic equations, disk equilibria, cooling and viscosity models are 
+given in \S\ref{basic}. The linear stability problem is defined in 
+\S\ref{linear}. We present results with parameterized `beta' cooling in 
 \S\ref{result_2d} and \S\ref{3ddisk} for 2D and 3D disks,

### 2016-01-23 paper/results2d.tex
reg update
model version 2

@@
 Routh-Hurwitz criteria imply that stability is ensured if 
-\begin{align}
+\begin{align}\label{stable_condition}
   \gamma > \theta \quad \text{and} \quad 
@@
 
-\subsubsection{$\theta = 0$}  
+\subsubsection{$\theta = 0$}\label{theta0}
 %Beta cooling with $\theta=0$ is commonly employed in 
@@
 
-\subsubsection{$\theta = 1$}
+\subsubsection{$\theta = 1$}\label{theta1}
 Beta cooling with $\theta=1$ corresponds to 

### 2016-01-23 paper/paper3d.tex
minor edits

@@
 dispersion relation are still widely applied to viscous, cooling
-accretion disk models \citep[e.g.][]{cossins09, kimura12}, 
+accretion disk models \citep[e.g.][]{cossins09, kimura12}, {\bf more refs?}
 even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
@@
 occurs in dusty fluids where the required frictional forces are 
-provided by dust-gas drag \citep{ward00, takahashi14}. In fact, this
+provided by dust-gas drag \citep{ward00, takahashi14}. 
+In fact, this
 is a mechanism to enhance particle clumping for planetesimational

### 2016-01-23 paper/results2d.tex
minor edits

@@
 $\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
-pressure, and set $\Gamma=1$ without loss of generality. 
+pressure, and set $\Gamma=1$. % without loss of generality. 
 The gravitational potential perturbation remains 3D and its midplane
@@
 where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}. 
+We use this generalized dispersion relation to investigate GI driven
+solely by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity
+in \S\ref{2dvisc}.
 
 \subsection{Inviscid limit}\label{2d_inviscid}
-Here we neglect viscosity by setting 
-$\alpha = \alpha_b = 0$. This eliminates viscous GI and 
-allows us to quantify the \emph{sole} effect of cooling on the
+We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
+eliminates viscous effects and allows us to quantify the 
+\emph{sole} effect of cooling on the 
 perturbations. An external, time-independent heat source should be
@@
                                 %as proxy for turbulence
-
-Eq. \ref{thindisk} simplifies to  
+                                %maybe turb heating actually behaves
+                                %like background heat source only 
+Eq. \ref{thindisk} becomes 
 \begin{align}\label{inviscid}
@@
 \end{align}
-similar to the classic Lin-Shu dispersion relation, which may be
-obtained by taking the limit $|\beta S|\to\infty$. 
-The first term on
-the right-hand-side represents destabilization by self-gravity; 
+similar to the classic Lin-Shu dispersion relation
+(i.e. Eq. \ref{classic_dr}), which may be obtained by taking the limit
+$|\beta S|\to\infty$. %with gamma=1
+The first term on the right-hand-side represents destabilization by self-gravity; 
 the second and third terms represent stabilization by rotation and

### 2016-01-23 paper/paper3d.tex
revising sec 4, minor edits to prev secs

@@
 cooling time  
-$\tcool$ is the time-scale over which the disk temperature $T$ is relaxed
-to some floor value $T_0$, which may be zero. That is, $\p_t T =
--(T-T_0)/\tcool$. It is often written as 
+$\tcool$ is the timescale over which the disk temperature $T$ is relaxed
+to some floor value $T_0$, which may be zero. 
+%That is, $\p_t T =
+%-(T-T_0)/\tcool$. 
+ It is often written as 
 \begin{align} \label{beta_def}
@@
 where $\beta$ is the corresponding dimensionless cooling time. This
-type of cooling function, first applied by \cite{gammie01}, is a
-simple parameterization that allows the full range of thermodynamic
-response to be explored. 
+type of parameterized cooling, first applied by \cite{gammie01},
+allows the full range of thermodynamic response to be explored. 
 {\bf physical origin of cooling in PPDs}
@@
 \end{align}
-where $\gamma$ is the adiabatic index. This equation shows that
-cooling increases the growth rate by reducing the magnitude of the
-pressure term. In fact, Eq. \ref{cool_dr} implies that instability is
-possible for any finite $Q$. %any finite pair of Q and beta 
-This shows that cooling can change the
-qualitative nature of GI compared to the simple Toomre condition.
+where $\gamma$ is the adiabatic index. Cooling increases the growth
+rate by reducing the magnitude of the 
+pressure term. In fact, for $t_c\in(0,\infty)$ the formal condition
+for instability is  
+\begin{align*}
+  2\pi G \Sigma |k| > \kappa^2,
+\end{align*}
+which is what would be obtained from Eq. \ref{classic_dr} with 
+pressure neglected ($c_s\to 0$). This condition does not actually
+depend on the cooling rate, and instability is
+possible for any finite $Q$. %at sufficiently small scales 
+This shows that the mere presence of
+cooling changes the qualitative nature of GI compared to the simple
+Toomre condition.   
+%{\bf numerical implications? if using standard beta cooling, then as
+%  long as it's there then sufficiently small scales will be
+ % unstbale. can access those scales with high resolution. defer to
+ % discussion section}
 
@@
 \end{align*}
-provided that $\nu k^2\neq0$. The above condition for viscous GI is 
+provided that $\nu k^2\neq0$. This condition for viscous GI is 
 identical that would be obtained from Eq. \ref{classic_dr} with

### 2016-01-23 paper/results2d.tex
revising sec 4, minor edits to prev secs

@@
 where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}. 
-We use this generalized dispersion relation to investigate GI driven
-solely by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity
+We use this generalized dispersion relation, which does not assume any
+$\alpha$-$\beta$ relation, %dispersion relation treats alpha/beta as
+                           %independent 
+to investigate GI driven
+by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity 
 in \S\ref{2dvisc}.
@@
 \emph{sole} effect of cooling on the 
-perturbations. An external, time-independent heat source should be
+perturbations. A time-independent heat source should be
 invoked to balance the imposed cooling to allow an equilibrium to be
@@
 similar to the classic Lin-Shu dispersion relation
-(i.e. Eq. \ref{classic_dr}), which may be obtained by taking the limit
-$|\beta S|\to\infty$. %with gamma=1
+(Eq. \ref{classic_dr}), which may be obtained by taking the limit
+$|\beta S|\to\infty$. %then setting gamma=1. other params fixed
+                      %(finite theta)
 The first term on the right-hand-side represents destabilization by self-gravity; 
 the second and third terms represent stabilization by rotation and
-pressure, respectively. The imposed cooling only affects the
-pressure response.
+pressure, respectively. The imposed cooling/irradiation only affects the
+pressure response. {\bf do we need to explain the terms again?}
 
@@
 \end{align}
-are both satisfied. Thus, whether or not the system is stable is
-independent of the cooling rate $\beta$ (provided it is non-zero and
-finite). The system becomes more unstable with decreasing $\theta$.
+are both satisfied. If only real growth rates are considered, then the
+second condition alone is sufficient for stability. %from property of
+                                %cubic crossing the real axis 
+%gamma/theta condition is related to some kind of thermal instability
+%oscilitory convection?? 
+%second condition concerning Q is associated with self-g 
+Whether or not the system is stable does not 
+depend on the actual cooling time $\beta$ (provided it is non-zero and
+finite). The system becomes more unstable with decreasing 
+irradiation levels $\theta$, i.e. as the floor temperature, towards which
+perturbations are cooled, drops.
 
 Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p |K| =
-0$ and $S = 
-S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
+0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
 \begin{align}\label{kstar}
@@
 %numerical simulations of self-gravitating disks \citep{gammie01}. 
-For beta cooling with $\theta=0$, the disk is unconditionally
-unstable for finite $Q$. The most unstable wavelength decreases with
+For beta cooling with $\theta=0$, the disk is unconditionally unstable
+for finite $Q$: instability is always possible at sufficiently small
+scales (recall the discussion in \S\ref{cool_gi}). At fixed $Q$,  
+the most unstable wavelength decreases with
 decreasing $\beta$. The $\beta\to0$ limit corresponds to a

### 2016-01-24 paper/model3d.tex
minor edits
halfway through version 2 of 2d results

@@
 where the first term represents viscous heating. For the viscous
-problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$. However,
+problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$ and
+obtain a relation between $\alpha$ and $\Lambda$
+(e.g. Eq. \ref{alpha_beta_relation} below). However,
 if we wish to neglect viscosity but include cooling, we must invoke
@@
 
-We will formulate the linear problem with beta cooling using the
-functional form for $\Lambda$ as given by
-Eq. \ref{beta_cool}. However, any explicit cooling function, say
+We shall formulate the linear problem with beta cooling of the
+functional form given by Eq. \ref{beta_cool}. However, any explicit cooling function, say
 $\Lambda=\Lambda_1(\rho,T)$, can be mapped to that obtained from
@@
 Eq. \ref{thermal_eq} to infer the required viscosity profile for
-thermal equilibrium,   
+thermal equilibrium. If  $\mathcal{H}_\mathrm{ext}=0$, 
 \begin{align}\label{alpha_beta_relation}
@@
 \end{align}
+%This relation applies if 
+%For generality, we will formulate the linear problem without assuming
+%Eq. \ref{alpha_beta_relation} (i.e. treat $\alpha$ and $\beta$ as
+%independent) and only apply it in numerical
+%calculations.
+
+%{\bf prob formulation treats alpha-beta independent. this relation
+%  only needed if hext is zero}
 
@@
 for optically-thin disks. However, this complication is beyond the scope of
-this work.
+this work. {\bf ref for flux lim diff?}

### 2016-01-24 paper/results2d.tex
halfway through version 2 of 2d results

@@
 with previous studies. The disk
-material is assumed to be confined to the midplane so $\delta v_z=0$. 
+material is assumed to be confined to the midplane and $\delta v_z=0$. 
 We make the replacement  
@@
 pressure, and set $\Gamma=1$. % without loss of generality. 
+%big gamma only controls vertical structure
+%can absorb big gamma into def of cs
 The gravitational potential perturbation remains 3D and its midplane
@@
 \end{align}
-where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}. 
-We use this generalized dispersion relation, which does not assume any
-$\alpha$-$\beta$ relation, %dispersion relation treats alpha/beta as
-                           %independent 
-to investigate GI driven
+where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp},
+which do not impose any $\alpha$-$\beta$ relation. We use this generalized
+dispersion relation  to investigate GI driven 
 by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity 
@@
 We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
-eliminates viscous effects and allows us to quantify the 
+eliminates viscous heating and forces, allowing us to quantify the 
 \emph{sole} effect of cooling on the 
@@
                                 %like background heat source only
+
 Eq. \ref{thindisk} becomes 
@@
 \end{align}
-similar to the classic Lin-Shu dispersion relation
+similar to the classic dispersion relation
 (Eq. \ref{classic_dr}), which may be obtained by taking the limit
@@
 \end{align}
-are both satisfied. If only real growth rates are considered, then the
-second condition alone is sufficient for stability. %from property of
-                                %cubic crossing the real axis 
+are both satisfied. 
 %gamma/theta condition is related to some kind of thermal instability
@@
 Whether or not the system is stable does not 
-depend on the actual cooling time $\beta$ (provided it is non-zero and
-finite). The system becomes more unstable with decreasing 
-irradiation levels $\theta$, i.e. as the floor temperature, towards which
-perturbations are cooled, drops.
+depend on the cooling time $\beta\in(0,\infty)$, 
+but on the irradiation level $\theta$.  
+The disk becomes more unstable with decreasing $\theta$, i.e. as the
+floor temperature, towards which perturbations are cooled, drops.
 
-Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p |K| =
+If only real growth rates are considered, then violating the second
+condition in Eq. \ref{stable_condition} alone is sufficient for instability. 
+%from property of cubic crossing the real axis 
+In that case the wavenumbers satisfying
+\begin{align}
+  \theta K^2 - \frac{2}{Q}|K| + 2(2-q) < 0 
+\end{align}
+are unstable. Thus the range of unstable wavenumbers increases with
+decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
+$(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
+upper limit to unstable wavenumbers, which could have important
+implications for numerical simulations (see \S\ref{prev_works}).
+
+Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p|K| =
 0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
@@
 \end{align}
-General solutions to this cubic for $S_*$ are unwieldy, but simple
-results may be obtained in special cases, discussed next.
+Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
+S_*<0 $ for $\gamma>\theta$. That is, as cooling becomes more rapid
+the disk becomes more unstable at shorter wavelengths. 
+%show dK*/dbeta < 0 
+%by diff K* wrt beta and diff S* wrt beta from the equations
+%and elim dS*/d beta
+
+Next, we consider two special cases: $\theta=0$, so perturbations are
+cooled towards zero temperature \citep[as typically employed in
+numerical simulations, e.g.][]{gammie01}; and $\theta=1$, where the
+equilibrium disk temperature equals the irradiation temperature.
 
 \subsubsection{$\theta = 0$}\label{theta0}
-%Beta cooling with $\theta=0$ is commonly employed in 
-%numerical simulations of self-gravitating disks \citep{gammie01}. 
 For beta cooling with $\theta=0$, the disk is unconditionally unstable
-for finite $Q$: instability is always possible at sufficiently small
-scales (recall the discussion in \S\ref{cool_gi}). At fixed $Q$,  
-the most unstable wavelength decreases with
-decreasing $\beta$. The $\beta\to0$ limit corresponds to a
+for finite $Q$.  The $\beta\to0$ limit corresponds to a 
 pressureless disk (not merely isothermal).
 
+%for definiteness 
 Let us consider a disk with 
@@
 $\theta=0$, we find 
-%In this case the coefficient of 
-%$S_*$ in Eq. \ref{inviscid_max} vanish, so that  
 \begin{align}\label{sstar}
@@
 The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
-increases with decreasing $\beta$. Instability does not require the
-cooling time to be below some critical value.
+increases with decreasing $\beta$. Instability formally occurs for
+arbitrarily long but finite cooling times. 
+%Instability does not require the
+%cooling time to be below some critical value.
 
@@
 \end{align}
-%Table \ref{bstar_compare} compares Eq. \ref{betastar} to 
-%previously reported values of $\beta_c$ 
-%for which fragmentation
-%occurs.
-This $\beta_*$ is similar to values of the cooling times, $\beta_\mathrm{c}$,
-below which fragmentation occur on dynamical timescales in
-numerical simulations of self-gravitating disks
-\citep{gammie01,rice05,rice11}, which employ the same beta cooling prescription.    
+It is interesting to note that this equation reproduce similar values 
+of the cooling times below which 
+%dynamical disk 
+%fragmentation occurs in numerical simulations
+numerical simulations show dynamical disk fragmentation 
+\citep{gammie01,rice05,rice11}. These simulations employ the same beta  
+cooling prescription with $\theta=0$, and determine the fragmentation
+boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
 We compare these in Table \ref{bstar_compare}, showing rough
-agreement; and further discuss the potential significance of this result in
-\S\ref{prev_works}.
-
-%{\bf brief physical interpretation. gravito-turbulent heating is a
-%  constant background heating - it does not affect perturbations about
-%  the turbulent state.  
-%  imposed beta cooling works on all
-%  scales. but `gravito-turbulence' only makes sense on scales larger
-%  than averaging procedure (probably about $H$). }
-%{\bf wild guess: fragmentation had high resolution because internal
-%  structure of forming-clumps become resolved. internal turbulence
-%  helps collapse. 
-%}
+agreement.
 

### 2016-01-24 paper/appendix3d.tex
halfway through version 2 of 2d results
ready to revise applications

@@
 \end{align}
+Note that these equations treat $\alpha$ and $\beta$ as independent
+parameters. 
+%If beta cooling is entirely balanced by viscous heating in
+%the background state, then the $\alpha$-$\beta$ relation,
+%Eq. \ref{alpha_beta_relation}, should be applied.
 
@@
 
-\section{Numerical method for the 3D eigenvalue problem}
+\section{Numerical method for the 3D eigenvalue problem}\label{3d_method}
 We use a pseudo-spectral method to solve the set of ordinary
@@
 %At the upper disk boundary we also apply reflection,
-We explicitly apply a reflecting upper disk boundary, 
+For simplicity we apply a reflecting upper disk boundary, %free
+                                %boundary give similar results
 \begin{align}

### 2016-01-24 paper/results2d.tex
ready to revise applications

@@
 Next, we consider two special cases: $\theta=0$, so perturbations are
-cooled towards zero temperature \citep[as typically employed in
+cooled towards zero temperature \citep[typically employed in
 numerical simulations, e.g.][]{gammie01}; and $\theta=1$, where the
@@
 \end{align}
-It is interesting to note that this equation reproduce similar values 
+This equation reproduce similar values 
 of the cooling times below which 
@@
 boundary, $\beta_c$, as a function of the adiabatic index $\gamma$.  
-We compare these in Table \ref{bstar_compare}, showing rough
-agreement.
+Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
+$\beta_c$.
 
@@
 \subsection{Viscous disk}\label{2dvisc}
-We now consider a viscous disk with 
-$\mu=-1,\,\lambda=0$ as discussed in \S\ref{visc_model}. We first
-check in Appendix \ref{gammie_check} that our dispersion relation
-reduces previous results in the adiabatic/isothermal limit.
+We now consider a viscous disk with parameters $\mu=-1,\,\lambda=0$ in our
+adopted viscosity law, Eq. \ref{visc_law}. In the 2D case this implies
+$\nu\Sigma$ is constant.   
+%\S\ref{visc_model}. 
+We check in Appendix \ref{gammie_check} that our dispersion relation
+reduces to previous results for viscous GI in the isothermal limit (by
+taking $|\beta S|\to~\infty$ and $\gamma=1$).
 
-To see the effect of finite cooling, we simplify the dispersion relation,
-Eq. \ref{thindisk}, by assuming $|\beta S|\ll 1$. Then  
-for $|K| \to 0$ we find
+To see the effect of cooling and irradiation, we simplify the
+dispersion relation, Eq. \ref{thindisk}, by assuming $|\beta S|\ll
+1$. Then for $|K| \to 0$ we find
 \begin{align}\label{gammie_smallk}
@@
 \end{align}
+%with $Q_\mathrm{crit}$ given by 
 to mimic a basic state maintained by gravito-turbulence where one
@@
 \citep{lin87}.   
+%should really only use this Q-alpha relation for order-unity values
+%of Q
 
@@
 \ref{gammie_bigk}. Comparing the two panels shows that increasing the
-irradiation level ($\theta$) surpresses small-scale perturbations.
+irradiation level ($\theta$) surpresses small-scale
+perturbations. %(This also occurs in the inviscid case.)
+
+%The case $\beta=100$ is shown for completeness but is not physically
+%consistent with gravito-turbulence, as the assumed $Q(\alpha)$
+%relation (Eq. \ref{Qalpha}) imply $Q\gg 1$, whereas $Q$ of order unity
+%is expected for a gravito-turbulent background. Nevertheless, it
+%illustrates that viscous GI is qualitatively different to classic
+%GI. Here, instability can occur for high $Q$.
 
@@
 growth timescales are long ($>10$ orbits) for $\beta \gtrsim
-20$. This region is marked by the vertical dashed-dotted line in
+20$. %increase beta -> drop alpha -> increase Q. latter two expected
+     %to stablize
+This region is marked by the vertical dashed-dotted line in
 Fig. \ref{gammie_maxrate_plot}. The optimum wavenumber decreases with
@@
 
-{\bf maybe move following to discussion}
-Numerical simulations of self-gravitating disks show there is a
+%{\bf maybe move following to discussion}
+Numerical simulations of gravito-turbulent disks show there is a
 maximum $\alpha$ ($\sim 0.06$) that can be sustained before

### 2016-01-24 paper/results3d.tex
new eigenvector fig
ready to revise applications

@@
 \ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
-avoid such uncertainties. 
+avoid such uncertainties. Our numerical approach is outlined in Appendix
+\ref{3d_method}.
+
 %We will find that while softening the
@@
 %quantitative match is  
-
 %{\bf note: fast beta-cooling in 3D destablizes vertical modes (enables
 %convection)}
-
 %take k>0 wlog
@@
 ($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
-constant (Eq. \ref{alpha_beta_relation}). We use $N=65$  and apply
-numerical disk surface such that $\rho(\zmax)=0.05\rho_0$.
+constant (Eq. \ref{alpha_beta_relation}). We consider even modes about
+the midplane, ane apply a numerical disk surface at $z=\zmax$ such
+that $\rho(\zmax)=0.05\rho_0$.
 
@@
 
-Fig. \ref{3d_inviscid} show growth rates and the most unstable wavenumber
-obtained for a 3D inviscid disk with $\qthree=0.71$
-($Q=2Q_\mathrm{crit}$) and $\theta=0$. We also plot 2D
+We consider an inviscid disk with $\qthree=0.71$ and $\theta=0$. 
+Recall the 3D gravity parameter $\qthree$ is defined by 
+Eq. \ref{Q3d_def}. This corresponds to a Toomre parameter of
+$Q=2Q_\mathrm{crit}$; where $Q_\mathrm{crit}$, defined by
+Eq. \ref{qcrit_def}, is the Toomre parameter value such that the 2D
+disk would be marginally stable in the absence of cooling.
+
+Fig. \ref{3d_inviscid} show growth rates and the most unstable
+wavenumber obtained for this model. We also plot 2D
 results with the 3D correction as described in Appendix
@@
 growth rates, but the most unstable wavenumber in 3D is somewhat smaller. 
-This reflects self-gravity being weakened in the verical 
+This reflects self-gravity being weakened in the vertical 
 direction: a larger horizontal scale is required to achieve the same
@@
 a factor of two.
-%A smaller $H_\mathrm{sg}$ is needed with faster 
-%cooling (larger viscosity) and vice versa. 
-%because most unstable wavenumber is smaller with high viscosity
 
-%smaller scales with increasing beta
-
-%, because viscosity makes the
-%disk behave more two-dimensional.
-
-%by eliminating vertical velocity gradients.
+%cooling helps vertical modes (convection) 
+%slower variation wrt z for low beta because of associated high
+%viscosity
+%smooth out vertical gradients 
+Fig. \ref{3d_visc_vz} shows the magnitude of the vertical velocity
+$|\delta v_z|$ scaled by the total horizontal velocity for $\beta =
+1,\,10$ and $100$. Cooling generally enhances vertical motion
+(convection). However, even in the highly unstable case with $\beta=1$,
+vertical speeds are sub-dominant at $\sim 30\%$ of the horizontal
+speeds. This suggests that it is not essential to consider 3D models
+when considering gravitational instabilities.
 
@@
 \end{figure}
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0.23cm
+    0.0cm]{figures/eigenvec_vz}
+  \caption{Magnitude of vertical velocities, normalized by the
+    magnitude of the total horizontal velocity, of the viscous GI in 
+    Fig. \ref{3d_visc}, for three cooling times: $\beta=1,\,10$, and $100$. \label{3d_visc_vz}} 
+\end{figure}

### 2016-01-24 paper/application.tex
finished revising 6.2

@@
   \item $\alpha_m$, the dimensionless viscosity associated with other
-    sources of turbulence.
+    sources of turbulence, such as magneto-rotational instabilities
+    (MRI), see also \S\ref{visc_gi} and \S\ref{MHD}. 
 \end{itemize} 
@@
 Note that we are considering the local stability at each radius. We
-are thus neglecting the slow, background viscous radial flow in the
-linear problem.
-
-%proper treatment of cooling  in 3d
-%vertically isothermal basic state for simplicity (also appropriate for
-%irradiated disks) --- this formally requires zero viscosity  
-%or have viscosity but assume it does not affect basic state (dominated
-%by irradiation) but affect perturbations
+are thus neglecting the slow, background viscous radial flow; as well
+as any global gravitational instabilities \citep{lodato05}. {\bf more refs?}
 
 \subsection{Effective $\alpha$}
-In a steady, viscously accreting Keplerian disk we have,
+In a steady, viscously accreting Keplerian disk 
+with constant $\dot{M}$ we have,
 approximately, $\nu\Sigma = \dot{M}/3\pi$. Hence
@@
 
-\subsection{PPD cooling}
-Energy loss in \citetalias{rafikov15}'s model is given by 
+\subsection{PPD beta cooling}
+Energy loss in \citetalias{rafikov15} is given by 
 \begin{align}\label{real_cool}
@@
 is the optical depth and recall Eq. \ref{opacity_law} 
-is our opacity model. Eq. \ref{ftau} accounts for cooling in the
-optically-thin ($\tau\ll 1$) and optically-thick ($\tau\gg1$)
-regimes.
+is our opacity model where $\kappa_d\propto T^b$ with $b=2$. 
+Eq. \ref{ftau} accounts for cooling in the optically-thin ($\tau\ll
+1$) and optically-thick ($\tau\gg1$) regimes.
 
-Since we are interested in the linear response of the system, the
-corresponding cooling parameters $\beta$ and $\theta$ are found by
-comparing the linearized form of the cooling function
-(Eq. \ref{real_cool}) to that of the model function
-(Eq. \ref{beta_cool}).  
+Note that Eq. \ref{real_cool} is a beta cooling prescription because
+it is an explicit function of the thermodynamic states. However, we
+formulated the linear problem with beta cooling of the functional form
+given by Eq. \ref{beta_cool}. In order to adapt the developed
+framework to the above PPD beta cooling function, we
+need to identify the $\beta$ and $\theta$ parameters that enter the 
+linearized equations (Eq. \ref{linear_beta}).
+
+%Since we are interested in the linear response of the system, the
+%corresponding cooling parameters $\beta$ and $\theta$ are found by
+%comparing the linearized form of the cooling function
+%(Eq. \ref{real_cool}) to that of the model function
+%(Eq. \ref{beta_cool}).
+
 %{\bf as opposed to comparing their unperturbed forms, as done in literature}
@@
   g(\tau, T) &= \left( \frac{\tau^2-1}{\tau^2+1}\right)\left(1 -
-  \frac{T_\mathrm{irr}^4}{T^4}\right). 
+  \frac{T_\mathrm{irr}^4}{T^4}\right). \label{g_def}
 \end{align}
-Comparing Eq. \ref{linear_real_cool} with Eq. \ref{beta_cool}, we identify
+Comparing Eq. \ref{linear_real_cool} with the linearized form of the
+standard cooling function, Eq. \ref{linear_beta}, we identify
 \begin{align}
-  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\\
-  &\theta = \frac{C_2}{C_1}, 
+  &\beta = \frac{f(\tau)c_s^2\Sigma\Omega}{2\sigma(\gamma-1)C_1T^4},\label{real_beta}\\
+  &\theta = \frac{C_2}{C_1},\label{real_theta} 
 \end{align}
-to be used in the 2D linear framework.
+to be used in the 2D dispersion relation (Eq. \ref{thindisk}).
 
-This `realistic' beta cooling prescription can behave differently to 
-our previous, simple beta cooling (Eq. \ref{beta_cool}) for the same
-physical system. In particular, for a non-irradiated disk
-($T_\mathrm{irr}=0$), simple beta 
-cooling has  $\theta=0$, but realistic beta cooling has $\theta\neq
-0$. However, fully-irradiated disks with $T=T_\mathrm{irr}$ can be
-modeled by simple beta cooling since in both cases $\theta=1$.
+Eq. \ref{real_beta} still represents a physical (dimensionless) cooling
+time, and is consistent with previous definitions of realistic cooling
+times within factors of order unity \citep[e.g.][their
+Eq. 2]{kratter10}.
+
+However, here $\theta$ is an effective or generalized
+irradiation level. 
+%no longer a direct measure of
+%irradiation. 
+By Eq. \ref{real_theta}, $\theta$ is now related to the true
+irradation temperature $\tirr$ through the function $g$ given by 
+Eq. \ref{g_def}. This is
+important because a physically non-irradiated disk with $\tirr=0$ will
+still have a non-zero effective irradiation, $\theta\neq0$. 
+%On the
+%other hand, fully irradiated disks with  $T=T_\mathrm{irr}$ have 
+%$\theta=1$, as is in the previous problem formulation
+%(Eq. \ref{tirr_def}).
+
+%the gammie-type cooling function does not represent realistic
+%cooling. because real cooling depends on sigma and pressure but gammie
+%cooling only depends on pressure
 

### 2016-01-24 paper/model3d.tex
finished revising 6.2

@@
 
-\subsubsection{Beta cooling}
+\subsubsection{Beta cooling}\label{beta_cool_model}
 In this case the energy loss per unit volume is specified 
@@
 the cooling timescale with $\beta$ a constant input parameter. Here, 
-\begin{align}
+\begin{align}\label{tirr_def}
   \tirr = \theta T_\mathrm{eq}

### 2016-01-24 paper/paper3d.tex
halfway through version 2 of 2d results
finished revising 6.2

@@
 derived from inviscid disks without cooling. 
+%e.g. using classic DR to determine what scales need to be resolved
 
@@
 rate by reducing the magnitude of the 
-pressure term. In fact, for $t_c\in(0,\infty)$ the formal condition
+pressure term. In fact, for $t_c\in[0,\infty)$ the formal condition
 for instability is  
@@
 rotation neglected ($\Omega\to 0$). That is, a classically stable
-disk can be made unstable by viscosity as it effectively removes
+disk can be destabilized by viscosity as it effectively removes
 rotational stabilization \citep{lynden-bell74}.

### 2016-01-24 paper/application.tex
revised application

@@
 
-However, here $\theta$ is an effective or generalized
+However, $\theta$ is now an effective 
 irradiation level. 
@@
 %irradiation. 
-By Eq. \ref{real_theta}, $\theta$ is now related to the true
+Eq. \ref{real_theta} shows that $\theta$ is related to the true
 irradation temperature $\tirr$ through the function $g$ given by 
@@
 still have a non-zero effective irradiation, $\theta\neq0$. 
+%This results from the fact that PPD cooling generally depends on both
+%$P$ and $\Sigma$, but in the absence of irradiation standard beta
+%cooling only depends on $P$.
+
 %On the
@@
 \subsection{Inviscid stability condition}
-We note that for the adopted opacity law, Eq. \ref{opacity_law}, 
+We note that for our adopted opacity law, Eq. \ref{opacity_law}, 
 \begin{align*}
@@
 implying $5/6<\theta<3/2$. 
-%For a non-irradiated disk, $T_\mathrm{irr}=0$, $\theta\to5/6$ as
-%$\tau\to0$ and $\theta\to3/2$ as $\tau\to\infty$.  
-From the results in \S\ref{2d_inviscid}, we conclude that this
-system cannot be unstable anywhere without
-viscous effects acting on the perturbations if 
+From the discussion in \S\ref{2d_inviscid} and applying
+Eq. \ref{stable_condition},  we conclude that without viscous effects 
+the disk is stable everywhere if  
 \begin{align} 
@@
 
-PPDs become optically-thin ($\tau\to 0$) with 
-increasing distance from the star. Then $\theta$ decreases, and
-the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to $\gamma >
-5/6$ in the outer disk, which is always satisfied.  
+PPDs become irradiation-dominated at large distances from the star,
+where $T\to\tirr$, or $\theta\to 1$. {\bf ref?}
+Then the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to
+$\gamma > 1$ in the outer disk, which is always satisfied.  
 On the other hand, numerical simulations of gravito-turbulence
-show that the second inequality is generally satisfied \citep{rice11}.  
-Taken together, this means that in the outer regions of a realistic PPD,
-cooling itself cannot cause instability.
+show that $1\lesssim Q \lesssim 2$ \citep{gammie01,rice11}, and 
+the second inequality is generally satisfied. Taken together, this
+means that in the outer regions of a realistic PPD, cooling itself
+cannot cause instability.
 
@@
 5\times10^{-4}\mathrm{cm}^2\,\mathrm{g}^{-1}\,\mathrm{K}^{-2}$  as in
-\citetalias{rafikov15}. We use 
-$\gamma=1.6$ so the inviscid stability condition
-(Eq. \ref{ppd_invisc_cond}) is satisfied
-everywhere. Fig. \ref{rafikov_model} shows the equilirbium 
+\citetalias{rafikov15}. We use $\gamma=1.6$, approximately applicable
+to molecular gas at low temperatures and used by other authors
+\citep{rice11,baehr15}. Then the global inviscid stability condition
+(Eq. \ref{ppd_invisc_cond}) is satisfied. 
+%for outer disk choice of gamma doesn't matter 
+%avoid complications from over-stable thermal modes 
+Fig. \ref{rafikov_model} shows the equilirbium 
 disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
@@
 \citet[][their Eq. 19 and 21, respectively]{sterzik95}, and identical if one
-takes $\theta=1$.
+takes $\theta=1$. %showing that irradiation makes the disk behave like
+                  %isothermal disk
 
@@
 \subsection{3D PPD with radiative diffusion}
-%{\bf radiative diffusion only for optically thick disk. how to connect
-%to optically thin regions?} 
+%appropriate for not-so-large distances where irrad can be neglected 
 We briefly consider 3D PPDs with explicit radiative diffusion
@@
 additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
-We use the same disk model as in \S\ref{pp2d_example} but without 
-irradiation ($T_\mathrm{irr}=0$), and a slightly smaller vertical
+We use the same disk model as in \S\ref{pp2d_example} but 
+with $T_\mathrm{irr}=0$, since our simple radiative diffusion model
+does not include irradiation (\S\ref{rad_cool}). 
+%physically, irrad is negligible for opt thick regions, and we are
+%restricted to such regions if applying rad diff
+We use a slightly smaller vertical
 domain with $\rho(\zmax)=0.1\rho_0$.  %to avoid very low density
@@
 becomes optically-thin, radiative diffusion (the 3D curve) is not
-valid and under-estimates the growth rates. Nevertheless, the
-comparison shows that the transition radius $R_t\sim60$AU can be
-calculated within the 2D framework. 
+valid and under-estimates the growth rates. %; while the $f(\tau)$
+%function approximately captures opticall-thin cooling in the 2D case.  
+Nevertheless, the comparison shows that the transition radius
+$R_t\sim60$AU can be calculated within the 2D framework.  
 %This is not surprising since the

### 2016-01-25 paper/summary.tex
revising summary

--- 2016-01-19 paper/summary.tex
@@
 In this paper, we develop the linear theory of cooling, irradiated,
-and turbulent accretion disks in order to understand gravitational
+and viscous accretion disks in order to understand gravitational
 instability (GI) in realistic protoplanetary disks (PPDs). 
-We model turbulence as an effective, 
-Navier-Stokes viscosity. This viscosity provides a 
-background heating to balance the imposed cooling,  
-and may also be active in the perturbed state.  
+We use a Navier-Stokes viscosity to mimic the effects of turbulence. 
+This viscosity provides a background heating to balance the imposed cooling,  
+and may also act on linear perturbations. %thermo and dynamic 
 We suggest that disk fragmentation, observed in numerical simulations, 
-can be understood as the non-linear outcome of a secondary
-GI of the gravito-turbulent basic state, driven by cooling and/or viscosity.   
+can be understood as the eventual outcome of secondary
+instabilities of the gravito-turbulent base state, driven by cooling
+and/or viscosity.     
 %This %allows us to assess realistic disk fragmentation without using
@@
 %move to subsection?
-While viscosity and cooling are related through thermal
-balance, they independently enhance GI: cooling reduces thermal
-stabilization; and viscous forces compromise rotational
+While viscosity and cooling can be related through thermal
+balance, they independently enhance growth rates: cooling reduces
+thermal stabilization; and viscous forces compromise rotational
 stabilization. This provides a physical explanation as to why
@@
 Eq. \ref{inviscid}. The sufficient condition for stability 
-depends on the irradiation level (Eq. \ref{stable_condition}), and 
-\emph{independent} of the cooling rate, provided it is non-zero and
+depends on the irradiation level (Eq. \ref{stable_condition}), and is 
+\emph{independent} of the cooling time, provided it is non-zero and
 finite. This means that long cooling times can still  
@@
 related ($\alpha\propto Q^{-2}$) to model a
-gravito-turbulent background, viscous GI occurs on orbital  
+gravito-turbulent background, this instability occurs on orbital  
 timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
-of a maximum stress sustainable by gravito-turbulence established in
-previous numerical simulations \citep{rice05}. Because viscosity affects 
+of a maximum stress sustainable by gravito-turbulence established
+through numerical simulations \citep{rice05}. Because viscosity affects 
 rotational support, viscous GI occurs on large scales, $kH\lesssim
 O(1)$.  Furthermore, irradiation preferentially
-stabilizes small-scale perturbations. %maybe that's why irradiated disks produce smaller alpha
+stabilizes small-scale perturbations. %also true for inviscid case
+%maybe that's why irradiated disks produce smaller alpha
 
@@
 We apply our linear framework to protoplanetary disks with 
-realistic models for cooling and gravito-turbulence
-\citep{rafikov15}. We show that, for linear dynamics, realistic 
-cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
-actually modeled by `beta' cooling with a  
-non-zero irradiation parameter ($\theta\neq0$), in contrast to 
-numerical simulations which typically employ beta cooling with  
-$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
-are such simulations to realistic PPDs must then be further 
-investigated.
-
-The important implication of having 
-$\theta\sim 1$ in a PPD is that cooling alone cannot 
-destabilize the disk at large radii. Instead, viscous GI occurs on
+realistic models for cooling and gravito-turbulence. 
+%\citep{rafikov15}. 
+%We show that, for linear dynamics, realistic 
+%cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
+%actually modeled by `beta' cooling with a  
+%non-zero irradiation parameter ($\theta\neq0$), in contrast to 
+%numerical simulations which typically employ beta cooling with  
+%$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
+%are such simulations to realistic PPDs must then be further 
+%investigated. 
+%The important implication of having 
+%$\theta\sim 1$ in a PPD is that cooling alone cannot 
+%destabilize the disk at large radii. 
+We show that with a realistic cooling model for PPDs,
+cooling alone cannot lead to gravitational instabilities. This is due
+to stabilization by an effective irradiation associated with the PPD
+cooling function, which is present even if 
+there is no physical irradiation. %due to matching functional forms of
+                                %cooling 
+Instead, viscous GI occurs on  
 dynamical timescales for $R\gtrsim 60$AU because 
@@
 \subsection{Relation to numerical simulations}\label{prev_works}
-Our framework may shed light on 
-%may offer an alternative understanding of 
-some recent numerical results concerning disk
-fragmentation.  \cite{paardekooper12} found that fragmentation can occur for 
+%frag in sims because of : cooling, viscosity, or high den
+%flucuations? don't know 
+Our results may shed light on some numerical simulations 
+of disk fragmentation. Table 
+\ref{bstar_compare} shows a close match between the characteristic
+cooling time for cooling-driven GI (Eq. \ref{betastar}) 
+and that for disk fragmentation observed in simulations 
+\citep{gammie01,rice05,rice11}. This suggests that, at least for those
+simulations, fragmentation is physically due to the removal of
+pressure stabilization by cooling on a radial scale of the disk scale
+height. 
+%i.e. cool-driven g 
+{\bf don't know if we want to dwell on this table}
+
+More recent simulations have raised the issue of numerical
+convergence. \cite{paardekooper12} found that fragmentation can occur
+for   
 slowly-cooled disks, $\tcool\Omega\gg O(1)$, but that this requires   
@@
 that for either cooling-driven or viscous GI, there is no critical 
-cooling rate (viscosity) beyond (below) which the disk is  
+cooling rate/viscosity below  which the disk is  
 formally stable. Instead, growth rates smoothly decrease with
-increasing (decreasing) $\tcool$ ($\alpha$), implying that
-instabilities take longer to develop for slowly-cooled disks.        
+increasing $\tcool$ (decreasing $\alpha$), implying that
+instabilities, and hence fragmentation, simply take longer to develop
+for slowly-cooled disks.         
 %hopkins paper
@@
 %resolution convergence 
-Another issue is convergence with numerical resolution. Notably, 
-\cite{baehr15} found that for fixed cooling rates, disks eventually 
-fragment with sufficient spatial resolution. %This could be due to 
-%small-scale, turbulent angular momentum transport being enabled at
-%high resolutions, which aids clump formation. 
-This might be related to viscous GI in the following sense: high
-resolutions enable small-scale turbulent angular momentum transport
-away from over-dense regions, aiding clump formation. This 
-mechanism is unavailable at low resolutions, for which one can only
-resort to rapid cooling for destabilization.
+Another issue is convergence with numerical resolution. For example,
+\cite{meru11} found that the critical cooling time for 
+disk fragmentation increases with numerical resolution. %sph particles in
+% their case 
+Follow-up studies attributed this to decreasing numerical viscous
+heating at higher resolution, which helps fragmentation
+\citep{lodato11,meru12}. %also global effects in
+                         %these. (paardekooper/meru/baruteau) 
+%note that these sims do not have a floor temp
+%is there an effective floor temp related to num visc?
+%high res -> smaller effective floor temp -> easier to frag?
 
-We note that power in gravito-turbulence is dominated by scales with
-$kH\sim 1$ \citep{cossins09}, and this is always  
-well-resolved by modern numerical simulations. However, this does not
-necessarily imply small-scale dynamics are unimportant \citep[especially for non-linear evolution, see][]{young15}, 
+Here, we highlight that most numerical experiments, including those
+above, employ the standard beta cooling function, Eq. \ref{beta_cool},
+\emph{without} a floor temperature. We show in \S\ref{2d_inviscid}
+(see also \S\ref{cool_gi}) that this implies cooling-driven GI can
+occur at any sufficiently small scales. Then as the numerical
+resolution increases, simulations can access a wider range of unstable
+scales.  %no upper bound 
+In this respect, it is not surprising to find non-convergence with 
+increasing resolution. %simple beta cooling will not give a
+                       %well-imposed problem
+
+We thus suggest that having a floor temperature in the standard beta
+cooling prescription is a necessary condition for  
+convergence with respect to numerical resolution. This limits the
+relevant scales to a finite range.  %for cool gi 
+%physically irradiation 
+Furthermore, without a floor temperature, standard beta cooling only
+depends on the pressure/energy density field. It is not clear how
+results obtained from such models can be applied to PPDs, where
+cooling depends on two thermodynamic states (e.g. $P$ and
+$T$). A floor temperature permits a mapping between beta
+cooling and realistic cooling (\S\ref{ppd_cooling}).
+
+%ultra-small scale cooling-GI may have slow growth rates
+%but will eventually matter if you sim for a long time 
+%
+%limit the relevant scales to a finite range 
+%we suggest that having a floor temperature (irradiation) is necessary
+%(though may not sufficient) 
+%floor temp may be physical: irradiation, or numerical (sph particles
+%can't have zero jitter)
+
+%point out viscosity can have a dynamic role in helping frag
+
+However, \cite{baehr15} included a floor temperature (or irradiation)
+in their simulations but still find that at fixed cooling rates, disks
+eventually fragment with sufficient spatial resolution.
+In light of our results on viscous GI, 
+we here highlight a possible contribution: %in addition to num visc (meru lodato)
+high resolutions enable small-scale turbulent angular momentum transport 
+% physical, cf numerical  
+to aid clump formation (see also \S\ref{MHD} below). 
+This mechanism is unavailable at low resolutions, for which one must
+rely on rapid cooling.
+
+%This, in fact, has been suggested for
+
+We note that, although power in gravito-turbulence is dominated by scales with
+$kH\sim 1$ \citep{cossins09}, which is  
+well-resolved by modern numerical simulations, that this does not
+necessarily imply small-scale dynamics/thermodynamics are unimportant
+\citep[especially for non-linear evolution, see][]{young15},  
 as is evident from the non-converging simulations described above.
@@
 
-\subsection{MHD turbulence}%turbulence and viscosity 
-It is important to note that while we consider gravito-turbulent disks  
-as an application, the linear theory itself does not assume any 
-specific origin of turbulence.
+\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
+We emphasize the linear theory we have developed does not assume
+
+%It is important to note that while we consider gravito-turbulent disks  
+%as an application, the linear theory itself does not assume any 
+%specific origin of turbulence.
 

### 2016-01-25 paper/summary.tex
rough version 2, old figs

@@
 
-However, \cite{baehr15} included a floor temperature (or irradiation)
+On the other hand, \cite{baehr15} included a floor temperature (or irradiation)
 in their simulations but still find that at fixed cooling rates, disks
@@
 
-We note that, although power in gravito-turbulence is dominated by scales with
-$kH\sim 1$ \citep{cossins09}, which is  
-well-resolved by modern numerical simulations, that this does not
+We comment that although modern simulations resolve the dominant
+scale associated with gravito-turbulence, $kH\sim 1$
+\citep{cossins09}, this does not
 necessarily imply small-scale dynamics/thermodynamics are unimportant 
@@
 We emphasize the linear theory we have developed does not assume  
+a particular origin for the turbulence represented by the imposed
+viscosity. For example, the 2D dispersion relation, Eq. \ref{thindisk}
+and Appendix \ref{2ddisp}, treats $\alpha$ as an independent
+input parameter.
 
-%It is important to note that while we consider gravito-turbulent disks  
-%as an application, the linear theory itself does not assume any 
-%specific origin of turbulence.
+%We only considered gravito-turbulence as an application.  
+%In
+%practice, this amounts to relating the $\alpha$ parameter to $Q$. 
+%most equations treat alpha/Q/beta as independent
 
 %effect of MHD turbulence on GI 
-Our models may thus apply to self-gravitating disks dominated by, for
-example, magneto-hydrodynamic (MHD) turbulence.   
+Our models may thus apply to self-gravitating disks dominated by MHD
+turbulence.     
 %In fact, the viscous approach may be more
@@
 %and frag 
-True disk fragmentation is a non-linear process characterized by clumps   
-reaching densities orders of magnitude above the ambient value, and
+True disk fragmentation is a non-linear process characterized by
+clumps reaching densities orders of magnitude above the ambient value, and
 which survives tidal shear and shocks \citep{paardekooper12,young16}.  
@@
 gravito-turbulent state becomes dynamically unstable, and thus should
-not persist, when cooling (viscosity) is too rapid (large).  
-However, numerical simulations show that there are only two
-possible outcomes of cooling, self-gravitating disks: steady
-gravito-turbulence or fragmentation. It seems then reasonable to 
-speculate that the non-existence of a gravito-turbulent state, here due to
-dynamical instability, implies disk fragmentation.
+not persist, when cooling is too rapid or when the associated viscous
+stresses are too large.  However, numerical simulations show that
+there are only two possible outcomes of cooling, self-gravitating
+disks: steady gravito-turbulence or fragmentation. It seems then
+reasonable to  speculate that the non-existence of a stable
+gravito-turbulent state, here due to dynamical instability, implies
+disk fragmentation.
 
-{\bf comment on stochasitc frag. we only consider 
-how to enhance GI by reducing rotation and pressure. but a third route is 
-increasing sigma. according to toomre Q these are the 3 effects one should consider
-for SG. turbulent flucations may occasionally produce high sigma dens perts. 
-possible relation to current framework? high sigma perts arise because of viscous GI? 
-high alpha $\to$ small-scale transport $\to$ enhance over-dense regions}
+%Our goal is to highlight how non-ideal effects, such as viscosity and
+%cooling, can promote GI. However, the Toomre parameter
+%(Eq. \ref{toomreQ_criterion}) suggest a third possibility: increasing
+%the surface density.
+
+Our deterministic approach cannot model `stochastic fragmentation'
+\citep{paardekooper12,hopkins13}. In this interpretation,  
+fragmentation is attributed to the occurence (and survival) 
+of non-linear density perturbations, which arise from the
+gravito-turbulent state simply by chance. 
+%Whether these high
+%density perturbations actually lead to a clump formation, 
+This might be considered as a secondary instability by lowering the
+Toomre $Q$ through a (random) increase in density
+(Eq. \ref{toomreQ_criterion}). By contrast, the viscosity/cooling
+mechanisms discussed in this paper lowers $Q$ by reducing pressure and
+Coriolis forces. %we have a physical way of lowering pressure/rotation
+
+%stochasitc frag -> it just happens by chance. no physics
 
@@
 %non-axisymmetry and transient dynamics 
+%While our linear problem has been formulated with some 
 Another possible generalization is non-axisymmetric
 disturbances. In barotropic, inviscid disks  
-non-axisymmetric GI can develop for $Q$ somewhat larger than unity
-\citep{papaloizou89,papaloizou91}. It would be interesting to study
+non-axisymmetric global GI can develop for $Q$ somewhat larger than unity
+\citep{papaloizou89,adams89,papaloizou91}. It would be interesting to study
 how non-axisymmetric perturbations are affected by cooling and

### 2016-01-25 paper/application.tex
revising summary
version 2 in text, about to redo figs

@@
 \section{Application to protoplanetary disks}\label{2dppd}
-We now apply our linear framework to assess the stability of realistic 
+We now apply our linear framework to assess the stability of  
 protoplanetary disks. We consider the gravito-turbulent disk models recently
@@
     sources of turbulence, such as magneto-rotational instabilities
-    (MRI), see also \S\ref{visc_gi} and \S\ref{MHD}. 
+    (MRI, see also \S\ref{visc_gi} and \S\ref{MHD}). 
 \end{itemize} 
@@
 
-\subsection{PPD beta cooling}
+\subsection{PPD beta cooling}\label{ppd_cooling}
 Energy loss in \citetalias{rafikov15} is given by 
@@
 \end{align}
-is the optical depth and recall Eq. \ref{opacity_law} 
+is the optical depth. Recall Eq. \ref{opacity_law} 
 is our opacity model where $\kappa_d\propto T^b$ with $b=2$. 
@@
 it is an explicit function of the thermodynamic states. However, we
-formulated the linear problem with beta cooling of the functional form
-given by Eq. \ref{beta_cool}. In order to adapt the developed
+formulated the linear problem with the standard beta cooling function 
+given by Eq. \ref{beta_cool}. In order to adapt the existing  
 framework to the above PPD beta cooling function, we 
@@
 
-Eq. \ref{real_beta} still represents a physical (dimensionless) cooling
-time, and is consistent with previous definitions of realistic cooling
-times within factors of order unity \citep[e.g.][their
-Eq. 2]{kratter10}.
+Eq. \ref{real_beta} still represents a physical cooling 
+time, and is consistent with previous definitions within factors of
+order unity \citep[e.g.][their Eq. 2]{kratter10}.
 
@@
 irradiation level. 
-%no longer a direct measure of
-%irradiation. 
 Eq. \ref{real_theta} shows that $\theta$ is related to the true
@@
 %(Eq. \ref{tirr_def}).
-
-%the gammie-type cooling function does not represent realistic
-%cooling. because real cooling depends on sigma and pressure but gammie
-%cooling only depends on pressure
 
@@
 PPDs become irradiation-dominated at large distances from the star,
-where $T\to\tirr$, or $\theta\to 1$. {\bf ref?}
+where $T\to\tirr$ and $\theta\to 1$. {\bf ref?}
 Then the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to
@@
 \subsection{Example 2D calculation}\label{pp2d_example}
-We consider a model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
+We consider a disk  model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
 $\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. 
@@
 \citep{rice11,baehr15}. Then the global inviscid stability condition
-(Eq. \ref{ppd_invisc_cond}) is satisfied. 
+(Eq. \ref{ppd_invisc_cond}) is satisfied. %which simplifies the problem
 %for outer disk choice of gamma doesn't matter 
@@
 Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) 
-which gives the optimum wavenumber as
+which gives the optimum wavenumber and growth rates as
 \begin{align}
-  |K| = \frac{3}{2\theta Q},
-\end{align}
-with growth rate
-\begin{align} 
+  |K| = \frac{3}{2\theta Q}, \quad
+%\end{align}
+%with growth rate
+%\begin{align} 
   S = \frac{27\alpha}{16\theta^3Q^4}. 
@@
 occurs is physically due to large turbulent stresses.
-
-%{\bf despite apply beta cooling results to realistic cooling disks}
-%Our result suggest an alternative interpretation of disk fragmentation
-%is where the gravito-turbulent state becomes dynamically unstable due to
-%both viscosity and self-gravity. 
-%{\bf note: disk might be stablized by extremely large alpha at large
-%  distances but `turbulent' framework no longer valid }
 
@@
 %function approximately captures opticall-thin cooling in the 2D case.  
-Nevertheless, the comparison shows that the transition radius
-$R_t\sim60$AU can be calculated within the 2D framework.  
+However, the comparison shows that the transition radius
+$R_t\sim60$AU can be correctly calculated within the 2D framework.  
+%notice the small softenings used 
 %This is not surprising since the

### 2016-01-25 paper/linear3d.tex
regular update
revising sec 4, minor edits to prev secs
ready to revise applications
version 2 in text, about to redo figs

@@
-%On the other hand, if a time-independent heat source is invoked, then
-%we may choose $\alpha, \beta$ and $\theta$ separately.  
 \section{Linear problem}\label{linear}
@@
 \end{align}
-and similarly for other variables. Here, $k$ 
-is a real horizontal wavenumber and $s$ is a (generally) complex
-growth rate. 
-%The gravitational potential perturbation $\delta \Phi$ depends on $z$
-%but only its midplane value $\delta \Phi(z=0)$ is
-%used in the dynamical equations, which has the same form as above. 
-For clarity, hereafter we drop the tilde.  
-%and note that equilibrium
-%quantities without the $\delta$ prefix.
+and similarly for other variables. Here, $k$  is an input real horizontal
+wavenumber and $s$ is a (generally) complex growth rate.  
+For simplicity, hereafter we drop the tilde.
 
@@
 \end{align}
-with
+where $^\prime$ denotes $d/dz$. The perturbed viscous forces are 
 \begin{align}
@@
     \frac{2}{3}\left(\ln{\rho\nu}\right)^\prime\delta v_x\right]\notag\\
-  &+\ii \nu_b k \left[\delta v_x^\prime + \left(\ln{\rho\nu_b}\right)^\prime\delta v_x\right],\\
+  &+\ii \nu_b k \left[\delta v_x^\prime + \left(\ln{\rho\nu_b}\right)^\prime\delta v_x\right],
 %  &\delta f_x = -\frac{4}{3} \nu k^2\Sigma \delta v_x,\\
@@
 %  v_y. 
+\end{align}
+and the perturbed viscous heating is given by 
+\begin{align}
   \frac{\delta\mathcal{H}_\mathrm{visc}}{\rho}=& \nu (q\Omega)^2
@@
   \delta\ln{\rho\nu} =& (1+\mu)\frac{\delta\rho}{\rho} +
-  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}, \label{linear_beta_cool}
+  \frac{\lambda}{c_s^2}\frac{\delta P}{\rho}. \label{linear_beta_cool}
 \end{align}
-where $^\prime$ denotes $d/dz$, and the 3D self-gravity parameter is
-\begin{align}
+In Eq. \ref{lin_gravity}, the 3D self-gravity parameter is
+\begin{align}\label{Q3d_def}
   Q_\mathrm{3D} \equiv \frac{\Omega^2}{4\pi G \rho_0}. 

### 2016-01-25 paper/model3d.tex
version 2 in text, about to redo figs

--- 2016-01-23 paper/model3d.tex
@@
 where $\rho$ is the density field and $\bm{v} = (v_x, v_y, v_z)$ is
-the velocity field.% $q$ is the shear parameter and $\Omega_z$ is the
-%local vertical oscillation frequency. 
+the velocity field. 
 We assume an ideal gas so that the pressure $P$ and thermal energy
@@
 \end{align}
-%where $G$ is the gravitational constant.
 
@@
 Our numerical calculations use $\mu=-1,\lambda=0$ so 
-that $\rho\nu$ is time-indpendent, folling previous 
+that $\rho\nu$ is time-indpendent, following previous 
 studies of viscous GI  
 \citep{lynden-bell74,hunter83,willerding92,gammie96}.  
-This choice also conveniently eliminates viscous over-stability
+This choice eliminates viscous over-stability
 \citep{schmit95,latter06}, which is unrelated to self-gravity, that
@@
 not well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made
-to bring out the physical process of interest --- viscous GI. %expect
-                                %to 
-%be generally present, but will 
-%be obscured by viscous over-stability with
-%other indices
+to bring out the physical process of interest --- viscous GI. 
 Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
@@
 This is perhaps not inconsistent with numerical simulations of disk
-fragmentation which show the internal structure of clumps, which
-represent increased density, is laminar \citep{gammie01}.  
+fragmentation which show the internal flow of high-density clumps
+is laminar \citep{gammie01}.   
 {\bf is this true?}
-
-%Furthermore, \cite{paardekooper12} measured
-%an increase in the effective viscosity when their gravito-turbulent
-%disk models underwent fragmentation (i.e. as $\Sigma$ increases).  
-%{\bf note paardekooper 2012: alpha goes down as Sigma goes up}
 
@@
 where the first term represents viscous heating. For the viscous
-problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$. However,
+problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$ to 
+obtain a relation between viscous heating and cooling 
+(e.g. Eq. \ref{alpha_beta_relation} below). However,
 if we wish to neglect viscosity but include cooling, we must invoke
@@
 considered in this work.
-%It is possible to have both models present - but too complicated
 
-\subsubsection{Beta cooling}
-In this case the energy loss per unit volume is specified 
+\subsubsection{Beta cooling}\label{beta_cool_model}
+By beta cooling we mean the energy loss per unit volume is specified  
 as an explicit function of the thermodynamic variables. 
@@
 \begin{align}\label{beta_cool}
-  \Lambda =
+  \Lambda(\rho, T) =
   %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
@@
 \end{align}
-Recall from Eq. \ref{beta_def} that 
-%\begin{align}
-  $\tcool = \beta\Omega^{-1}$ is 
-%\end{align}
+Recall  $\tcool = \beta\Omega^{-1}$ is 
 the cooling timescale with $\beta$ a constant input parameter. Here, 
-\begin{align}
+\begin{align}\label{tirr_def}
   \tirr = \theta T_\mathrm{eq}
@@
 $T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
-for example, stellar irradiation. We will thus refer to $\theta$ as
-the `irradiation level'. %stellar illum or env rad
+for example, stellar or background irradiation. We will thus refer to
+$\theta$ as the `irradiation level'. %stellar illum or envelope rad
 
-We will formulate the linear problem with beta cooling using the
-functional form for $\Lambda$ as given by
-Eq. \ref{beta_cool}. However, any explicit cooling function, say
-$\Lambda=\Lambda_1(\rho,T)$, can be mapped to that obtained from
-Eq. \ref{beta_cool} by equating their linearized forms, i.e. $\delta
-\Lambda \equiv \delta \Lambda_1$, which then defines the $\beta$ and
-$\theta$ parameters to be used in the framework we develop. This is
-done in \S\ref{2dppd} where we adopt a more realistic beta cooling 
-prescription for 2D PPDs. In this case, $\theta$ may or may not
-represent a physical irradiation.
-
-For models with beta cooling given by Eq. \ref{beta_cool} we assume an 
+For beta cooling given by Eq. \ref{beta_cool} we assume an 
 equilibrium polytropic relation 
@@
 
-Beta cooling is widely applied in 2D and 3D
-numerical simulations of self-gravitating disks \citep{gammie01,
-  rice05,baehr15}. It permits numerical experiments to be carried out in a
-controlled manner as a function of the cooling time $\beta$. An
-adiabatic disk corresponds to $\beta\to \infty$. The
+Beta cooling of the form Eq. \ref{beta_cool} is widely applied in 2D
+and 3D numerical simulations of self-gravitating disks \citep{gammie01,
+  rice05,rice11}. For this reason we will refer to it as the
+`standard' beta cooling. It permits numerical experiments to be
+carried out in a controlled manner as a function of the cooling time
+$\beta$. An adiabatic disk corresponds to $\beta\to \infty$. The
 physical meaning of the limit $\beta\to0$ depends on $\theta$, as
 discussed in \S\ref{theta0}---\ref{theta1}.
+
+Note that, although we assume standard beta cooling in formulating the
+linear problem later, any explicit 
+cooling function, say $\Lambda_1(\rho,T)$, can be mapped to
+that obtained from Eq. \ref{beta_cool} by equating their linearized
+forms, i.e. $\delta \Lambda \equiv \delta \Lambda_1$. This then 
+defines the $\beta$ and $\theta$ parameters to be used in the 
+framework we develop. We do this in \S\ref{2dppd} where we adopt a
+more realistic beta cooling function for PPDs. In this case, $\theta$
+may or may not represent a physical irradiation.
 
@@
 While radiative cooling is arguably more realistic than
-parametrized beta cooling, it generally implies a vertically
+beta cooling, it generally implies a vertically
 non-isothermal equilibrium disk, and increases the order of the

### 2016-01-25 paper/paper3d.tex
revising summary
version 2 in text, about to redo figs

@@
 or turbulent viscosity is too large, resulting in disk fragmentation.        
-% `alternative' or `physical' here?
-%quantify or develop
-We motivate and quantify an alternative interpretation of disk
+We motivate and quantify an interpretation of disk % a physical
+                                % interpretation 
 fragmentation as the inability to maintain gravito-turbulence due to  
-secondary gravitational instabilities driven by:  
+secondary instabilities driven by:  
 \begin{inparaenum}[1)] 
@@
 as well as full radiative diffusion in 3D. We show that 
-cooling/viscosity can destabilize the disk even when the classic
-Toomre stability condition is satisfied; and generally there is
-no critical cooling rate/viscosity below which the disk is formally
-stable. %there is no special/precise cooling/visc that marks the
-        %stable/unstable boundary
+%cooling/viscosity can destabilize the disk even when the classic
+%Toomre stability condition is satisfied; 
+generally there is no critical cooling rate/viscosity below which 
+the disk is formally stable. 
 We apply linear theory to 
@@
 fragmentation by removing rotational stabilization against 
-self-gravity. {\bf shorten abstract?}
+self-gravity. 
 \end{abstract}
@@
 \section{Introduction}\label{intro} 
-The gravitational instability (GI) of rotating disks    
-plays a fundamental role in many astrophysical problems. In the context of 
-gaseous protoplanetary disks (PPDs), GI has two applications. It can  
-provide gravitational torques to transport angular momentum outwards
-and thus enable mass accretion \citep{armitage10,turner14}.   
-GI may also lead to disk fragmentation, which has been invoked to 
-explain giant planet formation at large radii
-\citep{helled14}. Studying these non-linear phenomena require direct      
-numerical simulations. Nevertheless, physical insight can be  
-obtained through analytical understanding of GI.
+Understanding the gravitational stability of rotating disks    
+is central to many astrophysical problems. In the context of 
+gaseous protoplanetary disks (PPDs), gravitational instability (GI)
+has two applications. It can provide gravitational torques to
+transport angular momentum outwards and thus enable mass accretion
+\citep{armitage10,turner14}. GI may also lead to disk fragmentation,
+which has been invoked to explain giant planet formation at large
+radii \citep{helled14}. Studying these non-linear phenomena require
+direct numerical simulations. Nevertheless, physical insight can be    
+obtained through analytical modeling.
 
@@
 frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
-density and $G$ is the gravitational constant. The Toomre $Q$ is a
+density and $G$ is the gravitational constant. The Toomre parameter is a
 measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
@@
 
-However, as we emphasize in this work, the inclusion of non-ideal
-physics, such as cooling and viscosity, modifies the dispersion
-relation, Eq. \ref{classic_dr}, and hence the condition for GI. This 
-means that the Toomre condition alone is insufficient to assess GI
-in realistic disks. The goal of this work is to generalize the
-analytic treatment of disk GI to allow a self-consistent discussion of
-GI in realistic accretion disk models. Specifically, we will examine
-how  cooling and viscosity can lead to GI even when $Q>1$, by 
-removing pressure and rotational support, respectively.
-
-%Specifically, we will examine the effect of cooling and  
-%viscosity on GI in PPDs.
+However, as we emphasize in this work, including non-ideal
+physics, such as cooling and viscosity, modifies the classic dispersion
+relation, and hence the condition for GI. Thus the Toomre condition
+alone is insufficient to assess GI in realistic disks. The goal of
+this work is to generalize the analytic treatment of disk GI to allow
+a self-consistent discussion of GI in realistic accretion disk
+models. Specifically, we will examine how  cooling and viscosity can
+lead to GI even when $Q>1$, by removing pressure and rotational
+support, respectively.
 
@@
 
-We will quantify how cooling removes pressure support to enable GI
+We will quantify how cooling enables GI
 even when $Q>1$. For example, if perturbations are
@@
 rate by reducing the magnitude of the 
-pressure term. In fact, for $t_c\in[0,\infty)$ the formal condition
+pressure term. In fact, for $t_c\in[0,\infty)$ a formal condition
 for instability is  
@@
 Viscous disks can also develop GI even when $Q>1$. This is because, as 
-demonstrated below, viscosity compromises rotational support against  
+demonstrated below, viscosity eliminates rotational support against  
 self-gravity for long-wavelength disturbances 
@@
 In this work, we use viscosity to model such a 
-physical effect due to turbulence, but we note that viscous 
-GI does not assume any origin for the viscosity. We will generalized
-the theory of viscous GI to include an energy equation  with viscous
+physical effect due to turbulence, but we emphasize that viscous 
+GI does not assume a particular origin for the viscosity. We will
+generalize the theory of viscous GI to include an energy equation  with viscous
 heating, irradiation, explicit cooling, as well as
-three-dimensionality (3D). This 
-allows us to consider viscous GI in more realistic PPD models.
+three-dimensionality (3D). This allows us to consider viscous GI in
+more realistic PPD models.
 
@@
 We develop a general framework for viscous disks without  
-assuming a specific origin of turbulence that this viscosity 
-represents. We will, however, apply the theory to `gravito-turbulent'
+assuming a specific origin of the turbulence that this viscosity 
+represents. %without assuming the nature of the turbulence represented
+            %by this viscosity  
+We will, however, apply the theory to `gravito-turbulent'
 disks in which the viscosity associated with some underlying (classic)
@@
 thermal equilibrium \citep{gammie01}. 
-The dynamic and thermodynamic effect of this turbulence will be
-represented by a viscosity. Then the $\alpha$ and $\beta$ parameters
-defined above will be inversely related.
+Represent the dynamic and thermodynamic effect of this turbulence by a
+viscosity. Then the $\alpha$ and $\beta$ parameters defined above will
+be inversely related.
 
@@
 \S\ref{cool_gi}---\ref{visc_gi}, these effects can reduce
-gravitational stability. These considerations motivate a physical    
+gravitational stability. This motivates a physical    
 interpretation of disk fragmentation as the inability to maintain a
@@
 In \S\ref{2dppd} we consider PPDs with realistic viscosity/cooling   
-models, including radiative diffusion in 3D, and determine where such
-disks are gravitationally unstable. We summarize our results in
+models, including radiative diffusion in 3D, and determine where PPDs
+are gravitationally unstable and why. We summarize our results in
 \S\ref{summary} with a discussion of how our models may aid the 

### 2016-01-25 paper/results2d.tex
finished revising 6.2
version 2 in text, about to redo figs

@@
 \section{Two-dimensional disks with beta cooling}\label{result_2d}
-We begin in the 2D limit with beta cooling  to make connection 
+We begin in the 2D limit with standard beta cooling  to make connection 
 with previous studies. The disk
@@
 \end{align}
-where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp},
-which do not impose any $\alpha$-$\beta$ relation. We use this generalized
-dispersion relation  to investigate GI driven 
+where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}. % Note
+% that these functions, as given, do not assume any $\alpha$-$\beta$
+% relation.
+We use this generalized dispersion relation  to investigate GI driven 
 by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity 
@@
 %second condition concerning Q is associated with self-g 
-Whether or not the system is stable does not 
-depend on the cooling time $\beta\in(0,\infty)$, 
-but on the irradiation level $\theta$.  
-The disk becomes more unstable with decreasing $\theta$, i.e. as the
-floor temperature, towards which perturbations are cooled, drops.
+Notice these conditions do not actually
+depend on the cooling time. 
+The disk becomes more unstable with decreasing $\theta$, 
+i.e. as the floor temperature, towards which perturbations are cooled, drops.
 
@@
 \end{align}
-are unstable. Thus the range of unstable wavenumbers increases with
-decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
+are unstable. The range of unstable wavenumbers
+increases with decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
 $(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
@@
 time $\beta_*$ as that which removes pressure support against
-self-gravity over the natural lengthscale in the problem, $H$. We
-thus set $|K_*|=1$ and find, for the Keplerian disk,
+self-gravity over the natural lengthscale in the problem, the
+scale-height $H$. We thus set $|K_*|=1$ and find, for the Keplerian
+disk,
 
@@
 \subsubsection{$\theta = 1$}\label{theta1}
-Beta cooling with $\theta=1$ corresponds to 
+Standard beta cooling with $\theta=1$ corresponds to 
 `thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
@@
 shows that such a system would be dynamically unstable with 
-growth rate $s = O(\Omega)$. This due to the direct effect of viscous 
+growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
 stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 

### 2016-01-25 paper/results3d.tex
version 2 in text, about to redo figs

@@
 constant (Eq. \ref{alpha_beta_relation}). We consider even modes about
-the midplane, ane apply a numerical disk surface at $z=\zmax$ such
+the midplane, and apply a numerical disk surface at $z=\zmax$ such
 that $\rho(\zmax)=0.05\rho_0$.
@@
 
-We consider an inviscid disk with $\qthree=0.71$ and $\theta=0$. 
-Recall the 3D gravity parameter $\qthree$ is defined by 
-Eq. \ref{Q3d_def}. This corresponds to a Toomre parameter of
-$Q=2Q_\mathrm{crit}$; where $Q_\mathrm{crit}$, defined by
+We consider a 3D inviscid disk with $\qthree=0.71$ and $\theta=0$. 
+The 3D gravity parameter $\qthree$ is defined by 
+Eq. \ref{Q3d_def}. For this case the corresponding Toomre parameter 
+$Q=2Q_\mathrm{crit}$. Recall $Q_\mathrm{crit}$, defined by
 Eq. \ref{qcrit_def}, is the Toomre parameter value such that the 2D

### 2016-01-25 paper/summary.tex
version 2 in text, about to redo figs

@@
 instability (GI) in realistic protoplanetary disks (PPDs). 
-We use a Navier-Stokes viscosity to mimic the effects of turbulence. 
+We use a Navier-Stokes viscosity to mimic the effects of
+turbulence. %which may be of any origin, though we specialize to
+            %gravito-turb 
 This viscosity provides a background heating to balance the imposed cooling,  
@@
 can be understood as the eventual outcome of secondary
-instabilities of the gravito-turbulent base state, driven by cooling
+instabilities of a gravito-turbulent base state, driven by cooling
 and/or viscosity.     
-%This %allows us to assess realistic disk fragmentation without using
-%uncertain empirical conditions based on numerical simulations.   
-%{\bf cooling removes pressure support, viscosity removes rotational support}
-%{\bf our conclusions do require input of emperical resuls from numerical
-%sims}
-%going beyond toomre 
-%move to subsection?
 While viscosity and cooling can be related through thermal
@@
 dispersion relation 
-Eq. \ref{inviscid}. The sufficient condition for stability 
-depends on the irradiation level (Eq. \ref{stable_condition}), and is 
-\emph{independent} of the cooling time, provided it is non-zero and
-finite. This means that long cooling times can still  
+Eq. \ref{inviscid}. We find a sufficient condition for stability which  
+depends on the irradiation level (Eq. \ref{stable_condition}) but 
+\emph{independent} of the cooling time. 
+This means that long cooling times can still  
 formally lead to instability. However, growth timescales may be 
@@
 \citep{lynden-bell74,willerding92,gammie96}, to include 
-cooling, viscous heating and irradiation. In our fiducial disk,  
-where viscosity and self-gravity are inversely
+cooling, viscous heating and irradiation. %and 3D 
+We consider a disk with viscosity and self-gravity inversely
 related ($\alpha\propto Q^{-2}$) to model a 
-gravito-turbulent background, this instability occurs on orbital  
+gravito-turbulent background, and find viscous GI occurs on orbital
 timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion 
 of a maximum stress sustainable by gravito-turbulence established 
-through numerical simulations \citep{rice05}. Because viscosity affects 
+by numerical simulations \citep{rice05}. Because viscosity affects 
 rotational support, viscous GI occurs on large scales, $kH\lesssim
@@
 realistic models for cooling and gravito-turbulence. 
-%\citep{rafikov15}. 
-%We show that, for linear dynamics, realistic 
-%cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
-%actually modeled by `beta' cooling with a  
-%non-zero irradiation parameter ($\theta\neq0$), in contrast to 
-%numerical simulations which typically employ beta cooling with  
-%$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
-%are such simulations to realistic PPDs must then be further 
-%investigated. 
-%The important implication of having 
-%$\theta\sim 1$ in a PPD is that cooling alone cannot 
-%destabilize the disk at large radii. 
 We show that with a realistic cooling model for PPDs, 
@@
 
-%we do not look at fragment survival
-%fragment due to cooling or viscosity?
-%they should produce diff size frags
 \subsection{Relation to numerical simulations}\label{prev_works}
@@
 Our results may shed light on some numerical simulations 
-of disk fragmentation. Table 
+concerning disk fragmentation. Table 
 \ref{bstar_compare} shows a close match between the characteristic
@@
 pressure stabilization by cooling on a radial scale of the disk scale
-height. 
-%i.e. cool-driven g 
+height. %i.e. cool-driven g 
 {\bf don't know if we want to dwell on this table}
+%negligible effect from viscous heating
+%makes sense? sims show clumps are not turbulent internally
+%then visc plays no role in clumping -> only a cooling effect
 
@@
 convergence. \cite{paardekooper12} found that fragmentation can occur 
-for   
-slowly-cooled disks, $\tcool\Omega\gg O(1)$, but that this requires   
+for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
 simulations to run for significantly longer than dynamical 
@@
 
-We thus suggest that having a floor temperature in the standard beta
+We suggest having a floor temperature in the standard beta
 cooling prescription is a necessary condition for  
-convergence with respect to numerical resolution. This limits the
+numerical convergence. This limits the
 relevant scales to a finite range.  %for cool gi 
 %physically irradiation 
-Furthermore, without a floor temperature, standard beta cooling only
-depends on the pressure/energy density field. It is not clear how
-results obtained from such models can be applied to PPDs, where
-cooling depends on two thermodynamic states (e.g. $P$ and
-$T$). A floor temperature permits a mapping between beta
-cooling and realistic cooling (\S\ref{ppd_cooling}).
+Furthermore, without a floor temperature, standard beta cooling is
+a function of the pressure/energy density field only. It is not clear
+how relevant are results obtained from this is to actual PPDs,
+where cooling depends on two thermodynamic states (e.g. pressure and
+temperature). A floor temperature permits a mapping between standard beta
+cooling and PPD cooling (\S\ref{ppd_cooling}).
 
@@
 
-On the other hand, \cite{baehr15} included a floor temperature (or irradiation)
+On the other hand, \cite{baehr15} included a floor temperature 
 in their simulations but still find that at fixed cooling rates, disks
@@
 rely on rapid cooling.  
-
-%This, in fact, has been suggested for
+%maybe there is an effective floor temp which depends on res
 
@@
 necessarily imply small-scale dynamics/thermodynamics are unimportant 
-\citep[especially for non-linear evolution, see][]{young15},  
+\citep[especially for non-linear evolution, ][]{young15},  
 as is evident from the non-converging simulations described above.
@@
 \subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
-We emphasize the linear theory we have developed does not assume  
-a particular origin for the turbulence represented by the imposed
-viscosity. For example, the 2D dispersion relation, Eq. \ref{thindisk}
-and Appendix \ref{2ddisp}, treats $\alpha$ as an independent
+We emphasize the linear framework we have developed does not assume  
+a particular origin for the turbulence that is represented by the imposed
+viscosity. For example, the 2D dispersion relation given by Eq. \ref{thindisk}
+with Eq. \ref{bigA}---\ref{bigF}, treats $\alpha$ as an independent
 input parameter.
@@
 by \cite{fromang05}. He finds disk fragmentation with increasing numerical
-resolution, attributing this to resolving the most unstable MRI
+resolution, and attributes this to resolving the most unstable MRI
 wavelength, which enables small-scale angular momentum removal by MHD
-turbulence to aid fragmentation. %this supports using viscosity to
-                                %mimic turbulence
+turbulence to aid fragmentation. %clumping/mass accretion onto clump
 This would be a physical mechnism represented by the viscous GI 
@@
 which survives tidal shear and shocks \citep{paardekooper12,young16}.  
-%cite the young papers 
 Clearly, our linear models cannot address fragmentation
@@
 not persist, when cooling is too rapid or when the associated viscous
-stresses are too large.  However, numerical simulations show that
-there are only two possible outcomes of cooling, self-gravitating
-disks: steady gravito-turbulence or fragmentation. It seems then
-reasonable to  speculate that the non-existence of a stable
+stresses are too large.  However, steady gravito-turbulence and
+fragmentation are the two \emph{only} possible outcomes of cooling, 
+self-gravitating disks in the local limit \citep[][]{gammie01}.    
+Then it seems reasonable to speculate that the non-existence of a stable
 gravito-turbulent state, here due to dynamical instability, implies
@@
 \citep{paardekooper12,hopkins13}. In this interpretation,  
-fragmentation is attributed to the occurence (and survival) 
-of non-linear density perturbations, which arise from the
-gravito-turbulent state simply by chance. 
-%Whether these high
-%density perturbations actually lead to a clump formation, 
+fragmentation is attributed to the occurence and survival 
+of large, non-linear density enhancements, which arise from the
+gravito-turbulent flucuations simply by chance. 
+%cooling affects survival rate 
 This might be considered as a secondary instability by lowering the
-Toomre $Q$ through a (random) increase in density
+Toomre parameter through a (random) increase in density 
 (Eq. \ref{toomreQ_criterion}). By contrast, the viscosity/cooling
-mechanisms discussed in this paper lowers $Q$ by reducing pressure and
-Coriolis forces. %we have a physical way of lowering pressure/rotation
+destabilizing mechanisms discussed in this paper lowers $Q$ by
+reducing pressure and Coriolis forces. %we have a physical way of lowering pressure/rotation
 
@@
 %this visc law is used in prev studies (by much smarter people)
-How to quantitatively model turbulence (be it due to self-gravity or
-MHD) as a viscosity, especially on dynamical timescales, should be
-clarified with direct numerical simulations. The present linear models
-should then be modified accordingly. 
+How to quantitatively model gravito- or MHD turbulence 
+as a viscosity, especially on dynamical timescales, should be 
+clarified with direct numerical simulations. The present viscosity
+models should then be modified accordingly. 
 %could consider other visc parameters mu and lambda (but this

### 2016-01-25 paper/summary.tex
version 2 in text, about to redo figs

@@
 not persist, when cooling is too rapid or when the associated viscous
-stresses are too large.  However, steady gravito-turbulence and
-fragmentation are the two \emph{only} possible outcomes of cooling, 
+stresses are too large.  However, steady gravito-turbulence or 
+fragmentation are the \emph{only} possible outcomes of cooling, 
 self-gravitating disks in the local limit \citep[][]{gammie01}.    

### 2016-01-26 paper/results2d.tex
updating figs, need to correct 3d beta results, need to re-comp bg if Q depends on alpha

@@
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.4cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0.cm
     0.0cm]{figures/viscsg_modes}\\
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.4cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
     0.0cm]{figures/viscsg_modes_theta0d3}
@@
 %    0.0cm]{figures/result2d_fixQ}\\
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0.cm
     0.0cm]{figures/result2d_gvisc}

### 2016-01-26 codes/2d/viscsg.f90
code update

@@ -172,9 +172,16 @@ program viscsg
            err(m) = zmag(det)
            
            !filter out root with too large error
+!           if(filter .eqv. .true.) then
+!              if(err(m) .gt. tol) roots(m) = dcmplx(-1d0,0d0)
+!           endif
+             
+            !filter out over-stable modes 
            if(filter .eqv. .true.) then
-              if(err(m) .gt. tol) roots(m) = dcmplx(-1d0,0d0)
+              if(abs(x(2)/x(1)) .gt. tol) roots(m) = dcmplx(-1d0,0d0)
            endif
+
+
         enddo
         
         !pick most unstable root  among all roots, and store the associated error

### 2016-01-26 codes/3d/viscsg3d.f90

@@ -33,9 +33,11 @@ program viscsg
   integer :: i,j, jmid, kcount, bcount, info, lwork
   integer :: loc(1) 
   integer, allocatable :: ipiv(:)
-  real*8 :: dk, dlogb, optimize 
+  real*8 :: dk, dlogb, optimize, amid, bcool
   real*8 :: k, zbar, m, T_l, dT_l, d2T_l, n 
   complex*16, allocatable :: work(:)
+  real*8, external :: gviscosity
+
   namelist /params/ Q3d, smallq, bgmma, gmma, bmin, bmax, nb, kmin, kmax, nk, mu, lambda, gvisc, fixalpha, bulkv, tirr
   namelist /grid/   nz, zmax, nzout, vbc 
   namelist /job/    maximize, no_decay, no_oscil 
@@ -68,14 +70,12 @@ program viscsg
   !set up parameter survey axis 
   allocate(bvals(nb))
   allocate(kvals(nk))
-
+  allocate(Qarr(nb))
   !wavenumber space
   if(nk .gt. 1) then
-!     dk = (kmax - kmin)/(dble(nk)-1d0)
-      dk = log10(kmax/kmin)/(dble(nk)-1d0)
+     dk = (kmax - kmin)/(dble(nk)-1d0)
      do i=1, nk 
-!        kvals(i) = kmin + dk*(dble(i)-1d0)
-         kvals(i) = 10d0**(log10(kmin)+dk*(dble(i)-1d0))
+        kvals(i) = kmin + dk*(dble(i)-1d0)
      enddo
   else
      kvals(1) = kmin
@@ -95,29 +95,29 @@ program viscsg
   else
      bvals(1) = bmin 
   endif
-
+  
   open(10,file='bvals.dat')
   do i=1, nb
      write(10,fmt='(e22.15)') bvals(i)
   enddo
   close(10)
   
-  !****************************************!
-  !***********BASIC STATE SETUP************!
-  !****************************************!
-  write(6,fmt='(A)') '************BASIC STATE SETUP***********'
+  if(gvisc.eqv..true.) then !reset Q to depend on midplane alpha 
+     print*, 'using gviscosity, only OK for big gamma=1!'
+     do i=1, nb
+        bcool = bvals(i) 
+        amid    = 1d0/(bgmma*(gmma-1d0)*bcool*smallq**2d0)*(1d0 - tirr)
+        Qarr(i) = gviscosity(amid) 
+     enddo
+  else 
+     Qarr(:) = Q3d 
+  endif
   
[…]

### 2016-01-26 paper/application.tex
KMK edits

@@
 \end{itemize} 
-These are used in the statements of thermal equilibrium, mass and
-angular momentum conservation, together with an opacity law, 
-to construct a global disk model with surface density $\Sigma(R)$ and
+These properties serve as inputs for calculating thermal equilibrium, and mass and
+angular momentum conservation. Together with an opacity law, 
+they allow us to construct a global disk model with surface density $\Sigma(R)$ and
 temperature $T(R)$ where $R$ is the global cylindrical radius from the
-star. \citepalias[See][for details.]{rafikov15} This gives the $Q(R)$
-value required for input into our linear framework. We derive other
+star. \citepalias[See][for details.]{rafikov15} These profiles give $Q(R)$,
+required for input into our linear framework. We derive other
 dimensionless parameters below.
 
-Note that we are considering the local stability at each radius. We
-are thus neglecting the slow, background viscous radial flow; as well
-as any global gravitational instabilities \citep{lodato05}. {\bf more refs?}
+Although our disk is global in extent, here we consider the local stability at each radius.
+%KMK I disagree with neglect of viscous flow in that there is a global Mdot. Revised version:
+We are thus neglecting any global instabilities \citep{adams89,lodato05,kratter10} as well as any 
+ evolution of the disk properties in response to GI.
+% We are thus neglecting the slow, background viscous radial flow; as well
+%as any global gravitational instabilities \citep{lodato05}. {\bf more refs?}
 
@@
 \end{align} 
-which sets the viscosity coefficient to be used at each radius.
+which sets the viscosity coefficient to be used at each radius. Note that the disk profiles 
+employed here give constant $\dot{M}$ rather than constant $\alpha$.
 
@@
 \end{align}
-per unit area, 
+per unit area, where
 \begin{align}
@@
 
-Note that Eq. \ref{real_cool} is a beta cooling prescription because
+Note that Eq. \ref{real_cool} falls within our definition of a beta cooling prescription, because
 it is an explicit function of the thermodynamic states. However, we
 formulated the linear problem with the standard beta cooling function 
-given by Eq. \ref{beta_cool}. In order to adapt the existing  
-framework to the above PPD beta cooling function, we 
-need to identify the $\beta$ and $\theta$ parameters that enter the 
+given by Eq. \ref{beta_cool}, which has a different (less realistic) dependence on disk temperature.
+ In order to adapt the existing  framework to the above PPD cooling function, we 
+need to identify the equivalent $\beta$ and $\theta$ parameters that are required for the 
 linearized equations (Eq. \ref{linear_beta}).
@@
 
-However, $\theta$ is now an effective 
-irradiation level. 
+%KMK
+However, $\theta$ is no longer a good proxy for the irradiation level
+as it was in the standard beta cooling model.
+%is now an effective irradiation level. 
 Eq. \ref{real_theta} shows that $\theta$ is related to the true
-irradation temperature $\tirr$ through the function $g$ given by 
-Eq. \ref{g_def}. This is
-important because a physically non-irradiated disk with $\tirr=0$ will
-still have a non-zero effective irradiation, $\theta\neq0$. 
+irradiation temperature $\tirr$ through the function $g$ given by 
+Eq. \ref{g_def}, and is therefore a only a weak function of the irradiation temperature. 
+A physically non-irradiated disk with $\tirr=0$ will
+have a non-zero value of the `irradiation parameter', $\theta\neq0$. More specifically $\theta = O(1)$ 
+for all $\tirr <T$, and for our adopted opacity law, Eq. \ref{opacity_law}, 
+\begin{align*}
+  \theta = \frac{4-g}{4-2g},
+\end{align*}
+implying $5/6<\theta<3/2$. . %%KMK 
 %This results from the fact that PPD cooling generally depends on both
@@
 %cooling only depends on $P$. 
+{\bf KMK: we should discuss}
 
@@
 \subsection{Inviscid stability condition}
-We note that for our adopted opacity law, Eq. \ref{opacity_law}, 
-\begin{align*}
-  \theta = \frac{4-g}{4-2g},
-\end{align*}
-implying $5/6<\theta<3/2$. 
-From the discussion in \S\ref{2d_inviscid} and applying
+%We note that for our adopted opacity law, Eq. \ref{opacity_law}, 
+%\begin{align*}
+%  \theta = \frac{4-g}{4-2g},
+%\end{align*}
+%implying $5/6<\theta<3/2$. 
+With our new linearized cooling function in hand, from the discussion in \S\ref{2d_inviscid} and by applying
 Eq. \ref{stable_condition},  we conclude that without viscous effects 
@@
 PPDs become irradiation-dominated at large distances from the star,
-where $T\to\tirr$ and $\theta\to 1$. {\bf ref?}
+where $T\to\tirr$ and $\theta\to 1$ \citep{dalessio97,kratter11}.
 Then the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to
@@
 show that $1\lesssim Q \lesssim 2$ \citep{gammie01,rice11}, and 
-the second inequality is generally satisfied. Taken together, this
-means that in the outer regions of a realistic PPD, cooling itself
-cannot cause instability.
+the second inequality is generally satisfied. 
+%KMK
+Taken together, this means that in the outer regions of a realistic PPD, cooling itself
+is not responsible for a secondary instability of a gravito-turbulent disk,
+which might lead to fragmentation. This leaves viscous GI as the only possible
+culprit, as we illustrate below.
 
 \subsection{Example 2D calculation}\label{pp2d_example}
-We consider a disk  model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
+We relax the inviscid assumption and consider a disk  model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
 $\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. 
+%KMK
+Such a high accretion rate is consistent with those expected for 
+young protostellar disks \citep{shu77,enoch09}. Similarly,
+$10\mathrm{K}$ is a conservatively low background irradiation l
+evel consistent with cloud temperatures in star forming regions \citep{plume97,johnstone01}.
+Stellar irradiation will typically elevate $\tirr$ in addition to adding a radial dependence \citep{kratter08}
 We adopt the opacity scale $\kappa_{d0} =
@@
 \citetalias{rafikov15}. We use $\gamma=1.6$, approximately applicable
-to molecular gas at low temperatures and used by other authors
-\citep{rice11,baehr15}. Then the global inviscid stability condition
-(Eq. \ref{ppd_invisc_cond}) is satisfied. %which simplifies the problem
+to an ideal molecular gas at low temperatures 
+%KMK this is just true...and used by other authors
+%\citep{rice11,baehr15}. 
+This choice of $\gamma$ satisfies the global inviscid stability condition
+(Eq. \ref{ppd_invisc_cond}). %which simplifies the problem
 %for outer disk choice of gamma doesn't matter 
 %avoid complications from over-stable thermal modes 
-Fig. \ref{rafikov_model} shows the equilirbium 
+Fig. \ref{rafikov_model} shows the equilibrium 
 disk profile in terms of $Q$, $\alpha$, $\beta$, and $\theta$.  
-These are set as input to the 2D dispersion relation
+These profiles serve as input to the 2D dispersion relation
 (Eq. \ref{thindisk}).
 
-Fig. \ref{rafikov_growth} show growth timescales and
+Fig. \ref{rafikov_growth} shows growth timescales and
 optimum wavenumbers for viscous GI in this disk model.  
@@
 
-The most unstable wavelength is a few times the disk thickness. The
+The most unstable wavelength is a few times the disk thickness. 
+%KMK
+So long as $H<<R$, this result is consistent with our use of the local approximation. For
+our example disk $H/R \approx ??$ at $R=?$. The
 increase in $|K|$ from $\sim 10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is
-due to the decrease in $Q$; while that from $\sim 60\mathrm{AU}$ to
-$\sim 100\mathrm{AU}$ occurs as the disk transitions from
-optically-thick to optically-thin regimes. The mismatch at large
-distances is  
-not surprising since the above expressions assume $|K|\ll
-1$. Nevertheless the analytic estimates reproduce the correct
+due to the decrease in $Q$, while that from $\sim 60\mathrm{AU}$ to
+$\sim 100\mathrm{AU}$ occurs as the disk transitions from the
+optically-thick to optically-thin regime. The mismatch between the
+numerical and analytic solutions at large
+distances is expected since the above expressions assume $|K|\ll
+1$. Nevertheless the analytic estimates reproduce qualitatively correct
 behavior.
@@
 criteria for disk fragmentation \citep[e.g.][]{rafikov15}. 
-This suggests that fragmentation in the outer parts of a realistic PPD
-occurs is physically due to large turbulent stresses.
+Thus viscous GI may be responsible for the transition between gravito-turbulence
+and fragmentation due to the removal of rotational support by viscous (turbulent) stresses.
+%KMK
+%This suggests that fragmentation in the outer parts of a realistic PPD
+%occurs is physically due to large turbulent stresses.
 
@@
 We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
-additional boundary condition $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
+additional boundary condition {\bf KMK: that the disk surface temperature is fixed},
+ $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
 We use the same disk model as in \S\ref{pp2d_example} but 
@@
 
-Fig. \ref{rafikov_growth3d} show the growth rates and most unstable
-wavenumber for $R\in[10,100]$AU; along with corresponding 2D 
-results with softened self-gravity. There is a close match for
+Fig. \ref{rafikov_growth3d} shows the growth rates and most unstable
+wavenumber for $R\in[10,100]$AU; along with the corresponding 2D 
+results matched with softened self-gravity. There is good agreeement for
 $R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim
-1$) and both models apply. However, beyond $60$AU where the disk
+1$) and thus both models apply. However, beyond $60$AU where the disk
 becomes optically-thin, radiative diffusion (the 3D curve) is not
@@
 %function approximately captures opticall-thin cooling in the 2D case.  
-However, the comparison shows that the transition radius
+Nevertheless, the transition radius
 $R_t\sim60$AU can be correctly calculated within the 2D framework.  

### 2016-01-26 paper/linear3d.tex
KMK edits

@@
 \section{Linear problem}\label{linear}
-We consider infinitestimal axisymmetric Eulerian perturbations in the
+We consider infinitestimal axisymmetric Eulerian perturbations of the
 form \begin{align}
@@
 \end{align}
-and similarly for other variables. Here, $k$  is an input real horizontal
+and equivalent form for other variables. Here, $k$  is an input real horizontal
 wavenumber and $s$ is a (generally) complex growth rate.  
@@
 \end{align}
-where $^\prime$ denotes $d/dz$. The perturbed viscous forces are 
+where $^\prime$ denotes $d/dz$. The perturbed viscous forces are {\bf F is also radiative flux....}
 \begin{align}
@@
 \end{align}
-Since Eq. \ref{linear_rad_cool} contain vertical derivatives of the
-perturbations, it is generally not possible to map radiative cooling 
+Since Eq. \ref{linear_rad_cool} contains vertical derivatives of the
+perturbations, it is not generically possible to map radiative cooling 
 to the beta cooling prescription, except for special problems
-\citep[e.g.][]{lin15}.
+\citep[e.g.][]{lin15}. %KMK
+We now consider the behavior of two and three dimensional disks in the
+non-ideal physics, linear GI framework {\bf or some other transition sentence}.

### 2016-01-26 paper/model3d.tex
KMK edits

@@
 co-rotates with a fiducial point in the unperturbed disk at angular
-frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ corresponds
-to the radial, azimuthal and verical directions in the global disk. 
+frequency $\Omega$. The Cartesian co-ordinates $(x,y,z)$ correspond
+to the radial, azimuthal and vertical directions in the global disk. 
 The fluid equations are
@@
 \end{align}
-where $\mathcal{R}$ is the gas constant and $T$ is the temperature. We
-use $\gamma=1.4$ unless otherwise stated. The gas 
-gravitatational potential $\Phi$ is given via the Poisson equation,  
+where $\mathcal{R}$ is the gas constant and $T$ is the temperature. {\bf KMK cut because 
+mostly varied: We
+use $\gamma=1.4$ unless otherwise stated.} The gas 
+gravitational potential $\Phi$ is given via the Poisson equation,  
 \begin{align}
@@
 tidal, and viscous forces (see below), respectively. We consider    
-Keplerian disks with shear paramter $q=3/2$ and vertical oscillation 
+Keplerian disks with shear parameter $q=3/2$ and vertical oscillation 
 frequency $\Omega_z=\Omega$.
@@
 respectively, and $\mathcal{H}_\mathrm{ext}$ represents any
-time-independent heat source/sinks. We normally set
+time-independent heat source/sinks. We set
 $\mathcal{H}_\mathrm{ext}=~0$ unless otherwise stated.
@@
 \subsection{Viscosity and heating}\label{visc_model}
-The Cartesian components of the viscous stress tensor $\bm{T}$ is
+The Cartesian components of the viscous stress tensor $\bm{T}$ are
 defined by 
@@
 viscosity $\rho\nu_b$ for completeness, but will neglect it in
-numerical calculations. The associated viscous heating given by 
+numerical calculations. The associated viscous heating is given by 
 \begin{align}
@@
 
-We adopt the viscosity law  
+We adopt a viscosity law  
 \begin{align}\label{visc_law}
@@
 Our numerical calculations use $\mu=-1,\lambda=0$ so 
-that $\rho\nu$ is time-indpendent, following previous 
+that $\rho\nu$ is time-independent, following previous 
 studies of viscous GI  
@@
 This choice eliminates viscous over-stability
-\citep{schmit95,latter06}, which is unrelated to self-gravity, that
+\citep{schmit95,latter06}, which is unrelated to self-gravity, and
 would otherwise contaminate our results. %obscure
@@
 While steady-state viscosity values can be determined 
-analytically \citep[e.g.][]{rafikov15}, its time-dependent behavior is
+analytically or numerically \citep[e.g.][]{martin11,kratter08,rafikov15}, the time-dependent behavior is
 not well-explored. We emphasize the choice $\mu=-1,\lambda=0$ is made
@@
 Interestingly, though, \cite{laughlin96b} have suggested a $\nu \propto 1/\Sigma$
-dependence when modeling the evoluting of 2D self-gravitating disks 
+dependence when modeling the evolution of 2D self-gravitating disks 
 as a viscous process. We note that if $\rho\nu$ is constant in time
@@
 fragmentation which show the internal flow of high-density clumps
-is laminar \citep{gammie01}.   
-{\bf is this true?}
+is laminar \citep{gammie01}, although once clumps form they may effectively decouple
+from the background disk state, and thus no longer be described by the same prescription.   
+{\bf is this true? KMK: maybe caveat helps?}
 
@@
 where the first term represents viscous heating. For the viscous
-problem, $\nu\neq0$, we may set $\mathcal{H}_\mathrm{ext}=0$ to 
+problem, $\nu\neq0$, and we set $\mathcal{H}_\mathrm{ext}=0$ to 
 obtain a relation between viscous heating and cooling 
 (e.g. Eq. \ref{alpha_beta_relation} below). However,
-if we wish to neglect viscosity but include cooling, we must invoke
+if we wish to neglect viscosity (and the accompanying dissipation) but include cooling, we must invoke
 $\mathcal{H}_\mathrm{ext}\neq0$ to define an equilibrium state.  
@@
 
-\subsubsection{Beta cooling}\label{beta_cool_model}
-By beta cooling we mean the energy loss per unit volume is specified  
+\subsubsection{$\beta$ cooling}\label{beta_cool_model}
+In our $\beta$ cooling model, the energy loss per unit volume is specified  
 as an explicit function of the thermodynamic variables. 
@@
 for example, stellar or background irradiation. We will thus refer to
-$\theta$ as the `irradiation level'. %stellar illum or envelope rad
+$\theta$ as the `irradiation level'. As we discuss in Section ?, 
+the limits $\theta \rightarrow 0, \theta \rightarrow 1$ do not exactly 
+correspond to the limits of non-irradiated or irradiation-dominated disks
+as discussed in the literature \citep{kratter11,rice11}.%stellar illum or envelope rad
 
-For beta cooling given by Eq. \ref{beta_cool} we assume an 
+For $\beta$-cooling given by Eq. \ref{beta_cool} we assume an 
 equilibrium polytropic relation 
@@
 \subsubsection{Radiative cooling}\label{rad_cool}
-Here we assume energy transfer by radiative diffusion. Then
+A more realistic treatment of cooling considers energy transfer by radiative diffusion. Then
 \begin{align}
@@
 and take the constant index $b=2$ as appropriate for the cold outer regions
-of a realistic PPD \citep{bell94}, but retain the general notation $b$
+of a PPD with ISM-like dust grains \citep{bell94}, but retain the general notation $b$
 to keep track of the opacity.    
@@
 for optically-thin disks. However, this complication is beyond the scope of
-this work. {\bf ref for flux lim diff?}
+this work. {\bf ref for flux lim diff? \citep{levermore81}?}

### 2016-01-26 paper/paper3d.tex
KMK edits

@@
 \shorttitle{Non-ideal Gravitational instability}
-\shortauthors{M.-K.\ Lin \& K. Kratter}
+\shortauthors{M.-K.\ Lin \& K. M. Kratter}
 
@@
 Low mass, self-gravitating accretion disks admit quasi-steady,
-`gravito-turbulent' states in which cooling balances turbulent
+`gravito-turbulent' states in which cooling balances turbulent {\bf viscous?}
 heating. However, numerical simulations show that gravito-turbulence
-cannot be sustained beyond dynamical timescales when the cooling rate
-or turbulent viscosity is too large, resulting in disk fragmentation.        
+cannot be sustained beyond dynamical timescales when the cooling rate is too fast,
+or corresponding turbulent viscosity is too large. The result is is disk fragmentation.        
 We motivate and quantify an interpretation of disk % a physical
@@
 fragmentation as the inability to maintain gravito-turbulence due to  
-secondary instabilities driven by:  
+{\bf proper / formal / etc?} secondary instabilities driven by:  
 \begin{inparaenum}[1)] 
@@
 consider parameterized cooling functions in two and three dimensions,
-as well as full radiative diffusion in 3D. We show that 
+as well as %full 
+radiative diffusion in 3D. We show that 
 %cooling/viscosity can destabilize the disk even when the classic
@@
 generally there is no critical cooling rate/viscosity below which 
-the disk is formally stable. 
-We apply linear theory to 
+the disk is formally stable, although interesting limits appear when we examine the growth
+of modes of order the disk scale height. 
+We apply this new linear theory to 
 global models of protoplanetary disks (PPDs) subject to
 gravito-turbulence modeled as an effective 
-viscosity and cooling via dust opacity. 
+viscosity, and cooling via dust opacity. 
 %We find the sole effect of
@@
 fragmentation by removing rotational stabilization against 
-self-gravity. 
+self-gravity, and that the observed transition in behavior from gravito-turbulent to fragmenting may derive from
+instability of the gravito-turbulent state itself.
 \end{abstract}
@@
 which has been invoked to explain giant planet formation at large
-radii \citep{helled14}. Studying these non-linear phenomena require
+radii \citep{boss97,helled14} %% KMK need an original citation, not just review.
+Studying these non-linear phenomena requires
 direct numerical simulations. Nevertheless, physical insight can be    
@@
 but still order-unity values of $Q$
-\citep{papaloizou89,papaloizou91}. 
+\citep{lau_bertin78,papaloizou89,papaloizou91}. 
 %strictly speaking non axi *local* modes are stable -> only transient
@@
 
-It is important to remember that the popular dispersion relation, 
-Eq. \ref{classic_dr},  is derived from \emph{idealized} conditions: 
-the disk is laminar and does not experience any net thermal
+We emphasize that the oft-used dispersion relation and corresponding Toomre parameter, %It is important to remember that the popular dispersion relation, 
+(Eqs. \ref{toomreQ_criterion} and \ref{classic_dr}) are derived from \emph{idealized} conditions: 
+the base disk is laminar and does not experience any net thermal
 losses\footnote{Isothermal disks implicitly assume  
@@
 However, real disks can cool, for example, %radiative, convective cooling
-due to radiative losses. Accretion disks may also be turbulent, for
-which its dynamic and thermodynamic effect is often modeled through an
-effective viscosity \citep{shakura73,lin87,rafikov15}. {\bf 
-  more refs here?} 
+due to radiative losses. Accretion disks may also be turbulent, the dynamic and thermodynamic effect of which is often modeled through an
+effective viscosity \citep{shakura73,lin87,armitage01,rafikov15}. 
+%{\bf  more refs here?} 
 %in terms of AM transport and heating 
@@
 
-Despite this, the Toomre parameter and its associated
-dispersion relation are still widely applied to viscous, cooling
-accretion disk models \citep[e.g.][]{cossins09, kimura12}, {\bf more refs?}
-even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
-derived from inviscid disks without cooling. 
+Despite this, the Toomre parameter %and its associated
+%dispersion relation are
+ is still widely applied to viscous (turbulent), cooling
+accretion disk models \citep[e.g.][]{gammie01,cossins09, kimura12}, {\bf more refs?}
+%%KMK
+despite the mismatch in the included underlying physics.
+%even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
+%derived from inviscid disks without cooling. 
 %e.g. using classic DR to determine what scales need to be resolved
 
-However, as we emphasize in this work, including non-ideal
+We show in this work that including non-ideal
 physics, such as cooling and viscosity, modifies the classic dispersion
@@
 lead to GI even when $Q>1$, by removing pressure and rotational
-support, respectively.
+support, respectively.  We also review the origins for the transition from self-regulated GI 
+to fragmentation seen in numerical simulations. {\bf add sentence here more like s 1.4 laying
+out what you are about to talk about. at the very least a transition sentence or two?}
 
@@
 type of parameterized cooling, first applied by \cite{gammie01},
-allows the full range of thermodynamic response to be explored. 
-{\bf physical origin of cooling in PPDs}
+allows a range of thermodynamic responses to be explored. In reality, PPD cooling is controlled by radiation 
+from dust grains \citep{bell94,dalessio97,chiang97}. Although most PPDs subject to GI are optically thick, 
+cooling parameterized by the $\beta$ model above only captures optically thin cooling when used in numerical
+simulations.
+%{\bf physical origin of cooling in PPDs}
 
-We will quantify how cooling enables GI
-even when $Q>1$. For example, if perturbations are
-cooled towards zero tempertaure, we find the above dispersion relation is
+%%KMK 
+Previous work has quantified the role of cooling primarily as a means to reduce the sound speed term
+in $Q$ (though see e.g. \citealt{clarke07}). Here we will quantify how cooling enables GI
+even when $Q>1$. For example, if perturbations can
+cool towards zero temperature, we find the above dispersion relation is
 modified to read 
@@
 Viscous disks can also develop GI even when $Q>1$. This is because, as 
-demonstrated below, viscosity eliminates rotational support against  
+demonstrated below, viscosity removes rotational support against  
 self-gravity for long-wavelength disturbances 
@@
 formation %frag of dust 
-\citep{youdin11}. Then it is not unreasonable to expect an analogous
-fragmentation of gaseous disks due to viscosity.
+\citep{youdin11}. Therefore it is not unreasonable to expect analogous
+fragmentation in gaseous disks due to viscosity.
 
@@
 where $\alpha$ is the dimensionless viscosity
-coefficient. This
+coefficient \cite{shakura73}. This
 parameterization can be modified to include more complex dependencies on the
 fluid variables, but Eq. \ref{alpha_param} is the general form. 
-{\bf maybe no need to intro alpha param here?}
+%KMK I think this is sufficient here {\bf maybe no need to intro alpha param here?}
 
 For an isothermal, viscous, self-gravitating disk in 2D,
-\cite{gammie96} finds the approximate disperion relation   
+\cite{gammie96} finds the approximate dispersion relation   
 \begin{align}
@@
 provided that $\nu k^2\neq0$. This condition for viscous GI is 
-identical that would be obtained from Eq. \ref{classic_dr} with
+identical to what would be obtained from Eq. \ref{classic_dr} with
 rotation neglected ($\Omega\to 0$). That is, a classically stable
-disk can be destabilized by viscosity as it effectively removes
+disk can be destabilized by viscosity as it effectively eliminates
 rotational stabilization \citep{lynden-bell74}.
@@
 as an effective viscosity is a separate issue 
-\citep{balbus99}. However, it is reasonable to assume that turbulence
-frustates rotational support. (This is, after all, the physical
-mechanism to turbulent mass accretion.) 
-For example, \cite{fromang05} found numerical evidence that MHD
-turbulence can enhance GI and lead to disk fragmentation (see also
-\S\ref{MHD}).
+\citep{balbus99}. However, it is reasonable to assume that these mechanisms, which are observed in simulations 
+to transport angular momentum outwards (and mass inwards),
+frustate rotational support. %KMK (This is, after all, the physical mechanism to turbulent mass accretion.) 
+%%KMK this comment seems out of place here 
+%For example, \cite{fromang05} found numerical evidence that MHD
+%turbulence can enhance GI and lead to disk fragmentation (see also
+%\S\ref{MHD}).
 
-In this work, we use viscosity to model such a 
-physical effect due to turbulence, but we emphasize that viscous 
-GI does not assume a particular origin for the viscosity. We will
+In this work, we use viscosity to model two possible
+physical effects of turbulence: heating via dissipation and angular momentum transport. 
+%KMK
+We emphasize that our model for GI in the presence of viscosity (hearafter Viscous GI)
+does not assume a particular origin for the viscosity. GI itself or other forms of magnetic or hydrodynamic turbulence are allowed
+in this framework. We will
 generalize the theory of viscous GI to include an energy equation  with viscous
 heating, irradiation, explicit cooling, as well as
-three-dimensionality (3D). This allows us to consider viscous GI in
+three-dimensionality (3D) in order to consider viscous GI in
 more realistic PPD models.
@@
 We will, however, apply the theory to `gravito-turbulent'
-disks in which the viscosity associated with some underlying (classic)
+disks in which the viscosity is associated with some underlying (classic)
 GI, as described below.
 
-Consider a disk initially laminar without a heating mechanism and
-cools. Then the disk temperature will be lowered until $Q=O(1)$, whence
-non-axisymmetric modes grow and heats the disk through the dissipation
-of spiral shocks. %new rafikov 16 
-This permits a quasi-steady, turbulent state with
+Consider an initially laminar, $Q>>1$ disk, without an external heating mechanism, as it
+cools. The disk temperature will decline until $Q=O(1)$, whence
+non-axisymmetric modes grow and heat the disk through the dissipation
+of spiral shocks \citep{cossins09}. %new rafikov 16 
+This setup permits a quasi-steady, turbulent state with
 $Q =O(1)$ in which cooling is balanced by shock heating to maintain
 thermal equilibrium \citep{gammie01}. 
-Represent the dynamic and thermodynamic effect of this turbulence by a
+One may represent the dynamic and thermodynamic effect of this turbulence by a
 viscosity. Then the $\alpha$ and $\beta$ parameters defined above will
-be inversely related.
+be inversely related ({\bf KMK see foreward ref eq...?}).
 
-Experimentally, however, it is found that if the cooling rate or
+Numerical experiments, however, show that if the cooling rate or
 viscosity is too large, say,   
@@
 then the disk fragments \citep{gammie01,rice05,rice11}. In fact, in
-the absence of global effects, numerical simulations show there are
+the absence of global effects, numerical simulations show that there are
 two --- and only two --- possible outcomes for self-gravitating,
-cooling disks: gravito-turbulence or fragmentation.  
-%more accurately: given a total simulation time, there are two
+cooling disks: gravito-turbulence or fragmentation. %more accurately: given a total simulation time, there are two
 %outcomes 
@@
 and whether or not a critical cooling time can be defined at all 
-\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. 
+\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. {\bf KMK We defer a discussion of
+numerical convergence issues to Section \ref{sec-disc}.?}
 %mention numerical effects? 
@@
   exist for sufficiently rapid cooling or large viscosity}.  
-This intriguing because, as highlighted in
+This  is intriguing because, as highlighted in
 \S\ref{cool_gi}---\ref{visc_gi}, these effects can reduce
-gravitational stability. This motivates a physical    
+gravitational stability independently of their influence on the exact value of the classic $Q$. This motivates a physical    
 interpretation of disk fragmentation as the inability to maintain a
-gravito-turbulent state due to the action of cooling and/or viscosity.
+gravito-turbulent state due to %the action of 
+%KMK
+secondary instabilities driven by cooling and/or viscosity.
 

### 2016-01-26 paper/results2d.tex
KMK edits

--- 2016-01-25 paper/results2d.tex
@@
 \section{Two-dimensional disks with beta cooling}\label{result_2d}
-We begin in the 2D limit with standard beta cooling  to make connection 
+We begin in the 2D limit with standard beta cooling  to facilitate comparison 
 with previous studies. The disk
-material is assumed to be confined to the midplane and $\delta v_z=0$. 
+material is assumed to be confined to the midplane, and $\delta v_z=0$. 
 We make the replacement  
@@
 \emph{sole} effect of cooling on the 
-perturbations. A time-independent heat source should be
+perturbations. %KMK
+We emphasize that destabilization is independent of the effect of decreasing temperature on the instantaneous
+value of the classic-$Q$. A time-independent heat source should be
 invoked to balance the imposed cooling to allow an equilibrium to be
@@
                                 %like background heat source only 
-
+	%KMK interesting points. discuss later?
 Eq. \ref{thindisk} becomes 
@@
 pressure, respectively. The imposed cooling/irradiation only affects the
-pressure response. {\bf do we need to explain the terms again?}
+pressure response. %KMK no...? {\bf do we need to explain the terms again?}
 
@@
 i.e. as the floor temperature, towards which perturbations are cooled, drops. 
+{\bf KMK maybe more words here about apparent trend of $Q \rightarrow \inf$ for small $\theta$
+even forward ref. lengthscales below?}
 
@@
 upper limit to unstable wavenumbers, which could have important
-implications for numerical simulations (see \S\ref{prev_works}).
+implications for numerical simulations probing large wavenumbers and small scales
+at high resolution (see \S\ref{prev_works}).
 
@@
 S_*<0 $ for $\gamma>\theta$. That is, as cooling becomes more rapid
-the disk becomes more unstable at shorter wavelengths. 
+the disk becomes more unstable at shorter wavelengths.  {\bf KMK -- I don't follow}
 %show dK*/dbeta < 0 
@@
 For beta cooling with $\theta=0$, the disk is unconditionally unstable
-for finite $Q$.  The $\beta\to0$ limit corresponds to a 
+for finite $Q$. {\bf KMK some caveating about scales}  The $\beta\to0$ limit corresponds to a 
 pressureless disk (not merely isothermal).
@@
 \end{align} 
+{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
 so that it is marginally stable in the absence of cooling.  
@@
 \end{align}
-This equation reproduce similar values 
+This equation reproduces remarkably similar values 
 of the cooling times below which 
@@
 equivalent for $\gamma=1$.) The most unstable wavenumber and growth
-rates vary weakly with $\beta$.
+rates vary weakly with $\beta$.   {\bf KMK: summarize boring result? not sure 
+what this simplifies to in comparison to other limits?}.
 
@@
 
+%KMK moved fig ref. 
+Fig. \ref{gammie_rate_plot} shows growth rates as a function of the
+wavenumber $k$ obtained from the dispersion relation
+Eq. \ref{thindisk}.  To compute growth rates numerically, 
+ we consider a model with $\alpha_b=0$ and
+$\alpha=\alpha(\beta)$ given by thermal equilibrium (
+Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
+\begin{align}
+  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
+\end{align}
+%with $Q_\mathrm{crit}$ given by 
+to mimic a basic state maintained by gravito-turbulence where one
+might expect the dimensionless stress $\alpha \sim Q^{-2}$
+\citep{lin87}.
+
+It is useful to consider several limiting cases.
 To see the effect of cooling and irradiation, we simplify the
@@
 
-As a numerical example, we consider a model with $\alpha_b=0$ and
-$\alpha=\alpha(\beta)$ given by thermal equilibrium (
-Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
-\begin{align}
-  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
-\end{align}
-%with $Q_\mathrm{crit}$ given by 
-to mimic a basic state maintained by gravito-turbulence where one
-might expect the dimensionless stress $\alpha \sim Q^{-2}$
-\citep{lin87}.   
+%%KMK MOVED?
+%As a numerical example, we consider a model with $\alpha_b=0$ and
+%$\alpha=\alpha(\beta)$ given by thermal equilibrium (
+%Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
+%\begin{align}
+ % Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
+%\end{align}
+%%with $Q_\mathrm{crit}$ given by 
+%to mimic a basic state maintained by gravito-turbulence where one
+%might expect the dimensionless stress $\alpha \sim Q^{-2}$
+%\citep{lin87}.   
 %should really only use this Q-alpha relation for order-unity values
@@
 
-Fig. \ref{gammie_rate_plot} show growth rates as a function of the
-wavenumber $k$ obtained from the dispersion relation
-Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
+%Fig. \ref{gammie_rate_plot} shows growth rates as a function of the
+%wavenumber $k$ obtained from the dispersion relation
+%Eq. \ref{thindisk}. 
+The limiting behavior for small/large $K$ are
 well-captured by Eq. \ref{gammie_smallk} and
-\ref{gammie_bigk}. Comparing the two panels shows that increasing the
-irradiation level ($\theta$) surpresses small-scale
+\ref{gammie_bigk}. Comparing the two panels of Fig. \ref{gammie_rate_plots} shows that increasing the
+irradiation level ($\theta$) supresses small-scale
 perturbations. %(This also occurs in the inviscid case.)
@@
 \end{align*}
+{\bf KMK physically what is the limit $\beta \ll \alpha$?}
 where we have applied Eq. \ref{alpha_beta_relation}
@@
 the cooling time for $\beta\lesssim O(1)$ because larger scales are
-more resistant to the associated increase in viscous damping.
+more resistant to the associated increase in viscous damping.   {\bf KMK this is
+confusing b/c viscosity is in general acting on larger scales than pressure. also damping -> dissipation or 
+angular momentum transport?}
 
@@
 maximum $\alpha$ ($\sim 0.06$) that can be sustained before
-fragmentation \citep{rice11}. We can interpret this result in our
+fragmentation \citep{rice11} {\bf KMK rice 05a,b, no?}. We can interpret this result in our
 linear framework. 
@@
 and hence highly turbulent, self-gravitating disks to persist beyond
-dynamical timescales. This is consistent with previous numerical
+dynamical timescales.
+%KMK
+We interpret fragmentation as the response to the instability of the
+highly viscous state. This is consistent with previous numerical
 simulations performed by \cite{lodato05}. 

### 2016-01-26 paper/results3d.tex
updating figs, need to correct 3d beta results, need to re-comp bg if Q depends on alpha
KMK edits

@@
 We confirm the above results in 3D disks with vertical
-structure. This is expected to weaken gravitational instabilities
-because the disk mass is spread across the vertical dimension. It is
-possible to incorportate this effect in the previous 2D framework, but this
+structure. Accounting for the third dimension will weaken gravitational instabilities
+because the disk mass is spread across some vertical extent. It is
+possible to incorporate this effect in the previous 2D framework, but doing so
 introduces additional parameters as discussed in Appendix
@@
 ($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
-constant (Eq. \ref{alpha_beta_relation}). We consider even modes about
+constant (Eq. \ref{alpha_beta_relation}). We consider {\bf only} even modes about
 the midplane, and apply a numerical disk surface at $z=\zmax$ such
@@
 The 3D gravity parameter $\qthree$ is defined by 
-Eq. \ref{Q3d_def}. For this case the corresponding Toomre parameter 
+Eq. \ref{Q3d_def}. For such a disk, the corresponding Toomre parameter 
 $Q=2Q_\mathrm{crit}$. Recall $Q_\mathrm{crit}$, defined by
@@
 
-Fig. \ref{3d_inviscid} show growth rates and the most unstable
-wavenumber obtained for this model. We also plot 2D
+Fig. \ref{3d_inviscid} shows growth rates and the most unstable
+wavenumbers obtained for this model. We also plot 2D
 results with the 3D correction as described in Appendix
 \ref{3dcorr}. The (empirically) chosen value of
-$H_\mathrm{sg}=0.64H$ results in a close match between 2D and 3D
+$H_\mathrm{sg}=0.64H$, also defined in Appendix \ref{3dcorr}, results in a close match between 2D and 3D
 growth rates, but the most unstable wavenumber in 3D is somewhat smaller. 
-This reflects self-gravity being weakened in the vertical 
+This offset reflects self-gravity being weakened in the vertical 
 direction: a larger horizontal scale is required to achieve the same
@@
 
-Fig. \ref{3d_visc} show growth rates, maximized over $k$, as a
+Fig. \ref{3d_visc} shows growth rates, maximized over $k$, as a
 function of the cooling time $\beta$. We also plot 2D results with 3D
 corrections. Softening the self-gravity in 2D captures the correct 
-qualitative behavior as in the full 3D case, but it is clear that a
+qualitative behavior as in the full 3D case, but once again it is clear that a
 single, constant value of $H_\mathrm{sg}$ cannot re-produce 3D growth 
@@
 1,\,10$ and $100$. Cooling generally enhances vertical motion
-(convection). However, even in the highly unstable case with $\beta=1$,
+(convection) {\bf KMK really convection?}. However, even in the highly unstable case with $\beta=1$,
 vertical speeds are sub-dominant at $\sim 30\%$ of the horizontal
-speeds. This suggests that it is not essential to consider 3D models
-when considering gravitational instabilities.
+{\bf KMK  azimuthal or radial?} speeds. This suggests that it is not essential to consider 3D models
+when considering gravitational instabilities. {\bf KMK this is not true generically, so
+this statement needs to be more specific or cut}.
 

### 2016-01-26 paper/summary.tex
KMK edits

@@
 We use a Navier-Stokes viscosity to mimic the effects of
-turbulence. %which may be of any origin, though we specialize to
+turbulent angular momentum transport.
+%KMK
+. %which may be of any origin, though we specialize to
             %gravito-turb 
@@
 and/or viscosity.
+
+%KMK
+Previous work has focused on the impact of viscosity and 
+cooling on the equilibrium temperature of the disk, but merely used this 
+to measure the classic Toomre Q and thereby assess stability. We demonstrate
+through a more rigorous inclusion of these effects into the dispersion relation that
+they can drive secondary instabilities.
 While viscosity and cooling can be related through thermal
@@
 stabilization. This provides a physical explanation as to why
-rapidly-cooled, gravito-turbulent disks cannot exist.
+rapidly-cooled, gravito-turbulent disks cannot exist. Moreover, we discuss below how
+these models may lend support to the varied behavior observed in numerical simulations.
+%KMK
 
@@
 Eq. \ref{inviscid}. We find a sufficient condition for stability which  
-depends on the irradiation level (Eq. \ref{stable_condition}) but 
+depends on the irradiation level (Eq. \ref{stable_condition}) but is
 \emph{independent} of the cooling time. 
@@
 %maybe that's why irradiated disks produce smaller alpha     
+%%KMK this might be worth saying
 
@@
 realistic models for cooling and gravito-turbulence. 
-We show that with a realistic cooling model for PPDs, 
-cooling alone cannot lead to gravitational instabilities. This is due
-to stabilization by an effective irradiation associated with the PPD
+We show that with a physically motivated cooling model for PPDs, 
+cooling alone does not lead to gravitational instabilities. This is due
+to stabilization by an effective `irradiation' associated with the PPD
 cooling function, which is present even if 
-there is no physical irradiation. %due to matching functional forms of
+there is no physical irradiation. Physically, this  term
+represents the fact that disks do not cool towards zero temperature because
+radiative cooling is a strong (quartic) function of temperature  %%KMK?
+
+%due to matching functional forms of
                                 %cooling 
@@
 pressure stabilization by cooling on a radial scale of the disk scale
-height. %i.e. cool-driven g 
+height.
+%%KMK
+%Note that unlike our physical cooling function (Eq. \ref{real_cool}), the standard beta function in these
+%numerical simulations cools towards zero temperature, rather than a temperature floor or irradiation temperature 
+%%KMK some comment like that here, or too redundant with comment below??
+ %i.e. cool-driven g 
 {\bf don't know if we want to dwell on this table}
@@
 More recent simulations have raised the issue of numerical
-convergence. \cite{paardekooper12} found that fragmentation can occur 
+convergence. %KMK
+\cite{meru11} found that better resolved disks fragmented at longer cooling times.
+Follow-up studies attributed at least some of this effect to decreasing numerical viscous
+heating at higher resolution, which helps fragmentation
+\citep{lodato11,meru12}.
+
+ \cite{paardekooper12} also found that fragmentation can occur 
 for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
@@
 
+%%KMK moved up 
 %resolution convergence 
-Another issue is convergence with numerical resolution. For example,
-\cite{meru11} found that the critical cooling time for 
-disk fragmentation increases with numerical resolution. %sph particles in
+%Another issue is convergence with numerical resolution. For example,
+%cite{meru11} found that the critical cooling time for 
+%disk fragmentation increases with numerical resolution. %sph particles in
 % their case 
-Follow-up studies attributed this to decreasing numerical viscous
-heating at higher resolution, which helps fragmentation
-\citep{lodato11,meru12}. %also global effects in
+%Follow-up studies attributed this to decreasing numerical viscous
+%heating at higher resolution, which helps fragmentation
+%\citep{lodato11,meru12}. %also global effects in
                          %these. (paardekooper/meru/baruteau) 
@@
 (see also \S\ref{cool_gi}) that this implies cooling-driven GI can
-occur at any sufficiently small scales. Then as the numerical
+occur at any sufficiently small scales. Therefore as the numerical
 resolution increases, simulations can access a wider range of unstable
 scales.  %no upper bound  
-In this respect, it is not surprising to find non-convergence with 
+In this respect, it is perhaps not surprising to find non-convergence with 
 increasing resolution. %simple beta cooling will not give a
@@
 cooling and PPD cooling (\S\ref{ppd_cooling}). 
+%KMK
+Moreover, the standard beta cooling does 
+not account for optical depth effects, which force the midplane 
+and high density perturbations to cool more slowly.
 
@@
 
-On the other hand, \cite{baehr15} included a floor temperature 
+%KMK
+A temperature floor may be a necessary, but not sufficient condition
+for numerical convergence. \cite{baehr15} included a floor temperature 
 in their simulations but still find that at fixed cooling rates, disks
@@
 In light of our results on viscous GI, 
-we here highlight a possible contribution: %in addition to num visc (meru lodato)
-high resolutions enable small-scale turbulent angular momentum transport 
+we suggest another possible contribution to non-convergence: %in addition to num visc (meru lodato)
+high resolution enables small-scale turbulent angular momentum transport 
 % physical, cf numerical  
-to aid clump formation (see also \S\ref{MHD} below). 
-This mechanism is unavailable at low resolutions, for which one must
-rely on rapid cooling.  
+to aid clump formation via the removal of rotational support. (see also \S\ref{MHD} below). 
+If clumps are only marginally resolved, simulations may not have sufficient dynamic range for turbulent eddies
+to cascade down to these scales.
+%This mechanism is unavailable at low resolutions, for which one must
+%rely on rapid cooling.  
 %maybe there is an effective floor temp which depends on res
@@
 We comment that although modern simulations resolve the dominant
-scale associated with gravito-turbulence, $kH\sim 1$
+scale associated with gravito-turbulence very well, $kH\sim 1$
 \citep{cossins09}, this does not
@@
 as is evident from the non-converging simulations described above.   
-
+%KMK
+Even at very high resolution one might worry that the artificial dissipation scale
+imposed by the grid / smoothing length is too similar to scale of fragmentation.
 %it is true that most unstable wavelength is H is this is always
@@
 Explicit numerical simulations of magnetized, massive disks have been performed 
-by \cite{fromang05}. He finds disk fragmentation with increasing numerical
+by \cite{fromang05}. This study finds disk fragmentation with increasing numerical
 resolution, and attributes this to resolving the most unstable MRI
@@
 turbulence to aid fragmentation. %clumping/mass accretion onto clump
-This would be a physical mechnism represented by the viscous GI 
-discussed in this paper; which lends some support for  
+This physical mechanism is represented by the viscous GI 
+discussed in this paper, which lends some support for  
 the use of a viscosity to represent turbulence.
@@
 True disk fragmentation is a non-linear process characterized by
-clumps reaching densities orders of magnitude above the ambient value, and
-which survives tidal shear and shocks \citep{paardekooper12,young16}.  
+clumps reaching densities that are orders of magnitude above the ambient value. They must also survive
+disruption by tidal shear and shocks \citep{shlosman87,young16}.  
 Clearly, our linear models cannot address fragmentation
@@
 self-gravitating disks in the local limit \citep[][]{gammie01}.    
-Then it seems reasonable to speculate that the non-existence of a stable
+Thus it seems reasonable to speculate that the non-existence of a stable
 gravito-turbulent state, here due to dynamical instability, implies
@@
 of large, non-linear density enhancements, which arise from the
-gravito-turbulent flucuations simply by chance. 
+gravito-turbulent flucuations simply by chance. %KMK
+There is insufficient evidence that gravito-turbulence adequately samples
+the density power spectrum as assumed by \cite{hopkins13}.
 %cooling affects survival rate 
-This might be considered as a secondary instability by lowering the
+Nevertheless, one might consider this form of fragmentation as a secondary instability
+triggered by lowering the local
 Toomre parameter through a (random) increase in density 
@@
 destabilizing mechanisms discussed in this paper lowers $Q$ by
-reducing pressure and Coriolis forces. %we have a physical way of lowering pressure/rotation
+reducing pressure and Coriolis forces. 
+{\bf KMK: this last sentence sounds like it's just about changing classic Q}.
+%we have a physical way of lowering pressure/rotation
 
@@
 non-axisymmetric global GI can develop for $Q$ somewhat larger than unity
-\citep{papaloizou89,adams89,papaloizou91}. It would be interesting to study
+\citep{papaloizou89,adams89,papaloizou91,laughlin97}. It would be interesting to study
 how non-axisymmetric perturbations are affected by cooling and

### 2016-01-28 paper/model3d.tex
revision 3 by MKL

@@
 \end{align}
-where $\mathcal{R}$ is the gas constant and $T$ is the temperature. {\bf KMK cut because 
-mostly varied: We
-use $\gamma=1.4$ unless otherwise stated.} The gas 
+where $\mathcal{R}$ is the gas constant and $T$ is the temperature. The gas 
 gravitational potential $\Phi$ is given via the Poisson equation,  
@@
 as a viscous process. We note that if $\rho\nu$ is constant in time
-then increasing the density leads to reduction in the viscosity. 
+then increasing the density corresponds to reduction in the viscosity. 
 This is perhaps not inconsistent with numerical simulations of disk
 fragmentation which show the internal flow of high-density clumps
-is laminar \citep{gammie01}, although once clumps form they may effectively decouple
-from the background disk state, and thus no longer be described by the same prescription.   
-{\bf is this true? KMK: maybe caveat helps?}
+appear laminar \citep{gammie01}. However, once clumps form they may
+effectively decouple from the background disk state, and thus no 
+longer be described by the same prescription. 
+%we are not concerned with non-linear evolution    
+%{\bf is this true? KMK: maybe caveat helps?}
 
@@
 for example, stellar or background irradiation. We will thus refer to
-$\theta$ as the `irradiation level'. As we discuss in Section ?, 
-the limits $\theta \rightarrow 0, \theta \rightarrow 1$ do not exactly 
-correspond to the limits of non-irradiated or irradiation-dominated disks
-as discussed in the literature \citep{kratter11,rice11}.%stellar illum or envelope rad
+$\theta$ as the `irradiation level'. {\bf ASK} As we discuss in
+\S\ref{2d_inviscid}, the limits $\theta \rightarrow 0, \theta
+\rightarrow 1$ do not exactly correspond to the limits of
+non-irradiated or irradiation-dominated disks as discussed in the
+literature \citep{kratter11,rice11}. 
+%stellar illum or envelope rad
 
@@
 The vertical structure is first obtained from 
-Eq. \ref{vert_eq1}---\ref{vert_eq2} then inserted into
+Eq. \ref{vert_eq1}---\ref{vert_eq2}, %given strength of self-grav, via Q3d
+then inserted into 
 Eq. \ref{thermal_eq} to infer the required viscosity profile for
@@
 framework we develop. We do this in \S\ref{2dppd} where we adopt a
-more realistic beta cooling function for PPDs. In this case, $\theta$
+more realistic beta cooling function for PPDs. In that case, $\theta$
 may or may not represent a physical irradiation.
@@
 \begin{align}
-  \Lambda &= \nabla\cdot\bm{F},\label{rad_cool1}\\
-  \bm{F}   &= -\frac{16\sigma T^3}{3\kappa_d\rho}\nabla T, \label{rad_cool2}
+  \Lambda &= \nabla\cdot\bm{F}_\mathrm{rad},\label{rad_cool1}\\
+  \bm{F}_\mathrm{rad}   &= -\frac{16\sigma T^3}{3\kappa_d\rho}\nabla T, \label{rad_cool2}
 \end{align}
@@
 While radiative cooling is arguably more realistic than
-beta cooling, it generally implies a vertically
+$\beta$-cooling, it generally implies a vertically
 non-isothermal equilibrium disk, and increases the order of the
@@
 disks, but it is possible to modify the flux function to account
-for optically-thin disks. However, this complication is beyond the scope of
-this work. {\bf ref for flux lim diff? \citep{levermore81}?}
+for optically-thin disks \citep{levermore81}. However, this
+complication is beyond the scope of this work.

### 2016-01-28 paper/paper3d.tex
revision 3 by MKL

@@
   Low mass, self-gravitating accretion disks admit quasi-steady,
-`gravito-turbulent' states in which cooling balances turbulent {\bf viscous?}
+  `gravito-turbulent' states in which cooling balances turbulent viscous 
   heating. However, numerical simulations show that gravito-turbulence
   cannot be sustained beyond dynamical timescales when the cooling rate is too fast,
-or corresponding turbulent viscosity is too large. The result is is disk fragmentation.        
-We motivate and quantify an interpretation of disk % a physical
-                                % interpretation 
+  or when the corresponding turbulent viscosity is too large. The result is is disk fragmentation.         
+  We motivate and quantify an interpretation of disk 
   fragmentation as the inability to maintain gravito-turbulence due to  
-{\bf proper / formal / etc?} secondary instabilities driven by:  
+  formal secondary instabilities driven by:  
 \begin{inparaenum}[1)] 
@@
 consider parameterized cooling functions in two and three dimensions,
-as well as %full 
+as well as 
 radiative diffusion in 3D. We show that 
-%cooling/viscosity can destabilize the disk even when the classic
-%Toomre stability condition is satisfied; 
 generally there is no critical cooling rate/viscosity below which 
-the disk is formally stable, although interesting limits appear when we examine the growth
-of modes of order the disk scale height. 
+the disk is formally stable, although interesting limits appear 
+for unstable modes with lengthscales on the order of the disk scale height.   
+%when we examine the growth 
+%of modes of order the disk scale height. 
 We apply this new linear theory to 
@@
 viscosity, and cooling via dust opacity. 
-%We find the sole effect of
-%cooling is insufficient to 
-%destabilize a realistic PPD. 
 We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
@@
 fragmentation by removing rotational stabilization against 
-self-gravity, and that the observed transition in behavior from gravito-turbulent to fragmenting may derive from
+self-gravity, and that the observed transition in behavior from 
+gravito-turbulent to fragmenting may derive from  
 instability of the gravito-turbulent state itself.
@@
 axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
-adiabatic (including isothermal) disk. Such disturbances are formally
-unstable when $Q<1$. Non-axisymmetric modes can develop for larger,
-but still order-unity values of $Q$
-\citep{lau_bertin78,papaloizou89,papaloizou91}. 
+adiabatic (including isothermal) disk. 
+%Such disturbances are formally
+%unstable when $Q<1$.  
+When $Q<1$, there is a range of $k$ for which such disturbances are
+unstable.  
+Non-axisymmetric modes can develop for larger, but still order-unity
+values of $Q$ 
+\citep{lau78,papaloizou89,papaloizou91}. 
 %strictly speaking non axi *local* modes are stable -> only transient
@@
 
-We emphasize that the oft-used dispersion relation and corresponding Toomre parameter, %It is important to remember that the popular dispersion relation, 
-(Eqs. \ref{toomreQ_criterion} and \ref{classic_dr}) are derived from \emph{idealized} conditions: 
+We emphasize that the oft-used dispersion relation and corresponding
+Toomre parameter (Eqs. \ref{toomreQ_criterion} and \ref{classic_dr})
+are derived from \emph{idealized} conditions:  
 the base disk is laminar and does not experience any net thermal 
@@
 However, real disks can cool, for example, %radiative, convective cooling
-due to radiative losses. Accretion disks may also be turbulent, the dynamic and thermodynamic effect of which is often modeled through an
+due to radiative losses. Accretion disks may also be turbulent, the 
+dynamic and thermodynamic effect of which is often modeled through an 
 effective viscosity \citep{shakura73,lin87,armitage01,rafikov15}. 
@@
 
-Despite this, the Toomre parameter %and its associated
-%dispersion relation are
+However, the Toomre parameter  
 is still widely applied to viscous (turbulent), cooling
-accretion disk models \citep[e.g.][]{gammie01,cossins09, kimura12}, {\bf more refs?}
-%%KMK
+accretion disk models \citep[e.g.][]{gammie01,cossins09, kimura12}, 
 despite the mismatch in the included underlying physics. 
-%even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
-%derived from inviscid disks without cooling. 
-%e.g. using classic DR to determine what scales need to be resolved
+We show that including non-ideal physics, such as cooling and
+viscosity, in fact modifies the classic dispersion 
+relation, and hence the condition for GI. Thus the Toomre condition
+alone is insufficient to assess GI in realistic disks.
 
-We show in this work that including non-ideal
-physics, such as cooling and viscosity, modifies the classic dispersion
-relation, and hence the condition for GI. Thus the Toomre condition
-alone is insufficient to assess GI in realistic disks.  The goal of
-this work is to generalize the analytic treatment of disk GI to allow
+The goal of this work is to generalize the analytic treatment of disk
+GI, by including cooling and viscosity, to allow 
 a self-consistent discussion of GI in realistic accretion disk
-models. Specifically, we will examine how  cooling and viscosity can
-lead to GI even when $Q>1$, by removing pressure and rotational
-support, respectively.  We also review the origins for the transition from self-regulated GI 
-to fragmentation seen in numerical simulations. {\bf add sentence here more like s 1.4 laying
-out what you are about to talk about. at the very least a transition sentence or two?}
+models.  
+%Specifically, we will examine how  cooling and viscosity can
+%lead to GI even when $Q>1$, by removing pressure and rotational
+%support, respectively.  
+We first review in \S\ref{cool_gi} and \S\ref{visc_gi}, respectively,  
+how cooling or viscosity can lead to GI even when $Q>1$, by removing
+pressure or rotational support. We then 
+discuss in \S\ref{frag_intro} how these effects may relate   
+to the transition from self-regulated GI to fragmentation seen in
+numerical simulations. The rest of this paper is laid out in
+\S\ref{plan}.
+
+%{\bf add sentence here more like s 1.4 laying
+%out what you are about to talk about. at the very least a transition sentence or two?}
 
@@
 $\tcool$ is the timescale over which the disk temperature $T$ is relaxed
-to some floor value $T_0$, which may be zero. 
+to some floor value, which may be zero. 
 %That is, $\p_t T =
@@
 type of parameterized cooling, first applied by \cite{gammie01},
-allows a range of thermodynamic responses to be explored. In reality, PPD cooling is controlled by radiation 
-from dust grains \citep{bell94,dalessio97,chiang97}. Although most PPDs subject to GI are optically thick, 
-cooling parameterized by the $\beta$ model above only captures optically thin cooling when used in numerical
-simulations.
-%{\bf physical origin of cooling in PPDs}
+allows a range of thermodynamic responses to be explored.
+
+In reality, PPD cooling is controlled by radiation   
+from dust grains \citep{bell94,dalessio97,chiang97}. Although most
+PPDs subject to GI are optically thick, cooling parameterized by the
+$\beta$ model formally only captures optically thin cooling when
+used in numerical simulations. It is, however, possible to modify the
+standard cooling function to mimic optically-thick cooling (see,
+e.g. \S\ref{2dppd}).
 
@@
 Previous work has quantified the role of cooling primarily as a means to reduce the sound speed term
-in $Q$ (though see e.g. \citealt{clarke07}). Here we will quantify how cooling enables GI
+in $Q$ \citep[though see e.g.,][]{clarke07}. Here we will quantify how cooling enables GI
 even when $Q>1$. For example, if perturbations can
-cool towards zero temperature, we find the above dispersion relation is
+cool to zero temperature, we find the above dispersion relation is
 modified to read 
@@
 pressure neglected ($c_s\to 0$). This condition does not actually
-depend on the cooling rate, and instability is
+depend on the cooling time, and instability is
 possible for any finite $Q$. %at sufficiently small scales 
-This shows that the mere presence of
-cooling changes the qualitative nature of GI compared to the simple
-Toomre condition.   
+Thus, the mere presence of cooling changes the qualitative nature of
+GI compared to the simple Toomre condition. 
+%singular limit
+
 %{\bf numerical implications? if using standard beta cooling, then as
@@
 as an effective viscosity is a separate issue 
-\citep{balbus99}. However, it is reasonable to assume that these mechanisms, which are observed in simulations 
+\citep{balbus99}. However, it is reasonable to assume that these
+mechanisms, which are observed in simulations  
 to transport angular momentum outwards (and mass inwards),
-frustate rotational support. %KMK (This is, after all, the physical mechanism to turbulent mass accretion.) 
-%%KMK this comment seems out of place here 
-%For example, \cite{fromang05} found numerical evidence that MHD
-%turbulence can enhance GI and lead to disk fragmentation (see also
-%\S\ref{MHD}).
+frustate rotational support.
 
@@
 %KMK
-We emphasize that our model for GI in the presence of viscosity (hearafter Viscous GI)
-does not assume a particular origin for the viscosity. GI itself or other forms of magnetic or hydrodynamic turbulence are allowed
-in this framework. We will
-generalize the theory of viscous GI to include an energy equation  with viscous
-heating, irradiation, explicit cooling, as well as
-three-dimensionality (3D) in order to consider viscous GI in
-more realistic PPD models.
+We emphasize that our model for GI enabled by viscosity (hearafter
+viscous GI) does not assume a particular origin for the viscosity. GI
+itself or other forms of magnetic or hydrodynamic turbulence are 
+allowed in this framework. We will generalize the theory of viscous GI
+to include an energy equation  with viscous heating, irradiation,
+explicit cooling, as well as three-dimensionality (3D) in order to
+consider viscous GI in more realistic PPD models.
 
@@
 
-\subsection{Relevance to gravito-turbulent disk fragmentation}
+\subsection{Relevance to gravito-turbulent disk fragmentation}\label{frag_intro}
 We develop a general framework for viscous disks without  
 assuming a specific origin of the turbulence that this viscosity 
-represents. %without assuming the nature of the turbulence represented
-            %by this viscosity  
+represents. 
 We will, however, apply the theory to `gravito-turbulent'
-disks in which the viscosity is associated with some underlying (classic)
-GI, as described below.
+disks in which the viscosity is associated with some underlying
+(classic) GI, as described below.
 
-Consider an initially laminar, $Q>>1$ disk, without an external heating mechanism, as it
-cools. The disk temperature will decline until $Q=O(1)$, whence
-non-axisymmetric modes grow and heat the disk through the dissipation
-of spiral shocks \citep{cossins09}. %new rafikov 16 
+Consider an initially laminar, $Q\gg1$ disk, without an external
+heating mechanism, as it cools. The disk temperature will decline
+until $Q=O(1)$, whence non-axisymmetric modes grow and heat the disk
+through the dissipation of spiral shocks \citep{cossins09}. 
+%new rafikov 16  
 This setup permits a quasi-steady, turbulent state with
@@
 thermal equilibrium \citep{gammie01}. 
-One may represent the dynamic and thermodynamic effect of this turbulence by a
-viscosity. Then the $\alpha$ and $\beta$ parameters defined above will
-be inversely related ({\bf KMK see foreward ref eq...?}).
+If one represents this turbulent heating by a viscous heating, then
+the $\alpha$ and $\beta$ parameters defined above will  
+be inversely related (see, e.g. Eq. \ref{alpha_beta_relation}).
 
-Numerical experiments, however, show that if the cooling rate or
-viscosity is too large, say,   
+Numerical experiments, however, show that if the cooling time is too
+small (or the viscosity is too large), say,   
 \begin{align}\label{frag_cond}
@@
 two --- and only two --- possible outcomes for self-gravitating,
-cooling disks: gravito-turbulence or fragmentation. %more accurately: given a total simulation time, there are two
+cooling disks: gravito-turbulence or fragmentation. 
+%more accurately: given a total simulation time, there are two
 %outcomes 
@@
 and whether or not a critical cooling time can be defined at all
-\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. {\bf KMK We defer a discussion of
-numerical convergence issues to Section \ref{sec-disc}.?}
-%mention numerical effects? 
+\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. There are, in
+addition, numerical convergence issues when simulating disk  
+fragmentation, which we discuss in \S\ref{summary}.
+
 However, it is generally accepted that \emph{steady, gravito-turbulent
@@
 This is intriguing because, as highlighted in
-\S\ref{cool_gi}---\ref{visc_gi}, these effects can reduce
-gravitational stability independently of their influence on the exact value of the classic $Q$. This motivates a physical    
+\S\ref{cool_gi}---\ref{visc_gi}, cooling or viscosity can reduce 
+gravitational stability independently of their influence on the exact
+value of the classic $Q$. This motivates a physical      
 interpretation of disk fragmentation as the inability to maintain a
-gravito-turbulent state due to %the action of 
-%KMK
+gravito-turbulent state due to 
 secondary instabilities driven by cooling and/or viscosity.
@@
 literature \citep[e.g. $\beta_c\sim
-  3,\,\alpha_c\sim0.1$;][]{gammie01,rice05}.
+  3,\,\alpha_c\sim0.1$,][]{gammie01,rice05}.
 
-\subsection{Plan}
+\subsection{Plan}\label{plan}
 %plan of this paper 

### 2016-01-28 paper/results2d.tex
revision 3 by MKL

@@
 %second condition concerning Q is associated with self-g 
-Notice these conditions do not actually
-depend on the cooling time. 
-The disk becomes more unstable with decreasing $\theta$, 
-i.e. as the floor temperature, towards which perturbations are cooled, drops. 
-{\bf KMK maybe more words here about apparent trend of $Q \rightarrow \inf$ for small $\theta$
-even forward ref. lengthscales below?}
+A third criterion, $2(2-q)\gamma Q^2>1$, is formally required, but this
+is implied by Eq. \ref{stable_condition}. 
+Notice these conditions do not actually depend on the cooling time.
+
+%As perturbations are cooled to lower temperatures
+%(decreasing $\theta$), the minimum $Q$ required for stability
+%increases.
+
+If we define $Q_\mathrm{eff}=\sqrt{\theta}Q$, the second condition in
+Eq. \ref{stable_condition} is $Q_\mathrm{eff}>1$ (for
+the Keplerian case). In other words, if one  
+insists on applying a Toomre condition to a cooling disk, then the 
+Toomre parameter should be calculated with the floor
+temperature. %rather than the actual temperature 
+At fixed $Q$ this stability condition will fail for sufficiently small 
+$\theta$, i.e. as perturbations are allowed to cool to lower
+temperatures.
+
+%{\bf KMK maybe more words here about apparent trend of $Q \rightarrow \inf$ for small $\theta$
+%even forward ref. lengthscales below?}
 
@@
 $(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
-upper limit to unstable wavenumbers, which could have important
-implications for numerical simulations probing large wavenumbers and small scales
+upper limit to unstable wavenumbers, which could have implications for
+numerical simulations probing large wavenumbers and small scales 
 at high resolution (see \S\ref{prev_works}).
@@
 Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
-S_*<0 $ for $\gamma>\theta$. That is, as cooling becomes more rapid
-the disk becomes more unstable at shorter wavelengths.  {\bf KMK -- I don't follow}
+S_*<0 $ for $\gamma>\theta$. As cooling becomes more rapid the maximum 
+growth rate increases, and the most unstable wavelength decreases. 
 %show dK*/dbeta < 0 
@@
 %and elim dS*/d beta
+
+If $|S_*|\ll1$, Eq. \ref{inviscid_max} gives the simple solution 
+\begin{align}
+ \beta S_*\simeq \frac{1-2(2-q)Q^2\theta}{2(2-q)Q^2\gamma-1},
+\end{align}
+implying that the system can still be formally unstable for arbitarily 
+long cooling times. {\bf optional equation?}
 

### 2016-01-28 paper/results2d.tex
MKL version 3-ing

@@
 Notice these conditions do not actually depend on the cooling time. 
-
-%As perturbations are cooled to lower temperatures
-%(decreasing $\theta$), the minimum $Q$ required for stability
-%increases.
-
-If we define $Q_\mathrm{eff}=\sqrt{\theta}Q$, the second condition in
-Eq. \ref{stable_condition} is $Q_\mathrm{eff}>1$ (for
-the Keplerian case). In other words, if one  
-insists on applying a Toomre condition to a cooling disk, then the 
-Toomre parameter should be calculated with the floor
-temperature. %rather than the actual temperature 
-At fixed $Q$ this stability condition will fail for sufficiently small 
-$\theta$, i.e. as perturbations are allowed to cool to lower
-temperatures.
+At fixed $Q$, the second stability condition eventually fails for
+decreasing $\theta$, i.e. if perturbations are allowed
+to cool to sufficiently low temperatures.
 
@@
 Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
-S_*<0 $ for $\gamma>\theta$. As cooling becomes more rapid the maximum 
+S_*<0 $ for $\gamma>\theta$. Then as cooling becomes more rapid, the maximum 
 growth rate increases, and the most unstable wavelength decreases. 
@@
 %and elim dS*/d beta 
+For $|S_*|\ll1$, Eq. \ref{inviscid_max} gives the simple solution 
+\begin{align}
+  \beta S_*\simeq \frac{1-2(2-q)Q^2\theta}{2(2-q)Q^2\gamma-1}.  
+\end{align} 
+Assuming the denominator is positive definite (see below for the case
+when it vanishes),  then weakly-growing modes have
+$S_*\propto\beta^{-1}$, so $\beta\gg1 $ but finite. That is, the disk
+can be formally unstable for arbitrarily long cooling times.
 
-If $|S_*|\ll1$, Eq. \ref{inviscid_max} gives the simple solution 
-\begin{align}
- \beta S_*\simeq \frac{1-2(2-q)Q^2\theta}{2(2-q)Q^2\gamma-1},
-\end{align}
-implying that the system can still be formally unstable for arbitarily 
-long cooling times. {\bf optional equation?}
+%{\bf optional equation?}
 
@@
 \subsubsection{$\theta = 0$}\label{theta0}
-For beta cooling with $\theta=0$, the disk is unconditionally unstable
-for finite $Q$. {\bf KMK some caveating about scales}  The $\beta\to0$ limit corresponds to a 
+For $\beta$-cooling with $\theta=0$, the disk is unconditionally
+unstable for finite $Q$, although instability occur on smaller scales
+as $Q$ increases.  The $\beta\to0$ limit corresponds to a 
 pressureless disk (not merely isothermal).
@@
 \end{align} 
-{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
-so that it is marginally stable in the absence of cooling.  
+%{\bf KMK $\gamma = \theta$ from eq. 47? I'm missing something}
+which is the condition for marginal stability in an adiabatic disk. 
 How does finite cooling destabilize the disk?  
-Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} and setting
+Inserting Eq. \ref{qcrit_def} into Eq. \ref{inviscid_max} with 
 $\theta=0$, we find 
@@
 The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
-increases with decreasing $\beta$. Instability formally occurs for
-arbitrarily long but finite cooling times. 
+increases with decreasing $\beta$. 
+%Instability formally occurs for
+%arbitrarily long but finite cooling times. 
 %Instability does not require the
@@
 
-Nevertheless, we can define a characteristic cooling
+We can define a characteristic cooling
 time $\beta_*$ as that which removes pressure support against
@@
 \end{align}
-This equation reproduces remarkably similar values 
+This equation gives remarkably similar values 
 of the cooling times below which 
-%dynamical disk 
-%fragmentation occurs in numerical simulations
 numerical simulations show dynamical disk fragmentation 
@@
 Standard beta cooling with $\theta=1$ corresponds to 
-`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
-value over the cooling time \citep{lin15,mohandas15}. In this case 
-$\beta\to 0$ corresponds to an isothermal disk and $\beta
-\to \infty$ corresponds to an adiabatic disk. (These two limits are
-equivalent for $\gamma=1$.) The most unstable wavenumber and growth
-rates vary weakly with $\beta$.   {\bf KMK: summarize boring result? not sure 
-what this simplifies to in comparison to other limits?}.
+`thermal relaxation': the temperature is restored to its initial
+value over the cooling time \citep{lin15,mohandas15}. In this case no
+additional heat source need be invoked to define a steady state. From
+Eq. \ref{stable_condition}, the instability condition is $Q<1$. This
+is the same as the classic Toomre condition for an isothermal disk,  
+which is formally obtained in the limit  $\beta\to 0$
+(i.e. the disk temperature is constant in time). In this respect, a
+fully irradiated disk, in that the equilibrium temperature is set
+externally, behaves isothermally regardless of the cooling time.
+
+%The most unstable wavenumber and growth
+%rates vary weakly with $\beta$.  
+% {\bf KMK: summarize boring result? not sure 
+%what this simplifies to in comparison to other limits?}.
 

### 2016-01-28 paper/results2d.tex
MKL revision 3-ing

@@
 \end{align} 
-Assuming the denominator is positive definite (see below for the case
-when it vanishes),  then weakly-growing modes have
-$S_*\propto\beta^{-1}$, so $\beta\gg1 $ but finite. That is, the disk
-can be formally unstable for arbitrarily long cooling times.
-
-%{\bf optional equation?}
+%Assuming the denominator is positive definite (see below for the case
+%when it vanishes),  
+Thus $S_*\to0$ as $\beta\to\infty$, but growth rates are never zero. 
+That is, the disk can be formally unstable for arbitrarily long
+cooling times.
 
@@
 value over the cooling time \citep{lin15,mohandas15}. In this case no
-additional heat source need be invoked to define a steady state. From
-Eq. \ref{stable_condition}, the instability condition is $Q<1$. This
-is the same as the classic Toomre condition for an isothermal disk,  
-which is formally obtained in the limit  $\beta\to 0$
-(i.e. the disk temperature is constant in time). In this respect, a
+additional heat source need be invoked to define an inviscid steady state. From
+Eq. \ref{stable_condition}, the instability condition is $Q<1$. %no
+                                %beta dependence
+This is the same as the classic Toomre condition for an isothermal
+disk (which may be obtained from Eq. \ref{inviscid}  by taking
+$|\beta S|\to 0$). In this respect, a 
 fully irradiated disk, in that the equilibrium temperature is set
@@
 
-%KMK moved fig ref. 
-Fig. \ref{gammie_rate_plot} shows growth rates as a function of the
-wavenumber $k$ obtained from the dispersion relation
-Eq. \ref{thindisk}.  To compute growth rates numerically, 
- we consider a model with $\alpha_b=0$ and
-$\alpha=\alpha(\beta)$ given by thermal equilibrium (
-Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
-\begin{align}
-  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
-\end{align}
-%with $Q_\mathrm{crit}$ given by 
-to mimic a basic state maintained by gravito-turbulence where one
-might expect the dimensionless stress $\alpha \sim Q^{-2}$
-\citep{lin87}.
-
 It is useful to consider several limiting cases.
@@
 
+%KMK moved fig ref. 
+To compute growth rates numerically, 
+we consider a model with $\alpha_b=0$ and
+$\alpha=\alpha(\beta)$ given by thermal equilibrium (
+Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
+\begin{align}
+  Q = \frac{Q_\mathrm{crit}}{\sqrt{\alpha}},\label{Qalpha}
+\end{align}
+%with $Q_\mathrm{crit}$ given by 
+to mimic a basic state maintained by gravito-turbulence where one
+might expect the dimensionless stress $\alpha \sim Q^{-2}$
+\citep{lin87}.
+
 %%KMK MOVED?
@@
 %Eq. \ref{thindisk}.
-The limiting behavior for small/large $K$ are
+
+Fig. \ref{gammie_rate_plot} shows growth rates as a function of the
+wavenumber $k$ obtained from the dispersion relation 
+Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
 well-captured by Eq. \ref{gammie_smallk} and
-\ref{gammie_bigk}. Comparing the two panels of Fig. \ref{gammie_rate_plots} shows that increasing the
+\ref{gammie_bigk}. Comparing the two panels 
+%of Fig. \ref{gammie_rate_plot} 
+shows that increasing the
 irradiation level ($\theta$) supresses small-scale
@@
 limiting behavior for this case as   
-\begin{align*}
+\begin{align}
   S_*\propto \begin{cases}
@@
 %    \frac{1}{Q\sqrt{\alpha}}  
-    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha,    
+    \alpha^{-1/2}Q^{-1} = \mathrm{const.} & \beta \ll \alpha, \label{high_visc}   
   \end{cases}
-\end{align*}
-{\bf KMK physically what is the limit $\beta \ll \alpha$?}
+\end{align}
+%{\bf KMK physically what is the limit $\beta \ll \alpha$?}
 where we have applied Eq. \ref{alpha_beta_relation}
-and \ref{Qalpha}. The disk is formally unstable for all $\beta$, but
+and \ref{Qalpha}. %The second limit corresponds to a hypothetical
+%situation where rapid cooling is balanced by large viscous heating
+%(we discuss why this). 
+The disk is unstable for all $\beta$, but
 growth timescales are long ($>10$ orbits) for $\beta \gtrsim
@@
 the cooling time for $\beta\lesssim O(1)$ because larger scales are 
-more resistant to the associated increase in viscous damping.   {\bf KMK this is
+more resistant to the associated increase in viscous damping.     
+{\bf KMK this is
 confusing b/c viscosity is in general acting on larger scales than pressure. also damping -> dissipation or 
@@
 maximum $\alpha$ ($\sim 0.06$) that can be sustained before
-fragmentation \citep{rice11} {\bf KMK rice 05a,b, no?}. We can interpret this result in our
+fragmentation \citep{rice05}. We can interpret this result in our
 linear framework. 
-%We can use our case (\ref{gvisc}) to interpret this
 Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
 by generating a large gravito-turbulent heating rate
-($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
-shows that such a system would be dynamically unstable with 
+($\alpha\gtrsim 1$) through a small $Q$ (second case in
+Eq. \ref{high_visc}). Fig. \ref{gammie_maxrate_plot}  
+shows that such a disk would be dynamically unstable with 
 growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
@@
 %KMK
-We interpret fragmentation as the response to the instability of the
+We might interpret fragmentation as the response to the instability of the
 highly viscous state. This is consistent with previous numerical

### 2016-01-29 paper/application.tex
rough version 3  by MKL

@@
 protoplanetary disks. We consider the gravito-turbulent disk models recently
-developed by \cite[][hereafter \citetalias{rafikov15}]{rafikov15}. 
+developed by \citet[][hereafter \citetalias{rafikov15}]{rafikov15}. 
 This 2D, Keplerian disk orbits a Solar 
@@
 
-Although our disk is global in extent, here we consider the local stability at each radius.
+Although our disk is global in extent, here we consider the local
+stability at each radius. {\bf discuss: linear prob has no vrad in
+  base state} 
 %KMK I disagree with neglect of viscous flow in that there is a global Mdot. Revised version:
@@
 \begin{align*}
-  \theta = \frac{4-g}{4-2g},
+  \theta = \frac{4-g}{4-2g}. 
 \end{align*}
-implying $5/6<\theta<3/2$. . %%KMK 
+Thus for $T_\mathrm{irr} = 0$ we have $5/6<\theta<3/2$ by considering
+$\tau\to 0,\,\infty$. However, for $T=\tirr$ we have $\theta = 1$, as
+we expect intuitively.    
+%%KMK 
 %This results from the fact that PPD cooling generally depends on both
@@
 where $T\to\tirr$ and $\theta\to 1$ \citep{dalessio97,kratter11}.
-Then the first inequality in Eq. \ref{ppd_invisc_cond} relaxes to
-$\gamma > 1$ in the outer disk, which is always satisfied.  
-On the other hand, numerical simulations of gravito-turbulence
-show that $1\lesssim Q \lesssim 2$ \citep{gammie01,rice11}, and 
-the second inequality is generally satisfied. 
+Then Eq. \ref{ppd_invisc_cond} relaxes to 
+$\gamma,\, Q > 1$ in the outer disk. The condition on $\gamma$
+is then guarenteed. On the other hand, numerical 
+simulations of gravito-turbulence show that $1\lesssim Q \lesssim 2$
+\citep{gammie01,rice11}, and the second inequality is generally
+satisfied.   
 %KMK
-Taken together, this means that in the outer regions of a realistic PPD, cooling itself
-is not responsible for a secondary instability of a gravito-turbulent disk,
-which might lead to fragmentation. This leaves viscous GI as the only possible
-culprit, as we illustrate below.
+Taken together, this suggests that in the outer regions of a realistic
+PPD, cooling %by itself may not be not responsible 
+may not be the primary cause for a secondary 
+instability of a gravito-turbulent disk, which might have lead to
+fragmentation. This leaves viscous GI as the only possible culprit
+within our framework, as we illustrate below.
 
 \subsection{Example 2D calculation}\label{pp2d_example}
-We relax the inviscid assumption and consider a disk  model with $\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$, 
+We relax the inviscid assumption and consider a fiducial disk  model with
+$\dot{M} = 10^{-6}M_\sun\,\yr^{-1}$, $Q_0=1.5$,  
 $\tirr=10\mathrm{K}$, and $\alpha_m=10^{-3}$. 
@@
 young protostellar disks \citep{shu77,enoch09}. Similarly,
-$10\mathrm{K}$ is a conservatively low background irradiation l
-evel consistent with cloud temperatures in star forming regions \citep{plume97,johnstone01}.
-Stellar irradiation will typically elevate $\tirr$ in addition to adding a radial dependence \citep{kratter08}
+$10\mathrm{K}$ is a conservatively low background irradiation level
+consistent with cloud temperatures in star forming regions
+\citep{plume97,johnstone01}. 
+Stellar irradiation will typically elevate $\tirr$ in addition to
+adding a radial dependence \citep{kratter08} 
 We adopt the opacity scale $\kappa_{d0} =
@@
 These profiles serve as input to the 2D dispersion relation
-(Eq. \ref{thindisk}).
+(Eq. \ref{thindisk}, Eq. \ref{bigA}---\ref{bigF}).
 
 Fig. \ref{rafikov_growth} shows growth timescales and
-optimum wavenumbers for viscous GI in this disk model.  
-We also plot analytic estimates based on Eq. \ref{gammie_smallk} (instead of
-Eq. \ref{gammie_maxrate_simple} since here $\theta\sim 1$) 
-which gives the optimum wavenumber and growth rates as
+optimum wavenumbers for viscous GI in this fiducial model. For
+comparison we also plot a case with lower accretion rate, 
+$\dot{M}=10^{-7}M_\sun\,\yr^{-1}$; and analytic estimates based on
+Eq. \ref{gammie_smallk} (instead of Eq. \ref{gammie_maxrate_simple}
+since here $\theta\sim 1$) which gives the optimum wavenumber and
+growth rates as 
 \begin{align}
@@
 %KMK
-So long as $H<<R$, this result is consistent with our use of the local approximation. For
-our example disk $H/R \approx ??$ at $R=?$. The
-increase in $|K|$ from $\sim 10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is
-due to the decrease in $Q$, while that from $\sim 60\mathrm{AU}$ to
-$\sim 100\mathrm{AU}$ occurs as the disk transitions from the
+So long as $H\ll R$, this result is consistent with our use of the
+local approximation. For our fiducial disk, $H/R \sim 0.07$ around 
+$R\sim 100$AU, and $H/R <0.25$ througout the disk. 
+The increase in $|K|$ from 
+$\sim 10\mathrm{AU}$ to $\sim 20\mathrm{AU}$ is due to the decrease in
+$Q$, while that from $\sim 
+60\mathrm{AU}$ to $\sim 100\mathrm{AU}$ occurs as the disk transitions
+from the 
 optically-thick to optically-thin regime. The mismatch between the
@@
 
-The disk is subject to viscous GI on dynamical timescales ($\lesssim
-10$ orbits) for $R\gtrsim60$AU. Coincidentally, beyond this radius
-$\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is often quoted as 
-criteria for disk fragmentation \citep[e.g.][]{rafikov15}. 
-Thus viscous GI may be responsible for the transition between gravito-turbulence
-and fragmentation due to the removal of rotational support by viscous (turbulent) stresses.
-%KMK
-%This suggests that fragmentation in the outer parts of a realistic PPD
-%occurs is physically due to large turbulent stresses.
+The fiducial disk is subject to viscous GI on dynamical timescales
+($\lesssim 10$ orbits) for $R\gtrsim60$AU. We note this transition
+radius is also implied by Eq. 16 in \cite{kratter10}. Coincidentally,
+beyond this radius $\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is
+often quoted as criteria for disk fragmentation  
+\citepalias[e.g.][]{rafikov15}. Thus viscous GI may be responsible for 
+the transition between gravito-turbulence and fragmentation due to the
+removal of rotational support by viscous (turbulent) stresses.
+
+On the other hand, the lower $\dot{M}$ model also attain $\tcool\lesssim
+3\Omega^{-1}$ beyond $\sim 60$AU, but $\alpha\lesssim 0.03$. Applying
+empirical conditions for fragmentation may then lead to
+contradiction. Instead, if viscosity is the physical cause for
+fragmentation, then our result suggest the lower $\dot{M}$ disk should
+not fragment (at least much less likely than our fiducial case) 
+because the instability cannot develop on orbital timescales.
 
@@
     0.8cm]{figures/ppd_2d_maxk}
-  \caption{Growth timescales (top) of the most unstable
-    wavenumber (bottom) for viscous,
+  \caption{Black lines show growth timescales (top) of the most
+    unstable wavenumber (bottom) for viscous,
     self-gravitational modes in the 2D PPD model shown in
-    Fig. \ref{rafikov_model}. 
-    The black curve is obtained numerically
-    from the 2D dispersion Eq. \ref{thindisk}, while green curves 
+    Fig. \ref{rafikov_model}. The blue line are for the same disk
+    model but with lower accretion rate. Dotted lines are obtained
+    numerically from Eq. \ref{thindisk}, while green curves  
     are analytic results based on Eq. \ref{gammie_smallk}. 
@@
 We then solve the 3D eigenvalue problem as in \S\ref{3ddisk}, with the
-additional boundary condition {\bf KMK: that the disk surface temperature is fixed},
- $\delta T^\prime(0) = \delta T(\zmax) = 0$.    
-We use the same disk model as in \S\ref{pp2d_example} but 
-with $T_\mathrm{irr}=0$, since our simple radiative diffusion model
-does not include irradiation (\S\ref{rad_cool}). 
+additional boundary condition that the disk surface temperature is
+fixed, $\delta T(\zmax) = 0$. 
+We use the fiducial disk model as in \S\ref{pp2d_example} but 
+with $T_\mathrm{irr}=0$, since our simple radiative diffusion
+treatment does not include irradiation (\S\ref{rad_cool}). %This also
+                                %limits us 
+%to smaller radii than before.  
 %physically, irrad is negligible for opt thick regions, and we are
@@
 %function approximately captures opticall-thin cooling in the 2D case.  
-Nevertheless, the transition radius
-$R_t\sim60$AU can be correctly calculated within the 2D framework.  
+Nevertheless, the transition radius of  
+$\sim60$AU, beyond which growth timescales become dynamical,    
+can be correctly calculated within the 2D framework.  
 %notice the small softenings used 
@@
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
     0.0cm]{figures/ppd_3d_rates}\\

### 2016-01-29 paper/results2d.tex
rough version 3  by MKL

@@
 we consider a model with $\alpha_b=0$ and
-$\alpha=\alpha(\beta)$ given by thermal equilibrium (
-Eq. \ref{alpha_beta_relation}). Furthermore, we relate 
+$\alpha=\alpha(\beta)$ given by thermal equilibrium
+(Eq. \ref{alpha_beta_relation}). Furthermore, we relate   
 \begin{align}
@@
 Fig. \ref{gammie_rate_plot} shows growth rates as a function of the
-wavenumber $k$ obtained from the dispersion relation 
+wavenumber obtained from the dispersion relation 
 Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
@@
 the cooling time for $\beta\lesssim O(1)$ because larger scales are 
-more resistant to the associated increase in viscous damping.     
-{\bf KMK this is
-confusing b/c viscosity is in general acting on larger scales than pressure. also damping -> dissipation or 
-angular momentum transport?}
+more resistant to the associated increase in viscous damping. This is
+evident from the dispersion relation in the large wavenumber limit,
+Eq. \ref{gammie_bigk}, showing increasing viscosity weakens
+small-scale modes. 
+%{\bf KMK this is
+%confusing b/c viscosity is in general acting on larger scales than pressure. al%so damping -> dissipation or 
+%angular momentum transport?}
 

### 2016-01-29 paper/results3d.tex
rough version 3  by MKL

@@
 possible to incorporate this effect in the previous 2D framework, but doing so
-introduces additional parameters as discussed in Appendix
-\ref{3dcorr}. It is more direct to solve the 3D eigenvalue problem to
-avoid such uncertainties. Our numerical approach is outlined in Appendix
-\ref{3d_method}.
+introduces an additional `softening' parameter $H_\mathrm{sg}$ as discussed in
+Appendix \ref{3dcorr}. It is more direct to solve the 3D eigenvalue
+problem to avoid such uncertainties. Our numerical approach is
+outlined in Appendix \ref{3d_method}.
 
@@
 ($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
-constant (Eq. \ref{alpha_beta_relation}). We consider {\bf only} even modes about
-the midplane, and apply a numerical disk surface at $z=\zmax$ such
-that $\rho(\zmax)=0.05\rho_0$.
+constant (Eq. \ref{alpha_beta_relation}). We consider only even modes
+about the midplane, and apply a numerical disk surface at $z=\zmax$
+such that $\rho(\zmax)=0.05\rho_0$.
 
@@
 \ref{3dcorr}. The (empirically) chosen value of 
-$H_\mathrm{sg}=0.64H$, also defined in Appendix \ref{3dcorr}, results in a close match between 2D and 3D
+$H_\mathrm{sg}=0.64H$ results in a close match between 2D and 3D
 growth rates, but the most unstable wavenumber in 3D is somewhat smaller. 
@@
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.23cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0.cm
     0.0cm]{figures/3d_inviscid_rates}\\
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.23cm
-    0.88cm]{figures/3d_inviscid_kmax}
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
+    0.8cm]{figures/3d_inviscid_kmax}
   \caption{Growth rates (top) and optimal wavenumber (bottom) obtained
-    from the 3D eigenvalue problem (solid line). Asterisks are
-    corresponding values from the 2D dispersion relation
-    (Eq. \ref{thindisk}) but with a modified gravitational constant
-    as described in Appendix \ref{3dcorr}. \label{3d_inviscid}}
+    from the inviscid 3D eigenvalue problem (solid line). Asterisks
+    and diamonds are corresponding values from the 2D dispersion
+    relation (Eq. \ref{thindisk}) but with a softened gravity as
+    described in Appendix \ref{3dcorr}. \label{3d_inviscid}} 
 \end{figure}
@@
 (\S\ref{2dvisc}), but with 
-\begin{align*}
-  \qvert = \frac{Q_\mathrm{3D,crit}}{\sqrt{\alpha}}, 
-\end{align*}
+\begin{align}
+  \qvert = \frac{Q_\mathrm{3D,crit}}{\sqrt{\alpha}}, \label{Q3d_visc_model}
+\end{align}
 where $Q_\mathrm{3D,crit}\simeq0.36$ is the 3D equivalent to the 2D
-critical value, $Q_\mathrm{crit}$.
+critical value, $Q_\mathrm{crit}$. Note that the background vertical
+structure now varies with $\alpha$ through Eq. \ref{Q3d_visc_model},
+which in turn depends on the cooling time through thermal
+equilibrium (Eq. \ref{alpha_beta_relation}).
 
@@
 corrections. Softening the self-gravity in 2D captures the correct 
-qualitative behavior as in the full 3D case, but once again it is clear that a
-single, constant value of $H_\mathrm{sg}$ cannot re-produce 3D growth 
-rates for all $\beta$. 
-This suggests that the exact value of
-$H_\mathrm{sg}$ is problem-dependent, although taking
+qualitative behavior as in the full 3D case. For $\beta\gtrsim 1$
+choosing $H_\mathrm{sg}=0.8H$ produces a good match. However, it is
+clear that a single, constant value of $H_\mathrm{sg}$ cannot
+re-produce 3D growth rates for all $\beta$. This suggests that the
+exact value of $H_\mathrm{sg}$ is problem-dependent, although taking 
 $H_\mathrm{sg}\sim O(H)$ should give the correct 3D growth rate within 
 a factor of two. 
-
 %cooling helps vertical modes (convection) 
@@
 $|\delta v_z|$ scaled by the total horizontal velocity for $\beta =
-1,\,10$ and $100$. Cooling generally enhances vertical motion
-(convection) {\bf KMK really convection?}. However, even in the highly unstable case with $\beta=1$,
-vertical speeds are sub-dominant at $\sim 30\%$ of the horizontal
-{\bf KMK  azimuthal or radial?} speeds. This suggests that it is not essential to consider 3D models
-when considering gravitational instabilities. {\bf KMK this is not true generically, so
-this statement needs to be more specific or cut}.
+1,\,10$ and $100$. Vertical speeds are sub-dominant at $\lesssim 30\%$
+of the total horizontal speeds. 
+%This suggests that it is not essential 
+%to consider 3D models when considering gravitational instabilities. 
+%{\bf KMK this is not true generically, so
+%this statement needs to be more specific or cut}.
 
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0.23cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0.cm
     0.0cm]{figures/growth_visc3d}
@@
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0.23cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0.cm
     0.0cm]{figures/eigenvec_vz}

### 2016-01-29 paper/summary.tex
rough version 3  by MKL

@@
 %KMK
-. %which may be of any origin, though we specialize to
+%which may be of any origin, though we specialize to
             %gravito-turb 
@@
 and/or viscosity. 
+%frag in sim and frag in real disk may or may not have the same
+%physical origin
 
@@
 Previous work has focused on the impact of viscosity and 
-cooling on the equilibrium temperature of the disk, but merely used this 
-to measure the classic Toomre Q and thereby assess stability. We demonstrate
-through a more rigorous inclusion of these effects into the dispersion relation that
-they can drive secondary instabilities.
+cooling on the equilibrium temperature and surface density of an 
+accretion disk, but merely used this to  the classic Toomre $Q$,
+thereby assess stability. %Disk fragmentation is assessed similarly,
+%but is usually based on empirically-derived conditions (see below). 
+We demonstrate by explicitly including these effects into the
+dispersion relation that they can drive secondary instabilities.
 While viscosity and cooling can be related through thermal
@@
 The effect of cooling and irradiation on GI is quantified by the 
-dispersion relation 
-Eq. \ref{inviscid}. We find a sufficient condition for stability which  
+dispersion relation Eq. \ref{inviscid}. 
+We find sufficient conditions for stability which  
 depends on the irradiation level (Eq. \ref{stable_condition}) but is 
@@
 We show that with a physically motivated cooling model for PPDs, 
-cooling alone does not lead to gravitational instabilities. This is due
-to stabilization by an effective `irradiation' associated with the PPD
-cooling function, which is present even if 
-there is no physical irradiation. Physically, this  term
-represents the fact that disks do not cool towards zero temperature because
-radiative cooling is a strong (quartic) function of temperature  %%KMK?
+cooling alone does not lead to gravitational instabilities. This is
+due to stabilization by an effective `irradiation' associated with the
+PPD cooling function, which is present even if there is no physical
+irradiation. 
+This effective irradiation arises because, for the stability problem,
+%cf eqm disk 
+a non-zero floor temperature is generally required in the cooling
+prescription (Eq. \ref{beta_cool}) in order to map it to PPD cooling,
+regardless of the actual irradiation temperature.
 
+%it's associated with density-dependence
+
+%is because PPD cooling depends on both pressure and
+%temperature. Mapping this 
+%to the standard beta cooling function requires a non-zero floor
+%temperature, otherwise standard beta c
+
+%This can only be mapped to the
+%standard beta cooling 
+%Physically, this  term
+%represents the fact that disks do not cool towards zero temperature because
+%radiative cooling is a strong (quartic) function of temperature  %%KMK?
 %due to matching functional forms of
                                 %cooling
-Instead, viscous GI occurs on  
-dynamical timescales for $R\gtrsim 60$AU because 
-$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
+
+Instead, viscous GI occur on dynamical timescales in a PPD for
+$R\gtrsim 60$AU because $\alpha\gtrsim0.1$ there. 
+This corresponds to a Toomre $Q\simeq 1.5$ and 
 a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
@@
 %flucuations? don't know 
-Our results may shed light on some numerical simulations 
+Our results may help understand some numerical simulations 
 concerning disk fragmentation. Table 
@@
 simulations, fragmentation is physically due to the removal of
-pressure stabilization by cooling on a radial scale of the disk scale
-height.
+thermal stabilization by cooling on radial lengthscales of the disk 
+thickness.   
+%i.e. there is no effective heating mechanism on the scale of
+        %H 
+%{\bf This would imply over-densities on this scale 
+%do not experience heating during its growth. This is not unreasonable
+%because in reality, gravito-turbulent heating only occurs when a 
+%spiral shock passes through said over-density \citep{young16}. Thus
+%there there is a time interval during which only cooling acts, whence
+%our inviscid theory can be applied. --- too much?}
+
 %%KMK
-%Note that unlike our physical cooling function (Eq. \ref{real_cool}), the standard beta function in these
-%numerical simulations cools towards zero temperature, rather than a temperature floor or irradiation temperature 
+%Note that unlike our physical cooling function (Eq. \ref{real_cool}),
+%the standard beta function in these 
+%numerical simulations cools towards zero temperature, rather than a
+%temperature floor or irradiation temperature  
 %%KMK some comment like that here, or too redundant with comment below??
  %i.e. cool-driven g
-{\bf don't know if we want to dwell on this table}
+
+%{\bf don't know if we want to dwell on this table}
 %negligible effect from viscous heating
@@
 We suggest having a floor temperature in the standard beta
-cooling prescription is a necessary condition for  
+cooling prescription is a necessary for  
 numerical convergence. This limits the
@@
 % physical, cf numerical  
-to aid clump formation via the removal of rotational support. (see also \S\ref{MHD} below). 
+to aid clump formation via the removal of rotational support. (See also \S\ref{MHD} below.) 
 If clumps are only marginally resolved, simulations may not have sufficient dynamic range for turbulent eddies
@@
 Even at very high resolution one might worry that the artificial dissipation scale
-imposed by the grid / smoothing length is too similar to scale of fragmentation.
+imposed by the grid/smoothing length is too similar to scale of fragmentation.
 %it is true that most unstable wavelength is H is this is always
@@
 a particular origin for the turbulence that is represented by the imposed
-viscosity. For example, the 2D dispersion relation given by Eq. \ref{thindisk}
-with Eq. \ref{bigA}---\ref{bigF}, treats $\alpha$ as an independent
-input parameter.
+viscosity. For example, our 2D dispersion 
+relation, Eq. \ref{thindisk} with Eq. \ref{bigA}---\ref{bigF}, treats 
+$\alpha$ as an independent input parameter.
 
@@
 True disk fragmentation is a non-linear process characterized by
-clumps reaching densities that are orders of magnitude above the ambient value. They must also survive
+clumps reaching densities that are orders of magnitude above the
+ambient value. They must also survive 
 disruption by tidal shear and shocks \citep{shlosman87,young16}.  
@@
 Toomre parameter through a (random) increase in density 
-(Eq. \ref{toomreQ_criterion}). By contrast, the viscosity/cooling
-destabilizing mechanisms discussed in this paper lowers $Q$ by
-reducing pressure and Coriolis forces. 
-{\bf KMK: this last sentence sounds like it's just about changing classic Q}.
+(Eq. \ref{toomreQ_criterion}).
+% By contrast, the viscosity/cooling
+%destabilizing mechanisms discussed in this paper lowers $Q$ by
+%reducing pressure and Coriolis forces. 
+%{\bf KMK: this last sentence sounds like it's just about changing classic Q}.
 %we have a physical way of lowering pressure/rotation
@@
 %this visc law is used in prev studies (by much smarter people)
-How to quantitatively model gravito- or MHD turbulence 
+How to quantitatively model the effect of gravito- or MHD turbulence  
 as a viscosity, especially on dynamical timescales, should be 

### 2016-01-29 paper/application.tex
version 3 by MKL done

@@
 to be used in the 2D dispersion relation (Eq. \ref{thindisk}). 
-
 Eq. \ref{real_beta} still represents a physical cooling 
@@
 %KMK
-However, $\theta$ is no longer a good proxy for the irradiation level
+However, $\theta$ is no longer a direct proxy for the irradiation level
 as it was in the standard beta cooling model.
@@
 $\tau\to 0,\,\infty$. However, for $T=\tirr$ we have $\theta = 1$, as
-we expect intuitively.    
+expected intuitively.    
 %%KMK 
@@
 On the other hand, the lower $\dot{M}$ model also attain $\tcool\lesssim
-3\Omega^{-1}$ beyond $\sim 60$AU, but $\alpha\lesssim 0.03$. Applying
-empirical conditions for fragmentation may then lead to
+3\Omega^{-1}$ beyond $\sim 60$AU, but $\alpha\lesssim 0.03$
+everywhere. 
+Applying empirical conditions for fragmentation may then lead to
 contradiction. Instead, if viscosity is the physical cause for
@@
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
     0.0cm]{figures/ppd_2d_growth}\\
   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
-    0.8cm]{figures/ppd_2d_maxk}
+    0.cm]{figures/ppd_2d_maxk}
   \caption{Black lines show growth timescales (top) of the most
@@
     self-gravitational modes in the 2D PPD model shown in
-    Fig. \ref{rafikov_model}. The blue line are for the same disk
-    model but with lower accretion rate. Dotted lines are obtained
-    numerically from Eq. \ref{thindisk}, while green curves  
+    Fig. \ref{rafikov_model}. Blue curves are for the same disk
+    model but with a lower accretion rate. Solid curves are obtained
+    numerically from Eq. \ref{thindisk}, and dotted curves 
     are analytic results based on Eq. \ref{gammie_smallk}. 
@@
   \caption{Growth timescales (top) of the most unstable
-    wavenumber (bottom) for viscous, 
-    self-gravitating modes. The black curves are obtained
-    from the 3D eigenvalue problem with radiative diffusion,
-    applicable to optically-thick regions ($R\lesssim 60$AU); 
-    while the green curves are obtained from the 2D corresponding
-    model, which is applicable to both optically-thick and
-    optically-thin regimes.  
+    wavenumber (bottom) for viscous GI in a 3D PPD with radiative
+    diffusion  (black lines). Asterisks are corresponding results
+    obtained from the corresponding 2D problem with softened gravity.   
+    % The black curves are obtained
+  %  from the 3D eigenvalue problem with radiative diffusion,
+   % applicable to optically-thick regions ($R\lesssim 60$AU); 
+   % while the green curves are obtained from the 2D corresponding
+   % model, which is applicable to both optically-thick and
+   % optically-thin regimes.  
     \label{rafikov_growth3d}}

### 2016-01-29 paper/paper3d.tex
version 3 by MKL done

@@
   heating. However, numerical simulations show that gravito-turbulence
-  cannot be sustained beyond dynamical timescales when the cooling rate is too fast,
-  or when the corresponding turbulent viscosity is too large. The result is is disk fragmentation.         
+  cannot be sustained beyond dynamical timescales when the cooling
+  rate or corresponding turbulent viscosity is too large. The result is disk fragmentation.         
   We motivate and quantify an interpretation of disk 
@@
 \end{inparaenum}
-We analyze the gravitational stability of accretion disks 
-including these non-ideal effects explicitly.    
-We derive a general dispersion relation for linear waves in a 
-self-gravitating, viscous, non-adiabatic disk with internal heating,
-external irradiation, and cooling. We  
-consider parameterized cooling functions in two and three dimensions,
+%We analyze the gravitational stability of accretion disks 
+%including these non-ideal effects explicitly.    
+%We derive a general dispersion relation for linear waves in a 
+%self-gravitating, viscous, non-adiabatic disk with internal heating,
+%external irradiation, and cooling. 
+We analyze the stability of self-gravitating, viscous, non-adiabatic
+disks with internal heating, external irradiation, and cooling.  
+We consider parameterized cooling functions in two and three dimensions,
 as well as 
@@
 the disk is formally stable, although interesting limits appear 
-for unstable modes with lengthscales on the order of the disk scale height.   
+for unstable modes with lengthscales on the order of the disk thickness.   
 %when we examine the growth 

### 2016-01-29 paper/results2d.tex
version 3 by MKL done

@@
 %when it vanishes),  
-Thus $S_*\to0$ as $\beta\to\infty$, but growth rates are never zero. 
-That is, the disk can be formally unstable for arbitrarily long
-cooling times.
+Thus $S_*\to0$ as $\beta\to\infty$, but growth rates are never zero
+for any finite $\beta$. That is, the disk can be formally unstable for
+arbitrarily long cooling times.
 
@@
 To compute growth rates numerically, 
-we consider a model with $\alpha_b=0$ and
+we consider a model with $\gamma=1.4$, $\alpha_b=0$ and 
 $\alpha=\alpha(\beta)$ given by thermal equilibrium
@@
 \begin{figure}
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.4cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
     0.0cm]{figures/viscsg_modes}\\
-  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.4cm
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
     0.0cm]{figures/viscsg_modes_theta0d3}
@@
     Eq. \ref{gammie_smallk} and \ref{gammie_bigk}. Top: without
-    irradiation, bottom: partially irradiated disk. 
+    a floor temperature ($\tirr = 0$); bottom: with a floor
+    temperature $\tirr$ set to $30\%$ of the equilibrium temperature.   
     \label{gammie_rate_plot}}
@@
     viscous GI as a function of cooling time for the
-    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
-    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
-    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
-    line marks the region with $\alpha > 1$, and the vertical
-    dashed-dot line marks the region with growth timescales longer
-    than dynamical. 
+    case shown in the top panel of Fig. \ref{gammie_rate_plot}. Black curves 
+    computed from the dispersion relation (Eq. \ref{thindisk}), and
+    the orange curves are estimates based on
+    Eq. \ref{gammie_smallk}---\ref{gammie_bigk}. 
+    The dashed
+    line marks the region with $\alpha > 1$, and the 
+    dashed-dot line marks the region where growth timescales exceed
+    the dynamical time.   
     \label{gammie_maxrate_plot}}

### 2016-01-29 paper/results3d.tex
version 3 by MKL done

@@
 %in practice eliminate density with poisson
-\subsection{Inviscid disk}
+\subsection{Inviscid 3D disk}
 
-We consider a 3D inviscid disk with $\qthree=0.71$ and $\theta=0$. 
+We consider a 3D inviscid disk with $\gamma=1.4$, $\qthree=0.71$, and $\theta=0$. 
 The 3D gravity parameter $\qthree$ is defined by 
@@
 
-\subsection{Viscous disk} %growth rates, kmax, eigenfunction
+\subsection{Viscous 3D disk} %growth rates, kmax, eigenfunction
 For the 3D viscous problem we use the same set up as that in 2D
@@
   \caption{Growth rates from the viscous 3D eigenvalue problem (black solid
-    line). The other curves are obtained from the 2D dispersion
-    relation (Eq. \ref{thindisk}) but with a modified gravitational
-    constant as described in Appendix \ref{3dcorr}. \label{3d_visc}}
+    line). Asterisks and diamonds are obtained from the 2D dispersion
+    relation (Eq. \ref{thindisk}) with softened gravity as described
+    in Appendix \ref{3dcorr}. \label{3d_visc}} 
 \end{figure}

### 2016-01-29 paper/summary.tex
version 3 by MKL done

@@
 triggered by lowering the local 
-Toomre parameter through a (random) increase in density 
-(Eq. \ref{toomreQ_criterion}).
+Toomre parameter through a (random) increase in density. 
+%(Eq. \ref{toomreQ_criterion}).
 % By contrast, the viscosity/cooling

### 2016-01-30 paper/paper3d.tex
version 3 by MKL done
minor edit

@@
 %external irradiation, and cooling. 
-We analyze the stability of self-gravitating, viscous, non-adiabatic
-disks with internal heating, external irradiation, and cooling.  
+We analyze the gravitational stability of viscous, non-adiabatic
+accretion disks with internal heating, external irradiation, and cooling.  
 We consider parameterized cooling functions in two and three dimensions,
@@
 \end{align}
-\citep{toomre64}. Here, $c_s$ is the sound-speed,
+\citep{toomre64}. Here, $c_s$ is the isothermal sound-speed,
 $\kappa$ is the epicyclic frequency (which equals the rotation
@@
 axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
-adiabatic (including isothermal) disk. 
+isothermal\footnote{Eq. \ref{toomreQ_criterion}---\ref{classic_dr} also apply to adiabatic disks if one takes $c_s$ as the adiabatic sound-speed.} disk.  
 %Such disturbances are formally

### 2016-01-31 paper/summary.tex
minor edit
minor summary edits, new alpha(gamma) relation based on gammie's alpha and our betastar

@@
 thermal stabilization by cooling on radial lengthscales of the disk 
-thickness.   
+thickness. In this interpretation, gravito-turbulence only
+provides a background heating. %not active in frag 
+Our characteristic cooling time, given by 
+Eq. \ref{betastar}, corresponds to a dimensionless viscosity 
+as defined in the above studies\footnote{This differs from our
+  definition of $\alpha$ by a factor of $\gamma^{-1}$.}  
+\begin{align}
+  \alpha =
+  \frac{4}{9}\frac{1}{\gamma}\frac{\left(\sqrt{\gamma}-1\right)^{1/2}}{\left(\sqrt{\gamma}+1\right)}
+  \simeq \begin{cases}
+    0.062 & \gamma = 7/5, \\
+    0.063 & \gamma = 5/3,\\
+    0.059 & \gamma = 2.
+  \end{cases}
+\end{align}
+%where we have inserted a factor of $\gamma^{-1}$ to fit the
+%viscosity definition used by \cite{gammie01}.
+
 %i.e. there is no effective heating mechanism on the scale of
@@
 We suggest having a floor temperature in the standard beta
-cooling prescription is a necessary for  
+cooling prescription is  necessary for  
 numerical convergence. This limits the

### 2016-02-01 paper/results2d.tex
last minute fig add

@@
 equilibrium disk temperature equals the irradiation temperature.
+
+\begin{figure}
+  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
+    0.0cm]{figures/2dinvisc_theta}
+  \caption{
+    Growth rates for the 2D inviscid problem as a function of
+    the cooling time $\beta$ for two irradiation levels: $\theta=0.1$ (black) and
+    $\theta=0.33$ (orange). For each case the vertical dashed-dotted lines mark the cooling times beyond which
+    growth rates are longer than 10 orbits. 
+    \label{invisc_theta}}
+\end{figure}
 

### 2016-02-02 paper/application.tex
to be submitted version done

@@
 Although our disk is global in extent, here we consider the local
-stability at each radius. {\bf discuss: linear prob has no vrad in
-  base state} 
-%KMK I disagree with neglect of viscous flow in that there is a global Mdot. Revised version:
-We are thus neglecting any global instabilities \citep{adams89,lodato05,kratter10} as well as any 
+stability at each radius. We are thus neglecting any global
+instabilities \citep{adams89,lodato05,kratter10} as well as any
 evolution of the disk properties in response to GI.
+
 % We are thus neglecting the slow, background viscous radial flow; as well
@@
 
-Note that Eq. \ref{real_cool} falls within our definition of a beta cooling prescription, because
+Note that Eq. \ref{real_cool} falls within our definition of a beta
+cooling prescription, because 
 it is an explicit function of the thermodynamic states. However, we
 formulated the linear problem with the standard beta cooling function 
-given by Eq. \ref{beta_cool}, which has a different (less realistic) dependence on disk temperature.
+given by Eq. \ref{beta_cool}, which has a different (less realistic)
+dependence on disk temperature. 
  In order to adapt the existing  framework to the above PPD cooling function, we 
 need to identify the equivalent $\beta$ and $\theta$ parameters that are required for the 
-linearized equations (Eq. \ref{linear_beta}).
+linearized equations (see \S\ref{linear_bcool}).
 
@@
 to be used in the 2D dispersion relation (Eq. \ref{thindisk}). 
-Eq. \ref{real_beta} still represents a physical cooling 
-time, and is consistent with previous definitions within factors of
-order unity \citep[e.g.][their Eq. 2]{kratter10}.
+Eq. \ref{real_beta} represents a physical cooling \emph{for the
+  perturbations}, and is consistent with previous definitions within
+factors of order unity \citep[e.g.][their Eq. 2]{kratter10}.
 
 %KMK
-However, $\theta$ is no longer a direct proxy for the irradiation level
-as it was in the standard beta cooling model.
+%However, $\theta$ is no longer a direct proxy for the irradiation level
+%as it was in the standard beta cooling model.
 %is now an effective irradiation level. 
-Eq. \ref{real_theta} shows that $\theta$ is related to the true
-irradiation temperature $\tirr$ through the function $g$ given by 
+Eq. \ref{real_theta} shows that the re-defined $\theta$ is related to
+the true irradiation temperature $\tirr$ through the function $g$ given by 
 Eq. \ref{g_def}, and is therefore a only a weak function of the irradiation temperature. 
-A physically non-irradiated disk with $\tirr=0$ will
-have a non-zero value of the `irradiation parameter', $\theta\neq0$. More specifically $\theta = O(1)$ 
+%A physically non-irradiated disk with $\tirr=0$ will 
+%have a non-zero value of the `irradiation parameter', $\theta\neq0$. 
+More specifically $\theta = O(1)$ 
 for all $\tirr <T$, and for our adopted opacity law, Eq. \ref{opacity_law}, 
@@
 %cooling only depends on $P$. 
-{\bf KMK: we should discuss}
+%{\bf KMK: we should discuss}
 
@@
 %implying $5/6<\theta<3/2$. 
-With our new linearized cooling function in hand, from the discussion in \S\ref{2d_inviscid} and by applying
+With our new linearized cooling function in hand, from the discussion
+in \S\ref{2d_inviscid} and by applying 
 Eq. \ref{stable_condition},  we conclude that without viscous effects 
@@
 radius is also implied by Eq. 16 in \cite{kratter10}. Coincidentally,
-beyond this radius $\alpha\gtrsim 0.1$ and $\beta\lesssim 3$, which is
-often quoted as criteria for disk fragmentation  
+beyond this radius $\alpha\gtrsim 0.1$ (and 
+$\beta\lesssim 3$), which is often quoted as a condition for disk
+fragmentation   
 \citepalias[e.g.][]{rafikov15}. Thus viscous GI may be responsible for 

### 2016-02-02 paper/linear3d.tex
revision 3 by MKL
to be submitted version done

@@
 \end{align}
-where $^\prime$ denotes $d/dz$. The perturbed viscous forces are {\bf F is also radiative flux....}
+where $^\prime$ denotes $d/dz$. The perturbed viscous forces are %{\bf F is also radiative flux....}
 \begin{align}
@@
 
-\subsection{Linearized beta cooling}
+\subsection{Linearized beta cooling}\label{linear_bcool}
 For the standard beta cooling prescription, linearizing
@@
 the ideal gas law.
+
+Note that any linearized beta cooling function can be put into the
+form of Eq. \ref{linear_beta} with appropriate definitions of $t_c$
+and $\theta$ (see \S\ref{ppd_cooling} for an example).    
+For the stability problem we may simply regard $\theta$ as a parameter 
+for the density-dependence of any generic beta cooling function. If 
+we specifically consider standard beta cooling, then $\theta$ also
+represents physical irradiation.
 
@@
 to the beta cooling prescription, except for special problems
-\citep[e.g.][]{lin15}. %KMK
-We now consider the behavior of two and three dimensional disks in the
-non-ideal physics, linear GI framework {\bf or some other transition sentence}.
+\citep[e.g.][]{lin15}.
+
+We now consider the gravitational stability of two and three 
+dimensional disks in the presence of non-ideal
+physics: cooling and viscosity.  
+%{\bf or some other transition sentence}.

### 2016-02-02 paper/model3d.tex
to be submitted version done

@@
 $T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
-for example, stellar or background irradiation. We will thus refer to
-$\theta$ as the `irradiation level'. {\bf ASK} As we discuss in
-\S\ref{2d_inviscid}, the limits $\theta \rightarrow 0, \theta
-\rightarrow 1$ do not exactly correspond to the limits of
-non-irradiated or irradiation-dominated disks as discussed in the
-literature \citep{kratter11,rice11}. 
-%stellar illum or envelope rad
+for example, stellar or background irradiation. Note that $\theta$  
+should only be interpreted as an irradiation parameter when it is 
+defined through Eq. \ref{tirr_def} and 
+Eq. \ref{beta_cool} adopted as the cooling function.
 
@@
 
-Note that, although we assume standard beta cooling in formulating the
-linear problem later, any explicit 
-cooling function, say $\Lambda_1(\rho,T)$, can be mapped to
-that obtained from Eq. \ref{beta_cool} by equating their linearized
-forms, i.e. $\delta \Lambda \equiv \delta \Lambda_1$. This then 
-defines the $\beta$ and $\theta$ parameters to be used in the 
-framework we develop. We do this in \S\ref{2dppd} where we adopt a
+%We shall need to re-define $\theta$ and $\beta$  
+%when considering other cooling functions (see below).
+
+We assume standard beta cooling in formulating the linear
+problem. However, the corresponding linear problem for  
+any other explicit cooling function, say $\Lambda_1(\rho,T)$, can be  
+obtained by equating its linearized form to 
+that of Eq. \ref{beta_cool}, i.e. setting $\delta \Lambda \equiv
+\delta \Lambda_1$. This then defines the $\beta$ and $\theta$
+parameters to be used in the framework we develop later (see also 
+\S\ref{linear_bcool}). We do this in \S\ref{2dppd} where we adopt a  
 more realistic beta cooling function for PPDs. In that case, $\theta$ 
-may or may not represent a physical irradiation.
+may or may not directly represent a physical irradiation.
 

### 2016-02-02 paper/paper3d.tex
to be submitted version done

@@
 We apply this new linear theory to 
-global models of protoplanetary disks (PPDs) subject to
+global models of protoplanetary disks subject to
 gravito-turbulence modeled as an effective 
 viscosity, and cooling via dust opacity. 
-We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
+We find that viscosity renders the disk beyond $\sim 60$AU dynamically
 unstable on radial %this is for Mdot = 1d-6  
@@
 even when $Q>1$. For example, if perturbations can
-cool to zero temperature, we find the above dispersion relation is
+cool to arbitrarily low temperatures (which, in fact, is a common
+cooling prescription in numerical simulations), we find the above
+dispersion relation is  
 modified to read 
@@
 rotation neglected ($\Omega\to 0$). That is, a classically stable
-disk can be destabilized by viscosity as it effectively eliminates
+disk can be destabilized by viscosity as it reduces 
 rotational stabilization \citep{lynden-bell74}.
@@
 We develop a general framework for viscous disks without  
-assuming a specific origin of the turbulence that this viscosity 
-represents. 
+assuming a specific origin for the viscosity. 
 We will, however, apply the theory to `gravito-turbulent'

### 2016-02-02 paper/results2d.tex
to be submitted version done

@@
 
-Next, we consider two special cases: $\theta=0$, so perturbations are
-cooled towards zero temperature \citep[typically employed in
-numerical simulations, e.g.][]{gammie01}; and $\theta=1$, where the
-equilibrium disk temperature equals the irradiation temperature.
+Fig. \ref{invisc_theta} show growth rates in a $Q=1.7$ disk  
+as a function of the cooling time $\beta$ for two irradiation levels
+$\theta=0.1,\,0.33$. The vertical lines mark characterisic cooling
+times beyond which the growth timescale is long compared to the
+dynamical time. Increasing irradiation stablizes the
+disk, and faster cooling is required to achieve the same growth 
+rate as in a disk with weaker irradiation.
 
@@
     the cooling time $\beta$ for two irradiation levels: $\theta=0.1$ (black) and
-    $\theta=0.33$ (orange). For each case the vertical dashed-dotted lines mark the cooling times beyond which
+    $\theta=0.33$ (orange). For each case the vertical dashed-dotted
+    lines mark the cooling times beyond which 
     growth rates are longer than 10 orbits. 
@@
 \end{figure}
+
+Next, we consider the two limiting cases: $\theta=0$, so perturbations are
+cooled towards zero temperature \citep[typically employed in
+numerical simulations, e.g.][]{gammie01}; and $\theta=1$, where the
+equilibrium disk temperature equals the irradiation temperature.
 

### 2016-02-02 paper/summary.tex
minor summary edits, new alpha(gamma) relation based on gammie's alpha and our betastar
to be submitted version done

--- 2016-01-29 paper/summary.tex
@@
 thermal stabilization by cooling on radial lengthscales of the disk 
-thickness.   
+thickness. In this interpretation, gravito-turbulence only
+provides a background heating. %not active in frag 
+Our characteristic cooling time corresponds to a dimensionless background viscosity 
+as defined in the above studies\footnote{This differs from our
+  definition of $\alpha$ by a factor of $\gamma^{-1}$.}  
+\begin{align}
+  \alpha =
+  \frac{4}{9}\frac{\left(\sqrt{\gamma}-1\right)^{1/2}}{\gamma\left(\sqrt{\gamma}+1\right)}
+  \simeq \begin{cases}
+    0.062 & \gamma = 7/5, \\
+    0.063 & \gamma = 5/3,\\
+    0.059 & \gamma = 2.
+  \end{cases}
+\end{align}
+This $\alpha\sim0.06$ is roughly constant, consistent with \cite{rice05}.
+
+%where we have inserted a factor of $\gamma^{-1}$ to fit the
+%viscosity definition used by \cite{gammie01}.
+
 %i.e. there is no effective heating mechanism on the scale of
@@
 
-We suggest having a floor temperature in the standard beta
-cooling prescription is a necessary for  
+Note that there may be a `numerical floor temperature' in 
+simulations, e.g. due to numerical viscosity, which is expected to 
+decrease with resolution. Then we expect a disk can be
+destabilized at larger cooling times with increasing resolution
+because the effective floor temperature drops   
+\citep[cf.][see also Fig. \ref{invisc_theta}]{meru11}.
+
+We suggest having a physical floor temperature in the standard beta
+cooling prescription is  necessary for  
 numerical convergence. This limits the

### 2016-02-02 paper/application.tex
to be submitted version done

@@
 %is now an effective irradiation level. 
-Eq. \ref{real_theta} shows that the re-defined $\theta$ is related to
-the true irradiation temperature $\tirr$ through the function $g$ given by 
+Eq. \ref{real_theta} shows that $\theta$ 
+is related to the true irradiation temperature $\tirr$ through the function $g$ given by 
 Eq. \ref{g_def}, and is therefore a only a weak function of the irradiation temperature. 
@@
 beyond this radius $\alpha\gtrsim 0.1$ (and 
-$\beta\lesssim 3$), which is often quoted as a condition for disk
+$\tcool\Omega\lesssim 3$), which is often quoted as a condition for disk
 fragmentation   

### 2016-02-02 paper/summary.tex
to be submitted version done

@@
 due to stabilization by an effective `irradiation' associated with the 
-PPD cooling function, which is present even if there is no physical
+density-dependence of the PPD cooling function as it appears in the
+stability problem, which is present even if there is no physical
 irradiation.  
-This effective irradiation arises because, for the stability problem,
+%This effective irradiation arises because, for the stability problem,
 %cf eqm disk 
-a non-zero floor temperature is generally required in the cooling
-prescription (Eq. \ref{beta_cool}) in order to map it to PPD cooling,
-regardless of the actual irradiation temperature.
+%a non-zero floor temperature is generally required in the cooling
+%prescription (Eq. \ref{beta_cool}) in order to map it to PPD cooling,
+%regardless of the actual irradiation temperature.
 

### 2016-02-02 paper/appendix3d.tex
version 3 by MKL done
tiny edits

@@
 \end{align}
-Note that these equations treat $\alpha$ and $\beta$ as independent
-parameters. 
+Note that these equations treat all the parameters as independent
+(namely $\alpha$, $\alpha_b$, $\beta$, $\theta$, and $Q$).
+
 %If beta cooling is entirely balanced by viscous heating in
@@
 previous authors \citep{lynden-bell74,willerding92,gammie96} as 
-follows. We set $\mu=-1, \lambda=0$ to obtain the same viscosity
+follows. We set $\mu=-1, \lambda=0$ in Eq. \ref{visc_law} to obtain the same viscosity
 models. Next we consider $|\beta S|\to \infty$, i.e. no explicit

### 2016-02-02 paper/application.tex
tiny edits

@@
 to be used in the 2D dispersion relation (Eq. \ref{thindisk}). 
-Eq. \ref{real_beta} represents a physical cooling \emph{for the
+Eq. \ref{real_beta} represents a physical cooling time \emph{for the
   perturbations}, and is consistent with previous definitions within

### 2016-02-02 paper/summary.tex
tiny edits

@@
 cooling on the equilibrium temperature and surface density of an 
-accretion disk, but merely used this to  the classic Toomre $Q$,
+accretion disk, but merely used this to calculate  the classic Toomre $Q$,
 thereby assess stability. %Disk fragmentation is assessed similarly,
@@
 Note that there may be a `numerical floor temperature' in 
-simulations, e.g. due to numerical viscosity, which is expected to 
+simulations, e.g. due to numerical viscosity, which should  
 decrease with resolution. Then we expect a disk can be

### 2016-02-02 paper/appendix3d.tex
submitted version

@@
 models. Next we consider $|\beta S|\to \infty$, i.e. no explicit
-cooling on the perturbations. Then the condition $AD = BC$ imply 
+cooling on the perturbations. Then the condition $AD = BC$ implies 
 \begin{align}
@@
 \end{align}
-which agrees with \citet[Eq. 11]{willerding92} in the isothermal
-limit ($\gamma=1$). The non-isothermal term $\propto (\gamma-1)$
-originates from viscous dissipation which was excluded in the
+which agrees with the above studies in the isothermal
+limit \citep[$\gamma=1$; see also][their Eq. 28]{schmit95}.  
+The non-isothermal term $\propto (\gamma-1)$
+originates from viscous dissipation, which was excluded in the
 aforementioned works. Its effect is to increase the maximum wavenumber 
-for viscous GI. %it acts to stabilize
+for viscous GI.
 
@@
 differential equations, Eq. \ref{lin_mass}---\ref{lin_gravity}, on 
-the domain $ z\in[0,\zmax]$ with a parity condition at the midplane.
+the domain $ z\in[0,\zmax]$ with a parity condition at the mid-plane.
 We expand $\bm{U}=[\delta P,\delta\rho,\delta\Phi,
@@
 The basis functions are chosen to satisfy a reflecting boundary
-condition at the midplane, %GI has even symmetry across midplane 
+condition at the mid-plane, %GI has even symmetry across midplane 
 \begin{align}

### 2016-02-02 paper/application.tex
submitted version

@@
 We now apply our linear framework to assess the stability of  
-protoplanetary disks. We consider the gravito-turbulent disk models recently
+PPDs. We consider the gravito-turbulent disk models recently
 developed by \citet[][hereafter \citetalias{rafikov15}]{rafikov15}. 
@@
     gravito-turbulent;
-  \item $T_\mathrm{irr}$, the irradiation or floor temperature;
+  \item $T_\mathrm{irr}$, the irradiation temperature;
   \item $\alpha_m$, the dimensionless viscosity associated with other
@@
 $\gamma,\, Q > 1$ in the outer disk. The condition on $\gamma$
-is then guarenteed. On the other hand, numerical 
+is then guaranteed. On the other hand, numerical 
 simulations of gravito-turbulence show that $1\lesssim Q \lesssim 2$
@@
 may not be the primary cause for a secondary 
-instability of a gravito-turbulent disk, which might have lead to
+instability of a gravito-turbulent disk, leading to
 fragmentation. This leaves viscous GI as the only possible culprit
@@
 local approximation. For our fiducial disk, $H/R \sim 0.07$ around 
-$R\sim 100$AU, and $H/R <0.25$ througout the disk. 
+$R\sim 100$AU, and $H/R <0.25$ throughout the disk. 
 The increase in $|K|$ from 
@@
 everywhere. 
-Applying empirical conditions for fragmentation may then lead to
+Applying empirical cooling conditions for fragmentation may then lead to
 contradiction. Instead, if viscosity is the physical cause for
@@
 wavenumber for $R\in[10,100]$AU; along with the corresponding 2D 
-results matched with softened self-gravity. There is good agreeement for
+results matched with softened self-gravity. There is good agreement for
 $R\lesssim60$AU where the disk is optically thick ($\tau\gtrsim

### 2016-02-02 paper/model3d.tex
submitted version

@@
 then increasing the density corresponds to reduction in the viscosity. 
-This is perhaps not inconsistent with numerical simulations of disk
-fragmentation which show the internal flow of high-density clumps
-appear laminar \citep{gammie01}. However, once clumps form they may
+This is perhaps consistent with numerical simulations of disk
+fragmentation which show that the internal flow of high-density clumps
+is laminar \citep{gammie01}. However, once clumps form they may
 effectively decouple from the background disk state, and thus no 
@@
 
-\subsubsection{$\beta$ cooling}\label{beta_cool_model}
-In our $\beta$ cooling model, the energy loss per unit volume is specified  
+\subsubsection{Beta cooling}\label{beta_cool_model}
+In our beta cooling model, the energy loss per unit volume is specified  
 as an explicit function of the thermodynamic variables. 
@@
 
-For $\beta$-cooling given by Eq. \ref{beta_cool} we assume an 
+For beta cooling given by Eq. \ref{beta_cool} we assume an 
 equilibrium polytropic relation 
-\begin{align} 
+\begin{align}\label{poly_vert} 
   P  =
@@
 where $\Gamma$ is the constant polytropic index and $\rho_0 =
-\rho(z=0)$ is the equilibrium midplane density. 
+\rho(z=0)$ is the equilibrium mid-plane density. 
 The vertical structure is first obtained from 
@@
 While radiative cooling is arguably more realistic than
-$\beta$-cooling, it generally implies a vertically
+beta cooling, it generally implies a vertically
 non-isothermal equilibrium disk, and increases the order of the

### 2016-02-02 paper/paper3d.tex
submitted version

@@
 We apply this new linear theory to 
-global models of protoplanetary disks subject to
+models of protoplanetary disks subject to
 gravito-turbulence modeled as an effective 
-viscosity, and cooling via dust opacity. 
+viscosity, and cooling regulated by dust opacity. 
 We find that viscosity renders the disk beyond $\sim 60$AU dynamically
@@
 is central to many astrophysical problems. In the context of 
-gaseous protoplanetary disks (PPDs), gravitational instability (GI)
+gaseous protostellar or protoplanetary disks (PPDs), gravitational instability (GI)
 has two applications. It can provide gravitational torques to
@@
 \citep{armitage10,turner14}. GI may also lead to disk fragmentation,
-which has been invoked to explain giant planet formation at large
-radii \citep{boss97,helled14} %% KMK need an original citation, not just review.
+which has been invoked to explain the formation of stellar/sub-stellar companions or 
+giant planets 
+at large radii \citep{boss97,kratter06,stam09,helled14} %% KMK need an original citation, not just review.
 Studying these non-linear phenomena requires
@@
 measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
-the stabilzing effect of rotation ($\kappa$) and pressure 
+the stabilizing effect of rotation ($\kappa$) and pressure 
 ($c_s$). This is evident from the dispersion relation, 
@@
 are derived from \emph{idealized} conditions:  
-the base disk is laminar and does not experience any net thermal 
+the base disk is laminar, inviscid, and does not experience any net thermal 
 losses\footnote{Isothermal disks implicitly assume  
@@
 In fact, this
-is a mechanism to enhance particle clumping for planetesimational
+is a mechanism to enhance particle clumping for planetesimal
 formation %frag of dust 
@@
 to transport angular momentum outwards (and mass inwards),
-frustate rotational support.
+frustrate rotational support.
 
@@
 %KMK
-We emphasize that our model for GI enabled by viscosity (hearafter
+We emphasize that our model for GI enabled by viscosity (hereafter
 viscous GI) does not assume a particular origin for the viscosity. GI
@@
 
-Consider an initially laminar, $Q\gg1$ disk, without an external
-heating mechanism, as it cools. The disk temperature will decline
+Consider an initially laminar, $Q\gg1$ disk, without external 
+heating, as it cools. The disk temperature will decline
 until $Q=O(1)$, whence non-axisymmetric modes grow and heat the disk
@@
   disks do not  
-  exist for sufficiently rapid cooling or large viscosity}.  
+  exist for sufficiently rapid cooling or large viscosity} \citep{johnson03}.   
 This is intriguing because, as highlighted in

### 2016-02-02 paper/results2d.tex
submitted version

@@
 with previous studies. The disk
-material is assumed to be confined to the midplane, and $\delta v_z=0$. 
+material is assumed to be confined to the mid-plane, and $\delta v_z=0$. 
 We make the replacement  
@@
 %can absorb big gamma into def of cs
-The gravitational potential perturbation remains 3D and its midplane
+The gravitational potential perturbation remains 3D and its mid-plane
 value is given by    
@@
 We first simplify the problem by setting $\alpha = \alpha_b = 0$. This
-eliminates viscous heating and forces, allowing us to quantify the 
-\emph{sole} effect of cooling on the 
+eliminates viscous heating and forces in the linearized problem, 
+allowing us to quantify the \emph{sole} effect of cooling on the 
 perturbations. %KMK
@@
 invoked to balance the imposed cooling to allow an equilibrium to be
-defined ($\mathcal{H}_\mathrm{ext}\neq 0$). Alternatively, we are
-assuming viscosity only provides a background heating and does not
-play an active role in the perturbed state. %may or may not be true in a real
+defined ($\mathcal{H}_\mathrm{ext}\neq 0$). For example, we could assume
+that the viscosity only provides a background heating and does not
+play an active role in the perturbed state.
+
+ %may or may not be true in a real
                                 %disk because we are using viscosity
@@
 	%KMK interesting points. discuss later?
+
 Eq. \ref{thindisk} becomes 
@@
 
-Fig. \ref{invisc_theta} show growth rates in a $Q=1.7$ disk  
+Fig. \ref{invisc_theta} shows growth rates in a $Q=1.7$ disk  
 as a function of the cooling time $\beta$ for two irradiation levels
-$\theta=0.1,\,0.33$. The vertical lines mark characterisic cooling
+$\theta=0.1,\,0.33$. The vertical lines mark characteristic cooling
 times beyond which the growth timescale is long compared to the
-dynamical time. Increasing irradiation stablizes the
+dynamical time. Increasing irradiation stabilizes the
 disk, and faster cooling is required to achieve the same growth 
@@
 For $\beta$-cooling with $\theta=0$, the disk is unconditionally
-unstable for finite $Q$, although instability occur on smaller scales
+unstable for finite $Q$, although instability occurs on smaller scales
 as $Q$ increases.  The $\beta\to0$ limit corresponds to a 
@@
 $|\beta S|\to 0$). In this respect, a 
-fully irradiated disk, in that the equilibrium temperature is set
+fully irradiated disk, in which the equilibrium temperature is set
 externally, behaves isothermally regardless of the cooling time.
@@
 Eq. \ref{thindisk}. The limiting behavior for small/large $K$ are
-well-captured by Eq. \ref{gammie_smallk} and
+well-captured by Eqs. \ref{gammie_smallk} and
 \ref{gammie_bigk}. Comparing the two panels 
@@
 shows that increasing the
-irradiation level ($\theta$) supresses small-scale
+irradiation level ($\theta$) suppresses small-scale
 perturbations. %(This also occurs in the inviscid case.)
@@
   \caption{Growth rates for the 2D viscous problem as a function of
-    the radial wavenumber $k$ for a range of cooling times
-    $\beta$. The dashed and dotted lines correspond to asympototic
+    the radial wavenumber, $k$, for a range of cooling times, 
+    $\beta$. The dashed and dotted lines correspond to asymptotic
     behaviors for small and large $k$, respectively, computed from

### 2016-02-02 paper/results3d.tex
submitted version

@@
 In the following examples we consider a vertically isothermal disk
-($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
+($\Gamma=1$ in Eq. \ref{poly_vert}). Then in the viscous case $\alpha$ is vertically
 constant (Eq. \ref{alpha_beta_relation}). We consider only even modes
-about the midplane, and apply a numerical disk surface at $z=\zmax$
+about the mid-plane, and apply a numerical disk surface at $z=\zmax$
 such that $\rho(\zmax)=0.05\rho_0$.
@@
 corrections. Softening the self-gravity in 2D captures the correct 
-qualitative behavior as in the full 3D case. For $\beta\gtrsim 1$
+qualitative behavior of the full 3D case. For $\beta\gtrsim 1$
 choosing $H_\mathrm{sg}=0.8H$ produces a good match. However, it is

### 2016-02-02 paper/summary.tex
submitted version

@@
 and may also act on linear perturbations. %thermo and dynamic 
-We suggest that disk fragmentation, observed in numerical simulations, 
+We suggest that disk fragmentation observed in numerical simulations  
 can be understood as the eventual outcome of secondary
@@
 stability problem, which is present even if there is no physical
-irradiation.  
+irradiation. This captures the fact that density enhancements impede cooling.
+
 %This effective irradiation arises because, for the stability problem,
@@
 \citep{lodato11,meru12}. 
+We can expect this if numerical viscosity
+contributes to an effective irradiation, %then we do expect a disk can be
+%destabilized at longer cooling times with increasing resolution, 
+because then perturbations can cool to lower temperatures with
+increasing resolution, see Fig. \ref{invisc_theta}.
 
@@
 above, employ the standard beta cooling function, Eq. \ref{beta_cool},
-\emph{without} a floor temperature. We show in \S\ref{2d_inviscid}
+\emph{without} a physical floor temperature. We show in \S\ref{2d_inviscid}
 (see also \S\ref{cool_gi}) that this implies cooling-driven GI can
-occur at any sufficiently small scales. Therefore as the numerical
+occur at any sufficiently small scale. Therefore as the numerical
 resolution increases, simulations can access a wider range of unstable
-scales.  %no upper bound  
+scales. Although small-scale modes have weaker growth rates, they can 
+become important over long timescales.  
+%can access smaller, unstable scales 
 In this respect, it is perhaps not surprising to find non-convergence with 
-increasing resolution. %simple beta cooling will not give a
+increasing resolution and/or integration times. 
+%simple beta cooling will not give a
 %well-imposed problem
-
-Note that there may be a `numerical floor temperature' in 
-simulations, e.g. due to numerical viscosity, which should  
-decrease with resolution. Then we expect a disk can be
-destabilized at larger cooling times with increasing resolution
-because the effective floor temperature drops   
-\citep[cf.][see also Fig. \ref{invisc_theta}]{meru11}.
 
@@
 Furthermore, without a floor temperature, standard beta cooling is
-a function of the pressure/energy density field only. It is not clear
-how relevant are results obtained from this is to actual PPDs,
-where cooling depends on two thermodynamic states (e.g. pressure and
+a function of the pressure/energy density field only. 
+%It is not clear 
+%how relevant are results obtained from this is to actual PPDs,
+There may be some inconsistency in applying results obtained from 
+this to actual PPDs where cooling depends on two thermodynamic states (e.g. pressure and
 temperature). A floor temperature permits a mapping between standard beta
@@
 Moreover, the standard beta cooling does 
-not account for optical depth effects, which force the midplane 
+not account for optical depth effects, which force the mid-plane 
 and high density perturbations to cool more slowly.
@@
 Even at very high resolution one might worry that the artificial dissipation scale
-imposed by the grid/smoothing length is too similar to scale of fragmentation.
+imposed by the grid/smoothing length is too similar to the scale of fragmentation.
 %it is true that most unstable wavelength is H is this is always
@@
 \subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
-We emphasize the linear framework we have developed does not assume  
+We emphasize that the linear framework we have developed does not assume  
 a particular origin for the turbulence that is represented by the imposed
@@
 \citep{paardekooper12,hopkins13}. In this interpretation,  
-fragmentation is attributed to the occurence and survival 
+fragmentation is attributed to the occurrence and survival 
 of large, non-linear density enhancements, which arise from the
-gravito-turbulent flucuations simply by chance. %KMK
+gravito-turbulent fluctuations simply by chance. %KMK
 There is insufficient evidence that gravito-turbulence adequately samples

### 2016-03-22 paper/linear3d.tex
submitted version

@@
 \section{Linear problem}\label{linear}
-We consider infinitestimal axisymmetric Eulerian perturbations of the
+We consider infinitesimal axisymmetric Eulerian perturbations of the
 form \begin{align}
@@
 
-Note that any linearized beta cooling function can be put into the
+Note that any beta cooling function can be linearized in the
 form of Eq. \ref{linear_beta} with appropriate definitions of $t_c$

### 2016-03-23 paper/model3d.tex
revising, point 4

@@
 density $E$ are related by  
-\begin{align}
+\begin{align}\label{ideal_eos}
   P = (\gamma-1)E = \mathcal{R}\rho T, 
 \end{align}
-where $\mathcal{R}$ is the gas constant and $T$ is the temperature. The gas 
+where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
+{\bf For simplicity we refer to a single adiabatic index as that
+  appears in the equation of state for both 2D and 3D disk models 
+  \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
+                             %for the same 
+                             %disk in 2D 
+%johnson: specify 3D gamma, then translate into 2D gamma because disk
+%model is 2D 
+}
+The gas 
 gravitational potential $\Phi$ is given via the Poisson equation,  
@@
 Recall  $\tcool = \beta\Omega^{-1}$ is 
-the cooling timescale with $\beta$ a constant input parameter. Here, 
-\begin{align}\label{tirr_def}
-  \tirr = \theta T_\mathrm{eq}
-\end{align}
-is a reference temperature field parameterized as a constant 
-fraction $\theta$ of the equilibrium temperature
-$T_\mathrm{eq}$. Physically, $\tirr$ is the floor temperature set by,
-for example, stellar or background irradiation. Note that $\theta$  
+the cooling timescale with $\beta$ a constant input parameter. 
+{\bf Here, $T_\mathrm{irr}$}
+%Here, 
+%\begin{align}\label{tirr_def}
+%  \tirr = \theta T_\mathrm{eq}
+%\end{align}
+is a reference temperature field. 
+%parameterized as a 
+%fraction $\theta$ of the equilibrium temperature field 
+%$T_\mathrm{eq}(z)$. 
+Physically, $\tirr$ {\bf may} the floor temperature set by, 
+for example, stellar or background irradiation. {\bf
+
+}
+
+{\bf
+Note that $\theta$  
 should only be interpreted as an irradiation parameter when it is 
@@
 Eq. \ref{beta_cool} adopted as the cooling function. 
+}
 
@@
 where $\Gamma$ is the constant polytropic index and $\rho_0 =
-\rho(z=0)$ is the equilibrium mid-plane density. 
+\rho(z=0)$ is the equilibrium mid-plane density. {\bf Note that
+  $\Gamma$ is only relevant 
+  to the 3D problem as it determines the disk vertical structure.} 
 The vertical structure is first obtained from 

### 2016-03-23 paper/paper3d.tex
begin revision
updated aknowledge list
revising, point 4

@@
 %external irradiation, and cooling. 
-We analyze the gravitational stability of viscous, non-adiabatic
-accretion disks with internal heating, external irradiation, and cooling.  
-We consider parameterized cooling functions in two and three dimensions,
+We analyze the {\bf axisymmetric} gravitational stability of viscous, non-adiabatic
+accretion disks with internal heating, external irradiation, and
+cooling {\bf in the shearing box approximation}.   
+We consider parameterized cooling functions in {\bf 2D} and {\bf 3D}
+disks, %two and three dimensions,  
 as well as   
@@
 Understanding the gravitational stability of rotating disks    
-is central to many astrophysical problems. In the context of 
+is central to many astrophysical problems {\bf \citep{kratter16}}. In the context of 
 gaseous protostellar or protoplanetary disks (PPDs), gravitational instability (GI)
@@
 pressure or rotational support. We then 
-discuss in \S\ref{frag_intro} how these effects may relate   
-to the transition from self-regulated GI to fragmentation seen in
-numerical simulations. The rest of this paper is laid out in
-\S\ref{plan}.
+discuss in \S\ref{frag_intro}{\bf---\S\ref{gi_of_gi}} how these
+effects may relate to the transition from self-regulated GI to
+fragmentation seen in numerical simulations. The rest of this paper is
+laid out in \S\ref{plan}.
 
@@
 thermal equilibrium \citep{gammie01}. 
-If one represents this turbulent heating by a viscous heating, then
-the $\alpha$ and $\beta$ parameters defined above will  
-be inversely related (see, e.g. Eq. \ref{alpha_beta_relation}).
+%If one represents this turbulent heating and transport {\bf by a viscosity} 
+%heating by a viscous heating
+
+{\bf 
+  Global disk simulations \citep[e.g.][]{lodato04} show
+  that the transport and 
+  heating associated with this gravito-turbulence may be described as a
+  local viscous process provided that the disk-to-star mass ratio is
+  small ($\lesssim0.25$) and the disk is thin (aspect-ratio $\lesssim
+  0.1$).  
+}
+{\bf In this case} the $\alpha$ and $\beta$ parameters defined above
+will be inversely related (see, 
+e.g. Eq. \ref{alpha_beta_relation}). {\bf Recent 
+  vertically-extended shearing box simulations 
+  also confirm this relation in 3D disks \citep{shi14}.}
 
@@
 gravitational stability independently of their influence on the exact
-value of the classic $Q$. This motivates a physical      
+value of the classic $Q$. This 
+motivates a physical      
 interpretation of disk fragmentation as the inability to maintain a
@@
 secondary instabilities driven by cooling and/or viscosity.
+
+{\bf 
+\subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
+  In this work, we formally regard the gravito-turbulent
+  disk described above as an equilibrium state, to which we apply    
+  standard linear stability analysis. Then by `perturbations' we mean   
+  deviations away from this gravito-turbulent 
+  basic state\footnote{This should not be confused with 
+    flucuations with respect to the 
+    laminar disk that maintain the underlying
+    gravito-turbulence.}. 
+  We assume that instabilities defined this way  
+  signifies disk fragmentation, since numerical simulations  
+  show that the non-existance of steady gravito-turbulence imply    
+  fragmentation.   
+}
 

### 2016-03-23 paper/model3d.tex
revising

@@
 where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
-{\bf For simplicity we refer to a single adiabatic index as that
+{\bf For simplicity we only refer to the adiabatic index as that
   appears in the equation of state for both 2D and 3D disk models 
@@
   %  \frac{P}{(\gamma-1)T}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}, 
-  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c}.
+  \frac{\mathcal{R}\rho}{(\gamma-1)}\frac{\left(T-T_\mathrm{irr}\right)}{t_c},
 \end{align}
-Recall  $\tcool = \beta\Omega^{-1}$ is 
+{\bf where $T_\mathrm{irr}$ is a reference temperature field, and
+recall}  $\tcool = \beta\Omega^{-1}$ is 
 the cooling timescale with $\beta$ a constant input parameter.
-{\bf Here, $T_\mathrm{irr}$}
 %Here, 
@@
 %\end{align}
-is a reference temperature field. 
 %parameterized as a 
@@
 Physically, $\tirr$ {\bf may} the floor temperature set by, 
-for example, stellar or background irradiation. {\bf
-
-}
-
-{\bf
-Note that $\theta$  
-should only be interpreted as an irradiation parameter when it is 
-defined through Eq. \ref{tirr_def} and 
-Eq. \ref{beta_cool} adopted as the cooling function. 
-}
+for example, stellar or background irradiation.
 
@@
 \alpha(z) = \frac{1}{(\gamma-1)\beta
-   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right). 
+   q^2}\frac{c_s^2(z)}{c_{s0}^2}\left(1 - \theta\right),
 %  \alpha(z) = \frac{1}{\Gamma(\gamma-1)\beta
@@
 \end{align} 
+{\bf with 
+  \begin{align}\label{tirr_def}
+    \theta = \frac{\tirr}{T_\mathrm{eq}},
+  \end{align}
+   where $T_\mathrm{eq}(z)$ is the equilibrium temperature field.
+   We shall consider vertically isothermal disks with $\Gamma=1$, as
+  appropriate for the outer parts of irradiated protoplanetary disks
+   \citep{chiang97}. In this case  
+   $\alpha$, $T_\mathrm{eq}$ and $\theta$ are simply constants. 
+    Note that $\theta$  
+    should only be interpreted as an irradiation parameter when it is 
+    defined through Eq. \ref{tirr_def}, in conjunction with adopting  
+    Eq. \ref{beta_cool} as the cooling function. 
+ }
 

### 2016-03-23 paper/paper3d.tex
revising

@@
   disk described above as an equilibrium state, to which we apply    
-  standard linear stability analysis. Then by `perturbations' we mean   
+  standard linear stability analysis. Thus by `perturbations' we mean   
   deviations away from this gravito-turbulent 
@@
     gravito-turbulence.}. 
-  We assume that instabilities defined this way  
+  We assume instabilities defined this way  
   signifies disk fragmentation, since numerical simulations  
@@
 %A. Youdin 
+%paardekooper
+%cathie clarke 
+%lodato
+%gammie
 

### 2016-03-24 paper/model3d.tex
revising issue 8 now

@@
 {\bf For simplicity we only refer to the adiabatic index as that
-  appears in the equation of state for both 2D and 3D disk models 
+  appears in Eq. \ref{ideal_eos} for both 2D and 3D disk models 
   \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
@@
 \end{align}
-where $\Gamma$ is the constant polytropic index and $\rho_0 =
-\rho(z=0)$ is the equilibrium mid-plane density. {\bf Note that
-  $\Gamma$ is only relevant 
-  to the 3D problem as it determines the disk vertical structure.} 
+{\bf where $\rho_0 = \rho(z=0)$ is the equilibrium mid-plane density, and    
+$\Gamma$ is the constant polytropic index which determines the disk's
+  vertical structure. Thus $\Gamma$ is only relevant to the 3D
+  problem.}     
 The vertical structure is first obtained from 

### 2016-03-24 paper/results2d.tex
revising issue 8 now

@@
 perturbations. %KMK 
-We emphasize that destabilization is independent of the effect of decreasing temperature on the instantaneous
-value of the classic-$Q$. A time-independent heat source should be
-invoked to balance the imposed cooling to allow an equilibrium to be
-defined ($\mathcal{H}_\mathrm{ext}\neq 0$). For example, we could assume
-that the viscosity only provides a background heating and does not
-play an active role in the perturbed state.
+We emphasize that destabilization is independent of the effect of
+decreasing temperature on the instantaneous value of the
+classic-$Q$. A time-independent heat source should be invoked to
+balance the imposed cooling to allow an equilibrium to be 
+defined ($\mathcal{H}_\mathrm{ext}\neq 0$).
+
+{\bf
+For example, we could assume that the viscosity only provides a
+background heating and does not play an active role in the perturbed
+state. This is in fact done implicitly in the literature when
+discussing fragmentation or clump formation in cooling, 
+self-gravitating disk simulations \citep{gammie01}. 
+There, the effect of the ambient 
+gravito-turbulent viscosity on the forming-clump is neglected, as 
+one only compares adiabatic heating and the imposed cooling.  
+This comparison is encapsulated in the generalized dispersion relation
+below.}
 
@@
 \end{align}
-This equation gives remarkably similar values 
+This equation gives similar values 
 of the cooling times below which 
@@
 Table \ref{bstar_compare} shows rough agreement between $\beta_*$ and 
-$\beta_c$.
+$\beta_c$.  {\bf The match is remarkable, especially with the
+  global 3D simulations of \cite{rice05}, since Eq. \ref{betastar}
+  is derived for 2D disks in the local limit.}
 
@@
 (Eq. \ref{alpha_beta_relation}). Furthermore, we relate
+{\bf the strength of self-gravity and viscosity by}   
 \begin{align} 
@@
 %with $Q_\mathrm{crit}$ given by 
-to mimic a basic state maintained by gravito-turbulence where one
+{\bf to mimic a gravito-turbulent basic state,} where one
 might expect the dimensionless stress $\alpha \sim Q^{-2}$
@@
 %KMK
-We might interpret fragmentation as the response to the instability of the
-highly viscous state. This is consistent with previous numerical
+We might interpret fragmentation {\bf as a reflection of} instability
+of the highly viscous state. This is consistent with previous numerical
 simulations performed by \cite{lodato05}. 

### 2016-03-24 paper/application.tex
rough revision done

@@
 PPDs become irradiation-dominated at large distances from the star,
-where $T\to\tirr$ and $\theta\to 1$ \citep{dalessio97,kratter11}.
+where $T\to\tirr$ and $\theta\to 1$ {\bf \citep{chiang97,dalessio97,kratter11}}.
 Then Eq. \ref{ppd_invisc_cond} relaxes to 
@@
     are analytic results based on Eq. \ref{gammie_smallk}. 
+    {\bf 
+      For both accretion rates the cooling time drops below $3/\Omega$
+      beyond the vertical dashed line, but for the  
+      $\dot{M}=10^{-6}M_\sun\mathrm{yr}^{-1}$ disk $\alpha > 0.1$
+      beyond this radius, while $\alpha\lesssim
+      0.03$  for the
+      $\dot{M}=10^{-7}M_\sun\mathrm{yr}^{-1}$. 
+    }
     \label{rafikov_growth}}
@@
     obtained from the corresponding 2D problem with softened gravity.   
+    {\bf
+      Beyond the vertical dashed line $\alpha>0.1$, which corresponds to
+      cooling times $t_\mathrm{c}\lesssim 4/\Omega$.  
+    }
     % The black curves are obtained

### 2016-03-24 paper/paper3d.tex
rough revision done

@@
 occurs in dusty fluids where the required frictional forces are 
-provided by dust-gas drag \citep{ward00, takahashi14}. 
+provided by dust-gas drag {\bf \citep{goodman00,ward00, takahashi14}}. 
 In fact, this
@@
 formation %frag of dust 
-\citep{youdin11}. Therefore it is not unreasonable to expect analogous
+{\bf \citep{youdin05,youdin11}}. Therefore it is not unreasonable to expect analogous
 fragmentation in gaseous disks due to viscosity.
@@
 $Q =O(1)$ in which cooling is balanced by shock heating to maintain
-thermal equilibrium \citep{gammie01}. 
+thermal equilibrium {\bf \citep{gammie01,shi14}}. 
 %If one represents this turbulent heating and transport {\bf by a viscosity} 

### 2016-03-24 paper/results2d.tex
rough revision done

@@
 fully irradiated disk, in which the equilibrium temperature is set
-externally, behaves isothermally regardless of the cooling time.
+externally, behaves isothermally regardless of the cooling time 
+{\bf \citep{gammie01,johnson03}}.
 
@@
 %KMK
-We might interpret fragmentation {\bf as a reflection of} instability
-of the highly viscous state. This is consistent with previous numerical
+We might interpret fragmentation {\bf as an} instability
+of the highly viscous state {\bf (see \S\ref{gi_of_gi})}. This is
+consistent with previous numerical 
 simulations performed by \cite{lodato05}. 

### 2016-03-24 paper/summary.tex
rough revision done

@@
 because then perturbations can cool to lower temperatures with
-increasing resolution, see Fig. \ref{invisc_theta}.
+increasing resolution, see Fig. \ref{invisc_theta}.  {\bf
+  In global simulations, non-convergence has also been attributed to initial 
+  conditions that lead to internal edges \citep{paardekooper11b}, but
+  this cannot be modeled in our local setup. 
+  \cite{rice12} point out that the standard implementation of beta-cooling in
+  SPH simulations apply cooling on scales well-below the SPH smoothing
+  lengths, and that this inconsistency may contribute to
+  non-convergence. 
+}
 
- \cite{paardekooper12} also found that fragmentation can occur 
+{\bf However,} \cite{paardekooper12} also found {\bf in local grid-based simulations} that fragmentation can occur 
 for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   

### 2016-03-26 paper/model3d.tex
revising

@@
 the cooling timescale with $\beta$ a constant input parameter. 
-%Here, 
-%\begin{align}\label{tirr_def}
-%  \tirr = \theta T_\mathrm{eq}
-%\end{align}
-%parameterized as a 
-%fraction $\theta$ of the equilibrium temperature field 
-%$T_\mathrm{eq}(z)$. 
 Physically, $\tirr$ {\bf may} the floor temperature set by, 
@@
 
-For beta cooling given by Eq. \ref{beta_cool} we assume an 
+{\bf
+Beta cooling of the form Eq. \ref{beta_cool} is widely applied in 2D
+and 3D numerical simulations of self-gravitating disks \citep{gammie01,
+  rice05,rice11,paardekooper12}. In fact, for $T_\mathrm{irr}=0$
+Eq. \ref{beta_cool} the 
+cooling function $\Lambda = -E/t_\mathrm{c}$ is identical
+to that originally employed by \cite{gammie01}. 
+We will refer to Eq. \ref{beta_cool} as  
+`standard' beta cooling. It permits numerical experiments to be
+carried out in a controlled manner as a function of the cooling time
+$\beta$. An adiabatic disk corresponds to $\beta\to \infty$. The
+physical meaning of the limit $\beta\to0$ depends on $T_\mathrm{irr}$, as
+discussed in \S\ref{theta0}---\ref{theta1}.  
+}
+
+{\bf For standard beta cooling} we assume an 
 equilibrium polytropic relation 
@@
  }
-
-Beta cooling of the form Eq. \ref{beta_cool} is widely applied in 2D
-and 3D numerical simulations of self-gravitating disks \citep{gammie01,
-  rice05,rice11}. For this reason we will refer to it as the
-`standard' beta cooling. It permits numerical experiments to be
-carried out in a controlled manner as a function of the cooling time
-$\beta$. An adiabatic disk corresponds to $\beta\to \infty$. The
-physical meaning of the limit $\beta\to0$ depends on $\theta$, as
-discussed in \S\ref{theta0}---\ref{theta1}.
 

### 2016-03-26 paper/application.tex
revised version done

@@
     {\bf 
-      For both accretion rates the cooling time drops below $3/\Omega$
+      For both accretion rates the cooling time $\lesssim 3/\Omega$
       beyond the vertical dashed line, but for the 
@@
       beyond this radius, while $\alpha\lesssim
-      0.03$  for the
-      $\dot{M}=10^{-7}M_\sun\mathrm{yr}^{-1}$. 
+      0.03$ throughout the
+      $\dot{M}=10^{-7}M_\sun\mathrm{yr}^{-1}$ disk. 
     }

### 2016-03-26 paper/model3d.tex
revised version done

@@
 and 3D numerical simulations of self-gravitating disks \citep{gammie01,
-  rice05,rice11,paardekooper12}. In fact, for $T_\mathrm{irr}=0$
-Eq. \ref{beta_cool} the 
+  rice05,rice11,paardekooper12}. In fact, for $T_\mathrm{irr}=0$ the 
 cooling function $\Lambda = -E/t_\mathrm{c}$ is identical 
@@
 {\bf where $\rho_0 = \rho(z=0)$ is the equilibrium mid-plane density, and    
-$\Gamma$ is the constant polytropic index which determines the disk's
+$\Gamma$ is the constant polytropic index that determines the disk's
   vertical structure. Thus $\Gamma$ is only relevant to the 3D

### 2016-03-26 paper/paper3d.tex
revised version done

@@
 {\bf In this case} the $\alpha$ and $\beta$ parameters defined above
-will be inversely related (see, 
+{\bf are} inversely related (see, 
 e.g. Eq. \ref{alpha_beta_relation}). {\bf Recent 
@@
 \subsection{Instability of the gravito-turbulent state}\label{gi_of_gi}
-  In this work, we formally regard the gravito-turbulent
+  In this work, we formally treat the gravito-turbulent
   disk described above as an equilibrium state, to which we apply    
@@
   basic state\footnote{This should not be confused with 
-    flucuations with respect to the 
+    fluctuations with respect to the 
     laminar disk that maintain the underlying
@@
   signifies disk fragmentation, since numerical simulations  
-  show that the non-existance of steady gravito-turbulence imply    
+  show that the non-existence of steady gravito-turbulence implies 
   fragmentation.   
@@
 %\acknowlegement
+{\bf
+We thank the anonymous referee's prompt report that helped to improve
+the clarity of this work. We thank S. Stahler and A. Youdin for
+comments during the  course of this project; C. Clarke,
+C. Gammie, and S.-J. Paardekooper for feedback on the first version of
+this article, and G. Lodato for useful discussions.  
+}
 %S. Stahler 

### 2016-03-26 paper/results2d.tex
revising
revised version done

@@
 state. This is in fact done implicitly in the literature when
-discussing fragmentation or clump formation in cooling, 
+discussing fragmentation of cooling, 
 self-gravitating disk simulations \citep{gammie01}. 
@@
 The maximum growth rate, $S_*\propto \beta^{-1/3}$, smoothly
-increases with decreasing $\beta$. 
+increases with decreasing $\beta$. {\bf Notice for $\beta>1$ this growth
+  rate is faster than the imposed cooling rate $\beta^{-1}\Omega$}.
 %Instability formally occurs for

### 2016-03-26 paper/summary.tex
revising
revised version done

--- 2016-02-02 paper/summary.tex
@@
 because then perturbations can cool to lower temperatures with
-increasing resolution, see Fig. \ref{invisc_theta}.
+increasing resolution, see Fig. \ref{invisc_theta}.  {\bf
+  In global simulations, non-convergence has also been attributed to initial 
+  conditions that lead to internal edges \citep{paardekooper11b}, but
+  this cannot be modeled in our local setup. 
+  \cite{rice12} point out that the standard implementation of beta cooling in
+  SPH simulations apply cooling on scales well-below the SPH smoothing
+  lengths, and that this inconsistency may contribute to
+  non-convergence. 
+}
 
- \cite{paardekooper12} also found that fragmentation can occur 
+{\bf However,} \cite{paardekooper12} also found {\bf in local 2D
+  grid-based simulations} that fragmentation can occur  
 for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
@@
 increasing resolution and/or integration times. 
+{\bf
+  On the other hand, the convergence issue may be
+  less serious in 3D since small-scale modes are more stable in 3D
+  than in 2D (Appendix \ref{3dcorr}). 
+}
 %simple beta cooling will not give a
@@
 for numerical convergence. \cite{baehr15} included a floor temperature 
-in their simulations but still find that at fixed cooling rates, disks
+in their {\bf local 2D} simulations but still find that at fixed cooling rates, disks
 eventually fragment with sufficient spatial resolution.

### 2016-03-28 paper/model3d.tex
revised version done

@@
 where $\mathcal{R}$ is the gas constant and $T$ is the temperature. 
-{\bf For simplicity we only refer to the adiabatic index as that
-  appears in Eq. \ref{ideal_eos} for both 2D and 3D disk models 
+{\bf For simplicity we refer to the adiabatic index $\gamma$ as that
+  in Eq. \ref{ideal_eos} for both 2D and 3D disk models 
   \citep[cf.][]{johnson03}.  %should adopt diff values in 2D and 3D
@@
 
-\subsection{Steady states and cooling models}
+\subsection{Steady states and cooling models}\label{steady_state}
 We consider equilibrium solutions (here omitting the `eq' subscripts
@@
 the cooling timescale with $\beta$ a constant input parameter. 
-Physically, $\tirr$ {\bf may} the floor temperature set by, 
+Physically, $\tirr$ {\bf may be} the floor temperature set by, 
 for example, stellar or background irradiation.

### 2016-03-28 paper/paper3d.tex
revised version done

@@
 %of modes of order the disk scale height. 
-We apply this new linear theory to 
-models of protoplanetary disks subject to
+We apply this new linear theory to protoplanetary disks subject to
 gravito-turbulence modeled as an effective 
@@
 self-gravity, and that the observed transition in behavior from 
-gravito-turbulent to fragmenting may derive from  
+gravito-turbulent to fragmenting may {\bf reflect}   
 instability of the gravito-turbulent state itself.
@@
     gravito-turbulence.}. 
-  We assume instabilities defined this way  
-  signifies disk fragmentation, since numerical simulations  
-  show that the non-existence of steady gravito-turbulence implies 
+  In our model, we interpret instability of such perturbations  to 
+  signify fragmentation. Although we cannot formally demonstrate this, numerical
+  simulations suggest that gravito-turbulence and fragmentations are the
+  only outcomes of classic GI. Thus the failure to maintain the 
+  gravito-turbulent steady state seems a reasonable description of
   fragmentation. 
-}
+%  We assume instabilities defined this way  
+%  signifies disk fragmentation, since numerical simulations  
+%  show that the non-existence of steady gravito-turbulence implies 
+%  fragmentation.   
+%Indeed, we find that the region where a gravito-turbulent PPD is
+%dynamically unstable agrees with the current consensus on  where, and
+%the conditions under which, a PPD is expected to fragment based on
+%Eq. \ref{frag_cond}, with typical critical values cited in the
+%literature
 
-Indeed, we find that the region where a gravito-turbulent PPD is
-dynamically unstable agrees with the current consensus on  where, and
-the conditions under which, a PPD is expected to fragment based on
-Eq. \ref{frag_cond}, with typical critical values cited in the
-literature \citep[e.g. $\beta_c\sim
+Indeed, our linear analysis for perturbations with respect to the 
+gravito-turbulent state predicts that PPDs will fragment under similar
+conditions observed in numerical simulations} \citep[e.g. $\beta_c\sim
   3,\,\alpha_c\sim0.1$,][]{gammie01,rice05}.

### 2016-03-28 paper/summary.tex
revised version done

@@
   this cannot be modeled in our local setup. 
-  \cite{rice12} point out that the standard implementation of beta cooling in
-  SPH simulations apply cooling on scales well-below the SPH smoothing
+  \cite{rice12} point out that the standard implementation of beta
+  cooling in smoothed-particle hydrodynamics (SPH) applies cooling on scales well-below the SPH smoothing
   lengths, and that this inconsistency may contribute to
@@
 for slowly-cooled disks. 
+{\bf
+  However, to properly consider long timescales, it may be necessary
+  to account for
+  secular evolution in the global disk. 
+}        
 %hopkins paper
